---
title: "Differential Hydroxy-methylation Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: yeti
    code_folding: hide
---


```{r setup, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png", cache=TRUE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE, prompt=TRUE, comment='', fig.cap='', tidy.opts=list(keep.blank.line=FALSE, width.cutoff=200), fig.width = 16, fig.height = 14)
```

# Overview

RNAseq DE analysis for Ira Differential Ira Fetahu (IFETAHU@PARTNERS.ORG), Shi group.  

Contact John Hutchinson (jhutchin@hsph.harvard.edu) for additional details.

The most recent update of this html document occurred: `r date()`.

The sections below provide code to reproduce the included results and plots. 

---

# Setup

## Libraries and Variables

```{r vars}
library(broom)
library(tidyr)
library(stringr)
library(reshape2)
library(ggplot2)
library(CHBUtils)
library(dplyr)
library(biomaRt)
library(readr)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442","#0072B2", "#D55E00", "#CC79A7")
baseDir <- "~/projects/shi_wgbs/"
dataDir <- file.path(baseDir, "data")
resultsDir <- file.path(baseDir, "results")
metaDir <- file.path(baseDir, "meta")
pval.cutoff=0.05
```

---

# Data and Metadata Import 

- pull in Lorena's merged results file from BSMAP, MLML and custom scripts
- filter out rows with conflicts for MLML and negative MLML values

```{r import, cache.lazy=FALSE}
metadata = read.csv(file.path(metaDir, "Sample_IDs.csv"))
datafiles = file.path(dataDir, list.files(path=file.path(dataDir), pattern="merged"))
#load data
dataset <- lapply(datafiles, function(datafile) {
	id <- strsplit(basename(datafile), "_")[[1]][1]
	data <- read_delim(datafile, delim="\t")
	data <- filter(data, conflicts==0, hmC>0)
	data1 <- dplyr::select(data, chr, pos, strand, context)
	data2 <- dplyr::select(data, -chr, -pos, -strand, -context) 
	names(data2)<- paste(names(data2), id, sep="_")
	data <- cbind(data1, data2)
	tbl_df(data)
})  
```

Merge datasets to sites that are assayed in all samples

```{r merge}
merged.data <- Reduce(function(x,y) inner_join(x,y, by=c("chr", "pos", "strand", "context")), dataset) 
```

# Data Manipulations

Subset merged dataset to sites that have a  above zero level of hydroxymethylation in at least one sample in one group
(can modify this if necessary)

```{r hmC_fdr_filter}
merged.filtered.data <- filter(merged.data, (hmC_fdr_s345<pval.cutoff | hmC_fdr_s347<pval.cutoff) | (hmC_fdr_s409<pval.cutoff | hmC_fdr_s411<pval.cutoff | hmC_fdr_s413<pval.cutoff))
```

Subset data to columns of interest (C/T counts for bisulfite and oxy-bisulfite runs, pvalue fdrs for hmC presence, and estimated hydroxymethylation rates and qualities (conflicts) from MLML)

```{r subsetcols}
merged.filtered.subset.data <- dplyr::select(merged.filtered.data, chr, pos, strand, context, contains("CT_counts"), contains("hmC"),  -contains("pval"), -contains("fdr"), -contains("eff"))

merged.filtered.subset.data.m <- melt(merged.filtered.subset.data, id.vars=c("chr", "pos", "strand", "context")) %>% 
	tbl_df() %>%
       	separate(., variable, into=c("measure", "sampleid"), "_s") %>%
       	spread(., measure, value)
```

Add info about sample classes to dataset

```{r metadatamerge}
metadata <- metadata %>% tbl_df() %>% 
	dplyr::select(., phenotype, sampleid) %>% 
	distinct() %>% 
	mutate(., sampleid=as.character(sampleid))

merged.filtered.subset.data.m <- dplyr::left_join(merged.filtered.subset.data.m, metadata, by="sampleid") 
```

Drop any samples with negative MLMKL derived hMC levels (they mess up the regression otherwise)

```{r dropnegs}
merged.filtered.subset.data.m <- filter(merged.filtered.subset.data.m, hmC>=0)
```

Calculate mean and min counts for bisulfite and oxy-bisulfite runs for sample

```{r meanmincalcs}
merged.filtered.subset.data.m <- merged.filtered.subset.data.m %>% 
	rowwise() %>% 
	mutate(., meancounts=mean(c(CT_counts, ox_CT_counts)), mincounts=min(c(CT_counts, ox_CT_counts))) 
```

Run regression and extract statistics

```{r regressions}
glmsummaries <- merged.filtered.subset.data.m %>% 
	group_by(., chr, pos, strand, context) %>% 
	do(tidy(glm(hmC ~ phenotype, weights=meancounts, family=binomial, data=.)))
```

```{r filterregressions}
glmsummaries <- glmsummaries %>% filter(., !grepl("Intercept",term )) 
```
```{r ungroupregressions}
glmsummaries <- glmsummaries %>% ungroup()
```

Multiple test correction

```{r fdrregressions}
glmsummaries <- glmsummaries %>% mutate(., fdr=p.adjust(p.value, method="fdr"))
```

```{r saveimage}
save.image()
```


```{r sessioninfo}
sessionInfo()
```

