---
title: ""
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 5
  pdf_document:
    toc: yes
bibliography: references.bib

---
```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(knitcitations)
cleanbib()
options("citation_format" = "pandoc")

clientname=""
clientemail=""
labPI="Nicolas Stylopolous"
lablocation=""
analystname=""
analystemail=""


library(knitr)
opts_chunk$set(warning=FALSE, error=FALSE, message=FALSE, echo=FALSE,cache=FALSE, tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120), dev="svg")
options(width=200)
```

---

Array analysis for `r clientname` (`r clientemail`), `r labPI` group at `r lablocation`.  

Contact `r analystname` (`r analystemail`) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

---

# Methods Summary  

## Batch Analyses

All Affymetrix HTA 2.0 arrays were processed using the 'oligo' BioConductor package `r citep("10.1093/bioinformatics/btq431")`, quality-controlled with arrayQualityMetrics `r citep("10.1093/bioinformatics/btn647")` and normalized with RMA `r citep("10.1093/biostatistics/4.2.249")`. Differentially expressed genes were identified using limma `r citep("http://link.springer.com/chapter/10.1007%2F0-387-29362-0_23")`.

---

# Setup

## Variables
Working directories, files and other variables necessary to the analysis.

```{r variables}
## Setup Data and Results directory variables
baseDir <- getwd()
dataDir <- file.path(baseDir, "data")
metaDir <- file.path(baseDir, "meta")
resultsDir <- file.path(baseDir, "results")

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7") # colorblind friendly palette
covarsfilename="covars.desc" # tab delimited file describing samples
lowintensity.percentile=0.1
mad.quantile.cutoff=0.1
pvalue.cutoff=0.05
highlight.color="green"
lfc.cutoff=1
```

## Libraries

[Bioconductor](http://www.bioconductor.org) and [R](http://cran.r-project.org/) libraries used to process and visualize the data.

```{r libraries_variables, echo=TRUE}
library(knitr) # for simple tables
library(oligo) # array utilities
library(arrayQualityMetrics) # array quality control reports
library(limma) # array statistical analyses
library(CHBUtils) # some homegrown functions
library(reshape2) # data format utility
library(ggplot2) # pretty graphs
library(ggdendro) # for pretty dendrograms
library(RColorBrewer) # more colors
library(gridExtra) # for arranging multiple plots
library(pheatmap) # pretty heatmaps
library(corrgram)
library(pvca)
library(dplyr) # data format utility 
library(hta20sttranscriptcluster.db) #new package for annotation
library(sva) # Surrogate Variable Analysis (includes ComBat)
```

## Functions

```{r functions, echo=FALSE}
# for plotting amount of variation explained by principal components
PCAplot.sd.eset <- function(eset=NULL,  title=NULL){
  eset.core <- exprs(eset)
  myPca.core <- prcomp(t(eset.core))
  # SD of components
  sdevdf <- data.frame(cbind(as.numeric(myPca.core$sdev),c(1:length(myPca.core$sdev))))
  sdevdf$prop <-  sdevdf$X1/sum(sdevdf$X1)
  sdevdf$cum <- cumsum(sdevdf$prop)
  ggplot(sdevdf, aes(x=X2, y=prop)) + 
    geom_point(size=4, color="red") + 
    scale_x_continuous('Component') + 
    scale_y_continuous('Standard Deviation') +
    ggtitle(title) +
    geom_line(data=sdevdf, aes(x=X2, y=cum))
}

# used for formatting labels on ggplots
fmt <- function(){ 
  function(x) format(x,nsmall = 1,scientific = FALSE)
}

plot_dendro <- function(x, title="", labels.colname=NULL, colors.colname=NULL) {
  require(ggdendro)
  meta.x <- pData(x)
  # force the metadata into character format so you don't end up with gradient/continuous color schemes for numerical variables in the final plot  
  meta.x <- as.matrix(meta.x) 
  ## do the actual statistics and put into dendrogram 
  myDist <- dist(t(exprs(x)))
  myTree <-hclust(myDist)
  dhc <- as.dendrogram(myTree)
  ddata <- dendro_data(dhc, type="rectangle")
  # the labels of the dendrogram are pulled from the Expression set exprs column names, it's nice to rename them to something more intelligible if you haven't already, as well as match them up to the metadata for label coloring
  ## check to see if the column names of the expression set match anything in the metadata, or match the rownames
  if (identical(colnames(exprs(x)), row.names(meta.x))) {
    meta.x <- row2colnames(meta.x, "rownames")
    matchcol <- "rownames"
  } else if (any(apply(meta.x, 2, function(column) identical(as.character(unlist(column)), colnames(exprs(x)))))) {
    matchcol <- names(which(apply(meta.x, 2, function(column) identical(as.character(unlist(column)), colnames(exprs(x))))))
  } else {
    print("ExpressionSet sampleNames and pData row.names or pData column must match")
    stop()
  }
  ## merge the metadata with the dendrogram labels using the commmon column/rownames you just identified above
  ddata$labels <- merge(ddata$labels, meta.x, by.x="label", by.y=matchcol)
  # plot it like you mean it
  ggplot(segment(ddata)) +
    geom_segment(aes(x=x, y=y, xend=xend, yend=yend)) +
    theme_dendro() +
    geom_text(data=label(ddata), aes_string(x='x', y='y', label=labels.colname, color=colors.colname, hjust=-0.1), size=4)+
    scale_color_brewer(type = "seq", palette = "Set1")+
    coord_flip() + scale_y_reverse(expand=c(0.2, 50)) +
    theme(axis.text.x=element_blank(),
          axis.text.y=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank()) +
    ggtitle(title)
}
```

---

# Import Data and Metadata

## Data
- load in phenotypes and array names from metadata file (covars.desc) in "metadata" directory
  - this file contains the names and descriptions of CEL files contained in the data directory 

### Create new expression set objects to normalize

```{r dataload, results='hide'}
covars <- read.table(file.path(metaDir, covarsfilename),header=TRUE, sep="\t", row.names=1) 
# Load CEL files
celFiles <- file.path(dataDir, row.names(covars))
affyRaw <- read.celfiles(celFiles)
pData(affyRaw) <- covars
sampleNames(affyRaw) <- pData(affyRaw)$sampleID
```

## Sample metadata

```{r covars, results='asis', echo=FALSE}
# Sample information table
kable(pData(affyRaw))
```

---

# PreProcessing 

## RMA Normalized Data

- background correct and normalize each dataset with RMA `r citep("10.1093/bioinformatics/19.2.185")`

- summarize probesets on the gene ('core') level

```{r normalize, results='hide'}
affyNorm.core <- rma(affyRaw, target="core", background=TRUE, normalize=TRUE)
```

### Unsupervised Clustering of RMA Normalized Data

#### Hierarchical Clustering
The goal of these analyses are to naively evaluate the variability within the raw data and determine whether this variability can predict the different treatment groups. 

```{r cluster, out.width='75%'}
plot_dendro(affyNorm.core, title="", labels.colname="sampleID", colors.colname="treatment")
```

#### Principal Component Analysis (PCA)

This second approach is a dimension reduction and visualization technique that is used to project the multivariate (i.e.multiple genes) data vector of each array into a lower-dimensional plot, such that the spatial arrangement of the points in the plot reflects the overall data (dis)similarity between the arrays. 

```{r PCAsd, out.width='75%'}
pca <- prcomp(t(exprs(affyNorm.core)))
df <- data.frame(cbind(pca$x, pData(affyNorm.core)))
  
ggplot(df) +
  geom_point(aes(x=PC1, y=PC2, color=treatment), size=6) +
  geom_text(aes(x=PC1, y=PC2, label=blooddonor, vjust=-0.5), size=5) +
    theme_bw() +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=45, hjust=1, vjust=1),
          axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.25))) +
    ylab("PC2") + xlab("PC1") +
  ggtitle('PCA for Batch 1 samples')
```

## Annotate

So far we have only been working with the probesets,without reference to the genes they assay. Here we load in metadata about the probesets on the array (feature data), the gene symbols in particular.

```{r features, results='hide'}
featureData(affyNorm.core) <- getNetAffx(affyNorm.core, "transcript") # this will load the Affymetrix annotation, including the probeID, into the fData
# get gene symbols and entrezIDs for all probesets
fData(affyNorm.core)$symbol <- as.character(unlist(mget(featureNames(affyNorm.core), hta20sttranscriptclusterSYMBOL, ifnotfound=NA))) # curated annotations from Bioconductor 
fData(affyNorm.core)$entrezID <- as.character(unlist(mget(featureNames(affyNorm.core), hta20sttranscriptclusterENTREZID, ifnotfound=NA))) # curated annotations from Bioconductor 

```

## Statistical analyses
A linear model for microarray data analysis ([Limma][http://www.bioconductor.org/packages/release/bioc/html/limma.html]) was performed on the samples to identify differentially expressed genes for the comparison of the two treatment groups. Limma fits a linear model to the expression data for all samples for each gene and is designed to handle complex experiments involving comparisons between many RNA targets simultaneously.

### Design matrix 
To perform limma, we construct a design matrix which provides a representation of the different sample groups which have been analysed. _Remember that blooddonor is a not a continous variable even though they are numeric characters!_

* make a matrix with arrays as rows, sample groups as columns
* a one or a zero indicate respectively, that a sample either belongs or does not belong to the sample group

```{r}
# Make design matrix
pData(affyNorm.core)$blooddonor <- factor(pData(affyNorm.core)$blooddonor)
design <- model.matrix(~ 0 + treatment + blooddonor  ,  data=pData(affyNorm.core))
kable(design)
```

### Linear model

These matrices are used to fit a linear model to the data. The linear model is applied and pairwise comparisons are performed to identify differentially expressed genes. The comparisons are defined based on the contrasts, which select genes that show a significant expression change between the treated samples.

- first fit the linear model based on the design matrix for each gene based on the given series of arrays
- using the contrast matrix, compute estimated coefficients and standard errors for contrasts
- compute moderated t-statistics and log-odds of differential expression by empirical Bayes shrinkage of the standard errors towards a common value

```{r limma,warning=FALSE, message=FALSE}
# Setup constrasts
contrast <- makeContrasts(treatment=treatment125D3-treatmentDMSO, levels=colnames(design))

# Fit model
fit.core <- lmFit(affyNorm.core, design) 

# Compute cofficients for constrasts
fit2core <- contrasts.fit(fit.core, contrast) 

# Bayes shrinkage
fit2core <- eBayes(fit2core) 
```

## Probe-level Results: No filtering

```{r, fig.align='center'}
results <- topTable(fit2.core, coef=1, number=nrow(exprs(affyNorm.core)))
df <- cbind(results[,c('P.Value', 'logFC')])
ggplot(df) +
  geom_histogram(aes(x=P.Value)) +
  theme_bw() +
  xlab('p-value') 
```

#### Volcano plot

This is a type of scatter-plot that is used to quickly identify changes in large datasets composed of replicate data. It plots significance versus fold-change on the y- and x-axes, respectively. The **points in purple are genes that have an FDR < 0.05.** The fold changes are higher than those seen with the full dataset.

```{r, fig.align='center'}

df <- cbind(results[,c('adj.P.Val','P.Value', 'logFC')])
df <- cbind(df, threshold=as.logical(df$adj.P.Val < 0.05)) 
ggplot(data=df, aes(x=logFC, y=-log10(P.Value), colour=threshold)) +
  scale_color_manual(values = c("grey", "purple")) +
  geom_point(alpha=0.75, pch=16, size=2) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5)),
        axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.25))) +
  xlab("log2 fold change") + ylab("-log10 p-value")
```

#### Volcano plot

```{r}
df <- cbind(df, threshold=as.logical(df$adj.P.Val < 0.05)) 
ggplot(data=df, aes(x=logFC, y=-log10(P.Value), colour=threshold)) +
  scale_color_manual(values = c("grey", "purple")) +
  geom_point(alpha=0.75, pch=16, size=2) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5)),
        axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.25))) +
  xlab("log2 fold change") + ylab("-log10 p-value")
```

## Filter Probesets
Reducing the number of genes assayed reduces the multiple test correction and may allow us to identify more differentially expressed genes.

Starting  with `r nrow(fData(affyNorm.core))` probes,  remaining we can filter:

### By Annotation
- remove the probes without annotated genes

```{r filter1}
affyNorm.filt <- affyNorm.core[which(!is.na(fData(affyNorm.core)$symbol) & fData(affyNorm.core)$category=="main"),]
affyNorm.filt.b2 <- affyNorm.core.b2[which(!is.na(fData(affyNorm.core.b2)$symbol) & fData(affyNorm.core.b2)$category=="main"),]
```

`r nrow(fData(affyNorm.filt))` probes remaining

### By Low Expression Level
- remove probes with low expression levels (bottom `r lowintensity.percentile*100`% of all expression levels) in all samples

```{r filter3, cache=TRUE}
eset.core <- exprs(affyNorm.filt)
affyNorm.filt <- affyNorm.filt[!(apply(eset.core, 1, 
                                             function(x) all(x<quantile(exprs(affyNorm.filt), 0.1)))),]
```

`r nrow(fData(affyNorm.filt))` probes remaining

### By Low Variability
- remove probes with lower variation among all samples (without regard for group status) (dropped the bottom `r mad.quantile.cutoff*100`%) 

```{r filter4}
eset.core <- exprs(affyNorm.filt)
rowmads <- apply(eset.core, 1, mad)
mad.cutoff <- as.numeric(quantile(rowmads, mad.quantile.cutoff))
affyNorm.filt <- affyNorm.filt[rowmads>mad.cutoff,]
```

`r nrow(fData(affyNorm.filt))` probes remaining

### Linear model: Post-filtering
We will apply the same model fits to the reduced data matrix, after having applied several filters.

```{r}
fit.core <- lmFit(affyNorm.filt, design) 
fit2core <- contrasts.fit(fit.core, contrast) 
fit2core <- eBayes(fit2core) 
```

### Volcano plots post-filtering

```{r}
resultsfilt <- topTable(fit2core, coef=1, number=nrow(exprs(affyNorm.filt)))

df <- cbind(resultsfilt[,c('adj.P.Val','P.Value', 'logFC')])
df <- cbind(df, threshold=as.logical(df$adj.P.Val < 0.05)) 
ggplot(data=df, aes(x=logFC, y=-log10(P.Value), colour=threshold)) +
  scale_color_manual(values = c("grey", "purple")) +
  geom_point(alpha=0.75, pch=16, size=2) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5)),
        axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.25))) +
  xlab("log2 fold change") + ylab("-log10 p-value")
```

The **results table can be downloaded using the links below**. *Note that for all these files, values are not summarized for genes assayed by multiple probes (i.e. by taking the median value), so you may see multiple instances of the same gene in the results*

```{r write-table, echo=FALSE, eval=FALSE}

stats <- topTable(fit2core, coef=1, sort.by="P",adjust.method="BH",number=nrow(exprs(affyNorm.filt)), 
                 genelist=fData(affyNorm.filt)[,c("symbol", "entrezID", "mrnaassignment")])
stats$Passes.FDR.threshold  <-  as.factor(stats$adj.P.Val<pvalue.cutoff)
eset <- exprs(affyNorm.filt)
eset  <-  eset[match(row.names(stats), row.names(eset)),]
stats.eset <- cbind(stats, eset)

write.table(stats.eset, file="results_new/allGenes_Batch1_stats_exprs_Analysis2.xls", sep="\t", quote=F, col.names=NA)
```

* [results](results_new/allGenes_Batch1_stats_exprs_Analysis2.xls)

**The summary table above contains the following information:**

- logFC is the log2-fold change
- the AveExpr is the average expression value accross all arrays
- the moderated t-statistic (t) is the logFC to its standard error, the P.Value is the associated p-value
- the adj.P.Value is the p-value adjusted for multiple testing (by FDR) 
- the B-value (B) is the log-odds that a gene is differentially expressed (the-higher-the-better)
- the last 4 columns contain the log-transformed normalized expression levels for these genes in each sample


## Aggregate probesets
For any gene that has multiple probe mappings, aggregate expression data by taking a mean across all probes.

```{r aggregate, echo=TRUE}
df <- data.frame(exprs(affyNorm.filt))
symbol <- fData(affyNorm.filt)$symbol
df <- cbind(df, symbol)

# Average by Gene Symbol
genemeans <- aggregate(. ~ symbol, data=df, mean)
row.names(genemeans) <- genemeans$symbol
genemeans <- as.matrix(genemeans[,-1])
colnames(genemeans) <- colnames(exprs(affyNorm.filt))

# Create new expression set object 
affyNorm.gene <- ExpressionSet(genemeans)
pData(affyNorm.gene) <- pData(affyNorm.filt)                                                             
```

`r nrow(exprs(affyNorm.gene))` unique genes for differential expression analysis.

### Linear model: Gene-level analysis
We will apply the same model fits to the gene-level data matrix, after having aggregated probes for a single gene mapping (by taking an average).

```{r, echo=TRUE}

# Fit model for BATCH 1
fit.core <- lmFit(affyNorm.gene, design) 
fit2core <- contrasts.fit(fit.core, contrast) 
fit2core <- eBayes(fit2core) 

# Fit model for BATCH 2
fit.core <- lmFit(affyNorm.gene.b2, design.b2) 
fit2b2.core <- contrasts.fit(fit.core, contrast.b2)# Bayes shrinkage
fit2b2.core <- eBayes(fit2b2.core) 
```

### Results

For the gene-level analysis we expect to find fewer significant findings as the probes were collapsed down. 

```{r}
# Get results
resultsgene <- topTable(fit2core, coef=1, number=nrow(exprs(affyNorm.gene)))
```

The **results table for this gene-level analysis can be downloaded using the links below**. 

* [results](./results_new/uniqueGenes_Batch1_stats_exprs_Analysis2.xls)

---

# R Session Info

(useful if replicating these results)

```{r sessioninfo}
sessionInfo()
```

---

# References

```{r writebib, results='hide', echo=FALSE, message=FALSE}
write.bibtex(file="references.bib")
```

