---
title: Batch analysis of RRBS data
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: simplex
---


```{r projsetup, echo=FALSE, warning=FALSE, message=FALSE}
project="Fedulov - RRBS Batch analysis - alternative statistics"
clientname=" "
clientemail=""
labPI=""
lablocation=""
analystname="John Hutchinson"
analystemail="jhutchin@hsph.harvard.edu"

```


```{r knitrsetup, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png", cache=TRUE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE, prompt=TRUE, comment='', fig.cap='', tidy.opts=list(keep.blank.line=FALSE, width.cutoff=200), fig.width = 16, fig.height = 14)

```


---

# Fedulov methylation

Trying out different statistical approach for individual and pooled samples in RRBS data set to see if it helps with differences in hypo/hyper methylated numbers.

Contact `r analystname` (`r analystemail`) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

---

# Setup

## Libraries

R libraries needed to perform analysis

```{r setup}
library(stringr)
library(ggplot2)
library(CHBUtils)
library(RColorBrewer)
library(pheatmap)
library(devtools)
library(devtools)
install_git("https://github.com/druvus/methylKit")
library(methylKit)
library(readr)
library(dplyr)
library(bsseq)
library(DSS)
```

## Variables
Working directories, files and other variables necessary to the analysis.

```{r load}
## Setup Data and Results directory variables
if(file.exists("/home/jnh7/projects/kobzik-rrbs")){
	baseDir="/home/jnh7/projects/kobzik-rrbs"
} else{ 
baseDir="/Users/johnhutchinson/Work/projects/kobzik-rrbs/"
}
dataDir <- file.path(baseDir, "data/bed")
metaDir <- file.path(baseDir, "meta")
resultsDir <- file.path(baseDir, "results")

lo.count.cutoff <- 10
q.cutoff <- 0.01
alpha <- 1e-5 #(for logit transforms)

numcores <- parallel::detectCores()
```

# Preparations

## Import Data and Metadata

Previous analyses identified sample 14-b1-cntr as an outlier so it was not used for this analysis.

### Metadata

Code needed to load metadata

```{r metadataload, bootstrap.show.output=TRUE}
metadata <- read.table(file.path(metaDir, "meta.txt"),sep="\t",header=T)
# drop male sample
metadata <- metadata[-grep("14-b1-cntr", metadata$samplename),]

metadata.out <- metadata[,c("samplename", "exposure", "type" ,"cohort", "generation")]
kable(metadata.out)
``` 

### Data

Code needed to load methylation data
- will also format for bsseq and aggregate strands

```{r dataload, eval=FALSE}
setwd(dataDir)
for (n in 1:length(metadata$sampleid)) {
  sampleid <- metadata$sampleid[n]
  file <- paste("methylcall.CpG", sampleid, "mincov0.txt", sep=".")
  
  dat <- read_delim(file, delim="\t") 
  dat <- mutate(dat, methylated=round(coverage*freqC/100)) %>%
    mutate(., unmethylated=coverage-methylated) %>% 
    dplyr::select(., chr, base, strand, freqC,  methylated, unmethylated, coverage) %>%
    mutate(., methylated=as.integer(methylated), unmethylated=as.integer(unmethylated)) %>%
    arrange(., chr, base, strand)
  
  # collapsing strands
  ## find how far apart the CpG base calls are
  dat <- mutate(dat, diff=c(NA,diff(base)))
  dat.df <- as.data.frame(dat)
  ## Find indices of CpG dyads for which we have both positive and negative strand information
  ### rows should be one base apart (as based on data.dyad$diff), first on positive strand and next on negative strand
  dyad.neg.strand.indices=which(dat.df$diff==1 & dat$strand=="R")
  dyad.pos.strand.indices=dyad.neg.strand.indices-1
  ## get combined coverage from positive and negative strands at dyads
  dyad.combined.numreads <- dat.df[dyad.pos.strand.indices, "coverage"]+dat.df[dyad.neg.strand.indices, "coverage"]
  dyad.combined.numCs <- dat.df[dyad.pos.strand.indices, "methylated"]+dat.df[dyad.neg.strand.indices, "methylated"]
  dyad.combined.numTs <- dat.df[dyad.pos.strand.indices, "unmethylated"]+dat.df[dyad.neg.strand.indices, "unmethylated"]
  ## recalculate percent methylation
  dyad.combined.percmeth <- dyad.combined.numCs/(dyad.combined.numCs+dyad.combined.numTs)*100
  ## load combined coverage into the positive strand position of dyad into table
  dat.df[dyad.pos.strand.indices, "coverage"] <- dyad.combined.numreads
  dat.df[dyad.pos.strand.indices, "methylated"] <- dyad.combined.numCs
  dat.df[dyad.pos.strand.indices, "unmethylated"] <- dyad.combined.numTs
  dat.df[dyad.pos.strand.indices, "freqC"] <- dyad.combined.percmeth
  ## remove negative strand rows of the combined dyads above from table
  dat.df <- dat.df[-dyad.neg.strand.indices,]
  ## find dyads that only have info for the negative strand
  neg.strand.indices <- which(dat.df$strand=="R")
  ## adjust start and end of unpaired negative strand dyads to the positive strand position
  dat.df[neg.strand.indices,"base"] <- dat.df[neg.strand.indices,"base"]-1
  dat.df[neg.strand.indices, "strand"] <- "F"
  ## No need to adjust any dyads for which you only have info from the positive strand
  dat.df$strand <- ifelse(dat.df$strand=="F", "+", "-")
  if (n==1){
    BS <- BSseq( M = as.matrix(dat.df$methylated,ncol=1), Cov = as.matrix(dat.df$coverage,ncol=1), pos = dat.df$base, chr = dat.df$chr,  sampleNames = sampleid) 
  } else {
    BS.temp <- BSseq( M = as.matrix(dat.df$methylated,ncol=1), Cov = as.matrix(dat.df$coverage,ncol=1), pos = dat.df$base, chr = dat.df$chr,  sampleNames =sampleid)
    BS <- bsseq::combine(BS, BS.temp)
  }
}

identical(sampleNames(BS),as.character(unlist(metadata$sampleid)))
pData(BS) <- metadata
save(file=file.path(resultsDir, "BSeq.object"), list="BS")
```

```{r loadobject, echo=FALSE }
load(file=file.path(resultsDir, "BSeq.object"))
```


# Analysis

## Data Consolidation

### Filter by coverage 

- removes bases with very high coverage (above 99.9th percentile) to adjust for potential PCR biases
- also remove bases with very low coverage (<10)

```{r methcoveragefilter, eval=FALSE}
meth.quants.filtered <- filterByCoverage(meth.quants, lo.count = lo.count.cutoff , lo.perc = NULL, hi.count = NULL, hi.perc = 99.9)    # methylBase object
```

### Normalize 

Normalizes coverage values between samples using a scaling factor derived from differences between median of coverage distributions


```{r methnormalization, eval=FALSE}
meth.quants.filtered.norm <- normalizeCoverage(meth.quants.filtered )
```

### Merge samples

- only compare samples at sites where they all have coverage
- reduce data to only these sites 

```{r methmerge, eval=FALSE}
meth.quants.filtered.merged <- unite(meth.quants.filtered.norm, destrand = TRUE)
rm(meth.quants, meth.quants.filtered, meth.quants.filtered.norm)
save(file=file.path(resultsDir, "meth.quants.filtered.merged"), list="meth.quants.filtered.merged")
```

```{r methmergeload, echo=FALSE}
load(file=file.path(resultsDir, "meth.quants.filtered.merged"))
```


```{r batches}
#batch 1
meth.quants.b1F1cntr <- reorganize(meth.quants.filtered.merged ,sample.ids=c("15-b1-cntr", "16-b1-cntr", "17-b1-cntr", "1-b1-F1", "2-b1-F1", "3-b1-F1", "4-b1-F1"), treatment=c(0,0,0,1,1,1,1) )
meth.quants.b1F2cntr <- reorganize(meth.quants.filtered.merged ,sample.ids=c( "15-b1-cntr", "16-b1-cntr", "17-b1-cntr",  "5-b1-F2", "6-b1-F2", "7-b1-F2", "8-b1-F2"), treatment=c(0,0,0,1,1,1,1) )
meth.quants.b1F3cntr <- reorganize(meth.quants.filtered.merged ,sample.ids=c( "15-b1-cntr", "16-b1-cntr", "17-b1-cntr",  "10-b1-F3", "11-b1-F3", "12-b1-F3", "13-b1-F3"), treatment=c(0,0,0,1,1,1,1) )

#batch 2
meth.quants.b2F1cntr <- reorganize(meth.quants.filtered.merged  ,sample.ids=c("123-b2-cntr","124-b2-cntr", "125-b2-cntr", "126-b2-cntr", "60-b2-F1", "61-b2-F1", "62-b2-F1", "63-b2-F1"), treatment=c(0,0,0,0,1,1,1,1) )
meth.quants.b2F2cntr <- reorganize(meth.quants.filtered.merged  ,sample.ids=c("123-b2-cntr","124-b2-cntr", "125-b2-cntr", "126-b2-cntr", "90-b2-F2", "91-b2-F2", "92-b2-F2", "93-b2-F2"), treatment=c(0,0,0,0,1,1,1,1) )
meth.quants.b2F3cntr <- reorganize(meth.quants.filtered.merged  ,sample.ids=c("123-b2-cntr","124-b2-cntr", "125-b2-cntr", "126-b2-cntr", "119-b2-F3", "120-b2-F3", "121-b2-F3", "122-b2-F3"), treatment=c(0,0,0,0,1,1,1,1) )
```

## Differential methylation                           

### Logistic regression test

Significance wwas calculated by applying Logistic regression tests.
  
```{r logreg, eval=FALSE}
#batch1
b1F1.diffMeth <- calculateDiffMeth(meth.quants.b1F1cntr,slim=TRUE,  weighted.mean=TRUE,num.cores=numcores)
b1F2.diffMeth <- calculateDiffMeth(meth.quants.b1F2cntr,slim=TRUE,  weighted.mean=TRUE,num.cores=numcores)
b1F3.diffMeth <- calculateDiffMeth(meth.quants.b1F3cntr,slim=TRUE,  weighted.mean=TRUE,num.cores=numcores)

#batch2
b2F1.diffMeth <- calculateDiffMeth(meth.quants.b2F1cntr,slim=TRUE,  weighted.mean=TRUE,num.cores=numcores)
b2F2.diffMeth <- calculateDiffMeth(meth.quants.b2F2cntr,slim=TRUE,  weighted.mean=TRUE,num.cores=numcores)
b2F3.diffMeth <- calculateDiffMeth(meth.quants.b2F3cntr,slim=TRUE,  weighted.mean=TRUE,num.cores=numcores)

save(file=file.path(resultsDir, "batch.diffs"), list=ls()[grep("b.+diffMeth", ls())])
```

```{r loaddiffs, echo=FALSE}
load(file=file.path(resultsDir, "batch.diffs"))
```

### Data Exploration - why are the batches so different?

## Data Exploration

#### Principal Component Analysis

__All samples__

Overview of all included samples

The PCA plot does not show good separation of the two batches, soemthing you would want to see if you hoped to correct the batch effect.  RRBS methylation data is measured here in beta values (i.e. percent methylation). Unfortunately, these values are strongly heteroscedastic (i.e. the variance varies with the beta value, and is typically higher for the extreme values); PCA plots depend on variance so when using beta-values they can be dominated by a subset of measurements and not properly capture the variation. One way of addressing this heteroskedasticity is to logit transform the beta values into M values (log1(methylated/unmethylated) and run the PCA on these values. 

```{r PCA.dimension.plot.all, fig.cap="Plot of First Two Principals Components", fig.width = 8, fig.height = 8}
meth.pr.beta <- prepPCA(meth.quants.filtered.merged)
methylPCA(meth.pr.beta, meth.prext.metadata = metadata, metadata.merge.col = "samplename",  comp1=1, comp2=2, title="Beta values")
meth.pr.logit <- prepPCA(meth.quants.filtered.merged, logit=TRUE)
methylPCA(meth.pr.logit, meth.prext.metadata = metadata, metadata.merge.col = "samplename",  comp1=1, comp2=2, title="Logit values")
```

Using the logit transformed values, you can see that the samples separate better by cohort. This may have implications for which statistical approach to use for differential methylation detection.


### Comparing sites with discordant differential methhylation between individual and pooled samples

```{r }
# for F1 comparison
b1F1.diffMeth.mat <- getData(b1F1.diffMeth)
b1F1.diffMeth.mat$batch="one" 
b2F1.diffMeth.mat <- getData(b2F1.diffMeth)
b2F1.diffMeth.mat$batch="two"

F1.diffMeth.mat <- rbind(b1F1.diffMeth.mat, b2F1.diffMeth.mat)
F1.diffMeth.dt <- tbl_df(F1.diffMeth.mat)
```



```{r methdiffcompare_wholedist}

ggplot(F1.diffMeth.dt, aes(x=meth.diff, colour=batch)) + geom_density()+ggtitle("F1 methylation  differences  ")

# first look at the methylation distributions differences
filter(F1.diffMeth.dt, meth.diff>0) %>% ggplot(., aes(x=meth.diff, colour=batch)) + geom_density()+scale_x_log10()+ ggtitle("Methylation differences at 'over'-methylated sites") +xlab("log10 transformed hyper-methylation differences")

filter(F1.diffMeth.dt, meth.diff<0) %>% ggplot(., aes(x=abs(meth.diff), colour=batch)) + geom_density()+scale_x_log10()+ ggtitle("Methylation differences at 'under'-methylated sites")+xlab("log10 transformed absolute values of hypo-methylation differences")
```

It's clear from these plots that batch 1 (individual samples) shows more hypomethylation and batch 2 more  hypermethylation. (this largely repeats the results of the previous analyses, but doesn't involve cutoffs, so is a more robust way to look at the differences). That said, they both appear to be biased towards hypermethylation, but just vary in the degree.

These trends are even more apparent if you subset to only the significant hits. 

```{r methdiffcompare_bothsigdiffdist}
filter(F1.diffMeth.dt, qvalue<0.01) %>% ggplot(., aes(x=meth.diff, colour=batch)) + geom_density()+ggtitle("F1 methylation significant differences  ")

# first look at the methylation distributions differences
filter(F1.diffMeth.dt, meth.diff>0, qvalue<0.05) %>% ggplot(., aes(x=meth.diff, colour=batch)) + geom_density()+scale_x_log10()+ ggtitle("F1 methylation differences at significantly 'over'-methylated sites") +xlab("log10 transformed hyper-methylation differences")

filter(F1.diffMeth.dt, meth.diff<0, qvalue<0.05) %>% ggplot(., aes(x=abs(meth.diff), colour=batch)) + geom_density()+scale_x_log10()+ ggtitle("F1 methylation differences at significantly 'under'-methylated sites")+xlab("log10 transformed absolute values of hypo-methylation differences")
```

What does the counts data look like for sites that are hypermethylated/hypomethylated in batch 2 but not batch 1? (and vice versa)

```{r methdiffcompare_onesigdist}
filter(F1.diffMeth.dt, qvalue<0.01, meth.diff>25, batch=="two") %>% 
  dplyr::select(., chr, start, end, strand) %>% 
  dplyr::semi_join(F1.diffMeth.dt, ., by = c("chr", "start", "end")) %>% 
  ggplot(., aes(x=meth.diff, colour=batch)) + geom_density()+ ggtitle("Methylation differences at sites significant hypermethylated in batch2 F1")+xlab("methylation differences")

filter(F1.diffMeth.dt, qvalue<0.01, meth.diff<(-25), batch=="two") %>% 
  dplyr::select(., chr, start, end, strand) %>% 
  dplyr::semi_join(F1.diffMeth.dt, ., by = c("chr", "start", "end"))%>% 
  ggplot(., aes(x=meth.diff, colour=batch)) + geom_density()+ ggtitle("Distributions of methylation differences at sites hypomethylated in batch2 F1")+xlab("methylation differences")

filter(F1.diffMeth.dt, qvalue<0.01, meth.diff>25, batch=="one") %>% 
  dplyr::select(., chr, start, end, strand) %>% 
  dplyr::semi_join(F1.diffMeth.dt, ., by = c("chr", "start", "end")) %>% 
  ggplot(., aes(x=meth.diff, colour=batch)) + geom_density()+ ggtitle("Methylation differences at sites significant hypermethylated in batch1 F1")+xlab("methylation differences")

filter(F1.diffMeth.dt, qvalue<0.01, meth.diff<(-25), batch=="one") %>% 
  dplyr::select(., chr, start, end, strand) %>% 
  dplyr::semi_join(F1.diffMeth.dt, ., by = c("chr", "start", "end"))%>% 
  ggplot(., aes(x=meth.diff, colour=batch)) + geom_density()+ ggtitle("Distributions of methylation differences at sites hypomethylated in batch1 F1")+xlab("methylation differences")
```

```{r methdiffcompare_rawdatasetsup, eval=FALSE}
sampleids <- meth.quants.filtered.merged@sample.ids
meth.quant.mat <- getData(meth.quants.filtered.merged)
meth.quant.mat.m <- reshape2::melt(meth.quant.mat, id.vars=c("chr", "start", "end", "strand")) %>% tbl_df()
rm(meth.quant.mat)
meth.quant.mat.m <- mutate(meth.quant.mat.m, samplenum=unlist(str_extract_all(as.character(meth.quant.mat.m$variable), "(\\d)+$")))
meth.quant.mat.m <- mutate(meth.quant.mat.m, variable=str_replace_all(variable, "(\\d)+$", "" ))
meth.quant.mat.m <- mutate(meth.quant.mat.m, samplid=sampleids[as.numeric(samplenum)])
meth.quant.mat.m <- mutate(meth.quant.mat.m, batch=unlist(lapply(str_split(samplid, "-"), function(x) x[[2]])))
meth.quant.mat.m <- mutate(meth.quant.mat.m, generation=unlist(lapply(str_split(samplid, "-"), function(x) x[[3]])))


numCs.m <- filter(meth.quant.mat.m, variable=="numCs")
numTs.m <- filter(meth.quant.mat.m, variable=="numTs")

meth.quant.mat.betas.m <- inner_join(numCs.m, numTs.m, by=c("chr", "start", "end", "strand", "samplid", "samplenum", "batch", "generation"))
meth.quant.mat.betas.m <-  mutate(meth.quant.mat.betas.m, betaval=value.x/(value.x+value.y)) 
meth.quant.mat.betas.m <-  mutate(meth.quant.mat.betas.m, Mval=log2((value.x+1e-5)/(value.y+alpha))) 
meth.quant.mat.betas.m <- dplyr::select(meth.quant.mat.betas.m, -contains("variable"), -contains("value")) 
meth.quant.mat.betas.m <- reshape2::melt(meth.quant.mat.betas.m, id.vars=c("chr", "start", "end", "strand","samplid", "samplenum", "batch", "generation")) 
meth.quant.mat.betas.m <- tbl_df(meth.quant.mat.betas.m) 
meth.quant.mat.betas.m <- dplyr::select(meth.quant.mat.betas.m, chr, start, end, strand, variable, value, samplenum, samplid, batch, generation)

meth.quant.mat.m <- bind_rows(meth.quant.mat.betas.m, meth.quant.mat.m)

#save(file=file.path(resultsDir, "melted.methylkit.data"), list=c("meth.quant.mat.m"))
```

```{r methdiffcompare_rawdataplots}
ggplot(meth.quant.mat.m, aes(x=value, color=batch))+geom_density()+facet_wrap(~variable)+scale_x_log10()+ggtitle("all samples, all sites")
ggplot(filter(meth.quant.mat.m, generation=="cntr"), aes(x=value, color=batch))+geom_density()+facet_wrap(~variable)+scale_x_log10()+ggtitle("control samples, all sites")
ggplot(filter(meth.quant.mat.m, generation=="F1"), aes(x=value, color=batch))+geom_density()+facet_wrap(~variable)+scale_x_log10()+ggtitle("F1 samples, all sites")

filter(F1.diffMeth.dt, qvalue<0.01, meth.diff>25, batch=="two") %>% 
  dplyr::select(., chr, start, end, strand) %>% 
  dplyr::semi_join(meth.quant.mat.m, ., by = c("chr", "start", "end")) %>%
ggplot(., aes(x=value, color=batch, ))+geom_density()+facet_wrap(~variable+generation,ncol=4)+scale_x_log10()+ggtitle("batch2 hypermethylated sites")  

filter(F1.diffMeth.dt, qvalue<0.01, meth.diff<(-25), batch=="two") %>% 
  dplyr::select(., chr, start, end, strand) %>% 
  dplyr::semi_join(meth.quant.mat.m, ., by = c("chr", "start", "end")) %>%
ggplot(., aes(x=value, color=batch))+geom_density()+facet_wrap(~variable+generation, ncol=4)+scale_x_log10()+ggtitle("batch2 hypomethylated sites")  


filter(F1.diffMeth.dt, qvalue<0.01, meth.diff>(25), batch=="one") %>% 
  dplyr::select(., chr, start, end, strand) %>% 
  dplyr::semi_join(meth.quant.mat.m, ., by = c("chr", "start", "end")) %>%
ggplot(., aes(x=value, color=batch))+geom_density()+facet_wrap(~variable+generation, ncol=4)+scale_x_log10()+ggtitle("batch1 hyperethylated sites")  


filter(F1.diffMeth.dt, qvalue<0.01, meth.diff<(-25), batch=="one") %>% 
  dplyr::select(., chr, start, end, strand) %>% 
  dplyr::semi_join(meth.quant.mat.m, ., by = c("chr", "start", "end")) %>%
ggplot(., aes(x=value, color=batch))+geom_density()+facet_wrap(~variable+generation, ncol=4)+scale_x_log10()+ggtitle("batch1 hypomethylated sites")  

```

# R Session Info

(useful if replicating these results)

```{r sessioninfo}
sessionInfo()
```

---

