---
title: Batch analysis of RRBS data
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: simplex
---


```{r projsetup, echo=FALSE, warning=FALSE, message=FALSE}
project="Fedulov - RRBS Batch analysis - alternative statistics"
clientname=" "
clientemail=""
labPI=""
lablocation=""
analystname="John Hutchinson"
analystemail="jhutchin@hsph.harvard.edu"

```


```{r knitrsetup, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png", cache=TRUE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE, prompt=TRUE, comment='', fig.cap='', tidy.opts=list(keep.blank.line=FALSE, width.cutoff=200), fig.width = 16, fig.height = 14)

```


---

# Fedulov methylation

Trying out different statistical approach for individual and pooled samples in RRBS data set to see if it helps with differences in hypo/hyper methylated numbers.

Contact `r analystname` (`r analystemail`) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

---

# Setup

## Libraries

R libraries needed to perform analysis

```{r setup}
library(stringr)
library(ggplot2)
library(CHBUtils)
library(RColorBrewer)
library(pheatmap)
library(devtools)
library(readr)
library(dplyr)
library(bsseq)
library(DSS)
```

## Variables
Working directories, files and other variables necessary to the analysis.

```{r load}
## Setup Data and Results directory variables
if(file.exists("/home/jnh7/projects/kobzik-rrbs")){
	baseDir="/home/jnh7/projects/kobzik-rrbs"
} else{ 
baseDir="/Users/johnhutchinson/Work/projects/kobzik-rrbs/"
}
dataDir <- file.path(baseDir, "data/bed")
metaDir <- file.path(baseDir, "meta")
resultsDir <- file.path(baseDir, "results")

lo.count.cutoff <- 10
q.cutoff <- 0.01
alpha <- 1e-5 #(for logit transforms)

numcores <- parallel::detectCores()
```

# Preparations

## Import Data and Metadata

Previous analyses identified sample 14-b1-cntr as an outlier so it was not used for this analysis.

### Metadata

Code needed to load metadata

```{r metadataload, bootstrap.show.output=TRUE}
metadata <- read.table(file.path(metaDir, "meta.txt"),sep="\t",header=T)
# drop male sample
metadata <- metadata[-grep("14-b1-cntr", metadata$samplename),]

metadata.out <- metadata[,c("samplename", "exposure", "type" ,"cohort", "generation")]
kable(metadata.out)
``` 

### Data

Code needed to load methylation data
- will also format for bsseq and aggregate strands

```{r dataload, eval=FALSE}
setwd(dataDir)
for (n in 1:length(metadata$sampleid)) {
  sampleid <- metadata$sampleid[n]
  file <- paste("methylcall.CpG", sampleid, "mincov0.txt", sep=".")
  
  dat <- read_delim(file, delim="\t") 
  dat <- mutate(dat, methylated=round(coverage*freqC/100)) %>%
    mutate(., unmethylated=coverage-methylated) %>% 
    dplyr::select(., chr, base, strand, freqC,  methylated, unmethylated, coverage) %>%
    mutate(., methylated=as.integer(methylated), unmethylated=as.integer(unmethylated)) %>%
    arrange(., chr, base, strand)
  
  # collapsing strands
  ## find how far apart the CpG base calls are
  dat <- mutate(dat, diff=c(NA,diff(base)))
  dat.df <- as.data.frame(dat)
  ## Find indices of CpG dyads for which we have both positive and negative strand information
  ### rows should be one base apart (as based on data.dyad$diff), first on positive strand and next on negative strand
  dyad.neg.strand.indices=which(dat.df$diff==1 & dat$strand=="R")
  dyad.pos.strand.indices=dyad.neg.strand.indices-1
  ## get combined coverage from positive and negative strands at dyads
  dyad.combined.numreads <- dat.df[dyad.pos.strand.indices, "coverage"]+dat.df[dyad.neg.strand.indices, "coverage"]
  dyad.combined.numCs <- dat.df[dyad.pos.strand.indices, "methylated"]+dat.df[dyad.neg.strand.indices, "methylated"]
  dyad.combined.numTs <- dat.df[dyad.pos.strand.indices, "unmethylated"]+dat.df[dyad.neg.strand.indices, "unmethylated"]
  ## recalculate percent methylation
  dyad.combined.percmeth <- dyad.combined.numCs/(dyad.combined.numCs+dyad.combined.numTs)*100
  ## load combined coverage into the positive strand position of dyad into table
  dat.df[dyad.pos.strand.indices, "coverage"] <- dyad.combined.numreads
  dat.df[dyad.pos.strand.indices, "methylated"] <- dyad.combined.numCs
  dat.df[dyad.pos.strand.indices, "unmethylated"] <- dyad.combined.numTs
  dat.df[dyad.pos.strand.indices, "freqC"] <- dyad.combined.percmeth
  ## remove negative strand rows of the combined dyads above from table
  dat.df <- dat.df[-dyad.neg.strand.indices,]
  ## find dyads that only have info for the negative strand
  neg.strand.indices <- which(dat.df$strand=="R")
  ## adjust start and end of unpaired negative strand dyads to the positive strand position
  dat.df[neg.strand.indices,"base"] <- dat.df[neg.strand.indices,"base"]-1
  dat.df[neg.strand.indices, "strand"] <- "F"
  ## No need to adjust any dyads for which you only have info from the positive strand
  dat.df$strand <- ifelse(dat.df$strand=="F", "+", "-")
  if (n==1){
    BS <- BSseq( M = as.matrix(dat.df$methylated,ncol=1), Cov = as.matrix(dat.df$coverage,ncol=1), pos = dat.df$base, chr = dat.df$chr,  sampleNames = sampleid) 
  } else {
    BS.temp <- BSseq( M = as.matrix(dat.df$methylated,ncol=1), Cov = as.matrix(dat.df$coverage,ncol=1), pos = dat.df$base, chr = dat.df$chr,  sampleNames =sampleid)
    BS <- bsseq::combine(BS, BS.temp)
  }
}

identical(sampleNames(BS),as.character(unlist(metadata$sampleid)))
pData(BS) <- metadata
sampleNames(BS) <- pData(BS)$sampleid
save(file=file.path(resultsDir, "BSeq.object"), list="BS")
```

```{r loadobject, echo=FALSE }
load(file=file.path(resultsDir, "BSeq.object"))
```


# Analysis

## Separated by batch
- just F1 for now

```{r separatebybatch}
BS.b1 <- BS[,which(pData(BS)$cohort==1 & (pData(BS)$generation=="F1" | pData(BS)$generation=="cntr"))]
BS.b2 <- BS[,which(pData(BS)$cohort==2 & (pData(BS)$generation=="F1" | pData(BS)$generation=="cntr"))]
```

### Statistical Tests
- using DSS based Wald test

```{r DSStwogroup}
#use pData to subset and get samplenames 
dmlTest.b1 <- DMLtest(BS.b1, group1=as.character(pData(BS.b1)$sampleid[which(pData(BS.b1)$generation=="F1")]),group2=as.character(pData(BS.b1)$sampleid[which(pData(BS.b1)$generation=="cntr")]))
dmlTest.b2 <- DMLtest(BS.b2, group1=as.character(pData(BS.b2)$sampleid[which(pData(BS.b2)$generation=="F1")]), group2=as.character(pData(BS.b2)$sampleid[which(pData(BS.b2)$generation=="cntr")]))

diffs.b1 <- callDML(dmlTest.b1, p.threshold = 0.05, delta=0.1)
diffs.b1 <- subset(diffs.b1, fdr<0.01)
diffs.b2 <- callDML(dmlTest.b2, p.threshold = 0.05, delta=0.1)
diffs.b2 <- subset(diffs.b2, fdr<0.01)

callDMR(dmlTest.b1, )
```

```{r hypohypercompare}
hypo.b1 <- subset(diffs.b1, diff<0)
hyper.b1 <- subset(diffs.b1, diff>0)
hypo.b2 <- subset(diffs.b2, diff<0)
hyper.b2 <- subset(diffs.b2, diff>0)

diffs.b1.gr <- with(diffs.b1, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b2.gr <- with(diffs.b2, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))

length(diffs.b1.gr)
length(diffs.b2.gr)
length(intersect(diffs.b1.gr, diffs.b2.gr))
```


```{r allpairwise}
f1s.b1 <- as.character(pData(BS.b1)$sampleid[which(pData(BS.b1)$generation=="F1")])
cntrs.b1 <- as.character(pData(BS.b1)$sampleid[which(pData(BS.b1)$generation=="cntr")])
combos.b1 <- expand.grid(cntrs.b1, f1s.b1)

f1s.b2 <- as.character(pData(BS.b2)$sampleid[which(pData(BS.b2)$generation=="F1")])
cntrs.b2 <- as.character(pData(BS.b2)$sampleid[which(pData(BS.b2)$generation=="cntr")])
combos.b2 <- expand.grid(cntrs.b2, f1s.b2)

dmlTest.b1.15.1 <- DMLtest(BS.b1, group1="15", group2="1", smoothing=TRUE)
dmlTest.b1.16.1 <- DMLtest(BS.b1, group1="16", group2="1", smoothing=TRUE)
dmlTest.b1.17.1 <- DMLtest(BS.b1, group1="17", group2="1", smoothing=TRUE)
dmlTest.b1.15.2 <- DMLtest(BS.b1, group1="15", group2="2", smoothing=TRUE)
dmlTest.b1.16.2 <- DMLtest(BS.b1, group1="16", group2="2", smoothing=TRUE)
dmlTest.b1.17.2 <- DMLtest(BS.b1, group1="17", group2="2", smoothing=TRUE)
dmlTest.b1.15.3 <- DMLtest(BS.b1, group1="15", group2="3", smoothing=TRUE)
dmlTest.b1.16.3 <- DMLtest(BS.b1, group1="16", group2="3", smoothing=TRUE)
dmlTest.b1.17.3 <- DMLtest(BS.b1, group1="17", group2="3", smoothing=TRUE)
dmlTest.b1.15.4 <- DMLtest(BS.b1, group1="15", group2="4", smoothing=TRUE)
dmlTest.b1.16.4 <- DMLtest(BS.b1, group1="16", group2="4", smoothing=TRUE)
dmlTest.b1.17.4 <- DMLtest(BS.b1, group1="17", group2="4", smoothing=TRUE)

diffs.b1.15.1 <- callDML(dmlTest.b1.15.1, p.threshold = 0.05, delta=0.1)
diffs.b1.16.1 <- callDML(dmlTest.b1.16.1, p.threshold = 0.05, delta=0.1)
diffs.b1.17.1 <- callDML(dmlTest.b1.17.1, p.threshold = 0.05, delta=0.1)
diffs.b1.15.2 <- callDML(dmlTest.b1.15.2, p.threshold = 0.05, delta=0.1)
diffs.b1.16.2 <- callDML(dmlTest.b1.16.2, p.threshold = 0.05, delta=0.1)
diffs.b1.17.2 <- callDML(dmlTest.b1.17.2, p.threshold = 0.05, delta=0.1)
diffs.b1.15.3 <- callDML(dmlTest.b1.15.3, p.threshold = 0.05, delta=0.1)
diffs.b1.16.3 <- callDML(dmlTest.b1.16.3, p.threshold = 0.05, delta=0.1)
diffs.b1.17.3 <- callDML(dmlTest.b1.17.3, p.threshold = 0.05, delta=0.1)
diffs.b1.15.4 <- callDML(dmlTest.b1.15.4, p.threshold = 0.05, delta=0.1)
diffs.b1.16.4 <- callDML(dmlTest.b1.16.4, p.threshold = 0.05, delta=0.1)
diffs.b1.17.4 <- callDML(dmlTest.b1.17.4, p.threshold = 0.05, delta=0.1)

diffs.b1.15.1.gr <- with(diffs.b1.15.1, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b1.16.1.gr <- with(diffs.b1.16.1, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b1.17.1.gr <- with(diffs.b1.17.1, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b1.15.2.gr <- with(diffs.b1.15.2, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b1.16.2.gr <- with(diffs.b1.16.2, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b1.17.2.gr <- with(diffs.b1.17.2, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b1.15.3.gr <- with(diffs.b1.15.3, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b1.16.3.gr <- with(diffs.b1.16.3, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b1.17.3.gr <- with(diffs.b1.17.3, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b1.15.4.gr <- with(diffs.b1.15.4, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b1.16.4.gr <- with(diffs.b1.16.4, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))
diffs.b1.17.4.gr <- with(diffs.b1.17.4, GRanges(chr, IRanges(pos, pos), mu1=mu1, mu2=mu2, diff=diff, diff.se=diff.se, pval=pval, fdr=fdr))

length(diffs.b1.15.1.gr)
length(diffs.b1.16.1.gr)
length(diffs.b1.17.1.gr)
length(diffs.b1.15.2.gr)
length(diffs.b1.16.2.gr)
length(diffs.b1.17.2.gr)

length(diffs.b1.15.3.gr)
length(diffs.b1.16.3.gr)
length(diffs.b1.17.3.gr)
length(diffs.b1.15.4.gr)
length(diffs.b1.16.4.gr)
length(diffs.b1.17.4.gr)

length(intersect(diffs.b1.15.4.gr, diffs.b1.16.4.gr))
intersect(diffs.b1.15.1.gr, intersect(diffs.b1.16.1.gr, diffs.b1.17.1.gr))
intersect(diffs.b1.15.2.gr, intersect(diffs.b1.16.2.gr, diffs.b1.17.2.gr))
intersect(diffs.b1.15.3.gr, intersect(diffs.b1.16.3.gr, diffs.b1.17.3.gr))
intersect(diffs.b1.15.4.gr, intersect(diffs.b1.16.4.gr, diffs.b1.17.4.gr))


length(intersect(diffs.b1.1.gr, diffs.b1.3.gr))
length(intersect(diffs.b1.1.gr, diffs.b1.4.gr))
length(intersect(diffs.b1.2.gr, diffs.b1.3.gr))
length(intersect(diffs.b1.2.gr, diffs.b1.4.gr))
length(intersect(diffs.b1.3.gr, diffs.b1.4.gr))



```

# R Session Info

(useful if replicating these results)

```{r sessioninfo}
sessionInfo()
```

---

