##TODO
- rewrite compare function
- fix heatmap output to fit page 
- fix pathprint code
- do you need to specify plyr or can knitr pull from .Rprofile?

##General notes

- Store and run this script in the base directory of the consult 
- The script requires all .cel files to be in the same directory as the covars.desc (metadata) file
- An example covars.desc file might have columns with the original sample name ("Sample"), the new sample name ("Sample_new"), the group to which the sample belongs ("Group") and the Batch number in separate columns. The first column must have the .cel file names 

```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, cache=TRUE, highlight=TRUE, figalign="center", message=FALSE, warning=FALSE, fig.height=11, fig.width=11)
```

# R PACKAGES

```{r libraries, echo=TRUE, message=FALSE, warnings=FALSE}
library(affy)
library(arrayQualityMetrics)
library(RColorBrewer)
library(simpleaffy)
library(limma)
library(mouse430a2.db)
library(pheatmap)
library(pathprint)
library(GMAfunctions)
library(ape)
```

# VARIABLES
- get base directory for analyses
- specify data and results directories
- specify column headers used in covars.desc

```{r directories, echo=FALSE}
baseDir=getwd()
dataDir=paste(baseDir, "/data", sep="")
resultsDir=paste(baseDir, "/results", sep="")
sampleid_column_header="Sample_new"
groupid_column_header="Group"
```

# DATALOAD
## load in phenotype Data and CEL files
  
```{r loadcels}
mic.data <- read.affy('covars.desc', path=dataDir, verbose=T)
```

# ## QC on raw data 
# 
# ```{r QC_raw, echo=FALSE}
# arrayQualityMetrics(expressionset=mic.data,
#                     outdir='./results/report_raw',
#                     force=TRUE,
#                     do.logtransform=TRUE)
# ```
# 
# [Raw QC](./results/report_raw/index.html)

# Background Correct and Normalize
- using RMA normalization

```{r normalize, echo=FALSE}
mic.edata <- call.exprs(mic.data, "rma")
```

# ## QC on normalized data 
# 
# ```{r QC_RMA, echo=FALSE}
# arrayQualityMetrics(expressionset=mic.data,
#                     outdir='./results/report_RMA',
#                     force=TRUE,
#                     do.logtransform=TRUE)
# ```
# 
# [Normalized QC](./results/report_RMA/index.html)

# ANALYSIS
## Primary Clustering 
- are the samples clustering by sample type?

```{r expressionvalues, echo=FALSE}
### Extract expression values
###from normalized "Affybatch" object into a separate matrix
allData <- exprs(mic.edata) 
colnames(allData) <- pData(mic.edata)[,sampleid_column_header]
```

### Calculate the PCA

```{r calculate_PCA, echo=FALSE}
myPca <- prcomp(t(allData))
```

### Plot the SD of components
- this plot gives you an idea of how well a given number of primary components (plotted here as PCA on the x-axis) explains the variability (plotted here as stddev on the y-axis) in the data  

```{r plot_PCA_SDs, echo=FALSE}
plot(myPca$sdev, xlab="PCA", ylab="sddev")
```

### Plot all pairwise combinations of the first 5 primary components for the samples
- the more similar the samples are, the closer they will cluster in these plots
- sample groups are coded by color and letter

```{r plot_PCA, tidy=TRUE, echo=FALSE}
tmpPCAData <- as.data.frame(myPca$x[,1:5])
colors <- rainbow(length(unique(pData(mic.edata)$Group)))[factor(pData(mic.edata)$Group)]
plot(tmpPCAData, col=colors, pch=row.names(myPca$x))
```


# Pathprint
- how do these samples differ in terms of pathways?
- can we find any cell types in the GEO datasets that match subset of these samples well? 

### specify pathprint settings
- platform, species and sample names

```{r pathprint_settings, echo=FALSE}
mic.data.platform <- "GPL8321"
mic.data.species <- "mouse"
mic.data.names <- pData(mic.edata)$Sample_new
eset <- exprs(mic.edata)
dimnames(eset)[[2]] <- mic.data.names
````

### fingerprint all samples
```{r pathprint_exprs, echo=FALSE}
mic.data.fingerprint <- exprs2fingerprint(exprs = eset, platform = mic.data.platform, species = mic.data.species, progressBar = FALSE)
```

## HEATMAPs
- heatmap of top <100 pathways that vary the most across all samples

```{r pathprint_heatmap, echo=FALSE}
n=0.1
while (length(which(apply(mic.data.fingerprint, 1, sd) > n))>100) {
  n=n+0.1
  } 
pheatmap(mic.data.fingerprint[apply(mic.data.fingerprint, 1, sd) > n, ], labCol = mic.data.names, mar = c(10,10), col = c("blue", "white", "red"))
```


# ### Find the subset of varying pathways which are consistent within individual sample groups and make heatmap
# 
# ```{r}
# groups <- unique(pData(mic.edata)$Group)
# # get indices of samples within each group
# indices <- sapply(groups, function(group) {
#   grep(paste(group, "$", sep=""), pData(mic.edata)$Group)
# })
# agree.intra.grp <- mic.data.fingerprint[apply(mic.data.fingerprint, 1, function(x) {
#   x=as.vector(unlist(x))
#   all(as.vector(unlist(lapply(indices, function(y) {
#       y=as.vector(unlist(y))
#       compare(x[y])
#   }))))
# }),]
# pheatmap(agree.intra.grp[apply(agree.intra.grp, 1, sd) > 0.05, ], labCol = mic.data.names, mar = c(10,10), col = c("blue", "white", "red"), cellheight=10, cellwidth=20)
# ```
# 
# ##Find similar cell types within the GEOdatasets
# 
# ### first make consensus fingerprints for each sample group
# 
# ```{r consensus_fingerprint}
# mic.data.consensus.fingerprint <- t(ldply(indices, function(y) {
#   if (length(y)>1){
#     consensusFingerprint(mic.data.fingerprint[,y], threshold=0.9)
#   }
#   else {
#    mic.data.fingerprint[,y]
#   }
# }))
# dimnames(mic.data.consensus.fingerprint)[[2]] <- as.vector(unlist(mic.data.consensus.fingerprint[1,]))
# mic.data.consensus.fingerprint <- mic.data.consensus.fingerprint[-1,]
# class(mic.data.consensus.fingerprint) <- "numeric"
# ```
# 
# ###get similar geodata for each sample group
# 
# ```{r similar_geo}
# similar.GEO.data <- lapply(indices, function(y) {
#   if (length(y)>1){
#     consensus=consensusFingerprint(mic.data.fingerprint[,y], threshold=0.9)
#     geo.consensus.distance=consensusDistance(consensus, GEO.fingerprint.matrix)
#     table(GEO.metadata.matrix[match(head(row.names(geo.consensus.distance),50), GEO.metadata.matrix$GSM),"Source"])
#   }
#   else {
#     geo.consensus.distance=consensusDistance(mic.data.fingerprint[,y], GEO.fingerprint.matrix)
#     table(GEO.metadata.matrix[match(head(row.names(geo.consensus.distance),50), GEO.metadata.matrix$GSM),"Source"])
#   }
# })
# ```
# 
# 
# ##Tree building and informative pathways for nodes
# 
# ###find node pathways, output heatmap of their values  
# ```{r treebuilding}
# mic.data.dist <- dist(t(mic.data.fingerprint))
# mic.data.nj <- nj(mic.data.dist)
# mic.data.nj$node.label = letters[1:mic.data.nj$Nnode]
# plot(mic.data.nj, show.node.label = TRUE)
# mic.data.node.pathways <- nodePathways(tree = mic.data.nj, matrix = mic.data.fingerprint, threshold = 1.25)
# heatmap(mic.data.fingerprint[ mic.data.node.pathways[["InformPathways"]]$b, unlist(mic.data.node.pathways[["BranchTipNames"]]$a) ], scale = "none", col = c("blue", "white", "red"), mar = c(5,15), cexRow = 0.8, cexCol = 1)
# ```
# 
# ```{r treebuilding_nodes}
# ##tree building and informative pathways for nodes, using consensus fingerprints
# mic.data.dist <- dist(t(mic.data.consensus.fingerprint))
# mic.data.nj <- nj(mic.data.dist)
# mic.data.nj$node.label = letters[1:mic.data.nj$Nnode]
# plot(mic.data.nj, show.node.label = TRUE)
# mic.data.node.pathways <- nodePathways(tree = mic.data.nj, matrix = mic.data.consensus.fingerprint, threshold = 1.2)
# heatmap(mic.data.consensus.fingerprint[ mic.data.node.pathways[["InformPathways"]]$d, unlist(mic.data.node.pathways[["BranchTipNames"]]$a) ], scale = "none", col = c("blue", "white", "red"), mar = c(5,15), cexRow = 0.8, cexCol = 1)
# ```
# 
# 
# 
# ######
# #looking at 3 genes in topaz (top ) and mCherry (mCH) cells
# sub.mic.edata <- mic.edata
# exprs(sub.mic.edata) <- data.sub <- allData[,1:7]
# pData(sub.mic.edata) <- pData(mic.edata)[1:7,]
# 
# # limma
# design.pairs <- function(levels) {
#   n <- length(levels)
#   design <- matrix(0, n, choose(n, 2))
#   rownames(design) <- levels
#   colnames(design) <- 1:choose(n, 2)
#   k <- 0
#   for (i in 1:(n - 1))
#     for (j in (i + 1):n) {
#       k <- k+1
#       design[i, k] <- 1
#       design[j, k] <- -1
#       colnames(design)[k] <- paste(levels[i], "-", levels[j], sep="")
#     }
#   design
# }
# 
# # Create appropriate design matrix
# design.grp <- model.matrix(~ -1+factor(pData(sub.mic.edata)$Group))
# #!!!make sure the headings match!
# colnames(design.grp) <- sub("factor\\(pData\\(sub.mic.edata\\)\\$Group\\)","", colnames(design.grp))
# # Fits a linear model for each gene based on the given series of arrays
# sub.eset <- as.data.frame(exprs(sub.mic.edata))
# dimnames(eset)[[2]] <- pData(sub.mic.edata)$Sample
# fit.grp <- lmFit(sub.eset, design.grp) 
# # Creates appropriate contrast matrix to perform all pairwise comparisons
# contrast.matrix.grp <- design.pairs(levels(as.factor(pData(sub.mic.edata)$Group)))
# # Computes estimated coefficients and standard errors for contrasts
# fit2.grp <- contrasts.fit(fit.grp, contrast.matrix.grp) 
# # Computes moderated t-statistics and log-odds of differential expression 
# # by empirical Bayes shrinkage of the standard errors towards a common value.
# fit2.grp <- eBayes(fit2.grp) 
# 
# ###RESULTS
# sub.eset$Symbol <- unlist(mget(row.names(sub.eset),mouse430a2SYMBOL, ifnotfound=NA))
# sub.eset$GeneName <- unlist(mget(row.names(sub.eset),mouse430a2GENENAME, ifnotfound=NA))
# 
# ##stat tests and top results
# top.grp <- topTable(fit2.grp, coef=1, adjust="fdr", sort.by="B", p.value=1,  number=nrow(eset), genelist=fit2.grp$genes) 
# results <- decideTests(fit2.grp)
# 
# ##targets
# #activin A receptor,
# #BTK (Bruton tyrosine kinase) 
# #sclerostin
# target.IDs <- row.names(sub.eset)[c(grep("activin A", sub.eset$GeneName),grep("Btk", sub.eset$Symbol), grep("Sost$", sub.eset$Symbol))]
# 
# target.pvalues <- top.grp[top.grp$ID %in% target.IDs,]
# target.pvalues$Symbol <- unlist(mget(target.pvalues$ID,mouse430a2SYMBOL, ifnotfound=NA))
# target.pvalues$GeneName <- unlist(mget(target.pvalues$ID,mouse430a2GENENAME, ifnotfound=NA))
# 
# 
# target.sub.eset <- sub.eset[target.IDs,]
# target.sub.eset$ID <- row.names(target.sub.eset)
# row.names(target.sub.eset)=paste(row.names(target.sub.eset)," (", target.sub.eset$Symbol, ")", sep="")
# 
# pheatmap(target.sub.eset[,1:7], main=colnames(fit2.grp$coefficients)[2], color=brewer.pal(11, "RdBu"), fontsize_row=6,filename="./results/Topaz.v.mCherry.targetgenes.heatmap.pdf", cellheight=6, cellwidth=12)
# 
# write.table(merge(target.pvalues, target.sub.eset), file="./results/Topaz.v.mCherry.targetgenes.pvalues.and.expression.levels.xls", sep="\t", row.names=F, col.names=T)
