# Overview

- **PI:** Dennis Selkoe
- **Experiment:** Housing enrichment effects on microglia
- **Workflow:** `illumina_rnaseq`

Samples are purified microglia from mice subjected to either enriched (`EE`) or normal (`SH`) cage environments. Biological replicates are FACS sorts from brains of individual mice. Cells were sorted over the course of two days, which may have a confounding effect on the data. Male mice were purchased from Taconic in 4 cages, and each cage split for enriched or control environments. There are 6 replicates of both conditions. RNA-Seq was performed at [Harvard Biopolymers Facility][BPF] with 75 bp paired-end reads.

```{r bcbio}
project <- bcbioProject("illumina_rnaseq")
metadata <- bcbioMetadata(project,
                          group = "treatment",
                          laneSplit = TRUE)
summary <- bcbioSummary(project, metadata)
```

```{r metadata_table}
factor <- c("treatment",
            "cage",
            "replicate_number",
            "lane")
save(factor, file = "data/factor.rda")
metadata[, c("description", factor)] %>%
    printTable(caption = "Sample metadata")
```
