---
bibliography: "references.bib"
csl: "bioinformatics.csl"
title: "Anandasabpathy - Vaccina vaccination route effect on gene expression in immune derived cells"
output:
  html_document:
    code_folding: hide
    theme: yeti
    toc: yes
    toc_float: true
    toc_depth: 3
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library("knitcitations")
cleanbib()
options("citation_format" = "pandoc")

clientname="Niroshana Anandasabapathy"
clientemail="niroananda@gmail.com"
labPI="Niroshana Anandasabapathy"
lablocation="BWH"
analystname="John Hutchinson"
analystemail="jhutchin@hsph.harvard.edu"

library(knitr)
opts_chunk$set(warning=FALSE, error=FALSE, message=FALSE, cache=TRUE, tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120), dev="svg")
options(width=200)
```

---

Array analysis for `r clientname` (`r clientemail`), `r labPI` group at `r lablocation`.  

Contact `r analystname` (`r analystemail`) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

---

# Methods Summary  

All MoGene 2.0 ST arrays were processed using the 'oligo' BioConductor package `r citep("10.1093/bioinformatics/btq431")`, quality-controlled with arrayQualityMetrics `r citep("10.1093/bioinformatics/btn647")`, normalized with RMA `r citep("10.1093/biostatistics/4.2.249")`. Differentially expressed genes were identified using limma `r citep("http://link.springer.com/chapter/10.1007%2F0-387-29362-0_23")`.

---

# Setup

### Variables
Working directories, files and other variables necessary to the analysis.

```{r variables}
## Setup Data and Results directory variables
if(file.exists("~/Work/projects/na_vaccinia_scar/")) {
  baseDir <- "~/Work/projects/na_vaccinia_scar/"
} else {
  baseDir <- "~/projects/na_vaccinia_scar/"
}
dataDir <- file.path(baseDir, "data")
metaDir <- file.path(baseDir, "meta")
resultsDir <- file.path(baseDir, "results")

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7") # colorblind friendly palette
covarsfilename="metadata.csv" # tab delimited file describing samples
lowintensity.percentile=0.1
mad.quantile.cutoff=0.1
pvalue.cutoff=0.05
highlight.color="green"
lfc.cutoff=0.58496251
numpcs=4
```

### Libraries

Bioconductor `r citep("10.1038/nmeth.3252")` and R `r citep("http://cran.r-project.org/")` libraries used to process the data.

```{r libraries}
library(arrayQualityMetrics) # array quality control reports
library(oligo) # array utilities

library(pd.mogene.2.0.st) # array layout annotation
library(mogene20sttranscriptcluster.db) # array probe to gene annotations

library(ggdendro)
library(CHBUtils)
library(limma)
library(ggplot2)
library(pheatmap) # pretty heatmaps

library(readr)
library(dplyr)
library(tidyr)

library(biomaRt)

library(reshape2)

# use older version of biomart to allow homology changes
humanmart <- useDataset("hsapiens_gene_ensembl", useMart("ENSEMBL_MART_ENSEMBL", host="jul2016.archive.ensembl.org"))
mousemart <- useDataset("mmusculus_gene_ensembl", useMart("ENSEMBL_MART_ENSEMBL", host="jul2016.archive.ensembl.org"))

mouseattributes = listAttributes(mousemart)
mousefilters = listFilters(mousemart)

humanattributes = listAttributes(humanmart)
humanfilters = listFilters(humanmart)
```

### Functions
- some custom functions

```{r functions}
PCAplot.sd.eset <- function(eset=NULL,  title=NULL){
  eset.core <- exprs(eset)
  myPca.core <- prcomp(t(eset.core))
  # SD of components
  sdevdf <- data.frame(cbind(as.numeric(myPca.core$sdev),c(1:length(myPca.core$sdev))))
  sdevdf$prop <-  sdevdf$X1/sum(sdevdf$X1)
  sdevdf$cum <- cumsum(sdevdf$prop)
  ggplot(sdevdf, aes(x=X2, y=prop)) + 
    geom_point(size=4, color="red") + 
    scale_x_continuous('Component') + 
    scale_y_continuous('Standard Deviation') +
    ggtitle(title) +
    geom_line(data=sdevdf, aes(x=X2, y=cum))
}

```

---

# Import Data and Metadata

- load in phenotypes and array names from metadata file [metadata file](../meta/metadata.csv)
  - this file contains the names and descriptions of CEL files contained in the data directory 
- used these file names to load in the CEL files

```{r dataload, results='hide'}
covars <- read.csv(file.path(metaDir, covarsfilename)) ## simple comma separated  file with CEL file in second column  and sample metadata (i.e. sampleID, treatment group, batch etc.) in subsequent columns
covars <- covars %>% tbl_df() %>% unite(., category, location, celltype, treatment, remove=FALSE) %>% mutate(replicate = 1:n()) %>% unite(., sampleid, category, replicate, remove=FALSE) %>% as.data.frame()

setwd(dataDir)
affyRaw <- read.celfiles(as.character(covars$celfile))
pData(affyRaw) <- covars 
sampleNames(affyRaw) <- pData(affyRaw)$array_code
validObject(affyRaw)
```

### Sample metadata

```{r covars, results='asis'}
# Sample information table
output <- pData(affyRaw)
row.names(output) <- as.character(seq(1:nrow(output)))
kable(output, row.names = TRUE)
```

---

# Pre-Processing and Qualilty Control

### Raw Data QC

- using arrayQualityMetrics library `r citep("10.1093/bioinformatics/btn647")`

```{r rawQC, eval=FALSE}
affyRaw.eset <- ExpressionSet(assayData=exprs(affyRaw))
pData(affyRaw.eset) <- pData(affyRaw)
data.prepped <- prepdata(affyRaw.eset, do.logtransform = TRUE, intgroup=c("location","celltype", "treatment" ))

aqmheat <- aqm.heatmap(data.prepped)
aqmbox <- aqm.boxplot(data.prepped)
aqmdens <- aqm.density(data.prepped)
aqmma <- aqm.maplot(data.prepped)

aqm.writereport(list(aqmheat,  aqmbox, aqmdens, aqmma), outdir = file.path(resultsDir, "report_raw"), arrayTable = pData(affyRaw.eset), reporttitle = "Raw QC Report")
```

Arrays #32, 33, 21 and 57 are outliers from the boxplots.
57, 62 and 5 also seem off by density plot.
Many of these also look bad by MA-plot.

Dropped `r kable(output[c(5,21,32,33,57,62),])`

```{r dropoutliers}
outlierids <- pData(affyRaw)$array_code[c(5,21,32,33,57,62)]
affyRaw <- affyRaw[,which(!pData(affyRaw)$array_code %in% outlierids)]
```

**[Raw Data QC Report](../results/report_raw/index.html)**

### Background Correct and Normalize

- using RMA `r citep("10.1093/biostatistics/4.2.249")`

- summarize probesets on the gene ('core') level

```{r normalize, results='hide'}
affyNorm.core <- oligo::rma(affyRaw,  target="core", background=TRUE, normalize=TRUE)
```

### Normalized Data QC
- using arrayQualityMetrics library

```{r normQC, eval=FALSE}
data.prepped <- prepdata(affyNorm.core, do.logtransform = FALSE, intgroup=c("location","celltype", "treatment" ))

aqmheat <- aqm.heatmap(data.prepped)
aqmbox <- aqm.boxplot(data.prepped)
aqmdens <- aqm.density(data.prepped)
aqmma <- aqm.maplot(data.prepped)

aqm.writereport(list(aqmheat,  aqmbox, aqmdens, aqmma), outdir = file.path(resultsDir, "report_rma_core"), arrayTable = pData(affyNorm.core), reporttitle = "Normalized QC Report")
```

**[Normalized Data QC Report](../results/report_rma.core/index.html)**
 
As expected after normalization , the intensity distributions all look similar now, but clustering of the IFNgammaR1 samples is now poor.

---

# Unsupervised Clustering

### Hierarchical Clustering
The goal of these analyses are to naiively evaluate the variability within the raw data and determine whether this variability can predict the different sample groups

The first method produces a dendrogram by performing  
>  a hierarchical cluster analysis using a set of dissimilarities for the n objects being clustered

Sample IDs are shown and genotypes highlighted in red, green and blue. 

```{r cluster, out.width='75%'}
plot_dendro(affyNorm.core, title="Locations", labels.colname="location", colors.colname="location")
plot_dendro(affyNorm.core, title="Celltypes-Locations", labels.colname="location", colors.colname="celltype")
plot_dendro(affyNorm.core, title="Treatments-Locations", labels.colname="location", colors.colname="treatment")
```

There does appear to be clustering based on cell origin/location. While there may be some clustering based on celltype or treatment, these dendrograms are too simple to look at all the various covariates at the same time, including array batch, FACS date and RNA prep date.

### Principal Component Analysis (PCA)

This second approach is a dimension reduction and visualisation technique that is used to project the multivariate (i.e.multiple genes) data vector of each array into a lower-dimensional plot, such that the spatial arrangement of the points in the plot reflects the overall data (dis)similarity between the arrays. The data is typically reduced to a small number of dimensions (or components) which explain most of the sample variability. 

```{r PCAsd, out.width='75%'}
PCAplot.sd.eset(affyNorm.core, title="")
```

Here, 20 dimensions (i.e. the number of unique categories examined) explain only about ~45% of the variation observed in the samples.

As plots with more than 2 dimensions are difficult to visualize, we typically  split up the dimensions/components and plot them pairwise against each other; the plots here show scatterplots of the arrays along all dual combinations of the first `r numpcs` principal components. In the  plot, each genotype group is represented by a separate color. 

You can use these plots to explore if the arrays cluster, find outliers, and determine whether this is according to an intended experimental factor or according to unintended causes such as batch effects. In this plot,  color is determined by genotype.

```{r pca, fig.cap="Primary Component Analysis of samples - all combinations of the 5 first primary components", out.width='100%'}
library(GGally)
eset.core <- exprs(affyNorm.core)
pd <- pData(affyNorm.core)

myPca.core <- prcomp(t(eset.core))
tmpPCAData.core <- as.data.frame(myPca.core$x[, 1:6])

plotme <- cbind(tmpPCAData.core, pd) %>% tbl_df()
ggpairs(plotme, aes(color=location), columns=1:6, upper=list(continuous="density"))
ggpairs(plotme, aes(color=location), columns=1:6)
ggpairs(plotme, aes(color=celltype), columns=1:6, upper=list(continuous="density"))
ggpairs(plotme, aes(color=celltype), columns=1:6)
ggpairs(plotme, aes(color=treatment), columns=1:6, upper=list(continuous="density"))
ggpairs(plotme, aes(color=treatment), columns=1:6)
ggpairs(plotme, aes(color=treatment, shape=celltype), columns=1:6)


unite

```

PC1 - cell origin/location

PC4 - CD11B_plus/minus versus CD8 or CD103B_plus , celltype

PC3 - NT versus VACV_ID and VACB_SS

---

# Annotate

So far we have only been working with the probesets,without reference to the genes they assay. Here we load in metadata about the probesets on the array (feature data), the gene symbols in particular.

```{r features}
featureData(affyNorm.core) <- getNetAffx(affyNorm.core, "transcript") # this will load the Affymetrix annotation, including the probeID, into the fData
# get gene symbols and entrezIDs for all probesets
fData(affyNorm.core)$symbol <- as.character(unlist(mget(featureNames(affyNorm.core), mogene20sttranscriptclusterSYMBOL, ifnotfound=NA))) # curated annotations from Bioconductor 
fData(affyNorm.core)$entrezID <- as.character(unlist(mget(featureNames(affyNorm.core), mogene20sttranscriptclusterENTREZID, ifnotfound=NA))) # curated annotations from Bioconductor 
```

----
 
# Pre-Filter Data

Reducing the number of genes assayed reduces the multiple test correction and may allow us to identify more differentially expressed genes.
 
`r nrow(affyNorm.core)` initial probes

### By Annotation
- remove the control probes
- remove anything that doesn't have an annotated entrezID

```{r filter1}
affyNorm.core <- affyNorm.core[which(fData(affyNorm.core)$category=="main"),]
affyNorm.core <- affyNorm.core[which(!is.na(fData(affyNorm.core)$entrezID)),]
```

`r nrow(affyNorm.core)` probes remaining

### By Cross Hybridization
- some probes are annotated as potentially hybridizing to multiple targets

```{r filter2}
affyNorm.core <- affyNorm.core[which(fData(affyNorm.core)$crosshybtype=="1"),]
```

`r nrow(affyNorm.core)` probes remaining

### By Expression Level
- remove probes with low expression levels (bottom `r lowintensity.percentile*100`% of all expression levels) in all samples

```{r filter3}
eset.core <- exprs(affyNorm.core)
affyNorm.core <- affyNorm.core[!(apply(eset.core, 1, function(x) all(x<quantile(exprs(affyNorm.core), 0.1)))),]
```

`r nrow(affyNorm.core)` probes remaining

### By Variability
- remove probes with lower variation among all samples (without regard for group status) (dropped the bottom `r mad.quantile.cutoff*100`%) 

```{r filter4}
eset.core <- exprs(affyNorm.core)
rowmads <- apply(eset.core, 1, mad)
mad.cutoff <- as.numeric(quantile(rowmads, mad.quantile.cutoff))
affyNorm.core <- affyNorm.core[rowmads>mad.cutoff,]
```

`r nrow(affyNorm.core)` probes remaining

## Collapse probes
- the array has multiple probes per gene, here I'll collapse this information to have on data point per gene using methods from the collapseRows function `r citep("10.1186/1471-2105-12-322")` in the WGCNA package.
 - here I selected the probeset for each gene with with the maximum mean value for all samples

```{r WGCNA, results='hide'}
library(WGCNA)

eset.core.annot <- cbind(exprs(affyNorm.core), fData(affyNorm.core)) %>%  tbl_df() %>% dplyr::select(., contains("IL27"), contains("WT"), contains("IFNGR"), entrezID, transcriptclusterid)
eset.core <- dplyr::select(eset.core.annot, contains("IL27"), contains("WT"), contains("IFNGR")) %>% as.data.frame()
row.names(eset.core) <- eset.core.annot$transcriptclusterid
collapsed.data <- collapseRows(eset.core,rowID=eset.core.annot$transcriptclusterid, rowGroup=eset.core.annot$entrezID , method="MaxMean")

affyNorm.core <- affyNorm.core[which(collapsed.data$selectedRow),]
```

`r nrow(affyNorm.core)` genes remaining

---

# Statistical Analyses

A linear model for microarray data analysis Limma `r citep("http://link.springer.com/chapter/10.1007%2F0-387-29362-0_23")`  was performed on the samples to identify differentially expressed genes for all pair-wise comparisons of the sample groups. Limma fits a linear model to the expression data for all samples for each gene and is designed to handle complex experiments involving comparisons between many RNA targets simultaneously.

To perform limma, we construct two matrices. The design matrix provides a representation of the different sample groups which have been analysed. The contrast matrix allows the coefficients defined by the design matrix to be combined into contrasts of interest. Each contrast corresponds to a pairwise comparison between sample groups.

### Design matrix 
- make a matrix with arrays as rows, sample groups as columns
- a one or a zero indicate respectively, that a sample either belongs or does not belong to the sample group 

```{r design, results="asis"}
design <- model.matrix(~0+  genotype,  data=pData(affyNorm.core))
# make sure the headings match
colnames(design) <- sub("genotype", "", colnames(design))

kable(design)
```

These matrices are used to fit a linear model to the data. The linear model is applied and pairwise comparisons are performed to identify differentially expressed genes. The comparisons are defined based on the contrasts, which select genes that show a significant expression change between the treated samples.

- first fit the linear model based on the design matrix for each gene based on the given series of arrays
- using the contrast matrix, compute estimated coefficients and standard errors for contrasts
- compute moderated t-statistics and log-odds of differential expression by empirical Bayes shrinkage of the standard errors towards a common value

### Contrasts/Comparisons
- calculated statistics for 3 contrasts (i.e.comparisons)
1) WT vs. IFNgammaR1 KOs
2) WT vs. IL27R KOs
3) Differences between IL27R KOs and IFNgammaR1 KOs with respect to WTs
 (see [here](https://en.wikipedia.org/wiki/Difference_in_differences) for an explanation)


```{r limma,warning=FALSE, message=FALSE}
# Setup constrasts
contrast.matrix <- makeContrasts(WTvIFNGR1KO=WT-IFNGR1KO, WTvIL27RKO=WT-IL27RKO,  IFNGR1KO_IL27RKO_differences=(WT-IFNGR1KO)-(WT-IL27RKO), levels=colnames(design))
kable(contrast.matrix)

# Fit model
fit.core <- lmFit(affyNorm.core, design) 

# Compute cofficients for constrasts
fit2.core <- contrasts.fit(fit.core, contrast.matrix) 

# Bayes shrinkage
fit2.core <- eBayes(fit2.core) 
```


```{r allstats, results='hide'}
all.results <- lapply(seq(1:length(dimnames(contrast.matrix)$Contrasts)), function(num) {
  contrast <- dimnames(contrast.matrix)$Contrasts[num]
  stats <- topTable(fit2.core, coef=num, sort.by="B",adjust.method="BH",number=nrow(fData(affyNorm.core)), genelist=fData(affyNorm.core)[,c("probesetid", "symbol", "entrezID")])
  stats$Passes.FDR.threshold  <-  as.factor(stats$adj.P.Val<pvalue.cutoff)

  eset <- exprs(affyNorm.core)
  eset  <-  eset[match(stats$probesetid, row.names(eset)),]
  stats.eset <- cbind(stats, eset)
  return(list(contrast=contrast, stats.eset=stats.eset))
  })

# output all results to files
lapply(seq(1:length(dimnames(contrast.matrix)$Contrasts)), function(num) {
  contrast <- dimnames(contrast.matrix)$Contrasts[num]
  out.stats=as.data.frame(all.results[[num]]$stats.eset)
  write.table(out.stats, file=file.path(resultsDir, paste("all.genes.stats.exprs", contrast, "xls", sep=".")),  sep ="\t",, row.names=F, col.names=T)
})
```

### Results Summary

At an adjusted pvalue of less than `r pvalue.cutoff` and a log2 fold change cutoff of more than `r lfc.cutoff` we see the following number of significant genes for each comparison:

`r x=1`
####Wildtype versus IFNgammaR1 KO
`r nrow(subset(all.results[[x]]$stats.eset, abs(logFC)>lfc.cutoff & adj.P.Val<pvalue.cutoff))` differentially expressed genes.

`r x=x+1`
####Wildtype versus IL27R KO
`r nrow(subset(all.results[[x]]$stats.eset, abs(logFC)>lfc.cutoff & adj.P.Val<pvalue.cutoff))` differentially expressed genes.

`r x=x+1`
####Differences in differences for IFNgammaR1 KO and IL27R KO with respect to wildtype
`r nrow(subset(all.results[[x]]$stats.eset, abs(logFC)>lfc.cutoff & adj.P.Val<pvalue.cutoff))` differentially expressed genes.


---

# Visualizations

- some quick ways of looking at the results to see if they make sense

### Volcano plots

- great way to visualize significance and effect size at the same time

The plots represent each gene with a dot. The fold change (log2) is plotted on the x-axis and the adjusted p-value (log10, p-value adjusted/corrected for multiple testing) is on the y-axis. Density plots for these values are shown in the margins. 
The highlighted regions (in `r highlight.color`) cover the genes that are differentially expressed (i.e. p-value < `r pvalue.cutoff`, log2 foldchange > `r lfc.cutoff`). Tables below the plot shows the differentially expressed genes.

```{r volcanoplots, out.width='100%', dev='png'}
library(DT)

volcano_density_plot(as.data.frame(all.results[[1]]$stats.eset[,c("logFC", "adj.P.Val")]), pval.cutoff=pvalue.cutoff, lfc.cutoff=lfc.cutoff, shade.colour=highlight.color, title=dimnames(contrast.matrix)$Contrasts[1])
datatable(subset(all.results[[1]]$stats.eset, logFC>lfc.cutoff & adj.P.Val<pvalue.cutoff), rownames=FALSE, caption="Differentially Expressed Genes")

volcano_density_plot(as.data.frame(all.results[[2]]$stats.eset[,c("logFC", "adj.P.Val")]), pval.cutoff=pvalue.cutoff, lfc.cutoff=lfc.cutoff, shade.colour=highlight.color, title=dimnames(contrast.matrix)$Contrasts[2])
datatable(subset(all.results[[2]]$stats.eset, logFC>lfc.cutoff & adj.P.Val<pvalue.cutoff), rownames=FALSE, caption="Differentially Expressed Genes")

volcano_density_plot(as.data.frame(all.results[[3]]$stats.eset[,c("logFC", "adj.P.Val")]), pval.cutoff=pvalue.cutoff, lfc.cutoff=lfc.cutoff, shade.colour=highlight.color, title=dimnames(contrast.matrix)$Contrasts[3])
datatable(subset(all.results[[3]]$stats.eset, logFC>lfc.cutoff & adj.P.Val<pvalue.cutoff), rownames=FALSE, caption="Differentiallly Expressed Genes")
```

### Heatmaps

- quick way to look at how the differentially expressed genes look between the individual samples
- only plotted comparisons with differentially expressed genes
- only plotted up to the top 50 genes as determined by B-value (log odds of differential expression)
- colours represent the log transformed normalized intensitiesnor
- IDs are the gene symbol and entrezID for the differentially expressed gene

```{r heatmaps1, fig.height=8}
x=1
data.plot <- all.results[[x]]$stats.eset %>% 
  filter(., abs(logFC)>lfc.cutoff & adj.P.Val<pvalue.cutoff) %>%
  arrange(., desc(B)) %>% 
    dplyr::select(., symbol,entrezID,contains("IFNGR1"), contains("WT"), contains("IL27R")) %>%
  unite(., uniqueid, symbol, entrezID) %>%
  as.data.frame()

if(nrow(data.plot)>50){
  data.plot <- data.plot[1:50,]
}
data.plot <- col2rownames(data.plot, "uniqueid")  
pheatmap(data.plot, cluster_cols = TRUE, main=all.results[[x]]$contrast)
```

```{r heatmaps3, fig.height=10}
x=3
data.plot <- all.results[[x]]$stats.eset %>% 
  filter(., abs(logFC)>lfc.cutoff & adj.P.Val<pvalue.cutoff) %>%
  arrange(., desc(B)) %>%
  dplyr::select(., symbol,entrezID,contains("IFNGR1"), contains("WT"), contains("IL27R")) %>%
  unite(., uniqueid, symbol, entrezID) %>%
  as.data.frame()
if(nrow(data.plot)>50){
  data.plot <- data.plot[1:50,]
}
data.plot <- col2rownames(data.plot, "uniqueid")  
pheatmap(data.plot, cluster_cols = TRUE, main=all.results[[x]]$contrast)
```

----

# Downloads

### Statistics and expression levels of all genes for these comparisons

*Note that for all these files, I have not summarized values for genes assayed by multiple probes (i.e. by taking the median value), so you may see multiple instances of the same gene in the results*

`r x=1`
[wildtype versus IFNgammaR1 KO stats - all genes](../results/all.genes.stats.exprs.`r all.results[[x]]$contrast`.xls)

`r x=x+1`
[wildtype versus IL27R KO  stats - all genes](../results/all.genes.stats.exprs.`r all.results[[x]]$contrast`.xls)

`r x=x+1`
[differences in differences for IFNgammaR1 KO and IL27R KO with respect to wildtype stats - all genes](../results/all.genes.stats.exprs.`r all.results[[x]]$contrast`.xls)

**These summary tables contain the following information:**

- logFC is the log2-fold change
- the AveExpr is the average expression value accross all arrays
- the moderated t-statistic (t) is the logFC to its standard error, the P.Value is the associated p-value
- the adj.P.Value is the p-value adjusted for multiple testing (by FDR) 
- the B-value (B) is the log-odds that a gene is differentially expressed (the-higher-the-better)
- the last 9 columns contain the log-transformed normalized expression levels for these genes in each sample

---

# Signature Comparisons

## Load Signatures and Filter
- comparing Ifngr1 KO signatures to 
  - skin_DC/lymph_DC DEGs  (DCsig)
    - mix of human and mouse gene symbol IDs, needed mouse EntrezIDs
    - modify any gene symbols with two options to symbol accepted by NCBI
    - add homologs from mouse using Biomart, used EntrezIDs as common id between knockout and DC signature
    - some of the gene symbols are out of date, so after a first run I went back and re-checked any gene symbols missing mouse homologs and updated the gene symbol lto the currently accepted version
  - IFNgamma stimulation of macrophages (GSE18686) (IFNGsig)
    - human study, needed to convert to mouse EntrezIDs
    - dropped any probes that lacked an EntrezID (i.e. unannotated probes)
    - add EnsemblIDs and EntrezIDs, used EntrezID as the common id between the Ifngr1 knockout and the IFNGamma signature
    
**Note that because of id conversion mismatches between mouse and human, the number of genes in the converted signatures may not match up with the original signature**

```{r loadsigs}
# dropped non-accepted symbols prior to loading outside of R
# currently as human gene symbols
DCsig <- read_csv(file.path(dataDir, "20160901_Hutchinson_sigs.csv"))
DCsig <- dplyr::select(DCsig, hgnc_symbol)
genes = getLDS(attributes = c( "hgnc_symbol", "ensembl_gene_id", "entrezgene"), filters = "hgnc_symbol", values = as.character(unlist(DCsig$hgnc_symbol)), mart = humanmart, attributesL = c("mgi_symbol", "ensembl_gene_id", "entrezgene"), martL = mousemart, uniqueRows=T) %>% tbl_df()
names(genes) <- c("hgnc_symbol",  "human_ensemblid","human_entrezid", "mgi_symbol", "mouse_ensemblid", "mouse_entrezid")
DCsig.annot <- dplyr::full_join(DCsig, genes, by="hgnc_symbol")
datatable(DCsig.annot, caption = "Updated DC signature with human and mouse homology gene annotations")

IFNGsig <- read_csv(file.path(dataDir, "Results_GSE18686_IFNGsignature_DEG_byMSF.csv"))
# drop probes without EntrezIDs
IFNGsig <- dplyr::select(IFNGsig, SYMBOL, lgFCH_TrtIFNgIFNG, ENTREZID)
IFNGsig <- filter(IFNGsig, !is.na(ENTREZID))

genes <- getLDS(attributes = c( "hgnc_symbol","ensembl_gene_id", "entrezgene"), filters = "entrezgene", values = as.character(unlist(IFNGsig$ENTREZID)) ,mart = humanmart, attributesL = c("mgi_symbol", "ensembl_gene_id", "entrezgene"), martL = mousemart, uniqueRows=T)
names(genes) <- c("hgnc_symbol",  "human_ensemblid","human_entrezid", "mgi_symbol", "mouse_ensemblid", "mouse_entrezid")
IFNGsig.annot <- dplyr::full_join(IFNGsig, genes, by=c("ENTREZID"="human_entrezid"))
datatable(IFNGsig.annot, caption = "Updated IFNG signature with human and mouse homology gene annotations")


IFNGsig.annot.up <-filter(IFNGsig.annot, lgFCH_TrtIFNgIFNG>0)
IFNGsig.annot.down <-filter(IFNGsig.annot, lgFCH_TrtIFNgIFNG<0)
```    

Performed these analyses using the clusterProfiler `r citep("10.1089/omi.2011.0118")` package from Bioconductor, which lets you use custom gene sets to perform both hypergeometric anbd gene set enrichment analyses of gene expression data. To do so, I  first sannotate the genes analyzed in the Ifngr1 KO analysis as to whether they are part of one of the gene signatures. I used EntrezIDs to match up gene signatures to the Ifngr1 KO analysis.

For both hypergeometric overlap and GSEA-style analyses, I used the two gene signatures as lists of genes without rank. I also tried splitting the IFNGsig genes into those up/down regulated as well to see if anything interesting comes up.


```{r annotategenes}
set2gene <- all.results[[1]]$stats.eset %>% tbl_df() %>% dplyr::select(., entrezID) %>% 
  mutate(., DCsig=ifelse(entrezID %in% DCsig.annot$mouse_entrezid, "DC", 0)) %>%
  mutate(., IFNGsig=ifelse(entrezID %in% IFNGsig.annot$mouse_entrezid, "IFNG", 0)) %>% 
  mutate(., IFNGsigup=ifelse(entrezID %in% IFNGsig.annot.up$mouse_entrezid, "IFNGup", 0)) %>% 
   mutate(., IFNGsigdown=ifelse(entrezID %in% IFNGsig.annot.down$mouse_entrezid, "IFNGdown", 0)) %>% 
   gather(., set, value, DCsig, IFNGsig, IFNGsigup, IFNGsigdown) %>% 
  filter(., value!=0) %>%
  arrange(., entrezID) %>%
  dplyr::select(., set, entrezID) %>%
  as.data.frame()
```  

## Hypergeometric

This method uses hard cutoffs on the Ifngr1 knockout data to derive a list of genes to overlap with a gene signature. The statistical test calculates the likelihood of us seeing an overlap of the size observed given the background set of genes and the sizes of the two sets of genes. The list is treated as unordered, so each gene carries the same weight.

```{r hypergeometric}
library(VennDiagram)
library(clusterProfiler) 
background.genes <- all.results[[2]]$stats.eset$entrezID
deg <- all.results[[1]]$stats.eset %>% 
     tbl_df() %>% 
     filter(., abs(logFC)>lfc.cutoff & adj.P.Val<pvalue.cutoff) %>% 
     dplyr::select(., entrezID) %>%
     unlist() %>% 
     as.character()

# statistics uysing clusterprofiler
results = enricher(deg, TERM2GENE=set2gene, minGSSize=200, maxGSSize=1700, universe=background.genes, pvalueCutoff = 1, qvalueCutoff = 1)
kable(summary(results), caption="Overlapping gene signature statistics")   

kable(getBM(values=results@geneInCategory$DCsig, attributes=c("mgi_symbol", "entrezgene"), filters="entrezgene", mart=mousemart), caption="Overlapping genes")
```

### Venn Diagrams {.tabset}

#### DC signature

```{r dcsigoverlap}
overlap.DCsig <- calculate.overlap(x=list("Ifngr1_KO"=deg, "DCsig"=DCsig.annot$mouse_entrezid))
grid::grid.newpage()
draw.pairwise.venn(length(overlap.DCsig$a1), length(overlap.DCsig$a2), length(overlap.DCsig$a3), ext.text=FALSE, fill=c("red", "blue"), alpha=c(0.5, 0.5))
```

#### IFNgamma signature

```{r ifngsigoverlap}
overlap.IFNGsig <- calculate.overlap(x=list("Ifngr1_KO"=deg, "IFNGsig"=IFNGsig.annot$mouse_entrezid))
grid::grid.newpage()
draw.pairwise.venn(length(overlap.IFNGsig$a1), length(overlap.IFNGsig$a2), length(overlap.IFNGsig$a3), ext.text=FALSE, fill=c("red", "blue"), alpha=c(0.5, 0.5))
```


## Gene Set Enrichment

Gene set enrichment analysis tools like GAGE and clusterProfiler use ranked lists of genes (here ranked by log2FC) without using a threshold. This allows these gene set enrichment tools to use more information to identify enriched biological processes. The introduction to gene set enrichment analysis goes into more detail about some of the advantages of this approach: http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1239896/.

```{r gsea, results='hide'}
# get logfold  chnages and entrezids, sorted by logfoldchange
entrez_results <- all.results[[1]]$stats.eset
entrez_results <- entrez_results[order(entrez_results$logFC, decreasing = TRUE),]
foldchanges <- entrez_results$logFC
names(foldchanges) <- entrez_results$entrezID

# run clusterprofiler  
all.results.gsea <- all.results[[1]]$stats.eset
lfcs <- all.results.gsea$logFC
names(lfcs) <- all.results.gsea$entrezID
lfcs <- sort(lfcs, decreasing=TRUE)

results = GSEA(lfcs, TERM2GENE=set2gene, minGSSize=200, maxGSSize=1700) 
```

### Enrichment Plots {.tabset}

#### DC signature

```{r gseadcsigplot}
gseaplot(results, "DCsig")
```

#### IFNG signature

```{r gseaifngsigplot}
gseaplot(results, "IFNGsig")
```

#### IFNG down signature

```{r gseaifngsigdowngplot}
gseaplot(results, "IFNGsigdown")
```

#### IFNG up signature

```{r gseaifngsigupgplot}
gseaplot(results, "IFNGsigup")
```

### GSEA Results

```{r gagetable}
kable(results@result)
```  
  

---

# Compare to previous results

```{r comparechris}
library(readxl)
library(janitor)
results.chris <- read_excel("WT vs IFNgR KO Differentially Expressed Genes 1.5 and 2 fold Change P value 0.05.xlsx") %>% clean_names() %>%
  filter(.,pvalue_wt)_vs_ifngr1_ko<0.05, fold_change_wt_vs_ifngr1_ko>1.5 )
results.jh <- all.results[[1]]$stats.eset %>% tbl_df()


results.merged <- dplyr::inner_join(results.chris, results.jh, by=c("column_id"="probesetid")) %>%
  mutate(.,logfold_change_wt_vs_ifngr1_ko=log2(fold_change_wt_vs_ifngr1_ko) ) %>%
  mutate(., fc=2^logFC)

ggplot(results.merged, aes(x=logfold_change_wt_vs_ifngr1_ko, y=logFC))+geom_point()+xlab("log2 fold change - Chris")+ylab("log2 fold change - John")
ggplot(results.merged, aes(x=fold_change_wt_vs_ifngr1_ko, y=logFC))+geom_point()+xlab("fold change - Chris")+ylab("log2 fold change - John")
ggplot(results.merged, aes(x=fold_change_wt_vs_ifngr1_ko, y=fc))+geom_point()+xlab("fold change - Chris")+ylab("fold change - John")

ggplot(results.merged, aes(x=ratio_wt_vs_ifngr1_ko, y=fc))+geom_point()+xlab("ratio - Chris")+ylab("fold change - John")

ggplot(results.merged, aes(x=p_value_wt_vs_ifngr1_ko, y=adj.P.Val))+geom_point()+xlab("pvalue - Chris")+ylab("adjusted pvalue - John")
ggplot(results.merged, aes(x=p_value_wt_vs_ifngr1_ko, y=P.Value))+geom_point()+xlab("pvalue - Chris")+ylab("pvalue - John")



plot(density(table(results.merged$p_value_wt_vs_ifngr1_ko)))
plot(density(table(results.merged$adj.P.Val)))
plot(density(table(results.merged$P.Value)))
```

There are `r length(results.chris$gene_symbol))` probes in Chris's results, representing `r length(unique(results.chris$gene_symbol))` unique genes. My results only looked at one probeset per gene (the one with the highest average expression). Together, the two sets of results share `r nrow(results.merged)` genes/probes.

---
# R Session Info

(useful if replicating these results)

```{r sessioninfo}
sessionInfo()
```


---

# References

```{r writebib, results='hide', echo=FALSE, message=FALSE, cache=FALSE}
write.bibtex(file="references.bib")
```