---
output:
  knitrBootstrap::bootstrap_document:
    theme: readable
    highlight: zenburn
    theme.chooser: TRUE
    highlight.chooser: TRUE
  html_document:
    toc: true
    highlight: zenburn
---

```{r setup1, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='', bootstrap.show.code=FALSE)
```

```{r setup2}
library(knitr)

library(ggplot2)
library(reshape)
library(DESeq2)
library(genefilter)
library(CHBUtils)
library(gtools)
library(gridExtra)
library(devtools)
library(dplyr)
library(isomiRs)
library(stringr)

root_path = "/Users/rkhetani/Dropbox/HBC\ consults/Stankovic/"
root_file = paste0(root_path,"/report/")

condition = "condition"
```

```{r render,eval=FALSE, echo=FALSE}
library(rmarkdown)
library(knitrBootstrap)
# render("report.rmd")
```

```{r load}
setwd(root_path)
files = read.table("summary_re_exosome.csv", sep=",",header=T,colClasses = "character")

samples = files[,"sample_id"]

names_stats = files[,"size_stats"]
names(names_stats) = samples

groups = files[,"group"]
names(groups) = samples

summarydata = data.frame(row.names=samples,samples=samples,group=groups)
design <- data.frame(row.names=files$sample_id, condition=files$group)

```

# DE - exosomes_post-filtering
All differential expression analysis was performed on raw count matrices for miRNAs, isomirs and clusters, that had been filtered as follows:

Any row that did not have at least 3 samples with at least 3 counts, from at least one condition (poor hearing or good hearing) were removed (in the count matrices each row represents a distinct miRNA or isomir or cluster & each column represents a specific sample).


```{r de-setup}
 library(DESeq2)
 library(DEGreport)
 library(vsn)
```

```{r deseq2-handler}
 
 filter_handle <- function(res){
     res_nona <- res[!is.na(res$padj),]
     keep <- res_nona$padj < 0.1 
     res_nona[keep,]
 }

 handle_deseq2 = function(dds, summarydata, column) {
   all_combs = combn(levels(summarydata[,column]), 2, simplify=FALSE)
   all_results = list()
   contrast_strings = list()
   for(comb in all_combs) {
     contrast_string = paste(comb, collapse="_vs_")
     contrast = c(column, comb)
     res = results(dds, contrast=contrast)
     res = res[order(res$padj),]
     all_results = c(all_results, res)
     contrast_strings = c(contrast_strings, contrast_string)
   }
   names(all_results) = contrast_strings
   return(all_results)
 }

 plot_MA = function(res){
     for(i in seq(length(res))) {
         plotMA(res[[i]])
         title(paste("MA plot for contrast", names(res)[i]))
     }
 }

 plot_volcano = function(res){
     for(i in seq(length(res))) {
         stats = as.data.frame(res[[i]][,c(2,6)])
         p = volcano_density_plot(stats, title=names(res)[i], lfc.cutoff=1)
         print(p)
     }
 }

 do_de = function(raw, summarydata, condition){
     dss = DESeqDataSetFromMatrix(countData = raw[rowMeans(raw)>3,],
                        colData = summarydata,
                        design = ~ condition)
     dss = DESeq(dss)
     plotDispEsts(dss)
     dss
 }

 do_norm = function(dss, root_path, prefix){
     rlog_ma = assay(rlog(dss))
     # vst_ma = assay(varianceStabilizingTransformation(dss))
     count_ma = counts(dss, normalized=TRUE)
     raw = counts(dss, normalized=FALSE)
     fn_log = paste0(root_file, prefix, "log_matrix.txt")
     write.table(rlog_ma,fn_log,sep="\t")
     # write.table(vst_ma,fn_log,sep="\t")
     fn_count = paste0(root_file, prefix, "count_matrix.txt")
     write.table(count_ma,fn_count,sep="\t")
     fn_raw = paste0(root_file, prefix, "raw_matrix.txt")
     write.table(count_ma,fn_raw,sep="\t")
 }

```

## mirna
#### Dispersion-plot

```{r de}
exo.mirna <- read.table("/Users/rkhetani/Dropbox/HBC\ consults/Stankovic/filtered\ counts/exosome_count_atleast3gt3_mirna.txt", sep="\t",header=T, row.names=1)
exo.mirna <- exo.mirna[,c(1,7,8,2,3,4,9,10,5,11,12,6,13)]
dds = do_de(exo.mirna, design, condition)
```

```{r tables}
 do_norm(dds, root_path, "mirna_")
```

#### MA-plot
 
```{r DESeq-output, results='asis'}
 all_results = handle_deseq2(dds, design, condition)
 plot_MA(all_results)
```

#### Volcano-plot

```{r DESeq-volcano, results='hide'}
 plot_volcano(all_results)
```

#### Differentially expressed miRNAs
No differentially expressed miRNAs observed at an adjusted p-value threshold of < 0.1.
```{r DE-miRNA, results='hold'}
mirna.result <- results(dds)
summary(mirna.result)
```

## isomir
#### Dispersion-plot

```{r de-iso}
exo.isomir <- read.table("/Users/rkhetani/Dropbox/HBC\ consults/Stankovic/filtered\ counts/exosome_count_atleast3gt3_isomirs.txt", sep="\t",header=T, row.names=1)
exo.isomir <- exo.isomir[,c(1,7,8,2,3,4,9,10,5,11,12,6,13)]
dds = do_de(exo.isomir, design, condition)
```

```{r tables-iso}
 do_norm(dds, root_path, "isomirs_")
```

#### MA-plot
 
```{r DESeq-output-iso, results='asis'}
 all_results = handle_deseq2(dds, design, condition)
 plot_MA(all_results)
```

#### Volcano-plot

```{r DESeq-volcano-iso, results='hide'}
 plot_volcano(all_results)
```

#### Differentially expressed isomirs
No differentially expressed isomirs observed at an adjusted p-value threshold of < 0.1.
```{r DE-isomir, results='hold'}
isomir.result <- results(dds)
summary(isomir.result)
```

## clusters
#### Dispersion-plot

```{r de-c}
exo.clus <- read.table("/Users/rkhetani/Dropbox/HBC\ consults/Stankovic/filtered\ counts/exosome_count_atleast3gt3_clusters.txt", sep="\t",header=T, row.names=1)
exo.clus <- exo.clus[,c(1,7,8,2,3,4,9,10,5,11,12,6,13)]
dds = do_de(exo.clus, design, condition)
```

```{r tables-c}
 do_norm(dds, root_path, "clusters_")
```

#### MA-plot
 
```{r DESeq-output-c, results='asis'}
 all_results = handle_deseq2(dds, design, condition)
 plot_MA(all_results)
```

#### Volcano-plot

```{r DESeq-volcano-c, results='hide'}
 plot_volcano(all_results)
```

#### Differentially expressed clusters
No differentially expressed clusters observed at an adjusted p-value threshold of < 0.1.

```{r DE-cluster, results='hold'}
clus.result <- results(dds)
summary(clus.result)
```


