---
output:
  knitrBootstrap::bootstrap_document:
    theme: readable
    highlight: zenburn
    theme.chooser: TRUE
    highlight.chooser: TRUE
  html_document:
    toc: true
    highlight: zenburn
---

```{r setup1, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='', bootstrap.show.code=FALSE)
```

```{r setup2}
library(knitr)

library(ggplot2)
library(reshape)
library(DESeq2)
library(genefilter)
library(CHBUtils)
library(gtools)
library(gridExtra)
library(devtools)
library(dplyr)
library(isomiRs)
library(stringr)

root_path = "/Users/rkhetani/Dropbox/HBC\ consults/Stankovic/"
root_file = paste0(root_path,"/report/")

condition = "condition"
```

```{r render,eval=FALSE, echo=FALSE}
library(rmarkdown)
library(knitrBootstrap)
# render("report.rmd")
```

```{r load}
#setwd(root_path)
files = read.table("/Users/rkhetani/Dropbox/HBC\ consults/Stankovic/summary_re_cell.csv", sep=",",header=T,colClasses = "character")

samples = files[,"sample_id"]

names_stats = files[,"size_stats"]
names(names_stats) = samples

groups = files[,"group"]
names(groups) = samples

summarydata = data.frame(row.names=samples,samples=samples,group=groups)
design <- data.frame(row.names=files$sample_id, condition=files$group)

options(width = 100) 
```

# DE - cells_post-filtering

All differential expression analysis was performed on raw count matrices for miRNAs, isomirs and clusters, that were filtered as follows:

Any row that did not have at least 2 samples with at least 3 counts, from at least one condition (poor hearing or good hearing) were removed (in the count matrices each row represents a distinct miRNA or isomir or cluster & each column represents a specific sample).


```{r de-setup}
 library(DESeq2)
 library(DEGreport)
 library(vsn)
```

```{r deseq2-handler}
 
 filter_handle <- function(res){
     res_nona <- res[!is.na(res$padj),]
     keep <- res_nona$padj < 0.1 
     res_nona[keep,]
 }

 handle_deseq2 = function(dds, summarydata, column) {
   all_combs = combn(levels(summarydata[,column]), 2, simplify=FALSE)
   all_results = list()
   contrast_strings = list()
   for(comb in all_combs) {
     contrast_string = paste(comb, collapse="_vs_")
     contrast = c(column, comb)
     res = results(dds, contrast=contrast)
     res = res[order(res$padj),]
     all_results = c(all_results, res)
     contrast_strings = c(contrast_strings, contrast_string)
   }
   names(all_results) = contrast_strings
   return(all_results)
 }

 plot_MA = function(res){
     for(i in seq(length(res))) {
         plotMA(res[[i]])
         title(paste("MA plot for contrast", names(res)[i]))
     }
 }

 plot_volcano = function(res){
     for(i in seq(length(res))) {
         stats = as.data.frame(res[[i]][,c(2,6)])
         p = volcano_density_plot(stats, title=names(res)[i], lfc.cutoff=1)
         print(p)
     }
 }

 do_de = function(raw, summarydata, condition){
     dss = DESeqDataSetFromMatrix(countData = raw[rowMeans(raw)>3,],
                        colData = summarydata,
                        design = ~ condition)
     dss = DESeq(dss)
     plotDispEsts(dss)
     dss
 }

 do_norm = function(dss, root_path, prefix){
     rlog_ma = assay(rlog(dss))
     # vst_ma = assay(varianceStabilizingTransformation(dss))
     count_ma = counts(dss, normalized=TRUE)
     raw = counts(dss, normalized=FALSE)
     fn_log = paste0(root_file, prefix, "log_matrix.txt")
     write.table(rlog_ma,fn_log,sep="\t")
     # write.table(vst_ma,fn_log,sep="\t")
     fn_count = paste0(root_file, prefix, "count_matrix.txt")
     write.table(count_ma,fn_count,sep="\t")
     fn_raw = paste0(root_file, prefix, "raw_matrix.txt")
     write.table(count_ma,fn_raw,sep="\t")
 }

```

## mirna
#### Dispersion-plot

```{r de}
cell.mirna <- read.table("/Users/rkhetani/Dropbox/HBC\ consults/Stankovic/filtered\ counts/cell_count_atleast2gt3_mirna.txt", sep="\t",header=T, row.names=1)
cell.mirna <- cell.mirna[,c(1,6,2,3,4,5,7,8,9)]
dds = do_de(cell.mirna, design, condition)
```

```{r tables}
 do_norm(dds, root_path, "mirna_")
```

#### MA-plot
 
```{r DESeq-output, results='asis'}
 all_results = handle_deseq2(dds, design, condition)
 plot_MA(all_results)
```

#### Volcano-plot

```{r DESeq-volcano, results='hide'}
 plot_volcano(all_results)
```

#### Differentially expressed miRNAs
3 differentially expressed miRNAs observed at an adjusted p-value threshold of < 0.1.

```{r DE-miRNA,tidy=TRUE}
mirna.result <- results(dds)
summary(mirna.result)
mirna.result[order(mirna.result$padj)[1:3],]
```

```{r DE-miRNA-plot, results='hide'}
d <- plotCounts(dds, gene="hsa-miR-6798-3p", intgroup="condition", returnData=TRUE)
mirna1 <- ggplot(d, aes(x=condition, y=count)) +
labs(title="hsa-miR-6798-3p") +
geom_point(position=position_jitter(w=0.1,h=0)) +
theme(axis.text.x = element_text(size = rel(0.75))) +
theme(title = element_text(size = rel(0.90))) +
xlab("") +
scale_y_log10(breaks=c(1,10,15))

d <- plotCounts(dds, gene="hsa-miR-192-5p", intgroup="condition", returnData=TRUE)
mirna2 <- ggplot(d, aes(x=condition, y=count)) +
labs(title="hsa-miR-192-5p") +
geom_point(position=position_jitter(w=0.1,h=0)) +
theme(axis.text.x = element_text(size = rel(0.75))) +
theme(title = element_text(size = rel(0.90))) +
ylab("") +
scale_y_log10(breaks=c(2000, 3000, 4000))

d <- plotCounts(dds, gene="hsa-miR-431-5p", intgroup="condition", returnData=TRUE)
mirna3 <- ggplot(d, aes(x=condition, y=count)) +
labs(title="hsa-miR-431-5p") +
geom_point(position=position_jitter(w=0.1,h=0)) +
theme(axis.text.x = element_text(size = rel(0.75))) +
theme(title = element_text(size = rel(0.90))) +
xlab("") +
ylab("") +
scale_y_log10(breaks=c(250, 500, 1000))

grid.arrange(mirna1, mirna2, mirna3, ncol=3, main=str_wrap("Distribution of normalized counts between conditions for the 3 differentially expressed miRNAs", width=60))
dev.off()

```


## isomir
#### Dispersion-plot

```{r de-iso}
cell.isomir <- read.table("/Users/rkhetani/Dropbox/HBC\ consults/Stankovic/filtered\ counts/cell_count_atleast2gt3_isomirs.txt", sep="\t",header=T, row.names=1)
cell.isomir <- cell.isomir[,c(1,6,2,3,4,5,7,8,9)]
dds = do_de(cell.isomir, design, condition)
```

```{r tables-iso}
 do_norm(dds, root_path, "isomirs_")
```

#### MA-plot
 
```{r DESeq-output-iso, results='asis'}
 all_results = handle_deseq2(dds, design, condition)
 plot_MA(all_results)
```

#### Volcano-plot

```{r DESeq-volcano-iso, results='hide'}
 plot_volcano(all_results)
```

#### Differentially expressed isomirs
4 differentially expressed isomirs observed at an adjusted p-value threshold of < 0.1.

```{r DE-isomir,tidy=TRUE}
isomir.result <- results(dds)
summary(isomir.result)
isomir.result[order(isomir.result$padj)[1:4],]
```

```{r DE-isomir-plot, results='hide'}
d <- plotCounts(dds, gene="hsa-miR-224-5p.iso.t50.t3:d-T.ad:u-T.mm:0", intgroup="condition", returnData=TRUE)
isomir1 <- ggplot(d, aes(x=condition, y=count)) +
labs(title="hsa-miR-224-5p.iso.t50.t3:d-T.ad:u-T.mm:0") +
geom_point(position=position_jitter(w=0.1,h=0)) +
theme(plot.title = element_text(size = rel(0.75))) +
theme(title = element_text(size = rel(0.90))) +
xlab("") +
scale_y_log10(breaks=c(1,10,50))

d <- plotCounts(dds, gene="hsa-miR-431-5p.iso.t50.t3:u-A.ad:0.mm:0", intgroup="condition", returnData=TRUE)
isomir2 <- ggplot(d, aes(x=condition, y=count)) +
labs(title="hsa-miR-431-5p.iso.t50.t3:u-A.ad:0.mm:0") +
geom_point(position=position_jitter(w=0.1,h=0)) +
theme(plot.title = element_text(size = rel(0.75))) +
theme(title = element_text(size = rel(0.90))) +
xlab("") +
ylab("") +
scale_y_log10(breaks=c(200, 400, 600))

d <- plotCounts(dds, gene="hsa-miR-192-5p.iso.t50.t3:d-A.ad:0.mm:0", intgroup="condition", returnData=TRUE)
isomir3 <- ggplot(d, aes(x=condition, y=count)) +
labs(title="hsa-miR-192-5p.iso.t50.t3:d-A.ad:0.mm:0") +
geom_point(position=position_jitter(w=0.1,h=0)) +
theme(plot.title = element_text(size = rel(0.75))) +
theme(title = element_text(size = rel(0.90))) +
scale_y_log10(breaks=c(75, 100,150))

d <- plotCounts(dds, gene="hsa-miR-192-5p.ref.t50.t3:0.ad:0.mm:0", intgroup="condition", returnData=TRUE)
isomir4 <- ggplot(d, aes(x=condition, y=count)) +
labs(title="hsa-miR-192-5p.ref.t50.t3:0.ad:0.mm:0") +
geom_point(position=position_jitter(w=0.1,h=0)) +
theme(plot.title = element_text(size = rel(0.75))) +
theme(title = element_text(size = rel(0.90))) +
ylab("") +
scale_y_log10(breaks=c(1500,2000,2500))

grid.arrange(isomir1, isomir2, isomir3, isomir4, nrow=2, ncol=2, main=str_wrap("Distribution of normalized counts between conditions for the 4 differentially expressed isomirs", width=60))
dev.off()
```


## clusters
#### Dispersion-plot
```{r de-c}
cell.clus <- read.table("/Users/rkhetani/Dropbox/HBC\ consults/Stankovic/filtered\ counts/cell_count_atleast2gt3_clusters.txt", sep="\t",header=T, row.names=1)
cell.clus <- cell.clus[,c(1,6,2,3,4,5,7,8,9)]
dds = do_de(cell.clus, design, condition)
```

```{r tables-c}
 do_norm(dds, root_path, "clusters_")
```

#### MA-plot
 
```{r DESeq-output-c, results='asis'}
 all_results = handle_deseq2(dds, design, condition)
 plot_MA(all_results)
```

#### Volcano-plot

```{r DESeq-volcano-c, results='hide'}
 plot_volcano(all_results)
```

#### Differentially expressed clusters
3 differentially expressed clusters observed at an adjusted p-value threshold of < 0.1.

```{r DE-cluster,tidy=TRUE}
clus.result <- results(dds)
summary(clus.result)
clus.result[order(clus.result$padj)[1],]
# draw.heatmap.clus <- cell.clus[row.names(clus.result[order(clus.result$padj)[c(1:3)],]),]
# pheatmap(as.matrix(draw.heatmap.clus), color= cbPalette, annotation=design, cellheight=30, cellwidth = 30, breaks = c(100, 500, 25000, 50000, 60000))
```

```{r DE-cluster-plot, results='hide'}
d <- plotCounts(dds, gene="495", intgroup="condition", returnData=TRUE)
ggplot(d, aes(x=condition, y=count)) +
labs(title="cluster #495") +
geom_point(position=position_jitter(w=0.1,h=0)) +
theme(axis.text.x = element_text(size = rel(0.75))) +
theme(title = element_text(size = rel(0.90))) +
xlab("") +
scale_y_log10(breaks=c(25,100,400))
```