---
output:
  knitrBootstrap::bootstrap_document:
    theme: readable
    highlight: zenburn
    theme.chooser: TRUE
    highlight.chooser: TRUE
  html_document:
    toc: true
    highlight: zenburn
---

```{r setup, echo=FALSE, message=FALSE}
library(knitr)

library(ggplot2)
library(reshape)
library(DESeq2)
library(genefilter)
library(CHBUtils)
library(gtools)
library(gridExtra)
library(devtools)
library(dplyr)
library(isomiRs)

knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png", fig.width=6,fig.heigh=6,
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='', bootstrap.show.code=FALSE)

root_path = "/Users/rkhetani/Dropbox/HBC\ consults/Stankovic"
root_file = paste0(root_path,"/report/")

condition = "condition"
```

```{r render, eval=FALSE, echo=FALSE}
library(rmarkdown)
library(knitrBootstrap)
# render("report.rmd")
```

```{r load}
setwd(root_path)
files = read.table("summary_re_4groups.csv", sep=",",header=T,colClasses = "character")

samples = files[,"sample_id"]

names_stats = files[,"size_stats"]
names(names_stats) = samples

groups = files[,"group"]
names(groups) = samples

summarydata = data.frame(row.names=samples,samples=samples,group=groups)
design <- data.frame(row.names=files$sample_id, condition=files$group)

```

# Exploratory analysis

In this section we will see descriptive figures about quality of the data, 
reads with adapter, reads mapped to miRNAs, reads mapped to other small RNAs. 

## Size distribution

### After adapter removal, we can plot the size distribution of the small RNAs.

```{r adapter,fig.width=10}
tab = data.frame()
for (sample in samples){
    d = read.table(names_stats[sample], sep=" ")
    tab = rbind(tab, d %>% mutate(sample=sample, group=groups[sample]))
}

reads_adapter = tab %>% group_by(sample, group) %>% summarise(total=sum(V2))
ggplot(reads_adapter, aes(x=sample,y=total,fill=group)) +
    geom_bar(stat="identity", position = "dodge") +
    ggtitle("total number of reads with adapter") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggplot(tab, aes(x=V1,y=V2,fill=sample)) +
    geom_bar(stat="identity", position = "dodge") +
    facet_wrap(~group, ncol=2)+
    ggtitle("size distribution") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## miRNA

### Total miRNA expression annotated with mirbase

```{r mirna,results='hide'}
mi_files = files[,"miraligner"]
row.names(design) = samples

obj <- IsomirDataSeqFromFiles(files = mi_files, design = design ,header = T, cov = 1)
```

```{r mirna-mirbase}
ggplot( data.frame(sample=colnames(counts(obj)), total=colSums(counts(obj)))) +
    geom_bar(aes(x=sample,y=total), stat='identity')+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
mirna_step <- as.data.frame(colSums(counts(obj)))
```

### Distribution of mirna expression

```{r depth}
ggplot(melt(counts(obj))) +
    geom_boxplot(aes(x=X2,y=value))+
    scale_y_log10()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

### Cumulative distribution of miRNAs

```{r cum}

cs <- as.data.frame(apply(counts(obj),2,cumsum))
cs$pos <- 1:nrow(cs)

ggplot((melt(cs,id.vars = "pos")))+
    geom_point(aes(x=pos,y=value,color=variable))+
    scale_y_log10()
```

## Other small RNA

The data was analyzed with [seqcluster](http://seqcluster.readthedocs.org/)

This tools used all reads, uniquely mapped and multi-mapped reads. The first
step is to cluster sequences in all locations they overlap. The second step is to 
create meta-clusters: is the unit that merge all clusters that share the 
same sequences. This way the output are meta-clusters, common sequences that could
come from different region of the genome.


### Genome covered

```{r genome-covered, results='asis'}
cov_stats <- read.table(file.path(root_path, "align", "seqs_clean_rmlw.bam_cov.tsv"),sep="\t",check.names = F)

kable(cov_stats[cov_stats$V1=="genome",] %>% select(coverage=V2,ratio_genome=V5))

```

The normal value for data with strong small RNA signal is: 0.0002

### Classification

Number of reads in the data after each step

```{r reads-track}
reads_stats <- read.table(file.path(root_path, "seqcluster", "cluster", "read_stats.tsv"),sep="\t",check.names = F)
ggplot(reads_stats, aes(x=V2, y=V1, fill=V3)) + 
    geom_bar(stat = 'identity', position = 'dodge') +
    labs(list(x="samples", y="reads")) +
    scale_fill_brewer("steps", palette = 'Set1')+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


```{r cluster}
clus <- read.table(files[1,"clusters"],header=T,sep="\t",row.names=1, check.names = FALSE)
ann <- clus[,2]
toomany <- clus[,1]
clus_ma <- clus[,3:ncol(clus)]
clus_ma = clus_ma[,row.names(design)]
```


Check complex meta-clusters: This kind of event happens when there are small RNA over the whole genome, and all repetitive small RNAs map to thousands of places and sharing many sequences in many positions.

(If any meta-cluster is > 40% of the total data, maybe it is worth adding some filters like: minimum number of counts `-e` or `--min--shared` in `seqcluster prepare`)

```{r complex-meta-clusters}
clus_ma_norm = sweep(clus_ma, 2, colSums(clus_ma), "/")
head(clus_ma_norm[toomany>0.00001,])
```


#### Contribution by class:

rmsk (repeat masked) in the figure legend stands for genomic areas with Repeats 

```{r cluster_type,eval=TRUE}
rRNA <- colSums(clus_ma[grepl("rRNA",ann) & grepl("miRNA",ann)==F,])
miRNA <- colSums(clus_ma[grepl("miRNA",ann),])
tRNA <- colSums(clus_ma[grepl("tRNA",ann) & grepl("rRNA",ann)==F & grepl("repeat",ann)==F & grepl("miRNA",ann)==F,])
rmsk <- colSums(clus_ma[grepl("repeat",ann) & grepl("rRNA",ann)==F & grepl("miRNA",ann)==F,])
total <- colSums(clus_ma)

dd <- data.frame(samples=names(rRNA),
                 rRNA=rRNA,
                 miRNA=miRNA,
                 tRNA=tRNA,
                 rmsk=rmsk,
                 total=total)

ggplot(melt(dd)) +
    geom_bar(aes(x=samples,y=value,fill=variable),
             stat='identity',
             position="dodge")+
    scale_fill_brewer(palette = "Set1")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

dd_norm = dd
dd_norm[,2:5] = sweep(dd[,2:5],1,dd[,6],"/")
ggplot(melt(dd_norm[,1:5])) +
    geom_bar(aes(x=samples,y=value,fill=variable),
             stat='identity',
             position="dodge")+
    scale_fill_brewer(palette = "Set1")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
    labs(list(title="Relative proportion of small RNAs",y="% reads"))
```

