```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, cache=FALSE,  highlight=TRUE, figalign="center", echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120))
options(width=200)
```

# COMBINED MICROARRAY ANALYSES OF LEUKOCYTE DEVELOPMENT SYSTEMS
- HOXA9-ER induced time course
- Prostacyclin induced paired analyses
---

Analysis of microarray datasets for transcriptional signatures for [David Sykes](mailto:DBSYKES@PARTNERS.ORG).

Contact [John Hutchinson](mailto:jhutchin@hsph.harvard.edu) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

----

# GENERAL SETUP

## General purpose R libraries necessary for analysis

```{r general_libraries}
source("http://bioconductor.org/biocLite.R") # BioConductor script necessary for installing new BioC libraries with biocLite()
library(plyr) # library for iteratively working with data structures
library(ggplot2) # library for plotting 
library(RColorBrewer) # library for generating color palettes
library(googleVis) # library for presenting tables
```

## Locations of directories for data, metadata and results

```{r general_directories}
if (file.exists("/n/hsphS10/hsphfs1/chb/projects/ds_PTGIR_promoter/")) {
  baseDir <- "/n/hsphS10/hsphfs1/chb/projects/ds_PTGIR_promoter"
  }  else if (file.exists("/Volumes/home08/jhutchin/consults/ds_PTGIR_promoter")) {
    baseDir <- "/Volumes/home08/jhutchin/consults/ds_PTGIR_promoter"
    } else {
      baseDir <- "/Volumes/ody/consults/ds_PTGIR_promoter"
      }
dataDir <- file.path(baseDir, "data")
prostacylin_dataDir <- file.path(dataDir, "prostacylin_treated", "CEL_files")
normDev_dataDir <- file.path(dataDir, "normal_devo", "CEL_files")
extcontrol_dataDir <- file.path(dataDir, "extcontrol")
resultsDir <- file.path(baseDir, "results")
metaDir <- file.path(baseDir, "meta")
prostacylin_metaDir <- file.path(metaDir, "prostacylin_treated")
normDev_metaDir <- file.path(metaDir, "normal_devo")
extcontrol_metaDir <- file.path(metaDir, "extcontrol")
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")
dataDirs <- list(prostacylin=prostacylin_dataDir, normDev=normDev_dataDir,extcontrol=extcontrol_dataDir)
metaDirs <- list(prostacylin=prostacylin_metaDir, normDev=normDev_metaDir, extcontrol=extcontrol_metaDir)
````

----

# Microarray Analysis

## Load [Bioconductor][1] libraries for the microarray analysis 

```{r microarray_analysis_libraries}
library(virtualArray)
# to parse the CEL files and work with intensity values
library(affy) 
# for QC reports
library(arrayQualityMetrics)
# library to do stats 
library(limma) 
# pretty heatmaps
library(pheatmap) 
# transcript clusters <-> probes for mogene arrays
library(mogene10sttranscriptcluster.db)
library(mouse4302.db)
```

## Functions

```{r functions}
PCAplot <- function(eset=NULL, categories=NULL, title=NULL, colorpalette=NULL, alpha=1){
  alpha <- sprintf("%x", ceiling(alpha*255))
  colorpalette <- paste(colorpalette, alpha, sep="")
  eset.core <- exprs(eset) 
  myPca.core <- prcomp(t(eset.core))
  tmpPCAData.core <- as.data.frame(myPca.core$x[,1:4])
  pd <- pData(eset)
  colors <- colorpalette[factor(as.character(unlist(pd[,categories])))]
  legend_values=unique(cbind(colors, as.character(pd[,categories])))
  pairs(tmpPCAData.core, bg=colors, col="#606060", cex=2, pch=21, main=title, oma=c(8,5,5,14))
  legend("right", cex=0.7, col="#606060", pt.bg=legend_values[,1], pt.cex=1.5, legend=legend_values[,2],  pch=21, bty="n", x.intersp=1)
  }
rownames2col <- function(df, colname) {
  output <- cbind(row.names(df), df)
  colnames(output)[1] <- colname
  return(output)
  }
col2rownames <- function(df, colname, removecol=FALSE){
  row.names(df) <- df[,colname]
  if(removecol){df[,colname] <- NULL}
  return(df)
  }
```

## Data Import

### Datasets

#### HoxA9 Differentiated Time Course Dataset

#### Remodulin/Prostacyclin treated cells

#### External control dataset 
- bring in another completely different sample type on the same platform as the HoxA9 "normal" development samples
- want to be sure any differences we observe between the two datasets are real, and not just due to systematic differences between the arrays.

Used a Scadden dataset from GEO, GSE45430
Title:  Sox4 is a key oncogenic target in C/EBPα mutant Acute Myeloid Leukemia

```{r raw_import}
for(n in names(dataDirs)){
  covarfilename <- file.path(metaDirs[[n]], "covars.desc")
  covars <- read.delim(covarfilename)
  CELfiles <- file.path(dataDirs[[n]], as.vector(unlist(covars$CELfile)))
  mic.raw <- ReadAffy(filenames=CELfiles, phenoData=covars)
  sampleNames(mic.raw) <- pData(mic.raw)$sampleID
  assign(paste("mic.raw", n, sep="."), mic.raw)
  rm(mic.raw)
    }
```

## Raw Data QC

Initial Array quality control with [ArrayQualityMetrics][2]

### QC checks of raw data

```{r QC1, eval=F}
for(n in names(dataDirs)[1:2]){
  mic.raw <- get(paste("mic.raw", n, sep="."), mic.raw)
  arrayQualityMetrics(expressionset=mic.raw, outdir=file.path(resultsDir, "QCreport_raw", n), force=TRUE, don.logtransform=TRUE, intgroup="cell_line")
  rm(mic.raw)
  }
```

[QC report for prostacyclin treatment arrays]("./results/QCreport_raw/prostacyclin/index.html")
[QC report for HOXA9-ER arrays]("./results/QCreport_raw/normDev/index.html")

**Remove low quality arrays:**
*From prostacyclin data:*
377-3_1830F.CEL  failed RLE and NUSE tests.
*For HOXA9 normal development arrays:*
15_36h_1630D.CEL    failed both the RLE and NUSE tests.
15_4h_1630E.CEL failed 3 separate quality checks

HoxA9ER_63_0h also failed RLE but I decided to keep it for now, as its one of the samples that ties the experiments  together.

```{r filter_arrays}
filter.arrays <- c("377-3_1830F.CEL", "15_36h_1630D.CEL","15_4h_1630E.CEL")
for(n in names(dataDirs)[1:2]){
  mic.raw <- get(paste("mic.raw", n, sep="."))
  pd <- pData(mic.raw)
  mic.raw <- mic.raw[,which(!pd$CELfile %in% filter.arrays)]
  assign(paste("mic.raw", n, sep="."), mic.raw)
  rm(mic.raw, pd)
  }
```

## PCA Clustering - Raw Data, by study

```{r PCA1}
PCAplot(mic.raw.normDev, categories="cell_line", colorpalette=cbPalette[2:3], title="Raw Data - Normal Development - Cell lines")
PCAplot(mic.raw.normDev, categories="time", colorpalette=cbPalette, title="Raw Data - Normal Development - Time Points")

PCAplot(mic.raw.prostacylin, categories="cell_line", colorpalette=cbPalette[1:3], title="Raw Data - Remodulin Treatment - Cell lines")
PCAplot(mic.raw.prostacylin, categories="Treatment", colorpalette=cbPalette[2:3], title="Raw Data - Remodulin Treatment - Treatment")
```

Data looks pretty good, with decent clustering by the categories you'd expect.

Separately normalized data with RMA before continuing.

```{r RMA}
for(n in names(dataDirs)){
  mic.raw <- get(paste("mic.raw", n, sep="."))
  mic.RMA <- rma(mic.raw, normalize=T, background=T)
  assign(paste("mic.RMA", n, sep="."), mic.RMA)
  rm(mic.raw, mic.RMA)
  }
```

## PCA Clustering - Intra study normalized Data, by study

```{r PCA2}
PCAplot(mic.RMA.normDev, categories="cell_line", colorpalette=cbPalette[2:3], title="Normalized Data - Normal Development - Cell lines")
PCAplot(mic.RMA.normDev, categories="time", colorpalette=cbPalette, title="Normalized Data - Normal Development - Time Points")

PCAplot(mic.RMA.prostacylin, categories="cell_line", colorpalette=cbPalette[1:3], title="Normalized Data - Remodulin Treatment - Cell lines")
PCAplot(mic.RMA.prostacylin, categories="Treatment", colorpalette=cbPalette[2:3], title="Normalized Data - Remodulin Treatment - Treatment")
```

First we want to examine the levels of similarity between the gene expression values of the untreated samples for the HoxA9 time course and the Remodulin experiments. Ideally, the external control samples will be very different from the HoxA9 time course samples (which are on the same array) than the Remodulin experimental samples.

Use virtualArray library to reduce measurements to the set of genes common between the arrays. This will also have the effect of quantile normalizing across the different studies.

```{r virtualArray}
annotation(mic.RMA.prostacylin) <- "mogene10sttranscriptcluster"
esets <- paste("mic.RMA", names(dataDirs), sep=".")
virtualArrays <- virtualArrayExpressionSets(all_expression_sets=esets,removeBatcheffect = T )
# drop useless columns from pData
pData(virtualArrays)$Array.name <- NULL
pData(virtualArrays)$Covariate.1 <- NULL
# virtualArray seems to mess up sometimes and include some genes that are not actually on both arrays 
# remove these genes by finding and removing any rows that have NA's as expression values for any samples
eset.temp <- exprs(virtualArrays)
virtualArrays <- virtualArrays[which(!apply(eset.temp, 1, function(n) any(is.na(n)))),]
eset.temp <- exprs(virtualArrays)
eset.temp.m <- melt(eset.temp)
# confirm quantile normalization
ggplot(eset.temp.m, aes(x=value, color=X2)) + geom_density()

# merge in the pData from the datasets that virtualArray left behind
pd <- pData(virtualArrays)
pd2 <- merge(pd, pData(mic.raw.normDev), by.x="Sample.name", by.y="sampleID", all.x=T, all.y=T, sort=F)
pd3 <- merge(pd2, pData(mic.raw.prostacylin), by.x="Sample.name", by.y="sampleID", all.x=T, all.y=T, sort=F) 
pd3$CELfile = ifelse(is.na(pd3$CELfile.x),as.character(pd3$CELfile.y),as.character(pd3$CELfile.x))
pd3$CELfile.x <- NULL
pd3$CELfile.y <- NULL
pd3$cell_line = ifelse(is.na(pd3$cell_line.x),as.character(pd3$cell_line.y),as.character(pd3$cell_line.x))
pd3$cell_line.x <- NULL
pd3$cell_line.y <- NULL
pd3 <- pd3[order(pd3$Sample.name),]
virtualArrays <- virtualArrays[,order(pd$Sample.name)]
identical(pd3$Sample.name, pData(virtualArrays)$Sample.name)
pData(virtualArrays) <- pd3
rm(pd, pd2, pd3)
```

## PCA Clustering - Virtualarray Data

```{r PCA2}
PCAplot(virtualArrays, categories="Batch", colorpalette=cbPalette, title="Virtualarray - Studies")
PCAplot(virtualArrays[,which((pd$Treatment=="PBS" & pd$cell_line=="63HoxA9ER") | (pd$time==0 & pd$cell_line=="63HoxA9ER") | (pd$Batch=="mic.RMA.extcontrol" & grepl("Wildtype", pd$Sample.name)))], categories="Batch", colorpalette=cbPalette, title="Virtualarray - Studies")


PCAplot(virtualArrays, categories="cell_line", colorpalette=cbPalette, title="Virtualarray - Cell lines")
PCAplot(virtualArrays, categories="Treatment", colorpalette=cbPalette, title="Virtualarray - Remodulin Treatment")
```

PC1 v PC3 and PC1 v PC4 are concerning.
PC2 v PC4 and PC3 v PC4 look more like what we hope to see.

In any case, there are clear systematic platform differences in expression levels, would working with 

In any case, it looks like we would need to do a batch correction based on the zero time point/untreated samples before proceeding.

```{r combat}
library(sva)
 pheno = pData(virtualArrays)
  edata = exprs(virtualArrays)
  batch = pheno$Batch
  mod = model.matrix(~as.factor(cell_line), data=pheno)


## STOPPED here June.4.2013














```{r subset_data}
refined.metadata <- read.delim(file.path(metaDir,"unified-metadata-refined.04-17-13.tab"))
refined.metadata <- refined.metadata[which(!is.na(refined.metadata$age) & !is.na(refined.metadata$gender) & !is.na(refined.metadata$CEL_regex)),]
refined.metadata  <- refined.metadata[!(refined.metadata$study %in% c("GSE18897", "MEXP-2917", "TABM-940", "TABM-666")),]
refined.metadata  <- refined.metadata[refined.metadata$platform %in% c("GPL570", "A-AFFY-44", "GPL96", "A-AFFY-33", "GPL571", "GPL9188"),]
U133.meta <- refined.metadata
write.table(U133.meta, file.path(metaDir, "unified-metadata-refined_U133.tab"), quote=F, sep="\t", row.names=F, col.names=T)
write.table(U133.meta, file.path(metaDir, "unified-metadata-refined_U133.xls"), sep="\t", row.names=F, col.names=T)
```

That metadata file is [here](./meta/unified-metadata-refined_U133.xls)

## Setup

### Setup some more variables for the microarray analyses
- color palette for plots
- age ranges and names of the groups  
  - neonate = 0-4 years
  - child = 5-10 years
  - adult = 18-40 years

```{r microarray_analysis_variables}
# colorblind friendly palette
# age ranges to compare
child.age.range <- c(5,10)
adult.age.range <- c(18,40)
neonate.age.range <- c(0,4)
```

### Load and parse the unified, refined and subsetted metadata 
To make a covariate table that allows you to load the CEL files using the [affy][2] Bioconductor library
- this table has the CEL file location in the first column, without a header
- the other columns contain the sampleIDs, ages, genders,ethnicities and respective studies

- also limit samples to those wihtin the age ranges described above and label each with their "stage"

```{r covariatedataframe}
covartemplate.file <- "unified-metadata-refined_U133.tab" 
covartemplate <- read.table(file.path(metaDir,covartemplate.file ), header=T, colClasses="character", sep="\t")
# convert character ages to numeric to allow numeric subsetting 
covartemplate$age <- as.numeric(covartemplate$age) 
## limit samples to those within the described child, adult and neonate age ranges
covartemplate <- covartemplate[union(union(which(covartemplate$age<=max(child.age.range) & covartemplate$age>=min(child.age.range)), which(covartemplate$age<min(child.age.range))), which(covartemplate$age<=max(adult.age.range) & covartemplate$age>=min(adult.age.range))),] 
# convert age back to character value so you can use it in a character matrix
covartemplate$age <- as.character(covartemplate$age)
covartemplate <- as.matrix(covartemplate)
covars <- aaply(covartemplate, 1, function(n){
  # pull out the info you need piece by piece
  sampleID <- n[1] 
  age <- n[3]
  gender <- n[4]
  ethnicity=n[5]
  CELregex <- n[7]
  CELFTP <- n[6]
  study=n[9]
  platform=n[8]
  if(n[10]=="GEO"){
    # for GEO studies, get the CEL file name from the FTP locaiton in the metadata file
    CELfileloc <- file.path(dataDir,study, sub(".gz", "", basename(as.character(CELFTP)))) 
  } else {
    # for ArrayExpress studies, get the CEL file name directly from the CEL file regex in the metadata file
    CELfileloc <- file.path(dataDir, study, CELregex) 
    }
  if(!file.exists(CELfileloc)){
    # check if the CEL file is actually where its supposed to be, if not notify and move on
    print(paste(CELfileloc, "does not exist", sep=" ")) 
    break
    } else {
      # if CEL file is where its supposed to be, add location to covariate table
      return(list(CELfileloc=CELfileloc, ID=sampleID, age=age, gender=gender, ethnicity=ethnicity, study=study, platform=platform)) 
      }
  })
covars <- as.data.frame(covars)
covars$age <- as.numeric(covars$age)
# label samples with their stage as determined by age
# don't need to specify adult ranges, as we limited the dataset above to neonates, children and adults
covars$stage <- ifelse(covars$age<max(neonate.age.range), "neonate", ifelse(covars$age>=min(child.age.range) & covars$age<=max(child.age.range), "child", "adult")) 
```

### Load CEL files
- do this separately for each array type

```{r load_data}
GPL.groups <- list(U133.plus2=c("GPL570", "A-AFFY-44"), U133.A=c("GPL96", "A-AFFY-33"),U133.A.2=c("GPL571", "GPL9188"))
for(n in 1:length(GPL.groups)) {
  GPLs <- GPL.groups[[n]]
  covars.array <- covars[covars$platform %in% GPLs,]
  mic.raw <- ReadAffy(filenames=as.character(covars.array$CELfileloc), phenoData=covars.array)
  assign(paste("mic.raw", names(GPL.groups)[n], sep="."), mic.raw)
  rm(mic.raw)
}
```

### QC checks of raw data

```{r QC_raw, eval=FALSE}
for(n in 1:length(GPL.groups)){
  mic.raw <- get(paste("mic.raw", names(GPL.groups)[n], sep="."))
  arrayQualityMetrics(expressionset=mic.raw.temp, outdir=file.path(resultsDir, "QCreport_raw", names(GPL.groups)[n]), force=TRUE, do.logtransform=FALSE, intgroup=c("stage", "study"))
  rm(mic.raw)
  }
# had to log transform exprs outside of arrayQualityMetrics to finish the U133.A.2 analysis
```

[U133plus2 Raw Data QC Report](./results/QCreport_raw/U133.plus2/index.html)
[U133A Raw Data QC Report](./results/QCreport_raw/U133.A/index.html)
[U133A2 Raw Data QC Report](./results/QCreport_raw/U133.A.2/index.html)

### Outlier/Low quality array removal

For U133plus2, low quality studies and arrays have already been removed above (determined by the previous U133Plus2 analyses).

For U133A arrays, GSM768546 and GSM768549  (i.e. GSE31014 study) will be removed due to failed Relative Log Expression (RLE) outlier detection and Normalized Unscaled Standard Error (NUSE) tests. There also appear to be some visual artefacts on the GSM768549 array.

For U133A.2 arrays, all arrays from GSE17732 (GSM442648.CEL, GSM442649.CEL, GSM442650.CEL, GSM442651.CEL, GSM442652.CEL, GSM442653.CEL) failed both or either of the RLE and NUSE tests and will be removed. The individual arrays GSE17732.CEL, GSM474745.CEL, GSM474756.CEL, GSM539024.CEL, GSM539027.CEL, GSM539028.CEL, GSM758056.CEL, GSM758072.CEL also failed either of the RLE and NUSE tests and will be removed.

```{r filter_arrays}
filter_IDs <- list(U133.plus2="", U133.A=c("GSM768546","GSM768549"), U133.A.2=c("GSM442648", "GSM442649","GSM442650", "GSM442651", "GSM442652", "GSM442653", "GSE17732", "GSM474745", "GSM474756", "GSM539024", "GSM539027", "GSM539028", "GSM758056", "GSM758072"))
for (n in names(filter_IDs)){
  mic.raw <- get(paste("mic.raw", n, sep="."))
  mic.raw <- mic.raw[,which(!pData(mic.raw)$ID %in% filter_IDs[[n]])]
  assign(paste("mic.raw", n, sep="."), mic.raw)
  rm(mic.raw)
}
```

## Virtualarray
- subsets and combines different arrays by common gene symbols
- quantile normalizes data

```{r virtualarray}
# convert AffyBatch objects to ExpressionSet objects
# density plots of intensities
# also removed background
for (n in names(GPL.groups)){
  mic.raw <- mic.raw <- get(paste("mic.raw", n, sep="."))
  mic.raw.eset <- rma(mic.raw, verbose=F, normalize=F, background=T)
  sampleNames(mic.raw.eset) <- as.vector(unlist(pData(mic.raw.eset)$ID))
  assign(paste("mic.raw.eset", n, sep="."), mic.raw.eset)
  density(mic.raw.eset)
  rm(mic.raw, mic.raw.eset)
  }

# load into virtualarray
esets <- paste("mic.raw.eset", names(GPL.groups), sep=".")
virtualArrays <- virtualArrayExpressionSets(all_expression_sets=esets,removeBatcheffect = T )
# can't include additional stage covariates in batch removal because the U133.A dataset only contains one stage 
# without these additional stage covariates, the removeBatcheffect option makes no difference to the output, so left it as TRUE
pData(virtualArrays)$Array.name <- NULL
pData(virtualArrays)$Covariate.1 <- NULL
# setup pData
# setup sampleinfo dataframe
sampleinfo.df <- ldply(names(GPL.groups), function(n) {
  mic.raw.eset <- get(paste("mic.raw.eset", n, sep="."))
  pData(mic.raw.eset)$Batch <- n
  pData(mic.raw.eset)$CELfileloc <- NULL
  return(pData(mic.raw.eset))
  })
sampleinfo.df <- merge(pData(virtualArrays), sampleinfo.df, by.y="ID", by.x="Sample.name", sort=F)
names(sampleinfo.df) <- sub("Sample.name", "ID", names(sampleinfo.df))
row.names(sampleinfo.df) <- sampleinfo.df$ID
identical(as.character(unlist(pData(virtualArrays)$Sample.name)), as.character(unlist(sampleinfo.df$ID))) #check to make sure order is matched to current pData
identical(row.names(sampleinfo.df), sampleNames(virtualArrays)) #check to make sure order is matched to current pData
density(virtualArrays)
pData(virtualArrays) <- sampleinfo.df
validObject(virtualArrays) # secondary sanity check
pd <- pData(virtualArrays)
```

### QC checks on combined data with [ArrayQualityMetrics][3]

```{r virtualArrayQC, eval=FALSE}
arrayQualityMetrics(expressionset=virtualArrays, outdir=file.path(resultsDir, "QCreport_virtualarrays"), force=T, do.logtransform=F, intgroup=c("stage", "study"))
````

[Post combined and normalized QC Report](./results/QCreport_virtualarrays/index.html)

### Metadata for remaining studies
**Number of U133Plus2.0 arrays:**
Adult - `r length(which(pd$Batch.y=="U133.plus2" & pd$stage=="adult"))`
Child - `r length(which(pd$Batch.y=="U133.plus2" & pd$stage=="child"))`
Neonate - `r length(which(pd$Batch.y=="U133.plus2" & pd$stage=="neonate"))`
**Number of U133.A.2 arrays:**
Adult - `r length(which(pd$Batch.y=="U133.A.2" & pd$stage=="adult"))`
Child - `r length(which(pd$Batch.y=="U133.A.2" & pd$stage=="child"))`
Neonate - `r length(which(pd$Batch.y=="U133.A.2" & pd$stage=="neonate"))`
**Number of U133.A arrays:**
Adult - `r length(which(pd$Batch.y=="U133.A" & pd$stage=="adult"))`
Child - `r length(which(pd$Batch.y=="U133.A" & pd$stage=="child"))`
Neonate - `r length(which(pd$Batch.y=="U133.A" & pd$stage=="neonate"))`

```{r print_metadata, results='asis'}
pd <- pData(virtualArrays)
pd.gvis <- gvisTable(as.data.frame(apply(pd, 2, as.character)), options=list(width=960))  
print(pd.gvis, "chart")
```

### Primary Componenet Analysis Clustering
- the goal of this analysis is to blindly evaluate the variability within the normalized data and determine whether this variability could predict the different sample categories by itself
- in basic terms, we are checking whether the samples cluster by sample type
  - its a good sign if they do

```{r virtualarray.pca, fig.width=16, fig.height=16}
## All samples
## by platform
PCAplot(virtualArrays, categories="Batch.y", title="VirtualArrays - no batch correction", colorpalette=cbPalette, alpha=0.7) 
```

Very clearly clustering by platform type. Try batch correcting.

## BATCH CORRECTION

Correct for batch effects with ComBat from the [sva][4] package:

```{r batchcorrect, cache=TRUE}
library(sva)
# Create model with batch as factor variable
mod <- model.matrix(~as.factor(stage), data=pd)
batch <- pData(virtualArrays)$Batch.y
# Modify expression matrix
eset.core <- exprs(virtualArrays)
combat_edata <- ComBat(dat=eset.core,
                       batch=batch,
                       mod=mod,
                       numCovs=NULL,
                       par.prior=TRUE, 
                       prior.plots=TRUE)
virtualArrays.batch <- virtualArrays
exprs(virtualArrays.batch) <- combat_edata

eset.batch <- exprs(virtualArrays.batch)
colnames(eset.batch) <- pData(virtualArrays.batch)$ID
pd <- pData(virtualArrays.batch)
```

### Post Batch correction PCA

```{r virtualarray.PCA2, fig.height=18, fig.width=18}
PCAplot(virtualArrays.batch, categories="Batch.y", title="VirtualArrays - batch corrected, by platform", colorpalette=cbPalette, alpha=0.7)
PCAplot(virtualArrays.batch, categories="stage", title="VirtualArrays - batch corrected, by stage", colorpalette=cbPalette[5:7], alpha=0.7)
PCAplot(virtualArrays.batch, categories="study", title="VirtualArrays - batch corrected, by study", colorpalette=sort(c(brewer.pal(8,"Set1"),brewer.pal(8,"Set2"))), alpha=0.7)
```

These look much better. Still clustering by study, but with the number of single stage studies, it is difficult to conclusively say whether this is study or stage driven.

## Differential Expression Analyses

### Limma

A linear model for microarray data analysis [Limma][5] was performed on the samples to identify differentially expressed genes between all pair-wise comparisons of the sample groups. Limma fits a linear model to the expression data for all samples for each gene and is designed to handle complex experiments involving comparisons between many RNA targets simultaneously.

To perform limma, we construct two matrices. The design matrix provides a representation of the different sample groups which have been analysed. The contrast matrix allows the coefficients defined by the design matrix to be combined into contrasts of interest. Each contrast corresponds to a pairwise comparison between sample groups.

#### Design matrix 
- make a matrix with arrays as rows, sample groups as columns
- a one or a zero indicate respectively, that a sample either belongs or does not belong to the sample group 

```{r design, results="asis"}
design <- model.matrix(~ -1+factor(pData(virtualArrays)$stage))
# make sure the headings match
colnames(design) <- sub("factor\\(pData\\(virtualArrays\\)\\$stage\\)", "", colnames(design))
design.gvis <- gvisTable(as.data.frame(apply(design, 2, as.character)), options=list(width=240))  
print(design.gvis, "chart")
```

#### Contrast matrix
- to perform specified pairwise comparisons
- in this table, columns are contrasts/comparisons and rows are sample groups
-  a zero denotes that the sample group is not involved in the contrast, a 1 denotes that it has higher expression in the contrast and a -1 denotes lower expression in the contrast

```{r contrastmatrix, results='asis'}
contrast.matrix <- makeContrasts(adult-child,adult-neonate,neonate-child, levels=dimnames(design)[[2]])
contrast.gvis <- gvisTable(as.data.frame(apply(contrast.matrix, 2, as.character)), options=list(width=240, height=120))  
print(contrast.gvis, "chart")
```

These matrices are used to fit a linear model to the data. The linear model is applied and pairwise comparisons are performed to identify differentially expressed genes.

- first fit the linear model based on the design matrix for each gene based on the given series of arrays
- using the contrast matrix, compute estimated coefficients and standard errors for contrasts
- compute moderated t-statistics and log-odds of differential expression by empirical Bayes shrinkage of the standard errors towards a common value

#### Fit a linear model
- for each gene based on the given series of arrays

```{r linearmodel}
exprs.norm <- exprs(virtualArrays)
dimnames(exprs.norm)[[2]] <- as.character(pData(virtualArrays)$ID)
fit.exprs <- lmFit(exprs.norm, design) 
```

#### Compute estimated coefficients and standard errors for contrasts

```{r contrastfit}
fit2.exprs <- contrasts.fit(fit.exprs, contrast.matrix) 
```

#### Compute moderated t-statistics and log-odds of differential expression 
- by empirical Bayes shrinkage of the standard errors towards a common value

```{r bayes}
fit2.exprs <- eBayes(fit2.exprs) 
```

### Obtain top results
- generate a list of top 100 differentially expressed genes with at least 2 fold changes in expression and an adjusted pvalue <0.1,  as sorted by unadjusted pvalue

#### The summary table contains the following information: 
- logFC is the log2-fold change
- the AveExpr is the average expression value accross all arrays
- the moderated t-statistic (t) is the logFC to its standard error, the P.Value is the associated p-value
- the adj.P.Value is the p-value adjusted for multiple testing (by FDR) 
- the B-value (B) is the log-odds that a gene is differentially expressed (the-higher-the-better)

```{r virtualarray.toptable, results='hide'}
top.stats <- lapply(seq(1,3,1), function(n) {
  contrast <- dimnames(fit2.exprs$contrasts)$Contrasts[n]
  stats.top <- topTable(fit2.exprs, coef=n, adjust="fdr", sort.by="p", p.value=0.1, lfc=1, number=100, genelist=row.names(exprs.norm)) 
  # expression levels for these genes in all samples
  if(nrow(stats.top)>0) {
    exprs.norm.sub  <-  exprs.norm[stats.top[, "ID"], ] 
    stats.exprs.norm.top <- cbind(stats.top, exprs.norm.sub)
    } else { stats.exprs.norm.top <-"NA"}
  return(list(data=stats.exprs.norm.top, contrast=contrast))
  })
# write stats and expression levels to file
lapply(top.stats, function(n) {
  contrast <- gsub(" ", "",  n$contrast)
  output <- n$data
  write.table(output, file.path(resultsDir, paste("U133.virtualarray.top100.DE.genes", contrast, "xls", sep=".")), sep="\t", row.names=F, col.names=T)
  })
```

#### Excel files with statistics and normalized sample intensities for the top 100 differentially expressed probesets for each comparison

`r index=1`

[`r toupper(top.stats[[index]]$contrast)` comparison](./results/`r paste("U133.virtualarray.top100.DE.genes", top.stats[[index]]$contrast, "xls",sep=".")`)
`r index=index+1`

[`r toupper(top.stats[[index]]$contrast)` comparison](./results/`r paste("U133.virtualarray.top100.DE.genes", top.stats[[index]]$contrast, "xls",sep=".")`)
`r index=index+1`

[`r toupper(top.stats[[index]]$contrast)` comparison](./results/`r paste("U133.virtualarray.top100.DE.genes", top.stats[[index]]$contrast, "xls",sep=".")`)
`r index=index+1`

### Heatmaps of top 100 differentially expressed genes for each pairwise comparison
- there are three heatmaps for each comparison
  - both genders
  - female only
  - male only

*note that the differentially expressed genes clustered in these heatmaps are based on comparisons that include both genders* 

```{r virtualarray.heatmaps, fig.width=24, fig.height=24}
for(n in 1:3){
  # get the names of the stages that are being compared in this comparison
  contrast <- top.stats[[n]]$contrast
  stages <- names(which(abs(contrast.matrix[,contrast])>0))
  # subset the samples and the data to those in the stages compared in this comparison
  sample.indices <- which(design[,stages[1]]>0 | design[,stages[2]]>0)
  sample.pd <- pData(virtualArrays)[sample.indices,]
  top.exprs.contrast <- top.stats[[n]]$data
  row.names(top.exprs.contrast) <- top.exprs.contrast$ID
  top.exprs.contrast <- top.exprs.contrast[,as.character(unlist(sample.pd$ID))]
    # heatmap annotations
  heatmap.annots <- sample.pd[,c("ID", "study", "stage", "gender")]
  heatmap.annots <- as.data.frame(apply(heatmap.annots, 2, unlist))
  heatmap.annots <- col2rownames(heatmap.annots, "ID", removecol=T)
  # heatmap annotation colors
  study_colors <- c(brewer.pal(8, "Set1"), brewer.pal(8, "Set2"))
  names(study_colors) <- unique(unlist(pd$study))
  stage_colors <- c("white", "darkgrey", "black")
  names(stage_colors) <- unique(unlist(pd$stage))
  gender_colors <- c("cyan", "pink")
  names(gender_colors) <- unique(unlist(pd$gender))
  ann_colors = list(study = study_colors, stage = stage_colors, gender=gender_colors)
  ## Heatmaps
  # ALL genders 
  pheatmap(as.matrix(top.exprs.contrast), annotation=heatmap.annots, color=rev(brewer.pal(11,"RdBu")), main=contrast, show_colnames=F, fontsize=24, fontsize_row=10,annotation_colors=ann_colors)  
  # FEMALE gender  
  top.exprs.contrast.female <- top.exprs.contrast[,which(sample.pd$gender=="FEMALE")]
  pheatmap(as.matrix(top.exprs.contrast.female), annotation=subset(heatmap.annots,gender=="FEMALE"), color=rev(brewer.pal(11,"RdBu")), main=paste(contrast,"(FEMALE)", sep=" "), show_colnames=F,fontsize=24, fontsize_row=10,annotation_colors=ann_colors)  
  # MALE gender  
  top.exprs.contrast.male <- top.exprs.contrast[,which(sample.pd$gender=="MALE")]
  pheatmap(as.matrix(top.exprs.contrast.male), annotation=subset(heatmap.annots,gender="MALE"), color=rev(brewer.pal(11,"RdBu")), main=paste(contrast,"(MALE)", sep=" "), show_colnames=F, fontsize=24,fontsize_row=10,annotation_colors=ann_colors)  
  }
```

[1]: http://www.bioconductor.org (BioC)
[2]: http://www.bioconductor.org/packages/release/bioc/html/affy.html (affy)
[3]: http://www.bioconductor.org/packages/2.10/bioc/html/arrayQualityMetrics.html (arrayQualityMetrics)
[4]: http://www.bioconductor.org/packages/release/bioc/html/sva.html (sva)
[5]: http://www.bioconductor.org/packages/release/bioc/html/limma.html (limma)

