```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, cache=FALSE,  highlight=TRUE, figalign="center", 
               echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, 
               tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120), 
               fig.path="figure/WB-pathprint_multi-", 
               cache.path="cache/WB-pathprint_multi-")
options(width=200)
```

# PATHPRINT ANALYSES OF STUDIES ON MULTIPLE PLATFORMS
- using GEO normalized data

---

Analysis of public datasets for differential gene expression signatures between
adults and children for [Les Kobzik](mailto:LKOBZIK@hsph.harvard.edu) as part of
the DARPA 7-day Biodefense Program.

Contact [John Hutchinson](mailto:jhutchin@hsph.harvard.edu) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

----

# GENERAL SETUP

## General purpose R libraries necessary for analysis

```{r general_libraries, results='hide'}
source("http://bioconductor.org/biocLite.R") # BioConductor script necessary for installing new BioC libraries with biocLite()
library(plyr) # library for iteratively working with data structures
library(ggplot2) # library for plotting 
library(RColorBrewer) # library for generating color palettes
library(googleVis) # library for presenting tables
source("http://dl.dropboxusercontent.com/u/4253254/Resources/functions.r")
#library(devtools) # install libraries from github
#install_git("git://github.com/hbc/CHBUtils.git") # misc personal utilities
#library(dplyr)
```

## Locations of directories for data, metadata and results

```{r general_directories}
if (file.exists("/n/hsphS10/hsphfs1/chb/projects/lk_FOY/sepsis")) {
  baseDir <- "/n/hsphS10/hsphfs1/chb/projects/lk_FOY/sepsis"
  }  else if (file.exists("/Volumes/home08/jhutchin/consults/lk_FOY/sepsis")) {
    baseDir <- "/Volumes/home08/jhutchin/consults/lk_FOY/sepsis"
    } else {
      baseDir <- "/Volumes/ody/consults/lk_FOY/sepsis"
      }
dataDir <- file.path(baseDir, "data", "WB")
resultsDir <- file.path(baseDir, "results", "WB", "U133Plus2")
metaDir <- file.path(baseDir, "meta", "WB")
````

----

# Pathprint Analysis

## Setup

### Load libraries for Pathprint analysis and visualization

```{r pathprint_libraries, results='hide'}
library(GEOquery)
library(ArrayExpress)
library(pathprint)
library(limma)
library(ape)
library(pheatmap)
library(affy)
```

### Setup some more variables for the  analyses

```{r microarray_analysis_variables}
# colorblind friendly palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")
# age ranges to compare
child.age.range <- c(5,19)
adult.age.range <- c(20,60)
neonate.age.range <- c(0,4)
if (any(c(any(seq(adult.age.range[1], adult.age.range[2]) %in% seq(child.age.range[1], child.age.range[2])), 
any(seq(adult.age.range[1], adult.age.range[2]) %in% seq(neonate.age.range[1], neonate.age.range[2])),
any(seq(neonate.age.range[1], neonate.age.range[2]) %in% seq(child.age.range[1], child.age.range[2]))))) {
  stop("age ranges cannot overlap")
}
```

### Age ranges and names of the groups  
  - neonate = `r neonate.age.range[1]`-`r neonate.age.range[2]` years
  - child =`r child.age.range[1]`-`r child.age.range[2]` years
  - adult = `r adult.age.range[1]`-`r adult.age.range[2]` years

## Subset Data
- subsetted the unified metadata file to this platforms that are actually covered by Pathprint
- subsetted to samples for which we have ages

```{r subset_data}
refined.metadata <- read.delim(file.path(metaDir,"unified.metadata.refined.tab"))
refined.metadata <- refined.metadata[!is.na(refined.metadata$age),]
refined.metadata <- refined.metadata[refined.metadata$platform %in% names(platform.thresholds),]

write.table(refined.metadata, 
            file.path(metaDir, "unified.metadata.refined.pathprint_plaforms.tab"), 
            quote=F, 
            sep="\t", 
            row.names=F, 
            col.names=T)
```

That revised metadata file is [here](../meta/WB/unified.metadata.refined.pathprint_plaforms.tab)

### Parse the unified, refined and subsetted metadata file 
- to make a covariate table that allows you to download the data using the [GEOquery][2] Bioconductor library
- limit samples to those within the age ranges described above and label each with their "stage"

```{r covariatedataframe}
covars <- refined.metadata
# convert character ages to numeric to allow numeric subsetting 
covars$age <- as.numeric(as.character(covars$age))

## limit samples to those within the described child, adult and neonate age ranges
children.indices <- which(covars$age<=max(child.age.range) & 
  covars$age>=min(child.age.range))
neonate.indices <- which(covars$age<=max(neonate.age.range))
adult.indices <- which(covars$age<=max(adult.age.range) & 
  covars$age>=min(adult.age.range))
sample.indices <- c(neonate.indices, children.indices, adult.indices)
covars <- covars[sample.indices,]

# label samples with their stage as determined by age
# don't need to specify adult ranges, as we limited the dataset above to neonates, children and adults
covars$stage <- ifelse(covars$age<=max(neonate.age.range), "neonate", 
                       ifelse(covars$age>=min(child.age.range)
                              & covars$age<=max(child.age.range), "child", 
                              ifelse(covars$age>=min(adult.age.range)
                                     & covars$age<=max(adult.age.range), "adult", NA)))
covars <- covars[order(covars$stage),]
#remove unclassified samples
covars <- covars[!is.na(covars$stage),]

write.table(covars, 
            file.path(metaDir, "unified.metadata.refined.pathprint_plaforms.targetted.samples.tab"), 
            quote=F, 
            sep="\t", 
            row.names=F, 
            col.names=T)

```

That revised metadata file is [here](../meta/WB/"unified.metadata.refined.pathprint_plaforms.targetted.samples.tab)

### Download Expression data from GEO
- for each GEO study
  - pull down a study's expression data
    - subset study's expression data to samples of interest (using covars)
    - pathprint study
- combine pathprinted data

GSE13015 and GSE3284 have more than one platform per experiment, so need to handle those instances
Also processed GSE54514 separately for now (data not into GEO as of `r date()`).

```{r pathprintdata, results='hide'}
pathprint.GPLs <- names(platform.thresholds)
covars.GEO <- covars[which(covars$database=="GEO"),]

fingerprints.GEO.l <- lapply(unique(covars.GEO$study), function(GSE){
  print(GSE)
  GEOdata <- getGEO(GSE, GSEMatrix=TRUE)

  #GSE13015 and GSE3284 have more than one platform (GPL) per GSE, so need to handle those
  # if more than one platform in GSE, subset to platform that is covered by pathprint
  if (length(GEOdata)==1) {# only one GPL within the GSE
    GEOdata <- GEOdata[[1]]
    
    } else { # more than one platform in the GSE
      GPLs <-  sapply(GEOdata, function(study) annotation(study))
      GPL.in.pathprint <- which(GPLs %in% pathprint.GPLs) 
      
      if(length(GPL.in.pathprint)==1){ # more than one platform in GSE but only one covered by pathprint
        GEOdata <- GEOdata[[GPL.in.pathprint]]
        
        } else if (length(GPL.in.pathprint)==0) {
          stop("no valid pathprint platforms in this study")
          
          } else { # more than one platform covered by pathprint in the GSE 
            # which GSE indexes (platforms) have samples you want?
            GPLS.with.samples <- which(unlist(lapply(GEOdata, function(study){
              any(pData(study)$geo_accession %in% covars.GEO$arrayID)
              })))
            
            if (length(GPLS.with.samples)>1){ # samples split across more than one pathprinted platform            
              stop("more than one valid pathprinted platform found in this GSE that you want samples from; script can only handle one at a time")
              } else { # samples are only on one platform
                GEOdata <- GEOdata[[GPLS.with.samples]]
                }
            }
      }
  
  # get intensity values
  eset <- exprs(GEOdata)
  # subset to our samples of interest
  eset <- eset[,as.character(covars.GEO$arrayID[covars.GEO$study %in% GSE])]
  # write data to file for Les
  write.table(eset, file=file.path(dataDir, GSE, paste(GSE, "GEO.eset.tab", sep=".")))
  # run pathprint on subsetted data
  mic.platform <- annotation(GEOdata)
  mic.species <- "human"
  mic.names <- names(eset)
  mic.f.default <- exprs2fingerprint(exprs = eset, platform = mic.platform, species = mic.species, progressBar = FALSE)
  return(mic.f.default)
  })

fingerprints.GEO <- do.call(cbind, fingerprints.GEO.l)
```

### Download Expression data from ArrayExpress and Pathprint
- for the single ArrayExpress (AE) study
  - pull down study's raw expression data (as an AffyBatch)
    - RMA normalize into an ExpressionSet
    - subset study's expression data to samples of interest (using covars)
    - pathprint study
- combine pathprinted data

```{r pathprintAE, results='hide'}
covars.AE <- covars[which(covars$database=="AE"),]
fingerprints.AE.l <- lapply(unique(covars.AE$study), function(AEid){
    ae.dataDir=file.path(dataDir, AEid)
  if (!(file.exists(ae.dataDir))) {
    dir.create(ae.dataDir)
  }
  AEdata <-   try(ArrayExpress(accession=paste("E", AEid, sep="-"), path=ae.dataDir, save=TRUE))
  AEdata <- rma(AEdata)
  # get intensity values
  eset.AE <- as.data.frame(exprs(AEdata))
  # subset to our samples of interest
  eset.AE <- eset.AE[,as.character(covars.AE$arrayID[covars.AE$study %in% AEid])]
  # run pathprint on subsetted data
  mic.platform.AE <- as.character(unique(covars.AE$platform[covars.AE$study %in% AEid]))
  mic.species <- "human"
  mic.names <- names(eset.AE)
  mic.f.default <- exprs2fingerprint(exprs = eset.AE, platform = mic.platform.AE, species = mic.species, progressBar = FALSE)
  return(mic.f.default)
  })

fingerprints.AE <- do.call(cbind, fingerprints.AE.l)
```

### Normalize data from BD and Pathprint

```{r pathprintBD, results='hide'}
covars.BD <- covars[which(covars$database=="BD"),]
row.names(covars.BD) <- covars.BD$arrayID
celfiles.BD <- as.character(covars.BD$CEL_regex)
# load CEL files 
affyraw.BD <- ReadAffy(filenames=file.path(dataDir, "BD", celfiles.BD), phenoData=covars.BD)
# change names to match covars$arrayID
sampleNames(affyraw.BD) <- as.character(covars.BD$arrayID)
# normalize
affynorm.BD  <- affy::rma(affyraw.BD)
# extract intensities
eset.BD <- exprs(affynorm.BD)
# fingerprint
mic.platform.BD <- as.character(unique(covars.BD$platform))
mic.species <- "human"
mic.names <- names(eset.BD)
fingerprints.BD <- exprs2fingerprint(exprs = eset.BD, platform = mic.platform.BD, species = mic.species, progressBar = FALSE)
```

#### Combine Pathprints of all samples

```{r unify}
fingerprints <- do.call(cbind, list(fingerprints.GEO, fingerprints.AE, fingerprints.BD))
save(list="fingerprints",file=file.path(resultsDir, "pathprints.RData"))
```

```{r loaddata, eval=FALSE, echo=FALSE}
## hack to get around caching issues with knitr
load(file.path(resultsDir, "pathprints.RData"))
```

## Default Pathway Differentiation

Find pathways with minimal intra-group variation and maximal inter-group variation.

```{r pathwaydiff_vars}
# cutoffs used to filter the data
ttest.cutoff=1e-10
sd.cutoff=0.5
```

For this, I set two filtering criteria:
1) maximize homogeneity within an age group - based on minimizing the standard deviation (as I have no idea how to calculate a trinary entropy), cutoff used was sd<`r sd.cutoff`
2) maximize differences between groups - using ttests, cutoff used here was p<`r ttest.cutoff`

### Setup design and contrast matrices 

```{r pathprint_design}
# reorder the fingerprints and metadata about the samples
fingerprints <- fingerprints[,order(colnames(fingerprints))]
pd <- covars[order(covars$arrayID),]

if(!identical(as.character(pd$arrayID),colnames(fingerprints))){
  stop("covariate and fingerprint samples are not identically ordered")
}
design <- model.matrix(~ 0 + stage,  data=pd)
# tweak the headings
colnames(design) <- sub("stage", "", colnames(design))

contrast.matrix <- makeContrasts(adult-child,adult-neonate,neonate-child, levels=colnames(design))
```

### Heatmaps
- clustered by both column (sample) and row (pathway)

#### Setup annotations

```{r pathprint_heatmap_colors}
# annotation colors
study_colors <- rep(cbPalette,2)[1:18]
names(study_colors) <- unique(unlist(pd$study))
stage_colors <- c("white", "darkgrey", "black")
names(stage_colors) <- unique(unlist(pd$stage))
gender_colors <- c("blue", "pink", "yellow")
genders <-as.vector(unique(unlist(pd$gender)))
genders[is.na(genders)] <- "unknown"
names(gender_colors) <- unique(unlist(genders))
ann_colors = list(study = study_colors, stage = stage_colors, gender = gender_colors)
```

#### Make comparisons and plot

```{r pathprint_default_contrasts, fig.width=12, fig.height=12}
# for each comparison
for (n in 1:3) {
  contrast <- dimnames(contrast.matrix)$Contrasts[n]
  contrast <- gsub(" ", "", contrast)
  stage.1 <- unlist(strsplit(contrast, "-"))[1]
  stage.2 <- unlist(strsplit(contrast, "-"))[2]
  # get subset of barcoded data and stages
  fingerprints.contrast <- fingerprints[,pd$stage %in% c(stage.1, stage.2)]
  pd.contrast <- pd[pd$stage %in% c(stage.1, stage.2),] 
  contrast.factor <- as.factor(pd.contrast$stage)
  sds.stage = adply(fingerprints.contrast, 1, function(x) {
    by(x, contrast.factor, function(x) sd(x))
    })
  row.names(sds.stage) <- sds.stage$X1
  sds.stage$X1 <- NULL
  sds.stage <- as.matrix(sds.stage)
  mode(sds.stage) <- "numeric"
  # t-tests between groups
  ttestpvals.contrast <- apply(fingerprints.contrast, 1, function(x) {
    try(t.test(x~contrast.factor)$p.value)
    })
  ttestpvals.contrast<-as.numeric(ttestpvals.contrast)
  
  # fisher's exact tests
  #   getcounts <- function(x){
  #     table(factor(x,lev=c(-1,0,1)))
  #     }
  #   fisherpvals.contrast <- apply(fingerprints.contrast, 1, function(x){
  #     counts.df <- aggregate(x, by=list(stage=pd.contrast$stage), getcounts, simplify=T)
  #     counts.df$stage <- NULL
  #     pval <- fisher.test(counts.df)$p.value
  #     return(pval)
  #     })
  
  # filter genes by intra-group variation and across group differences
  fingerprints.contrast.sub <- fingerprints.contrast[sds.stage[,stage.1]<=sd.cutoff  & sds.stage[,stage.2]<=sd.cutoff & ttestpvals.contrast<ttest.cutoff &  !is.na(ttestpvals.contrast),]
  #   fingerprints.contrast.sub <- fingerprints.contrast[sds.stage[,stage.1]<=sd.cutoff  & sds.stage[,stage.2]<=sd.cutoff & fisherpvals.contrast<ttest.cutoff &  !is.na(fisherpvals.contrast),]
  
  # remove any columsn with only NAs
  fingerprints.contrast.sub <- fingerprints.contrast.sub[!apply(fingerprints.contrast.sub, 1, function(n) all(is.na(n))),]
  dim(fingerprints.contrast.sub)
  
  ## Heatmaps
  # annotations for heatmaps
  heatmap.annots <- pd.contrast[,c("arrayID", "study", "stage", "gender")]
  heatmap.annots <- as.data.frame(apply(heatmap.annots, 2, unlist))
  row.names(heatmap.annots) <- heatmap.annots$arrayID
  heatmap.annots$arrayID <- NULL
  # getting the annotations and annotation colors to play nice
  heatmap.annots <- as.matrix(heatmap.annots)
  
  heatmap.annots[,"gender"][is.na(as.vector(heatmap.annots[,"gender"]))] <- "unknown"
  
  # both genders
  if(nrow(fingerprints.contrast.sub)>0){
    pheatmap(fingerprints.contrast.sub, color=c("blue", "white", "red")    , cluster_rows=T, cluster_cols=T, show_colnames=F, annotation=heatmap.annots,  legend_breaks=c(-1,0,1),main=paste("Pathways for ", contrast, "comparison - clustered",  sep=" "), fontsize=14, fontsize_row=10, annotation_colors=ann_colors)
    }
  }
```

### Remove Trauma studies

```{r removetrauma}
trauma.studies <- c("GSE5580","GSE11375","GSE36809")
pd <- pd[!(pd$study %in% trauma.studies),]
fingerprints <- fingerprints[,colnames(fingerprints) %in% pd$arrayID]
```

#### Setup annotations again

```{r pathprint_heatmap_colors2}
# annotation colors
study_colors <- rep(cbPalette,2)[1:15]
names(study_colors) <- unique(unlist(pd$study))
stage_colors <- c("white", "darkgrey", "black")
names(stage_colors) <- unique(unlist(pd$stage))
gender_colors <- c("blue", "pink", "yellow")
genders <-as.vector(unique(unlist(pd$gender)))
genders[is.na(genders)] <- "unknown"
names(gender_colors) <- unique(unlist(genders))
ann_colors = list(study = study_colors, stage = stage_colors, gender = gender_colors)
```

#### Make comparisons and plot

```{r pathprint_default_contrasts2, ,fig.width=12, fig.height=12, warning=FALSE}
# for each comparison
for (n in 1:3) {
  contrast <- dimnames(contrast.matrix)$Contrasts[n]
  contrast <- gsub(" ", "", contrast)
  stage.1 <- unlist(strsplit(contrast, "-"))[1]
  stage.2 <- unlist(strsplit(contrast, "-"))[2]
  # get subset of barcoded data and stages
  fingerprints.contrast <- fingerprints[,pd$stage %in% c(stage.1, stage.2)]
  pd.contrast <- pd[pd$stage %in% c(stage.1, stage.2),] 
  contrast.factor <- as.factor(pd.contrast$stage)
  sds.stage = adply(fingerprints.contrast, 1, function(x) {
    by(x, contrast.factor, function(x) sd(x))
    })
  row.names(sds.stage) <- sds.stage$X1
  sds.stage$X1 <- NULL
  sds.stage <- as.matrix(sds.stage)
  mode(sds.stage) <- "numeric"
  # t-tests between groups
  ttestpvals.contrast <- apply(fingerprints.contrast, 1, function(x) {
    try(t.test(x~contrast.factor)$p.value)
    })
  ttestpvals.contrast<-as.numeric(ttestpvals.contrast)
  
  # filter genes by intra-group variation and across group differences
  fingerprints.contrast.sub <- fingerprints.contrast[sds.stage[,stage.1]<=sd.cutoff  & sds.stage[,stage.2]<=sd.cutoff & ttestpvals.contrast<ttest.cutoff &  !is.na(ttestpvals.contrast),]
  
  # remove any columsn with only NAs
  fingerprints.contrast.sub <- fingerprints.contrast.sub[!apply(fingerprints.contrast.sub, 1, function(n) all(is.na(n))),]
  dim(fingerprints.contrast.sub)
  
  ## Heatmaps
  # annotations for heatmaps
  heatmap.annots <- pd.contrast[,c("arrayID", "study", "stage", "gender")]
  heatmap.annots <- as.data.frame(apply(heatmap.annots, 2, unlist))
  row.names(heatmap.annots) <- heatmap.annots$arrayID
  heatmap.annots$arrayID <- NULL
  # getting the annotations and annotation colors to play nice
  heatmap.annots <- as.matrix(heatmap.annots)
  
  heatmap.annots[,"gender"][is.na(as.vector(heatmap.annots[,"gender"]))] <- "unknown"
  
  # both genders
  if(nrow(fingerprints.contrast.sub)>0){
    pheatmap(fingerprints.contrast.sub, color=c("blue", "white", "red")    , cluster_rows=T, cluster_cols=T, show_colnames=F, annotation=heatmap.annots,  legend_breaks=c(-1,0,1),main=paste("Pathways for ", contrast, "comparison - clustered",  sep=" "), fontsize=14, fontsize_row=10, annotation_colors=ann_colors)
    }
  }
```


### Remove GSE54514


```{r removeGSE54514}
pd <- pd[!(pd$study %in% "GSE54514"),]
fingerprints <- fingerprints[,colnames(fingerprints) %in% pd$arrayID]
```

#### Setup annotations again

```{r pathprint_heatmap_colors3}
# annotation colors
study_colors <- rep(cbPalette,2)[1:14]
names(study_colors) <- unique(unlist(pd$study))
stage_colors <- c("white", "darkgrey", "black")
names(stage_colors) <- unique(unlist(pd$stage))
gender_colors <- c("blue", "pink", "yellow")
genders <-as.vector(unique(unlist(pd$gender)))
genders[is.na(genders)] <- "unknown"
names(gender_colors) <- unique(unlist(genders))
ann_colors = list(study = study_colors, stage = stage_colors, gender = gender_colors)
```

#### Make comparisons and plot

```{r pathprint_default_contrasts3, ,fig.width=12, fig.height=12, warning=FALSE}
# for each comparison
for (n in 1:3) {
  contrast <- dimnames(contrast.matrix)$Contrasts[n]
  contrast <- gsub(" ", "", contrast)
  stage.1 <- unlist(strsplit(contrast, "-"))[1]
  stage.2 <- unlist(strsplit(contrast, "-"))[2]
  # get subset of barcoded data and stages
  fingerprints.contrast <- fingerprints[,pd$stage %in% c(stage.1, stage.2)]
  pd.contrast <- pd[pd$stage %in% c(stage.1, stage.2),] 
  contrast.factor <- as.factor(pd.contrast$stage)
  sds.stage = adply(fingerprints.contrast, 1, function(x) {
    by(x, contrast.factor, function(x) sd(x))
    })
  row.names(sds.stage) <- sds.stage$X1
  sds.stage$X1 <- NULL
  sds.stage <- as.matrix(sds.stage)
  mode(sds.stage) <- "numeric"
  # t-tests between groups
  ttestpvals.contrast <- apply(fingerprints.contrast, 1, function(x) {
    try(t.test(x~contrast.factor)$p.value)
    })
  ttestpvals.contrast<-as.numeric(ttestpvals.contrast)
  
  # filter genes by intra-group variation and across group differences
  fingerprints.contrast.sub <- fingerprints.contrast[sds.stage[,stage.1]<=sd.cutoff  & sds.stage[,stage.2]<=sd.cutoff & ttestpvals.contrast<ttest.cutoff &  !is.na(ttestpvals.contrast),]
  
  # remove any columsn with only NAs
  fingerprints.contrast.sub <- fingerprints.contrast.sub[!apply(fingerprints.contrast.sub, 1, function(n) all(is.na(n))),]
  dim(fingerprints.contrast.sub)
  
  ## Heatmaps
  # annotations for heatmaps
  heatmap.annots <- pd.contrast[,c("arrayID", "study", "stage", "gender")]
  heatmap.annots <- as.data.frame(apply(heatmap.annots, 2, unlist))
  row.names(heatmap.annots) <- heatmap.annots$arrayID
  heatmap.annots$arrayID <- NULL
  # getting the annotations and annotation colors to play nice
  heatmap.annots <- as.matrix(heatmap.annots)
  
  heatmap.annots[,"gender"][is.na(as.vector(heatmap.annots[,"gender"]))] <- "unknown"
  
  # both genders
  if(nrow(fingerprints.contrast.sub)>0){
    pheatmap(fingerprints.contrast.sub, color=c("blue", "white", "red")    , cluster_rows=T, cluster_cols=T, show_colnames=F, annotation=heatmap.annots,  legend_breaks=c(-1,0,1),main=paste("Pathways for ", contrast, "comparison - clustered",  sep=" "), fontsize=14, fontsize_row=10, annotation_colors=ann_colors)
    }
  }
```



[1]: http://www.bioconductor.org (BioC)
[2]: http://www.bioconductor.org/packages/2.11/bioc/html/GEOquery.html (GEOquery_BioC_library)


