```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, cache=FALSE,  highlight=TRUE, figalign="center", echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120), fig.path="figure/WB-U133Plus2-pathprint-", cache.path="cache/WB-U133Plus2-pathprint-")
options(width=200)
```

# PATHPRINT ANALYSES OF U133PLUS2.0 STUDIES

---

Analysis of public datasets for differential gene expression signatures between adults and children for [Les Kobzik](mailto:LKOBZIK@hsph.harvard.edu) as part of the DARPA 7-day Biodefense Program.

Contact [John Hutchinson](mailto:jhutchin@hsph.harvard.edu) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

----

# GENERAL SETUP

## General purpose R libraries necessary for analysis

```{r pathprint_general_libraries_functions}
source("http://bioconductor.org/biocLite.R") # BioConductor script necessary for installing new BioC libraries with biocLite()
library(xtable) # table generation for reports
library(plyr) # library for iteratively working with data structures
library(ggplot2) # library for plotting 
library(RColorBrewer) # library for generating color palettes
library(googleVis) # library for presenting tables
library(pheatmap)# pretty heatmaps
library(Biobase)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")

# formula to count number of occurrences of each of the pathprint outcomes
count_trinary_outcomes <- function(vector){
  c(length(which(vector==-1)),length(which(vector==0)),length(which(vector==1)) )
  }
```

## Locations of directories for data, metadata and results

```{r pathprint_general_directories}
if (file.exists("/n/hsphS10/hsphfs1/chb/projects/lk_FOY/sepsis")) {
  baseDir <- "/n/hsphS10/hsphfs1/chb/projects/lk_FOY/sepsis"
  }  else if (file.exists("/Volumes/home08/jhutchin/consults/lk_FOY/sepsis")) {
    baseDir <- "/Volumes/home08/jhutchin/consults/lk_FOY/sepsis"
    } else {
      baseDir <- "/Volumes/ody/consults/lk_FOY/sepsis"
      }
dataDir <- file.path(baseDir, "data", "WB")
resultsDir <- file.path(baseDir, "results", "WB", "U133Plus2")
metaDir <- file.path(baseDir, "meta", "WB")
````

## Load in raw and RMA normalized  microarray data from generic microarray analysis.

```{r pathprint_load_affybatch, echo=FALSE}
load(file.path(resultsDir, "RDATA.raw_and_normalized_microarray.data_U133Plus2.0"))
pd <- pData(mic.norm.eset)
```

---

# Pathprint
## Load libraries

```{r pathprint_libraries}
library(pathprint)
library(limma)
library(ape)
# annotations for the hgU1332.0Plus array
library(hgu133plus2.db)
```

## Pathprint samples - default settings

```{r pathprint, cache=TRUE}
mic.platform <- "GPL570"
mic.species <- "human"
mic.names <- pData(mic.norm.eset)$arrayID
mic.f.default <- exprs2fingerprint(exprs = exprs(mic.norm.eset), platform = mic.platform, species = mic.species, progressBar = FALSE)
dimnames(mic.f.default)[[2]] <- pData(mic.norm.eset)$arrayID
```

```{r pathprint_nexus, echo=FALSE}
Nexus <- function (matrixname, file, dataType = "discrete") {
  mat <- t(matrixname)
  manb <- dim(mat)[2]
  innb <- dim(mat)[1]
  indnames <- rownames(mat)
  charnames <- colnames(mat)
  print("Removing punctuation from taxa and character names")
  indnames <- gsub("[[:punct:]]", "", indnames)
  indnames <- gsub(" ", "", indnames)
  charnames <- gsub("[[:punct:]]", "", charnames)
  charnames <- gsub(" ", "", charnames)
  if (sum(duplicated(indnames) > 0)) 
    stop("duplicated taxa names")
  if (sum(duplicated(charnames) > 0)) 
    stop("duplicated character names")
  for (i in 1:length(charnames)) {
    charnames[i] <- paste(i, charnames[i], sep = " ")
    }
  charstring <- paste(charnames, collapse = ", ")
  matm <- mat
  symbols <- paste(unique(as.character(mat)), collapse = "")
  print(paste("compiling NEXUS file using", dataType, "characters", 
              sep = " "))
  cat("#NEXUS", "\n", "\n", "begin taxa;", "\n", "dimensions ntax=", 
      innb, ";", "\n", file = file, sep = "")
  cat("taxlabels", "\n", file = file, sep = "", append = TRUE)
  for (i in 1:innb) {
    cat(indnames[i], "\n", file = file, sep = "", append = TRUE)
    }
  cat(";", "\n", "end;", "\n", "\n", file = file, sep = "", 
      append = TRUE)
  cat("begin characters;", "\n", "dimensions nchar=", manb, 
      " ntax=", innb, ";", "\n", file = file, sep = "", append = TRUE)
  if (dataType == "discrete") {
    cat("Format datatype=Standard symbols=", "\"", symbols, 
        "\"", ";", "\n", file = file, sep = "", append = TRUE)
    }
  if (dataType == "continuous") {
    cat("Format datatype=continuous", ";", "\n", file = file, 
        sep = "", append = TRUE)
    }
  cat("CHARSTATELABELS", "\n", file = file, sep = "", append = TRUE)
  cat(charstring, " ;", "\n", file = file, sep = "", append = TRUE)
  cat("matrix", "\n", file = file, sep = "", append = TRUE)
  for (j in 1:innb) {
    if (dataType == "discrete") {
      cat(indnames[j], "\t", matm[j, ], "\n", file = file, 
          sep = "", append = TRUE)
      }
    if (dataType == "continuous") {
      cat(indnames[j], "\t", file = file, sep = "", append = TRUE)
      cat(matm[j, ], file = file, sep = " ", append = TRUE)
      cat("\n", file = file, sep = "", append = TRUE)
      }
    }
  cat(";", "\n", "end;", "\n", "\n", file = file, sep = "", 
      append = TRUE)
  }
mic.f.default.adult <- mic.f.default[,unlist(pd$arrayID[which(pd$stage=="adult")])]
colnames(mic.f.default.adult) <- paste("adult", colnames(mic.f.default.adult), sep="_")
mic.f.default.child <- mic.f.default[,unlist(pd$arrayID[which(pd$stage=="child")])]
colnames(mic.f.default.child) <- paste("child", colnames(mic.f.default.child), sep="_")
mic.f.default.adult.child <- cbind(mic.f.default.adult, mic.f.default.child)
Nexus((mic.f.default.adult.child+1), file.path(resultsDir, "fingerprints.adult.child.nxs"))
```

### Default Pathway Differentiation

Took a similar general approach as I took for the barcoding analyses.
i.e. minimize intra-group variation while maximizing inter-group variation.

```{r pathwaydiff_vars}
# cutoffs used to filter the data
ttest.cutoff=1e-8
sd.cutoff=0.5
```

For this, I set two filtering criteria:
1) maximize homogeneity within an age group - based on minimizing the standard deviation (as I have no idea how to calculate a trinary entropy), cutoff used was sd<`r sd.cutoff`
2) maximize differences between groups - using ttests, cutoff used here was p<`r ttest.cutoff`

#### Setup design and contrast matrices 

Same setup as for the previous limma analysis.

```{r pathprint_design}
design <- model.matrix(~ -1+factor(pd$stage))
# make sure the headings match
colnames(design) <- sub("factor.pd.stage.", "", colnames(design))
```

```{r pathprint_contrastmatrix}
contrast.matrix <- makeContrasts(adult-child,adult-neonate,neonate-child, levels=dimnames(design)[[2]])
```

#### Heatmaps
- clustered by both column (sample) and row (pathway)

```{r pathprint_heatmap_colors}
# annotation colors
study_colors <- c( cbPalette[1:7] )
names(study_colors) <- unique(unlist(pd$study))
stage_colors <- c("white", "darkgrey", "black")
names(stage_colors) <- unique(unlist(pd$stage))
ann_colors = list(study = study_colors, stage = stage_colors)
```


```{r pathprint_default_contrasts, fig.width=12, fig.height=12}
# for each comparison
for (n in 1:3) {
  contrast <- dimnames(contrast.matrix)$Contrasts[n]
  contrast <- gsub(" ", "", contrast)
  stage.1 <- unlist(strsplit(contrast, "-"))[1]
  stage.2 <- unlist(strsplit(contrast, "-"))[2]
  # get subset of barcoded data and stages
  mic.f.default.contrast <- mic.f.default[,pData(mic.norm.eset)$stage %in% c(stage.1, stage.2)]
  pData.contrast <- pData(mic.norm.eset)[pData(mic.norm.eset)$stage %in% c(stage.1, stage.2),] 
  contrast.factor <- as.factor(pData.contrast$stage)
  sds.stage = adply(mic.f.default.contrast, 1, function(x) {
    by(x, contrast.factor, function(x) sd(x))
    })
  row.names(sds.stage) <- sds.stage$X1
  sds.stage$X1 <- NULL
  sds.stage <- as.matrix(sds.stage)
  mode(sds.stage) <- "numeric"
  # t-tests between groups
  ttestpvals.contrast <- apply(mic.f.default.contrast, 1, function(x) {
    try(t.test(x~contrast.factor)$p.value)
    })
  ttestpvals.contrast<-as.numeric(ttestpvals.contrast)
  # filter genes by intra-group variation and across group differences
  mic.f.default.contrast.sub <- mic.f.default.contrast[sds.stage[,stage.1]<=sd.cutoff  & sds.stage[,stage.2]<=sd.cutoff & ttestpvals.contrast<ttest.cutoff &  !is.na(ttestpvals.contrast),]
  dim(mic.f.default.contrast.sub)
  ## Heatmaps
  # annotations for heatmaps
  heatmap.annots <- pData.contrast[,c("arrayID", "study", "stage")]
  heatmap.annots <- as.data.frame(apply(heatmap.annots, 2, unlist))
  row.names(heatmap.annots) <- heatmap.annots$arrayID
  heatmap.annots$arrayID <- NULL
  # both genders
  if(nrow(mic.f.default.contrast.sub))
  pheatmap(mic.f.default.contrast.sub, color=c("blue", "white", "red"), cluster_rows=T, cluster_cols=F, show_colnames=F, annotation=heatmap.annots,  legend_breaks=c(-1,0,1),main=paste("Pathways for ", contrast, "comparison",  sep=" "), fontsize=14, fontsize_row=10, annotation_colors=ann_colors)
  }
```
