---
output:
  knitrBootstrap::bootstrap_document:
    theme: readable
    highlight: zenburn
    theme.chooser: TRUE
    highlight.chooser: TRUE
  html_document:
    toc: true
    highlight: zenburn
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='', echo=FALSE)
```

```{r load-discordant-concordant, echo=FALSE}
library(ggplot2)
library(reshape2)
library(extrafont)
library(knitr)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442",
"#0072B2", "#D55E00", "#CC79A7")
cbPalette = c("#543005", "#8c510a", "#bf812d", "#dfc27d", "#f6e8c3",
    "#c7eae5", "#80cdc1", "#35978f", "#01665e", "#003c30")

load_tables = function(discordant, concordant) {
    discordant$class = "discordant"
    concordant$class = "concordant"
    concordant$QUAL = as.numeric(concordant$QUAL)
    discordant$QUAL = as.numeric(discordant$QUAL)
    df = rbind(discordant, concordant)
    df$EFF = NULL
    snps = unlist(lapply(df$REF, nchar)) == unlist(lapply(df$ALT, nchar))
    df$snp = ifelse(snps, "snp", "indel")
    return(df)
}
  

discordant = read.table('discordant-merged.tab', header=TRUE, sep="\t",
  stringsAsFactors=FALSE)
concordant = read.table('concordant-merged.tab', header=TRUE, sep="\t",
  stringsAsFactors=FALSE)
df = load_tables(discordant, concordant)
df = df[, c("CHROM", "POS", "REF", "ALT", "QUAL", "AC", "AN", "AF", "DP",
"class", "snp")]
df$square = "non-square"
```


```{r concordant-discordant-square}
discordant_square = read.table('square/discordant-merged-square.tab', header=TRUE, sep="\t",
  stringsAsFactors=FALSE)
concordant_square = read.table('square/concordant-merged-square.tab', header=TRUE, sep="\t",
  stringsAsFactors=FALSE)
df_square = load_tables(discordant_square, concordant_square)
df_square = df_square[, c("CHROM", "POS", "REF", "ALT", "QUAL", "AC", "AN", "AF", "DP",
"class", "snp")]
df_square$square = "square"
```

# Discordant SNPs are often low-depth heterozygote calls
```{r discordant-square-plot}
ggplot(df_square, aes(DP, QUAL, color=AF)) +
  geom_point(alpha=0.3) + 
  facet_wrap(~ snp + class) + scale_y_log10() + scale_x_log10()
```

There is a slight shift to lower qualities for discordant indels and SNPs and
overall there are more higher quality homozygous calls for concordant vs discordant
calls. 

```{r concordant-discordant-histogram-depth}
ggplot(df, aes(DP)) + geom_histogram() +
  facet_wrap(~ snp + class) + scale_y_log10() + scale_x_log10()
```

```{r concordant-discordant-histogram-quality}
ggplot(df, aes(QUAL)) + geom_histogram() +
  facet_wrap(~ snp + class) + scale_y_log10() + scale_x_log10()
```

# Squaring-off effect
```{r squaring-off-plot}
df_joined = rbind(df, df_square)
ggplot(df_joined, aes(DP, QUAL, color=square)) + geom_point(alpha=0.3) +
  facet_wrap(~ snp + class) + scale_y_log10() + scale_x_log10()
```

Squaring off causes many more reads to be assigned to indels and SNPs.

```{r summary-concordance-table}
get_summary = function(summary1_fn, summary2_fn, sep=",") {
    summary_qc1 = read.table(summary1_fn, sep=sep, header=TRUE)
    summary_qc1$comparison = "qc1_vs_qc2"
    summary_qc2 = read.table(summary2_fn, sep=sep, header=TRUE)
    summary_qc2$comparison = "qc2_vs_qc3"
    original_summary = rbind(summary_qc1, summary_qc2)
    return(original_summary)
}

summary_1 = '5210116_qc1-summary.csv'
summary_2 = '5210116_qc2-summary.csv'
original_summary = get_summary(summary_1, summary_2)
original_summary$version = "non-square"

summary_1 = 'square/5210116_qc1-summary.csv'
summary_2 = 'square/5210116_qc2-summary.csv'
square_summary = get_summary(summary_1, summary_2)
square_summary$version = "square"

summary_stats = rbind(original_summary, square_summary)

library(dplyr)
summary_subset = tbl_df(summary_stats[, c("version", "type", "concordant", "discordant1",
                                          "discordant2")])
summary_subset$type = factor(summary_subset$type)

summary_grouped = group_by(summary_subset, type)

z = summary_subset %>%
    group_by(version, type) %>%
        summarise(concordant=mean(concordant),
                  discordant=mean(discordant1 + discordant2))

z$ratio = z$concordant / (z$concordant + z$discordant)

grading_1 = '5210116_qc1-grading.tsv'
grading_2 = '5210116_qc2-grading.tsv'

original_summary = get_summary(grading_1, grading_2, " ")
original_summary$version = "non-square"

summary_1 = 'square/5210116_qc1-grading.tsv'
summary_2 = 'square/5210116_qc2-grading.tsv'
square_summary = get_summary(summary_1, summary_2, " ")
square_summary$version = "square"

summary_stats = rbind(original_summary, square_summary)
summary_stats$discordant = rowSums(summary_stats[, c("extra", "missing", "vardiff", "hethom")])
summary_stats$ratio = summary_stats$concordant / rowSums(summary_stats[, c("concordant", "extra", "missing", "vardiff", "hethom")])
```

# Concordance summary

```{r summary-table, echo=TRUE, results='asis'}
keep = c("type", "version", "concordant", "discordant", "ratio")
summary_stats = summary_stats[, keep]
summary_stats = subset(summary_stats, type != "mixed")

summarized = summary_stats %.%
    group_by(type, version) %.%
    summarize(concordant = mean(concordant),
                discordant = mean(discordant),
                ratio = mean(ratio))
kable(summarized, format="html")
 

```

The overall concordance between technical replicates is good and is in line with other
studies. Squaring off results in a small boost in concordance for SNPs and indels over
not squaring off the calls.

# SNP discordance by class

```{r clean-snp-data}
vcf = read.table("square/replicates-merged.tab", sep="\t", header=TRUE)
concordant = read.table("square/concordant-merged-square.tab", sep="\t", header=TRUE)
discordant = read.table("square/discordant-merged-square.tab", sep="\t", header=TRUE)
concordant_locs = concordant[, c("CHROM", "POS", "REF", "ALT")]
discordant_locs = discordant[, c("CHROM", "POS", "REF", "ALT")]
in_discordant = merge(vcf, discordant_locs, by=c("CHROM", "POS", "REF", "ALT"))
in_concordant = merge(vcf, concordant_locs, by=c("CHROM", "POS", "REF", "ALT"))
in_discordant$concordance = "discordant"
in_concordant$concordance = "concordant"

qc1 = read.table("square/5210116_qc1.tab", sep="\t", header=TRUE)
qc1$sample = "5210116_qc1"
qc2 = read.table("square/5210116_qc2.tab", sep="\t", header=TRUE)
qc2$sample = "5210116_qc2"
qc3 = read.table("square/5210116_qc3.tab", sep="\t", header=TRUE)
qc3$sample = "5210116_qc3"
colnames(qc1)[21:27] = c("AO", "RO", "DP", "QA", "QR", "GT", "AD")
colnames(qc2)[21:27] = c("AO", "RO", "DP", "QA", "QR", "GT", "AD")
colnames(qc3)[21:27] = c("AO", "RO", "DP", "QA", "QR", "GT", "AD")
vcf = rbind(qc1, qc2, qc3)
in_discordant = merge(vcf, discordant_locs, by=c("CHROM", "POS", "REF", "ALT"))
in_concordant = merge(vcf, concordant_locs, by=c("CHROM", "POS", "REF", "ALT"))
in_discordant$concordance = "discordant"
in_concordant$concordance = "concordant"
concordance = rbind(in_discordant, in_concordant)
concordance$REF = as.character(concordance$REF)
concordance$ALT = as.character(concordance$ALT)
concordance$QUAL = as.numeric(as.character(concordance$QUAL))
concordance$AB = as.numeric(as.character(concordance$AB))
concordance$AO = as.numeric(as.character(concordance$AO))
concordance$RO = as.numeric(as.character(concordance$RO))
concordance$ABP = as.numeric(as.character(concordance$ABP))
concordance$DP = as.numeric(as.character(concordance$DP))
new_concordance = concordance
snp_index = (unlist(lapply(new_concordance$REF, nchar)) == 1) & (unlist(lapply(new_concordance$ALT, nchar)) == 1)
snps = new_concordance[snp_index,]
allele1 = unlist(lapply(strsplit(as.character(snps$GT), "[/||]"), "[", 1))
allele2 = unlist(lapply(strsplit(as.character(snps$GT), "[/||]"), "[", 2))
phased = rep(NA, nrow(snps))
phased[grep("/", snps$GT)] = "unphased"
phased[-grep("/", snps$GT)] = "phased"
snps$phased = as.factor(phased)
genotypes = rep(NA, nrow(snps))
genotypes[allele1 == snps$REF & allele2 == snps$REF] = "hom_ref"
genotypes[allele1 != snps$REF & allele2 != snps$REF] = "hom_alt"
genotypes[is.na(genotypes)] = "het"
snps$genotype = as.factor(genotypes)
ad_ref = as.numeric(unlist(lapply(strsplit(as.character(snps$AD), ","), "[", 1)))
ad_alt = as.numeric(unlist(lapply(strsplit(as.character(snps$AD), ","), "[", 2)))
het_phased = snps$genotype == "het" & snps$phased == "phased"
het_unphased = snps$genotype == "het" & snps$phased == "unphased"
snps$ref_depth = ad_ref
snps$alt_depth = ad_alt
snps$ref_depth[is.na(snps$ref_depth)] = snps[is.na(snps$ref_depth),]$RO
snps$alt_depth[is.na(snps$alt_depth)] = snps[is.na(snps$alt_depth),]$AO

library(dplyr)
snps = tbl_df(snps)
positions = group_by(snps, CHROM, POS, REF, ALT)
positions$GT = as.character(positions$GT)
positions$genotype = as.character(positions$genotype)
genotypes = summarize(positions, gt = length(unique(genotype)),
    common_genotype = names(table(genotype)[which.max(table(genotype))]),
    minor_genotype = names(table(genotype)[which.min(table(genotype))]))
status_disagree = merge(snps, subset(genotypes, gt == 2),
    by=c("CHROM", "POS", "REF", "ALT"))

baseline_prop = table(snps$genotype) / sum(table(snps$genotype))



variants_disagree = subset(status_disagree, common_genotype != genotype)
variants_agree = subset(status_disagree, common_genotype == genotype)
variants_disagree$status = "minority"
variants_agree$status = "majority"
```

We have three technical replicates, looking at cases where two out of the three agree
is useful to try to understand where the differences come from. For the two that agree,
call them the 'majority' and the one that is different call it the 'minority'. This
table shows the number of 'majority -> minority' disagreements there are for each
genotype.

```{r variant-disagreements, echo=TRUE, results='asis'}
t = table(paste(variants_disagree$common_genotype, "->", variants_disagree$genotype))
kable(data.frame(t))
```

```{r variant-disagreement-setup}
d = rbind(variants_disagree, variants_agree)
d = subset(d, concordance == "discordant")
d$AO = as.numeric(as.character(d$AO))
d$RO = as.numeric(as.character(d$RO))
d$QA = as.numeric(d$QA)
d$transition = paste(d$common_genotype, "->", d$minor_genotype)
d$qratio = d$QA / d$QR
d$ratio = d$AO / d$RO
d$ratio = d$alt_depth / d$ref_depth
d$depth = d$alt_depth + d$ref_depth
```

```{r depth-vs-ratio-plot, echo=TRUE}
 ggplot(d, aes(depth, log2(ratio), color=status)) +
     scale_x_log10() + geom_point(alpha=0.3) + facet_wrap(~ transition) +
         ## scale_color_brewer(palette="Set3", name="maj -> min",
         ##                    limits=levels(d$transition)) +
ylab("log2(alt_depth/ref_depth)") +
             theme_bw()
```
For homozygous disagreements (**hom_ref->hom_alt**, etc) the overall
depth is very low, and the ratio of alt/ref trends towards 1,
indicating this might be a sampling issue and these variants are
undersampled hets.

For disagreements of heterozygotes and homozygotes they tend to be of
a higher depth but there is still a shift of the read depth in the
direction of the disagreement. For example **het->hom_ref**, the **het** calls
are closer to a ratio of 1 and the **hom_ref** calls are shifted towards
having more reference reads. Continuing to look at the **het->hom_ref**
panel, even the **het** calls are shifted to have more reference reads
instead of being centered around 1. These might be due to these being
**het** calls but there being a bias towards one allele which coupled with
random sampling results in a incongruent call.

```{r table-again, echo=TRUE, results='asis'}
kable(data.frame(t))
```

If all of these variants are true hets which have some sampling issues causing
an incorrect **hom_alt** or **hom_ref** call, we should expect there to be more
**het->hom_alt** and **het->hom_ref** disagreements than **hom_ref->het** and
**hom_alt->het** disagreements, since there has to be two samples disagreeing
with the truth for a **hom_ref->het** transition than for a **het->hom_ref**.


<!--
odd_variant_idx = snps$ref_depth | is.na(snps$alt_depth)
odd_variants = snps[odd_variant_idx,]

kosher_variants = snps[-odd_variant_idx,]

new_concordance = kosher_variants[, c("QUAL", "AC", "AN", "AF", "HRun", "AO", "QR", "RO", "ref_depth", "alt_depth", "phased", "genotype", "concordance", "depth")]
index = 1:nrow(new_concordance)
testindex = sample(index, trunc(length(index)/2))
testset = new_concordance[testindex,]
trainset = new_concordance[-testindex,]


library(rpart)
fit = rpart(concordance ~ ., data=trainset)
p = data.frame(predict(fit, testset))

pred = ifelse(p$concordant < p$discordant, "discordant", "concordant")
table(testset$concordance == pred)
table(testset[testset$concordance != pred,]$concordance)
table(testset$concordance)

post(fit, file="~/Downloads/tree.ps")

x = rownames(testset[pred == "concordant",])
locs = unique(kosher_variants[x,c("CHROM", "POS", "REF", "ALT")])
m = merge(locs, snps, by=c("CHROM", "POS", "REF", "ALT"))

all = rownames(testset)
all_locs = unique(kosher_variants[all,c("CHROM", "POS", "REF", "ALT")])

ggsave(filename="plots/quality-transition.png")

 ggplot(d, aes((alt_depth + ref_depth), fill=transition)) + geom_histogram() +
     scale_x_log10() + facet_wrap(~ transition) + xlab("total depth") +
         ## scale_fill_brewer(palette="Set3", name="maj -> min",
         ##                    limits=levels(d$transition)) +
             theme_bw()
ggsave(filename="plots/depth-by-transition.png")

indels = new_concordance[!snp_index,]
ind_disc = subset(indels, concordance == "discordant")
allele1 = unlist(lapply(strsplit(as.character(indels$GT), "[/||]"), "[", 1))
allele2 = unlist(lapply(strsplit(as.character(indels$GT), "[/||]"), "[", 2))
phased = rep(NA, nrow(indels))
phased[grep("/", indels$GT)] = "unphased"
phased[-grep("/", indels$GT)] = "phased"
indels$phased = as.factor(phased)
genotypes = rep(NA, nrow(indels))
genotypes[allele1 == indels$REF & allele2 == indels$REF] = "hom_ref"
genotypes[allele1 != indels$REF & allele2 != indels$REF] = "hom_alt"
genotypes[is.na(genotypes)] = "het"
indels$genotype = as.factor(genotypes)
ad_ref = as.numeric(unlist(lapply(strsplit(as.character(indels$AD), ","), "[", 1)))
ad_alt = as.numeric(unlist(lapply(strsplit(as.character(indels$AD), ","), "[", 2)))
het_phased = indels$genotype == "het" & indels$phased == "phased"
het_unphased = indels$genotype == "het" & indels$phased == "unphased"
indels$ref_depth = ad_ref
indels$alt_depth = ad_alt
indels$ref_depth[is.na(indels$ref_depth)] = indels[is.na(indels$ref_depth),]$RO
indels$alt_depth[is.na(indels$alt_depth)] = indels[is.na(indels$alt_depth),]$AO

library(dplyr)
indels = tbl_df(indels)
positions = group_by(indels, CHROM, POS, REF, ALT)
positions$GT = as.character(positions$GT)
positions$genotype = as.character(positions$genotype)
genotypes = summarize(positions, gt = length(unique(genotype)),
    common_genotype = names(table(genotype)[which.max(table(genotype))]),
    minor_genotype = names(table(genotype)[which.min(table(genotype))]))
status_disagree = merge(indels, subset(genotypes, gt == 2),
    by=c("CHROM", "POS", "REF", "ALT"))

baseline_prop = table(indels$genotype) / sum(table(indels$genotype))


variants_disagree = subset(status_disagree, common_genotype != genotype)
variants_agree = subset(status_disagree, common_genotype == genotype)
variants_disagree$status = "minority"
variants_agree$status = "majority"

# for disagreeing variants how many of each transition are there
t = table(paste(variants_disagree$common_genotype, "->", variants_disagree$genotype))

d = rbind(variants_disagree, variants_agree)
d = subset(d, concordance == "discordant")
d$AO = as.numeric(as.character(d$AO))
d$RO = as.numeric(as.character(d$RO))
d$QA = as.numeric(d$QA)
d$transition = paste(d$common_genotype, "->", d$minor_genotype)
d$qratio = d$QA / d$QR
d$ratio = d$AO / d$RO

 ggplot(d, aes(AO, RO, color=transition)) +
     scale_x_log10() + geom_jitter(alpha=0.9) + facet_wrap(~ status) +
  scale_y_log10() +
         scale_color_brewer(palette="Set3", name="maj -> min",
                            limits=levels(d$transition)) + 
             theme_bw()
  
 ggplot(d, aes(QUAL, log2(ratio), color=transition)) +
     scale_x_log10() + geom_jitter(alpha=0.9) + facet_wrap(~ status) +
         scale_color_brewer(palette="Set3", name="maj -> min",
                            limits=levels(d$transition)) + 
             theme_bw()
-->
