```{r PBMCsetup, echo=FALSE}
opts_chunk$set(tidy=TRUE, cache=FALSE,  highlight=TRUE, figalign="center", echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120))
options(width=200)
```

# BARCODE ANALYSES OF U133PLUS2.0 STUDIES OF PBMC SAMPLES

---

Analysis of public datasets for differential gene expression signatures between adults and children for [Les Kobzik](mailto:LKOBZIK@hsph.harvard.edu) as part of the DARPA 7-day Biodefense Program. PBMC.samples.

Contact [John Hutchinson](mailto:jhutchin@hsph.harvard.edu) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

----

# GENERAL SETUP

## General purpose R libraries necessary for analysis

```{r PBMCgeneral_libraries}
source("http://bioconductor.org/biocLite.R") # BioConductor script necessary for installing new BioC libraries with biocLite()
library(xtable) # table generation for reports
library(plyr) # library for iteratively working with data structures
library(ggplot2) # library for plotting 
library(RColorBrewer) # library for generating color palettes
library(googleVis) # library for presenting tables
library(xlsx) # library for outputting Excel files
options(java.parameters = "-Xmx1024m")
```

## Locations of directories for data, metadata and results
- also misc. variables

```{r PBMCgeneral_directories}
if (file.exists("/n/hsphS10/hsphfs1/chb/projects/lk_FOY")) {
  baseDir <- "/n/hsphS10/hsphfs1/chb/projects/lk_FOY"
  }  else if (file.exists("/Volumes/home08/jhutchin/consults/lk_FOY/")) {
    baseDir <- "/Volumes/home08/jhutchin/consults/lk_FOY"
    } else {
      baseDir <- "/Volumes/ody/consults/lk_FOY"
      }
dataDir <- file.path(baseDir, "data")
resultsDir <- file.path(baseDir, "results")
metaDir <- file.path(baseDir, "meta")
# colorblind friendly palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")
````

## Load in raw and [RMA][1] normalized AffyBatch microarray data, as well as the top differentially expressed genes from the [limma][2]  [analysis][6]

```{r PBMCload_affybatch, echo=FALSE}
load(file.path(resultsDir, "RDATA.raw_and_normalized_microarray.data.topstats.PBMC.U133Plus2.0"))
```

---

# Barcoding

The group of Rafael Irizarry introduced the concept of ’barcoding’ expression information in *McCall MN, Uppal K, Jaffee HA, Zilliox MJ, and Irizarry RA (2011). The Gene Expression Barcode: leveraging public data repositories to begin cataloging the human and murine transcriptomes, Nucleic Acids Research,39:D1011-5.*
  
From their site, [The Gene Expression Barcode][3]
>The barcode algorithm is designed to estimate which genes are expressed and which are unexpressed in a given microarray hybridization. The output of our algorithm is a vector of ones and zeros denoting which genes are estimated to be expressed (ones) and unexpressed (zeros). We call this a gene expression barcode.

## Load in necessary [Bioconductor][4] libraries

```{r PBMCbarcode_libraries}
# frozen RMA and barcoding library
library(frma) 
# previously analyzed dataset from same array
library(hgu133plus2frmavecs)
# for pretty dendrograms
library(ggdendro)
# for contrast matrix
library(limma)
# pretty heatmaps
library(pheatmap) 
# annotations for the hgU1332.0Plus array
library(hgu133plus2.db) 
```

## Run fRMA (frozen RMA) and Barcode expression data

Using the [frma][5] Bioconductor libraray. As described in *McCall MN, Bolstad BM, and Irizarry RA (2010). Frozen Robust Multi-Array Analysis (fRMA), Biostatistics, 11(2):242-253.* frozen RMA essentially normalizes against a previously analyzed ("frozen") dataset of samples run on the same array.

Used the same samples as the previous limma analysis.

```{r PBMCfrma_run, cache=TRUE}
mic.frma <- frma(mic.raw, summarize="random_effect")
save.image(file.path(resultsDir, "RDATA.frma.normalized.U133Plus2.0"))
bc <- barcode(mic.frma) 
pd <- pData(mic.frma)
dimnames(bc)[[2]] <- as.vector(unlist(pd[which(row.names(pd) %in% dimnames(bc)[[2]]),"ID"]))
```

## Compare hierarchal clustering  before and after barcoding
- very little change

```{r PBMCbarcode_clustering, fig.width=12, fig.height=12}
rma.exprs <- exprs(mic.norm.eset)
colnames(rma.exprs) <- pData(mic.frma)$stage
myDist.pre <- dist(t(rma.exprs))
myTree.pre <- hclust(myDist.pre)
dhc.pre <- as.dendrogram(myTree.pre)
ddata.pre <- dendro_data(dhc.pre)

bc.exprs <- bc
colnames(bc.exprs) <- pData(mic.frma)$stage
myDist.bc <- dist(t(bc.exprs))
myTree.bc <- hclust(myDist.bc)
dhc.bc <- as.dendrogram(myTree.bc)
ddata.bc <- dendro_data(dhc.bc, type="rectangle")
df$g <- factor(df$g, levels = letters[4:1])


p.pre <- ggplot(segment(ddata.pre)) +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend)) +
  theme_dendro() + 
  geom_text(data=label(ddata.pre), aes(x=x, y=y, label=label, hjust=-0.1, colour=factor(label, levels=c("adult", "neonate", "child")))) +
  coord_flip() + 
  scale_y_reverse(expand=c(0.2, 50)) +
  labs(title = "Original RMA Normalized Expression Data") +
  theme(legend.title=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())

p.bc <- ggplot(segment(ddata.bc)) +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend)) +
  theme_dendro() + 
  geom_text(data=label(ddata.bc), aes(x=x, y=y, label=label, hjust=-0.1, colour=factor(label,levels=c("adult", "neonate", "child")))) +
  coord_flip() + 
  scale_y_reverse(expand=c(0.2, 50)) +
  labs(title = "fRMA Normalized and Barcoded Expression Data") +
  theme(legend.title=element_blank(),
    axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())
p.pre 
p.bc 
## cleanup
rm(p.pre, p.bc, ddata.bc, dhc.bc, myTree.bc, myDist.bc, ddata.pre, dhc.pre, myTree.pre, myDist.pre, rma.exprs )
```

#### Setup design and contrast matrices 

Same setup as for the previous limma analysis.

```{r PBMCdesign}
design <- model.matrix(~ -1+factor(pData(mic.norm.eset)$stage))
# make sure the headings match
colnames(design) <- sub("factor\\(pData\\(mic.norm.eset\\)\\$stage\\)", "", colnames(design))
```

```{r PBMCcontrastmatrix}
contrast.matrix <- makeContrasts(adult-child,adult-neonate,neonate-child, levels=dimnames(design)[[2]])
```

## Heatmaps of barcoded expression data

### Top 100 differentially expressed genes from the previous standard non-barcoded, RMA normalized, limma  data
- for comparison 
- again for both genders together, and separate plots for female and male genders

```{r PBMCheatmap_annotation_colors}
   # annotation colors
  study_colors <- c("#FF0000","#00FF00", "#0000FF", cbPalette )
  names(study_colors) <- unique(unlist(pd$study))
  stage_colors <- c("white", "darkgrey", "black")
  names(stage_colors) <- unique(unlist(pd$stage))
  gender_colors <- c("cyan", "pink")
  names(gender_colors) <- unique(unlist(pd$gender))
  ann_colors = list(study = study_colors, stage = stage_colors, gender=gender_colors)
```

- heatmaps

```{r PBMCbarcode_heatmaps, fig.height=18, fig.width=18}
for(n in 1:3) {
  # get stages involved in this comparison, their sample ids, and barcode data
  contrast <- dimnames(contrast.matrix)$Contrasts[n]
  stages <- names(which(abs(contrast.matrix[,contrast])>0))
  sample.indices <- which(design[,stages[1]]>0 | design[,stages[2]]>0)
  pData.contrast <- pData(mic.norm.eset)[sample.indices,]
  top.IDs.contrast <- top.stats[[n]]$data$ID
  top.barcodes.contrast <- bc[which(row.names(bc) %in% top.IDs.contrast),]
  top.barcodes.contrast <- top.barcodes.contrast[,unlist(pData.contrast$ID)]
  # remove any AFFY control probes
  if (any(grepl("AFFX", row.names(top.barcodes.contrast)))){
    top.barcodes.contrast <- top.barcodes.contrast[-(grep("AFFX", row.names(top.barcodes.contrast))),]  
    }
  # prep heatmap row labels - add gene symbol to probeset id
  row.names(top.barcodes.contrast) <- paste(as.vector(unlist(mget(row.names(top.barcodes.contrast), hgu133plus2SYMBOL, ifnotfound=NA))), " (" ,row.names(top.barcodes.contrast), ")", sep="")
  # annotations
  heatmap.annots <- pData.contrast[,c("ID", "study", "stage", "gender")]
  heatmap.annots <- as.data.frame(apply(heatmap.annots, 2, unlist))
  row.names(heatmap.annots) <- heatmap.annots$ID
  heatmap.annots$ID <- NULL
  ## Heatmaps
  # both genders
  pheatmap(as.matrix(top.barcodes.contrast), color=c("blue", "red"), annotation=heatmap.annots, annotation_colors=ann_colors, legend_breaks=c(0,1), main=contrast, show_colnames=F, fontsize=18, fontsize_row=10, cluster_cols=FALSE, cluster_rows=TRUE) 
  # female gender
  top.barcodes.contrast.female <- top.barcodes.contrast[,which(pData.contrast$gender=="FEMALE")]
  pheatmap(as.matrix(top.barcodes.contrast.female), color=c("blue", "red"), annotation=subset(heatmap.annots, gender=="FEMALE"), main=paste(contrast, "(FEMALE)", sep=" "),show_colnames=F, fontsize=18, fontsize_row=10,legend_breaks=c(0,1), annotation_colors=ann_colors, cluster_cols=FALSE, cluster_rows=TRUE) 
  # male gender
  top.barcodes.contrast.male <- top.barcodes.contrast[,which(pData.contrast$gender=="MALE")]
  pheatmap(as.matrix(top.barcodes.contrast.male), color=c("blue", "red"), annotation=subset(heatmap.annots,gender=="MALE"), main=paste(contrast, "(MALE)", sep=" "),show_colnames=F, fontsize=18, fontsize_row=10, annotation_colors=ann_colors, legend_breaks=c(0,1), cluster_cols=FALSE, cluster_rows=TRUE) 
  }
# cleanup
rm(n, contrast, stages, sample.indices, pData.contrast, heatmap.annots, gender_colors, stage_colors, study_colors, top.IDs.contrast, top.barcodes.contrast, top.barcodes.contrast.female, top.barcodes.contrast.male)
```


As with the whole blood analyses, there are some genes that cluster well by age group, but also many that are homogenous across age groups. In cases that cluster by age group, it is difficult to separate study specific effects from age group effects.

Once again, independent analyses of the barcoded data should allow the barcoding process to better adjust for batch effects.

## Independent Analyses of Barcoded Data
We can try to identify probes with consistent differences between the cohorts by only keeping probes exhibiting low entropy within a group. Maximum entropy would be when a probe is expressed in exactly 50 % of the samples. 
 – but high entropy overall (i.e each group is internally consistent but the groups themselves are different)

### The function to calculate binary entropy

```{r PBMCbinary_entropy_function}
binaryEntropy <- function(v) {
  # Proportion of probes found to be expressed
  p <- sum(v) / length(v)
  if (p == 0 | p == 1) {
    return(0)
    }
  return(-p * log10(p) - (1 - p) * log10(1 - p))
  }
```

### Heatmaps of genes with consistent intra and inter-group expression patterns 

I tried out a couple of entropy cutoffs in an attempt to balance the number of resulting genes with the observed study effectys.
In the first passs I allowd relatively more intra-group entropy (<0.3), while allowing the entropy across both groups to be higher (>0.3).
  
You can see in these heatmaps that there are still some fairly strong study specific effects, so more stringent intra-group entropy cutoffs may be necessary.

```{r PBMCfilters1}
intra.entropycutoff <- 0.3

inter.entropycutoff <- 0.3
# prep data for output
class(bc) <- "numeric"
all(colnames(bc)==pData(mic.frma)$ID)
```


```{r PBMCbarcode_diff_heatmap0.3, fig.width=18, fig.height=18}
# filter and output heatmap for each contrast
for (n in 1:3) {
  contrast <- dimnames(contrast.matrix)$Contrasts[n]
  contrast <- gsub(" ", "", contrast)
  stage.1 <- unlist(strsplit(contrast, "-"))[1]
  stage.2 <- unlist(strsplit(contrast, "-"))[2]
  # get subset of barcoded data and stages
  bc.contrast <- bc[,pData(mic.frma)$stage %in% c(stage.1, stage.2)]
  pData.contrast <- pData(mic.frma)[pData(mic.frma)$stage %in% c(stage.1, stage.2),] 
  f.contrast <- as.factor(pData.contrast$stage)
  entropies.stage = adply(bc.contrast, 1, function(n) {
    by(n, f.contrast, function(x) binaryEntropy(x))
    })
  row.names(entropies.stage) <- entropies.stage$X1
  entropies.stage$X1 <- NULL
  entropies.stage <- as.matrix(entropies.stage)
  mode(entropies.stage) <- "numeric"
  entropies.contrast <- apply(bc.contrast, 1, binaryEntropy)
  # filter genes by intra-group and across group entropy
  bc.contrast.sub <- bc.contrast[entropies.stage[,stage.1]<=intra.entropycutoff  & entropies.stage[,stage.2]<=intra.entropycutoff & entropies.contrast>=inter.entropycutoff,]
  dim(bc.contrast.sub)
  # remove any AFFY control probes
  if (any(grepl("AFFX", row.names(bc.contrast.sub)))){
    bc.contrast.sub <- bc.contrast.sub[-(grep("AFFX", row.names(bc.contrast.sub))),]  
    }
  ## output to file
  output <- as.data.frame(bc.contrast.sub)
  probeID <- row.names(output)
  symbol <- unlist(mget(row.names(output), hgu133plus2SYMBOL, ifnotfound=NA))
  output.entropies.stage <- entropies.stage[entropies.stage[,stage.1]<=intra.entropycutoff  & entropies.stage[,stage.2]<=intra.entropycutoff & entropies.contrast>=inter.entropycutoff,]
  output.entropies.contrast <- entropies.contrast[entropies.stage[,stage.1]<=intra.entropycutoff  & entropies.stage[,stage.2]<=intra.entropycutoff & entropies.contrast>=inter.entropycutoff]
  output <- cbind(probeID, cbind(symbol, cbind(output.entropies.stage, cbind(output.entropies.contrast, output))))
  names(output)[3:5] <- c(paste("intra", c(stage.1,stage.2), "entropy", sep="_" ), "overall_entropy")
  write.table(output, file.path(resultsDir, "PBMC", paste("U133_2.0Plus.barcode.entropy.cutoff", intra.entropycutoff, contrast, "xls", sep=".")), sep="\t", row.names=F, col.names=T)
  ## Heatmaps
  # annotations for heatmaps
  heatmap.annots <- pData.contrast[,c("ID", "study", "stage", "gender")]
  heatmap.annots <- as.data.frame(apply(heatmap.annots, 2, unlist))
  row.names(heatmap.annots) <- heatmap.annots$ID
  heatmap.annots$ID <- NULL
  # row labels - add gene symbol to probeset id
  row.names(bc.contrast.sub) <- paste(as.vector(unlist(mget(row.names(bc.contrast.sub), hgu133plus2SYMBOL, ifnotfound=NA))), " (" ,row.names(bc.contrast.sub), ")", sep="")
  # decide if you are going to show row labels in heatamp
  if(nrow(bc.contrast.sub)>100){showrownames_var <-  FALSE}else{showrownames_var <- TRUE}
  # both genders
  pheatmap(bc.contrast.sub, color=c("blue", "red"), cluster_rows=TRUE, cluster_cols=FALSE, show_colnames=F, show_rownames=showrownames_var, fontsize=18, fontsize_row=10, annotation=heatmap.annots, annotation_colors=ann_colors, legend_breaks=c(0,1), main=paste("Barcoded Intensities for ", contrast, " comparison",  sep=""))
  # female only
  bc.contrast.sub.female <- bc.contrast.sub[,which(pData.contrast$gender=="FEMALE")]
  pheatmap(bc.contrast.sub.female, color=c("blue", "red"), cluster_rows=TRUE, cluster_cols=FALSE, show_colnames=F, show_rownames=showrownames_var, annotation=subset(heatmap.annots, gender=="FEMALE"), legend_breaks=c(0,1), main=paste("Barcoded Intensities for ", contrast, " comparison (FEMALE)",  sep=""), fontsize=18, fontsize_row=8, annotation_colors=ann_colors)
  # male only
  bc.contrast.sub.male <- bc.contrast.sub[,which(pData.contrast$gender=="MALE")]
  pheatmap(bc.contrast.sub.male, color=c("blue", "red"), cluster_rows=TRUE, cluster_cols=FALSE, show_colnames=F, show_rownames=showrownames_var,annotation=subset(heatmap.annots, gender=="MALE"), legend_breaks=c(0,1), main=paste("Barcoded Intensities for ", contrast, " comparison (MALE)",  sep=""), fontsize=18, fontsize_row=8, annotation_colors=ann_colors)
  }
```

#### Excel files with probeIDs, gene symbols, entropy values and barcoded expression values for each comparison

[Adult vs. Child](./results/PBMC/U133_2.0Plus.barcode.entropy.cutoff.0.3.adult-child.xls)  
[Adult vs. Neonate](./results/PBMC/U133_2.0Plus.barcode.entropy.cutoff.0.3.adult-neonate.xls)  
[Neonate vs. Child](./results/PBMC/U133_2.0Plus.barcode.entropy.cutoff.0.3.neonate-child.xls) 

For the second pass, I imposed stricter criteria for intra-group entropy (<0.275) and let the entropy across both groups be higher (>0.3) than this value.
- this is the actually about the lowest you can set the intra-group entropy and still get any results with higher entropy across the groups

You can see from the heatmaps that there are fewer study specific effects. The tradeoff is that you get fewer genes to work with.

```{r PBMCfilters2}
intra.entropycutoff <- 0.275
inter.entropycutoff <- 0.3
```

```{r PBMCbarcode_diff_heatmap0.25, ref.label='PBMCbarcode_diff_heatmap0.3', fig.width=18, fig.height=18}
```

#### Excel files with probeIDs, gene symbols, entropy values and barcoded expression values for each comparison

[Adult vs. Child](./results/PBMC/U133_2.0Plus.barcode.entropy.cutoff.0.275.adult-child.xls)  
[Adult vs. Neonate](./results/PBMC/U133_2.0Plus.barcode.entropy.cutoff.0.275.adult-neonate.xls)  
[Neonate vs. Child](./results/PBMC/U133_2.0Plus.barcode.entropy.cutoff.0.275.neonate-child.xls)

[1]: http://bioinformatics.oxfordjournals.org/content/19/2/185 (RMA)
[2]: http://www.bioconductor.org/packages/release/bioc/html/limma.html (limma)
[3]: http://barcode.luhs.org (Barcode)
[4]: http://www.bioconductor.org (BioC)
[5]: http://www.bioconductor.org/packages/2.6/bioc/html/frma.html (fRMA)

