```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, cache=FALSE,  highlight=TRUE, figalign="center", echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120))
options(width=200)
```

# PATHPRINT ANALYSES OF U133PLUS2.0 STUDIES

---

Analysis of public datasets for differential gene expression signatures between adults and children for [Les Kobzik](mailto:LKOBZIK@hsph.harvard.edu) as part of the DARPA 7-day Biodefense Program.

Contact [John Hutchinson](mailto:jhutchin@hsph.harvard.edu) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

----

# GENERAL SETUP

## General purpose R libraries necessary for analysis

```{r general_libraries_functions}
source("http://bioconductor.org/biocLite.R") # BioConductor script necessary for installing new BioC libraries with biocLite()
library(xtable) # table generation for reports
library(plyr) # library for iteratively working with data structures
library(ggplot2) # library for plotting 
library(RColorBrewer) # library for generating color palettes
library(googleVis) # library for presenting tables
library(pheatmap)# pretty heatmaps
library(Biobase)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")

# formula to count number of occurrences of each of the pathprint outcomes
count_trinary_outcomes <- function(vector){
  c(length(which(vector==-1)),length(which(vector==0)),length(which(vector==1)) )
  }
```

## Locations of directories for data, metadata and results

```{r general_directories}
if (file.exists("/n/hsphS10/hsphfs1/chb/projects/lk_FOY")) {
  baseDir <- "/n/hsphS10/hsphfs1/chb/projects/lk_FOY"
  }  else if (file.exists("/Volumes/home08/jhutchin/consults/lk_FOY/")) {
    baseDir <- "/Volumes/home08/jhutchin/consults/lk_FOY"
    } else {
      baseDir <- "/Volumes/ody/consults/lk_FOY"
      }
dataDir <- file.path(baseDir, "data")
resultsDir <- file.path(baseDir, "results")
metaDir <- file.path(baseDir, "meta")

## these are the arrays that showed problems in the raw data QC
exclude.IDs <-  c("GSM493776","GSM493778","GSM493779","GSM493780","GSM493788","GSM493790","GSM493794","GSM493798","GSM493799","GSM493802","GSM493815","GSM647538","GSM647539","GSM647566","GSM648632","GSM648694","T1D_C3_3","T1D_C2_3","T1D_C2_5","T1D_C2_6","T1D_C2_4","T1D_C3_2","T1D_C3_5","T1D_C1_5","T1D_C3_1","T1D_C3_4","GSM565090","GSM647608","GSM648630","GSM648631","GSM648691","GSM648692","GSM648693","T1D_C1_4","T1D_C1_1","T1D_C1_2","T1D_C1_3","T1D_C2_2","T1D_C2_1","GSM270838","GSM401075","GSM401110","GSM401130","GSM402322","GSM402328","GSM468380","GSM468381","GSM468390","GSM468391","GSM468392","GSM468397","GSM912849","GSM912841","GSM912838","HPBT105","HPBT101","HPBT108","HPBT106","HPBT107","HPBT102","HPBT103","HPBT104","153579hp133a11")
````

## Load in raw and RMA normalized  microarray data from generic microarray [analysis][1]

```{r load_affybatch, echo=FALSE}
load(file.path(resultsDir, "RDATA.raw_and_normalized_microarray.data_U133Plus2.0"))
pd <- pData(mic.norm.eset)
```

---

# Pathprint
## Load libraries

```{r pathprint_libraries}
library(pathprint)
library(limma)
library(ape)
# annotations for the hgU1332.0Plus array
library(hgu133plus2.db)
```

## Pathprint samples - default settings

```{r pathprint, cache=TRUE}
mic.platform <- "GPL570"
mic.species <- "human"
mic.names <- pData(mic.norm.eset)$ID
mic.f.default <- exprs2fingerprint(exprs = exprs(mic.norm.eset), platform = mic.platform, species = mic.species, progressBar = FALSE)
dimnames(mic.f.default)[[2]] <- pData(mic.norm.eset)$ID
```

```{r nexus, echo=FALSE}
Nexus <- function (matrixname, file, dataType = "discrete") {
  mat <- t(matrixname)
  manb <- dim(mat)[2]
  innb <- dim(mat)[1]
  indnames <- rownames(mat)
  charnames <- colnames(mat)
  print("Removing punctuation from taxa and character names")
  indnames <- gsub("[[:punct:]]", "", indnames)
  indnames <- gsub(" ", "", indnames)
  charnames <- gsub("[[:punct:]]", "", charnames)
  charnames <- gsub(" ", "", charnames)
  if (sum(duplicated(indnames) > 0)) 
    stop("duplicated taxa names")
  if (sum(duplicated(charnames) > 0)) 
    stop("duplicated character names")
  for (i in 1:length(charnames)) {
    charnames[i] <- paste(i, charnames[i], sep = " ")
    }
  charstring <- paste(charnames, collapse = ", ")
  matm <- mat
  symbols <- paste(unique(as.character(mat)), collapse = "")
  print(paste("compiling NEXUS file using", dataType, "characters", 
              sep = " "))
  cat("#NEXUS", "\n", "\n", "begin taxa;", "\n", "dimensions ntax=", 
      innb, ";", "\n", file = file, sep = "")
  cat("taxlabels", "\n", file = file, sep = "", append = TRUE)
  for (i in 1:innb) {
    cat(indnames[i], "\n", file = file, sep = "", append = TRUE)
    }
  cat(";", "\n", "end;", "\n", "\n", file = file, sep = "", 
      append = TRUE)
  cat("begin characters;", "\n", "dimensions nchar=", manb, 
      " ntax=", innb, ";", "\n", file = file, sep = "", append = TRUE)
  if (dataType == "discrete") {
    cat("Format datatype=Standard symbols=", "\"", symbols, 
        "\"", ";", "\n", file = file, sep = "", append = TRUE)
    }
  if (dataType == "continuous") {
    cat("Format datatype=continuous", ";", "\n", file = file, 
        sep = "", append = TRUE)
    }
  cat("CHARSTATELABELS", "\n", file = file, sep = "", append = TRUE)
  cat(charstring, " ;", "\n", file = file, sep = "", append = TRUE)
  cat("matrix", "\n", file = file, sep = "", append = TRUE)
  for (j in 1:innb) {
    if (dataType == "discrete") {
      cat(indnames[j], "\t", matm[j, ], "\n", file = file, 
          sep = "", append = TRUE)
      }
    if (dataType == "continuous") {
      cat(indnames[j], "\t", file = file, sep = "", append = TRUE)
      cat(matm[j, ], file = file, sep = " ", append = TRUE)
      cat("\n", file = file, sep = "", append = TRUE)
      }
    }
  cat(";", "\n", "end;", "\n", "\n", file = file, sep = "", 
      append = TRUE)
  }
mic.f.default.adult <- mic.f.default[,unlist(pd$ID[which(pd$stage=="adult")])]
colnames(mic.f.default.adult) <- paste("adult", colnames(mic.f.default.adult), sep="_")
mic.f.default.child <- mic.f.default[,unlist(pd$ID[which(pd$stage=="child")])]
colnames(mic.f.default.child) <- paste("child", colnames(mic.f.default.child), sep="_")
mic.f.default.adult.child <- cbind(mic.f.default.adult, mic.f.default.child)
Nexus((mic.f.default.adult.child+1), file.path(resultsDir, "fingerprints.adult.child.nxs"))
```

### Default Pathway Differentiation

Took a similar general approach as I took for the barcoding [analyses][1].
i.e. minimize intra-group variation while maximizing inter-group variation.

```{r pathwaydiff_vars}
# cutoffs used to filter the data
fisherpval.cutoff=1e-5
ttest.cutoff=1e-6
sd.cutoff=0.6
```

For this, I set two filtering criteria:
1) maximize homogeneity within an age group - based on minimizing the standard deviation (as I have no idea how to calculate a trinary entropy), cutoff used was sd<`r sd.cutoff`
2) maximize differences between groups - using ttests, cutoff used here was p<`r fisherpval.cutoff`

#### Setup design and contrast matrices 

Same setup as for the previous limma [analysis][2]

```{r design}
design <- model.matrix(~ -1+factor(pd$stage))
# make sure the headings match
colnames(design) <- sub("factor.pd.stage.", "", colnames(design))
```

```{r contrastmatrix}
contrast.matrix <- makeContrasts(adult-child,adult-neonate,neonate-child, levels=dimnames(design)[[2]])
```

#### Heatmaps
- clustered by both column (sample) and row (pathway)

```{r heatmap_colors}
# annotation colors
study_colors <- c("#FF0000","#00FF00", "#0000FF", cbPalette )
names(study_colors) <- unique(unlist(pd$study))
stage_colors <- c("white", "darkgrey", "black")
names(stage_colors) <- unique(unlist(pd$stage))
gender_colors <- c("cyan", "pink")
names(gender_colors) <- unique(unlist(pd$gender))
ann_colors = list(study = study_colors, stage = stage_colors, gender=gender_colors)
```


```{r default_pathprint_contrasts, fig.width=12, fig.height=12}
# for each comparison
for (n in 1:3) {
  contrast <- dimnames(contrast.matrix)$Contrasts[n]
  contrast <- gsub(" ", "", contrast)
  stage.1 <- unlist(strsplit(contrast, "-"))[1]
  stage.2 <- unlist(strsplit(contrast, "-"))[2]
  # get subset of barcoded data and stages
  mic.f.default.contrast <- mic.f.default[,pData(mic.norm.eset)$stage %in% c(stage.1, stage.2)]
  pData.contrast <- pData(mic.norm.eset)[pData(mic.norm.eset)$stage %in% c(stage.1, stage.2),] 
  contrast.factor <- as.factor(pData.contrast$stage)
  sds.stage = adply(mic.f.default.contrast, 1, function(x) {
    by(x, contrast.factor, function(x) sd(x))
    })
  row.names(sds.stage) <- sds.stage$X1
  sds.stage$X1 <- NULL
  sds.stage <- as.matrix(sds.stage)
  mode(sds.stage) <- "numeric"
  # t-tests between groups
  ttestpvals.contrast <- apply(mic.f.default.contrast, 1, function(x) {
    try(t.test(x~contrast.factor)$p.value)
    })
  ttestpvals.contrast<-as.numeric(ttestpvals.contrast)
  # filter genes by intra-group variation and across group differences
  mic.f.default.contrast.sub <- mic.f.default.contrast[sds.stage[,stage.1]<=sd.cutoff  & sds.stage[,stage.2]<=sd.cutoff & ttestpvals.contrast<ttest.cutoff &  !is.na(ttestpvals.contrast),]
  dim(mic.f.default.contrast.sub)
  ## Heatmaps
  # annotations for heatmaps
  heatmap.annots <- pData.contrast[,c("ID", "study", "stage", "gender")]
  heatmap.annots <- as.data.frame(apply(heatmap.annots, 2, unlist))
  row.names(heatmap.annots) <- heatmap.annots$ID
  heatmap.annots$ID <- NULL
  # both genders
  pheatmap(mic.f.default.contrast.sub, color=c("blue", "white", "red"), cluster_rows=T, cluster_cols=T, show_colnames=F, annotation=heatmap.annots,  legend_breaks=c(-1,0,1),main=paste("Pathways for ", contrast, "comparison",  sep=" "), fontsize=14, fontsize_row=10, annotation_colors=ann_colors)
  }
```

## Optimizing Pathprint thresholds

### Calculate SCEs and POEs

```{r POE_thresholds}
library(metaArray)
library(foreach)
library(multicore)
library(doMC)
registerDoMC(cores = 2)

gsdb=get(genesets[mic.species])
geo.ann=customCDFAnn(exprs(mic.norm.eset), chipframe[[mic.platform]]$ann)
SCE=single.chip.enrichment(exprs=geo.ann, geneset=gsdb,transformation = "squared.rank", statistic = "mean",normalizedScore = FALSE, progressBar = FALSE)

pb <- txtProgressBar(min = 0,max = length(pathprint.Hs.gs),style = 3)
POE<- foreach (i = names(pathprint.Hs.gs)) %dopar% {
  # load the SCG values for each pathway
  sample <- SCE[i,]
  # setup POE vector
  sample.POE<-sample
  sample.POE[]<-NA
  # subset to avoid NAs
  sample.valid<-sample[!(is.na(sample))]
  fit<-fit.em(sample.valid,cl = rep(0,length(sample.valid)))
  setTxtProgressBar(pb, match(i, names(pathprint.Hs.gs)))
  sample.POE[!(is.na(sample))]<-fit$expr
  }
# correct for NA rows
if(sum(sapply(POE, length) == 0) > 0){
  POE[[which(sapply(POE, length) == 0)]] = rep(0, ncol(SCE))
  }
POE.matrix<-as.matrix(t(as.data.frame(POE)))
rownames(POE.matrix)<-names(pathprint.Hs.gs)
colnames(POE.matrix)<-colnames(SCE)
dimnames(POE.matrix)[[2]] <- pData(mic.norm.eset)$ID
save(POE.matrix, file =file.path(resultsDir, "POE.matrix.RData"))
```

### Heatmap of POEs for Adult-Child comparison

Here I used the POEs for the adult-child comparison to construct a heatmap. Pathways were filtered based on t-test pvalues between the two groups.

```{r POE.ttests, fig.width=18, fig.height=18}
n=1
contrast <- dimnames(contrast.matrix)$Contrasts[n]
contrast <- gsub(" ", "", contrast)
stage.1 <- unlist(strsplit(contrast, "-"))[1]
stage.2 <- unlist(strsplit(contrast, "-"))[2]
contrast <- dimnames(contrast.matrix)$Contrasts[n]
contrast <- gsub(" ", "", contrast)
stage.1 <- unlist(strsplit(contrast, "-"))[1]
stage.2 <- unlist(strsplit(contrast, "-"))[2]
# subset metadata
pd.contrast <- pd[which(pd$stage %in% c(stage.1, stage.2)),]
contrast.factor <- as.factor(pd.contrast$stage)
# annotations for heatmaps
heatmap.annots <- pd.contrast[,c("ID", "study", "stage", "gender")]
heatmap.annots <- as.data.frame(apply(heatmap.annots, 2, unlist))
row.names(heatmap.annots) <- heatmap.annots$ID
heatmap.annots$ID <- NULL
# subset data
POE.matrix.contrast <- POE.matrix[,which(pd$stage %in% c(stage.1, stage.2))]
# t-tests for inter-group differences
POE.ttests <- apply(POE.matrix.contrast, 1, function(x) {
    try(t.test(x~contrast.factor)$p.value)
    })
pheatmap(POE.matrix.contrast[which(POE.ttests<1e-2) ,], color=rev(brewer.pal(11,"RdBu")), cluster_rows=T, cluster_cols=T, show_colnames=F, annotation=heatmap.annots,  main=paste("Pathways for ", contrast, "comparison",sep=" "), fontsize=14, fontsize_row=10, annotation_colors=ann_colors, show_rownames=F)
```

### Intra and inter group mean euclidean distances
- for different thresholds for the adult-child comparisons
- measure of how the different thresholds affect clustering of the sample groups
  - is there a threshold that separates the two groups while maximizing similarity within each group?

```{r threshold_euclideans}
library(fields) # for the euclidean distance calculations
calc_euclidean_mean <- function(x,y, fingerprint, contrasts) {
  dists <- rdist(t(fingerprint)[which(contrasts==x),],t(fingerprint)[which(contrasts==y),])
  dists <- dists[upper.tri(dists)]
  mean.dist <- mean(dists)
  return(mean.dist)
  }  
# subset data and metadata
pd.contrast <- pd[which(pd$stage %in% c("adult", "child")),]
contrast.factor <- as.factor(pd.contrast$stage)
POE.matrix.contrast <- POE.matrix[,which(pd$stage %in% c("child", "adult"))]

# calculate the mean euclidean distances for the three comparisons (intra-adult, intra-child and inter-child-adult)
distances <- sapply(seq(0,1,0.0005), function(threshold) {
  f.threshold<-(POE.matrix.contrast > threshold) - (POE.matrix.contrast < - threshold)
  adult_adult <- calc_euclidean_mean("adult", "adult", f.threshold, contrast.factor)
  child_child <- calc_euclidean_mean("child", "child", f.threshold, contrast.factor)
  adult_child <- calc_euclidean_mean("adult", "child", f.threshold, contrast.factor)
  return(list(threshold=threshold,intraadult=adult_adult, intrachild=child_child, inter=adult_child)    )
  })
#setup for ggplot
distances <- as.data.frame(t(distances))
library(reshape2)
library(scales)  
distances.m <- melt(distances, id="threshold")

# plot mean euclideand distances as a funciton of threshold used
ggplot(distances.m, aes(x=unlist(threshold), y=unlist(value), colour=variable)) + geom_line()+ ggtitle("Mean euclidean  distances within and between adult and child sample fingerprints") + xlab("threshold")  + ylab("mean euclidean distance")

ggplot(distances.m, aes(x=unlist(threshold), y=unlist(value), colour=variable)) + geom_line()+ scale_x_continuous(trans=log2_trans())+ggtitle("Mean euclidean  distances within and between adult and child sample fingerprints") + xlab("threshold (log2 scale)")  + ylab("mean euclidean distance")

# try to find the threshold where the mean inter-distance is largest relative to the intra-distance
distances <- apply(distances, 2, as.numeric)
distance.ratios <- distances[,"inter"]/rowMeans(distances[,c("intrachild", "intraadult")]) 
qplot(distances[,"threshold"],distance.ratios )+xlab("threshold")+ylab("inter/mean(intra)")
qplot(distances[,"threshold"],distance.ratios )+xlab("threshold (log2 scale)")+ylab("inter/mean(intra)")+ scale_x_continuous(trans=log2_trans())
```



[1]: https://dl.dropboxusercontent.com/u/4253254/Consults/lk_FOY/April.18.2013/barcode.microarray.U133Plus2.0.html (previous barcode analysis)
[2]: https://dl.dropboxusercontent.com/u/4253254/Consults/lk_FOY/April.18.2013/Generic.microarray.U133Plus2.0.html (previous microarray analysis)