```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, cache=FALSE,  highlight=TRUE, figalign="center", 
               echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, 
               tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120), 
               fig.path="figure/WB-pathprint_multi-", 
               cache.path="cache/WB-pathprint_multi-")
options(width=200)
```

# PATHPRINT ANALYSES OF STUDIES ON MULTIPLE PLATFORMS
- using GEO normalized data

---

Analysis of public datasets for differential gene expression signatures between
adults and children for [Les Kobzik](mailto:LKOBZIK@hsph.harvard.edu) as part of
the DARPA 7-day Biodefense Program. Normal phenotypes.

Contact [John Hutchinson](mailto:jhutchin@hsph.harvard.edu) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

----

# GENERAL SETUP

## General purpose R libraries necessary for analysis

```{r general_libraries, results='hide'}
source("http://bioconductor.org/biocLite.R") # BioConductor script necessary for installing new BioC libraries with biocLite()
library(plyr) # library for iteratively working with data structures
library(ggplot2) # library for plotting 
library(RColorBrewer) # library for generating color palettes
library(googleVis) # library for presenting tables
source("http://dl.dropboxusercontent.com/u/4253254/Resources/functions.r")
#library(devtools) # install libraries from github
#install_git("git://github.com/hbc/CHBUtils.git") # misc personal utilities
#library(dplyr)
library(knitr)
```

## Locations of directories for data, metadata and results

```{r general_directories}
if (file.exists("/n/hsphS10/hsphfs1/chb/projects/lk_FOY/")) {
  baseDir <- "/n/hsphS10/hsphfs1/chb/projects/lk_FOY/"
  }  else if (file.exists("/Volumes/home08/jhutchin/consults/lk_FOY/")) {
    baseDir <- "/Volumes/home08/jhutchin/consults/lk_FOY/"
    } else {
      baseDir <- "/Volumes/ody/consults/lk_FOY/"
      }
dataDir <- file.path(baseDir, "data")
resultsDir <- file.path(baseDir, "results", "WB", "multiplatform")
metaDir <- file.path(baseDir, "meta", "WB")
````

----

# Pathprint Analysis

## Setup

### Load libraries for Pathprint analysis and visualization

```{r pathprint_libraries, results='hide'}
library(GEOquery)
library(ArrayExpress)
library(pathprint)
library(limma)
library(ape)
library(pheatmap)
library(affy)
```

### Setup some more variables for the  analyses

```{r microarray_analysis_variables}
# colorblind friendly palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")
# age ranges to compare
child.age.range <- c(5,12)
adult.age.range <- c(18,50)
neonate.age.range <- c(0,4)
if (any(c(any(seq(adult.age.range[1], adult.age.range[2]) %in% seq(child.age.range[1], child.age.range[2])), 
any(seq(adult.age.range[1], adult.age.range[2]) %in% seq(neonate.age.range[1], neonate.age.range[2])),
any(seq(neonate.age.range[1], neonate.age.range[2]) %in% seq(child.age.range[1], child.age.range[2]))))) {
  stop("age ranges cannot overlap")
}
```

### Age ranges and names of the groups  
  - neonate = `r neonate.age.range[1]`-`r neonate.age.range[2]` years
  - child =`r child.age.range[1]`-`r child.age.range[2]` years
  - adult = `r adult.age.range[1]`-`r adult.age.range[2]` years

## Subset Data
- subsetted the unified metadata file to this platforms that are actually covered by Pathprint
- subsetted to samples for which we have ages

```{r subset_data}
refined.metadata <- read.delim(file.path(metaDir,"unified.metadata.refined.May2014.tab"))
refined.metadata <- refined.metadata[!is.na(refined.metadata$age),]
refined.metadata <- refined.metadata[refined.metadata$platform %in% names(platform.thresholds),]

write.table(refined.metadata, 
            file.path(metaDir, "unified.metadata.refined.pathprint_plaforms.May2014.tab"), 
            quote=F, 
            sep="\t", 
            row.names=F, 
            col.names=T)
```

That revised metadata file is [here](../meta/WB/unified.metadata.refined.pathprint_plaforms.May2014.tab)

### Parse the unified, refined and subsetted metadata file 
- to make a covariate table that allows you to download the data using the [GEOquery][2] Bioconductor library
- limit samples to those within the age ranges described above and label each with their "stage"
- removed studies removed in previous normal analyses (GSE18897, MEXP-2917, TABM-940 & TABM-666)

```{r covariatedataframe}
covars <- refined.metadata

# remove previously rejected studies
reject.studies <- c("GSE18897", "MEXP-2917", "TABM-940", "TABM-666")
covars <- covars[!(covars$study %in% reject.studies),]

# convert character ages to numeric to allow numeric subsetting 
covars$age <- as.numeric(as.character(covars$age))

## limit samples to those within the described child, adult and neonate age ranges
children.indices <- which(covars$age<=max(child.age.range) & 
  covars$age>=min(child.age.range))
neonate.indices <- which(covars$age<=max(neonate.age.range))
adult.indices <- which(covars$age<=max(adult.age.range) & 
  covars$age>=min(adult.age.range))
sample.indices <- c(neonate.indices, children.indices, adult.indices)
covars <- covars[sample.indices,]

# label samples with their stage as determined by age
# don't need to specify adult ranges, as we limited the dataset above to neonates, children and adults
covars$stage <- ifelse(covars$age<=max(neonate.age.range), "neonate", 
                       ifelse(covars$age>=min(child.age.range)
                              & covars$age<=max(child.age.range), "child", 
                              ifelse(covars$age>=min(adult.age.range)
                                     & covars$age<=max(adult.age.range), "adult", NA)))
covars <- covars[order(covars$stage),]
#remove unclassified samples
covars <- covars[!is.na(covars$stage),]

covars <- droplevels(covars)
write.table(covars, 
            file.path(metaDir, "unified.metadata.refined.pathprint_plaforms.targetted.samples.May2014.tab"), 
            quote=F, 
            sep="\t", 
            row.names=F, 
            col.names=T)

```

That revised metadata file is [here](../meta/WB/"unified.metadata.refined.pathprint_plaforms.targetted.samples.May2014.tab)


All studies have only one platform:

```{r platformtable, results='asis'}
platforms.table <- ddply(covars, "study", function(df) table(df$platform))
platforms.table <- col2rownames(platforms.table, "study", removecol = TRUE)
totals <- apply(platforms.table, 2, sum)

platforms.table <- rbind(platforms.table, totals)
row.names(platforms.table)[nrow(platforms.table)] <- "total"
kable(platforms.table)
```

### Download Expression data from GEO
- for each GEO study
  - pull down a study's expression data
    - subset study's expression data to samples of interest (using covars)
    - pathprint study
- combine pathprinted data

```{r pathprintdata, results='hide', eval=FALSE}
pathprint.GPLs <- names(platform.thresholds)

fingerprints.GEO.l <- lapply(unique(covars$study), function(GSE){
  print(GSE)
  GEOdata <- getGEO(GSE, GSEMatrix=TRUE)
  
  #GSE13015 and GSE3284 have more than one platform (GPL) per GSE, so need to handle those
  # if more than one platform in GSE, subset to platform that is covered by pathprint
  if (length(GEOdata)==1) {# only one GPL within the GSE
    GEOdata <- GEOdata[[1]]
    
    } else { # more than one platform in the GSE
      GPLs <-  sapply(GEOdata, function(study) annotation(study))
      GPL.in.pathprint <- which(GPLs %in% pathprint.GPLs) 
      
      if(length(GPL.in.pathprint)==1){ # more than one platform in GSE but only one covered by pathprint
        GEOdata <- GEOdata[[GPL.in.pathprint]]
        
        } else if (length(GPL.in.pathprint)==0) {
          stop("no valid pathprint platforms in this study")
          
          } else { # more than one platform covered by pathprint in the GSE 
            # which GSE indexes (platforms) have samples you want?
            GPLS.with.samples <- which(unlist(lapply(GEOdata, function(study){
              any(pData(study)$geo_accession %in% covars$sampleID)
              })))
            
            if (length(GPLS.with.samples)>1){ # samples split across more than one pathprinted platform            
              stop("more than one valid pathprinted platform found in this GSE that you want samples from; script can only handle one at a time")
              } else { # samples are only on one platform
                GEOdata <- GEOdata[[GPLS.with.samples]]
                }
            }
      }
  # get intensity values
  eset <- exprs(GEOdata)
  # subset to our samples of interest
  eset <-  eset[,as.character(covars$sampleID[covars$study %in% GSE]), drop=FALSE]
  # write data to file for Les
  write.table(eset, file=file.path(dataDir, GSE, paste(GSE, "GEO.eset.tab", sep=".")))
  # run pathprint on subsetted data
  mic.platform <- annotation(GEOdata)
  mic.species <- "human"
  mic.names <- colnames(eset)
  mic.f.default <- exprs2fingerprint(exprs = eset, platform = mic.platform, species = mic.species, progressBar = FALSE)
  return(mic.f.default)
  })

fingerprints <- do.call(cbind, fingerprints.GEO.l)
```


```{r unify, eval=FALSE, echo=FALSE}
save(list="fingerprints",file=file.path(resultsDir, "pathprints.RData"))
q()
```

```{r loaddata, eval=TRUE, echo=FALSE}
## hack to get around caching issues with knitr
load(file.path(resultsDir, "pathprints.RData"))
```

## Default Pathway Differentiation

Find pathways with minimal intra-group variation and maximal inter-group variation.

```{r pathwaydiff_vars}
# cutoffs used to filter the data
ttest.cutoff=1e-10
sd.cutoff=0.5
```

For this, I set two filtering criteria:
1) maximize homogeneity within an age group - based on minimizing the standard deviation (as I have no idea how to calculate a trinary entropy), cutoff used was sd<`r sd.cutoff`
2) maximize differences between groups - using ttests, cutoff used here was p<`r ttest.cutoff`

### Setup design and contrast matrices 

```{r pathprint_design}
# reorder the fingerprints and metadata about the samples
fingerprints <- fingerprints[,order(colnames(fingerprints))]
pd <- covars[order(covars$sampleID),]

if(!identical(as.character(pd$sampleID),colnames(fingerprints))){
  stop("covariate and fingerprint samples are not identically ordered")
}
design <- model.matrix(~ 0 + stage,  data=pd)
# tweak the headings
colnames(design) <- sub("stage", "", colnames(design))

contrast.matrix <- makeContrasts(adult-child,adult-neonate,neonate-child, levels=colnames(design))
```

### Heatmaps
- clustered by both column (sample) and row (pathway)

#### Setup annotations

```{r pathprint_heatmap_colors}
# annotation colors
study_colors <- rep(cbPalette,3)[1:27]
names(study_colors) <- unique(unlist(pd$study))
stage_colors <- c("white", "darkgrey", "black")
names(stage_colors) <- unique(unlist(pd$stage))
gender_colors <- c("blue", "pink", "yellow")
genders <-as.vector(unique(unlist(pd$gender)))
genders[is.na(genders)] <- "unknown"
names(gender_colors) <- unique(unlist(genders))
ann_colors = list(study = study_colors, stage = stage_colors, gender = gender_colors)
```

#### Make comparisons and plot

```{r pathprint_default_contrasts, fig.width=12, fig.height=12}
# for each comparison
for (n in 1:3) {
  contrast <- dimnames(contrast.matrix)$Contrasts[n]
  contrast <- gsub(" ", "", contrast)
  stage.1 <- unlist(strsplit(contrast, "-"))[1]
  stage.2 <- unlist(strsplit(contrast, "-"))[2]
  # get subset of barcoded data and stages
  fingerprints.contrast <- fingerprints[,pd$stage %in% c(stage.1, stage.2)]
  pd.contrast <- pd[pd$stage %in% c(stage.1, stage.2),] 
  contrast.factor <- as.factor(pd.contrast$stage)
  sds.stage = adply(fingerprints.contrast, 1, function(x) {
    by(x, contrast.factor, function(x) sd(x))
    })
  row.names(sds.stage) <- sds.stage$X1
  sds.stage$X1 <- NULL
  sds.stage <- as.matrix(sds.stage)
  mode(sds.stage) <- "numeric"
  # t-tests between groups
  ttestpvals.contrast <- apply(fingerprints.contrast, 1, function(x) {
    try(t.test(x~contrast.factor)$p.value)
    })
  ttestpvals.contrast<-as.numeric(ttestpvals.contrast)
    
  # filter genes by intra-group variation and across group differences
  fingerprints.contrast.sub <- fingerprints.contrast[sds.stage[,stage.1]<=sd.cutoff  & sds.stage[,stage.2]<=sd.cutoff & ttestpvals.contrast<ttest.cutoff &  !is.na(ttestpvals.contrast),]
  #   fingerprints.contrast.sub <- fingerprints.contrast[sds.stage[,stage.1]<=sd.cutoff  & sds.stage[,stage.2]<=sd.cutoff & fisherpvals.contrast<ttest.cutoff &  !is.na(fisherpvals.contrast),]
  
  # remove any columsn with only NAs
  fingerprints.contrast.sub <- fingerprints.contrast.sub[!apply(fingerprints.contrast.sub, 1, function(n) all(is.na(n))),]
  dim(fingerprints.contrast.sub)
  
  ## Heatmaps
  # annotations for heatmaps
  heatmap.annots <- pd.contrast[,c("sampleID", "study", "stage", "gender")]
  heatmap.annots <- as.data.frame(apply(heatmap.annots, 2, unlist))
  row.names(heatmap.annots) <- heatmap.annots$sampleID
  heatmap.annots$sampleID <- NULL
  # getting the annotations and annotation colors to play nice
  heatmap.annots <- as.matrix(heatmap.annots)
  
  heatmap.annots[,"gender"][is.na(as.vector(heatmap.annots[,"gender"]))] <- "unknown"
  
  # order by stage
  heatmap.annots <- heatmap.annots[order(heatmap.annots[,"stage"]),]
  fingerprints.contrast.sub <- fingerprints.contrast.sub[,row.names(heatmap.annots)]
  
  # both genders
  if(nrow(fingerprints.contrast.sub)>0){
    pheatmap(fingerprints.contrast.sub, color=c("blue", "white", "red")    , cluster_rows=T, cluster_cols=FALSE, show_colnames=F, annotation=heatmap.annots,  legend_breaks=c(-1,0,1),main=paste("Pathways for ", contrast, "comparison - unclustered",  sep=" "), fontsize=14, fontsize_row=10, annotation_colors=ann_colors)
    }
  }
```


[1]: http://www.bioconductor.org (BioC)
[2]: http://www.bioconductor.org/packages/2.11/bioc/html/GEOquery.html (GEOquery_BioC_library)


