```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, cache=FALSE,  highlight=TRUE, figalign="center", echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120), fig.path="figure/WB-unify-", cache.path="cache/WB-unify-")
options(width=200)
```

# UNIFICATION AND STANDARDIZATION OF STUDIES

---

Analysis of public datasets for differential gene expression signatures between adults and children for [Les Kobzik](mailto:LKOBZIK@hsph.harvard.edu) as part of the DARPA 7-day Biodefense Program.

Contact [John Hutchinson](mailto:jhutchin@hsph.harvard.edu) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

----

# GENERAL SETUP

## General purpose R libraries necessary for analysis

```{r general_libraries}
source("http://bioconductor.org/biocLite.R") # BioConductor script necessary for installing new BioC libraries with biocLite()
library(xtable) # table generation for reports
library(plyr) # library for iteratively working with data structures
library(ggplot2) # library for plotting 
library(RColorBrewer) # library for generating color palettes
library(googleVis) # library for presenting tables
```

## Locations of directories for data, metadata and results

```{r general_directories}
if (file.exists("/n/hsphS10/hsphfs1/chb/projects/lk_FOY")) {
  baseDir <- "/n/hsphS10/hsphfs1/chb/projects/lk_FOY"
  }  else if (file.exists("/Volumes/home08/jhutchin/consults/lk_FOY/")) {
    baseDir <- "/Volumes/home08/jhutchin/consults/lk_FOY"
    } else {
      baseDir <- "/Volumes/ody/consults/lk_FOY"
      }
dataDir <- file.path(baseDir, "data")
resultsDir <- file.path(baseDir, "results", "WB")
metaDir <- file.path(baseDir, "meta", "WB")
````

----

# INITIAL DOWNLOAD of METATDATA
## Obtaining metadata for all studies 
I downloaded the metadata for all the whole blood studies Les shared in his [file](../meta/WB/MoreNormalsForPathPrintMay2014.xlsx)
I used two [Bioconductor][1] libraries to interface with the [Gene Expression Omnibus][2] and [Array Express][3] databases ([GEOquery][4] and [ArrayExpress][5] respectively). 

```{r database_download_libraries}
library(GEOquery)
library(ArrayExpress)
```

### Load in the series IDs of the GEO studies
The IDs of the GEO series (GSEs) are:

```{r GEOIDs}
GSE.ids <- c("GSE10715","GSE11375","GSE16028","GSE16059","GSE19743","GSE3284","GSE37171","GSE14844","GSE26378","GSE26440","GSE8121","GSE9692","GSE35571","GSE46449","GSE16059","GSE25414","GSE54247","GSE53195","GSE34404","GSE42826","GSE38900","GSE47728","GSE33828","GSE53195","GSE40396","GSE45603","GSE35846","GSE48152","GSE36382","GSE33321","GSE20142","GSE40561","GSE42026","GSE22098","GSE52005","GSE3846","GSE45878","GSE41890","GSE22255","GSE47353","GSE53823","GSE21311","GSE46743","GSE42002","GSE47727","GSE16561","GSE55319","GSE48348")
GSE.ids=sub(" .*" , "", GSE.ids)
GSE.ids=sub(" .*" , "", GSE.ids)
print(GSE.ids)
```

**There are `r length(GSE.ids)` GEO series in total**

### Data for each GEO series was downloaded via GEOquery into separate directories by series ID
- for series with more than one dataset (GDS), I numbered the metadata files separately, according to their order in the series  metadata
- for series with only one dataset, the metadata file contains the number "1"

```{r GEOquerypDATA, eval=FALSE}
for (GSE.id in GSE.ids) {
  # name of the directory receiveing the files
  gse.dataDir=file.path(dataDir, GSE.id) 
  # check to see if this directory exists already 
  if (!(file.exists(gse.dataDir))) { 
    # if it does not, create it
    dir.create(gse.dataDir) 
    }
  # pull down the available series data, using exception handling (i.e. try())
  gse <- try(getGEO(GSE.id, destdir=gse.dataDir)) 
  # for every dataset within the series
  for(n in 1:length(gse)){ 
    # grab the metadata
    metadata <- pData(gse[[n]]) 
     # discard columns you aren't interested in
    metadata <- metadata[,!grepl("data_row_count|status|submission_date|last_update_date|channel_count|scan_protocol|data_processing|hyb_protocol|taxid_ch1|label|contact", names(metadata))]
    # filename to writout the metadata, contains the series id and the number of the dataset 
    file.out <- paste("metadata", GSE.id, n, "tab", sep=".") 
    # write out to tab-delimited file, retaining column names
    write.table(metadata, file=file.path(gse.dataDir, file.out), quote=F, sep="\t", col.names=T, row.names=F) 
    # sleep 5 seconds between downloads so you don't hammer the GEO server, as it seems to be a bit unstable
    Sys.sleep(5) 
    }
  }
```


### Load in the Array Express IDs of the Array Express studies
The IDs of the Array Express studies are: 

```{r AEids}
#AE.ids <- c("MEXP-2917", "MEXP-3567", "MEXP-884", "MTAB-25", "TABM-666","TABM-940")
AE.ids <- c("E-MTAB-1174","E-MTAB-1573","E-MTAB-1573","E-MTAB-1708","E-MTAB-2232","E-TABM-1036")
print(AE.ids)
```

### Data for each Array Express series was downloaded with ArrayExpress into separate directories by series ID

```{r AEquerypData, eval=FALSE}
for (AE.id in AE.ids) {
  ae.dataDir=file.path(dataDir, AE.id)
  if (!(file.exists(ae.dataDir))) {
    dir.create(ae.dataDir)
  }
  ae <-   try(ArrayExpress(accession=AE.id, path=ae.dataDir, save=TRUE,  ))
  Sys.sleep(5)
}
```

- for the Array Express studies, there was no need to export the metadata from R as the ArrayExpress library downloads all raw data files and metadata
  - metadata was saved in *sdrf files

These GEO and ArrayExpress metadata files were then individually hand examined by Les to extract, if available:  
1. sample infomation for:
- unique sample identifier (for GSE studies, this is the GSM id, for ArrayExpress studies, the uniquely assigned study id)
- sample type (multiple if necessary)
- age
- gender
- microarray platform
- series ID
- database

The file with the unified (but not consistent) metadata can be found [here](../meta/WB/unified.metadata.unrefined.May2014.xlsx)

The end goal is a file containing the metadata for all control samples in all studies. To generate a file that can be then be used to load in the metadata and its respective raw data later in the analysis, I also needed consistent labeling. 

To make this file, I ran the "unified.metadata.unrefined.May2014.xlsx" file through [Google refine][6], merging multiple terms for the same thing into a single term. For example, terms such as "F", "Female", "Fem", and "f" were merged into the single term FEMALE. All ages are in years based units. 

The unified, refined metadata file can be found [here](../meta/WB/unified.metadata.refined.May2014.tab)

---


[1]: http://www.bioconductor.org (BioC)
[2]: http://www.ncbi.nlm.nih.gov/gds/ (GEO)
[3]: http://www.ebi.ac.uk/arrayexpress/ (ArrayExpress)
[4]: http://www.bioconductor.org/packages/2.11/bioc/html/GEOquery.html (GEOquery_BioC_library)
[5]: http://www.bioconductor.org/packages/2.11/bioc/html/ArrayExpress.html (ArrayExpress_BioC_library)
[6]: https://code.google.com/p/google-refine/ (Google_Refine)


