---
title: "Whangbo - Treatment response RNAseq Analysis"
output:
  html_document:
    code_folding: hide
    theme: yeti
    toc: yes
    toc_float: true
    toc_depth: 3
    number_sections: true
---

```{r setup, echo=FALSE}
library("knitcitations")
cleanbib()
options("citation_format" = "pandoc")

clientname="Jennifer Whangbo"
clientemail="Jennifer.Whangbo@childrens.harvard.edus"
labPI="Ritz"
lablocation="Dana Farber"
analystname="John Hutchinson"
analystemail="jhutchin@hsph.harvard.edu"

knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",cache=TRUE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE, prompt=TRUE, comment='', fig.cap='')
```

---

RNAseq analysis for `r clientname` (`r clientemail`), `r labPI` group at `r lablocation`.  

Contact `r analystname` (`r analystemail`) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

---

Methods

Discarded any samples which were not matched pre and post treatment for the same patient.

---

# Setup

```{r check-library-functions, echo=FALSE}
check_install = function(packages) {
   not_installed = setdiff(packages, rownames(installed.packages()))
   if(length(not_installed) > 0) {
      write(paste("The libraries", not_installed, "are not installed, aborting.",sep=" "), stdout())
      stop()
   }
}
```

```{r load-libraries, echo=FALSE}
packages = c("ggplot2", "reshape", "gplots", "edgeR", "CHBUtils", "pheatmap","DESeq2", "tximport", "DT", "DEGreport", "dplyr")
check_install(packages)
installed = lapply(packages, library, character.only = TRUE)
```

```{r qc-setup}
if (file.exists("/home/jnh7/projects/jw_rnaseq")){
  baseDir <- "/home/jnh7/projects/jw_rnaseq"
} else {
baseDir <- "~/Work/projects/jw_rnaseq"
}

project_summary = file.path(baseDir, "data/2016-11-02_bcbio/project-summary.csv")
counts_file = file.path(baseDir, "data/2016-11-02_bcbio/combined.counts")
tx2genes_file = file.path(baseDir, "data/2016-11-02_bcbio/tx2gene.csv")
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
summarydata = read.table(project_summary, header=TRUE, sep=",")
summarydata = summarydata[,colSums(is.na(summarydata)) < nrow(summarydata)]
# handle newer bcbio-nextgen runs that use description as the key
if("description" %in% colnames(summarydata)) {
  rownames(summarydata) = summarydata$description
  summarydata$Name = rownames(summarydata)
  summarydata$description = NULL
} else {
  rownames(summarydata) = summarydata$Name
  # summarydata$Name = NULL
}
summarydata = summarydata[order(rownames(summarydata)),]
if (file.exists(tx2genes_file)) {
  sample_dirs = file.path(baseDir, "data", rownames(summarydata))
  salmon_files = file.path(sample_dirs, "salmon", "quant.sf")
  sailfish_files = file.path(sample_dirs, "sailfish", "quant.sf")
  new_sailfish = file.path(sample_dirs, "sailfish", "quant", "quant.sf")
  new_salmon = file.path(sample_dirs, "salmon", "quant", "quant.sf")
  if (file.exists(salmon_files[1])) {
    sf_files = salmon_files
  } else if (file.exists(sailfish_files[1])) {
    sf_files = sailfish_files
  }  else if (file.exists(new_sailfish[1])) {
    sf_files = new_sailfish
  }  else if (file.exists(new_salmon[1])) {
    sf_files = new_salmon
  }
  names(sf_files) = rownames(summarydata)
  tx2gene = read.table(tx2genes_file, sep=",", row.names=NULL, header=FALSE)
  txi.salmon = tximport(sf_files, type="salmon", tx2gene=tx2gene, reader=readr::read_tsv, countsFromAbundance="lengthScaledTPM")
  counts = round(data.frame(txi.salmon$counts, check.names=FALSE))
} else {
  counts = read.table(counts_file, header=TRUE, row.names="id", check.names=FALSE)
}
counts = counts[, order(colnames(counts)), drop=FALSE]
colnames(counts) = gsub(".counts", "", colnames(counts))
y = DGEList(counts=counts)
y = calcNormFactors(y)
normalized_counts = cpm(y, normalized.lib.sizes=TRUE)

# this is a list of all non user-supplied metadata columns that could appear
known_columns = c("Name", "X.GC", "Exonic.Rate", "Sequences.flagged.as.poor.quality",
    "rRNA_rate", "Fragment.Length.Mean", "Intronic.Rate", "Intergenic.Rate",
    "Mapping.Rate", "Quality.format", "Duplication.Rate.of.Mapped", "Mapped",
    "rRNA", "Sequence.length", "Transcripts.Detected", "Mean.Per.Base.Cov.",
    "Genes.Detected", "Unique.Starts.Per.Read", "unique_starts_per_read",
    "complexity", "X5.3.bias", "Duplicates.pct", "Duplicates", "Mapped.reads",
    "Average.insert.size", "Mapped.reads.pct", "Total.reads",
    "avg_coverage_per_region", "Mapped.Reads")
summarydata[,"Fragment.Length.Mean"] = summarydata$Average.insert.size
metadata = summarydata[, !colnames(summarydata) %in% known_columns, drop=FALSE]
metadata = metadata[, colSums(is.na(metadata)) < nrow(metadata), drop=FALSE]
metadata$samplecode <- NULL

exclude.samples <- c("nr_pre_tcon_bi","nr_pre_treg_bi","nr_pre_cd8_fh","nr_wk4_nk_fh","nr_wk4_cd8_mld","nr_pre_nk_rl","nr_wk4_tcon_rl","nr_wk4_treg_rl","r_pre_treg_ls","r_wk4_treg_ta")
metadata <- metadata[!(row.names(metadata) %in% exclude.samples),]
summarydata <- summarydata[!(row.names(summarydata) %in% exclude.samples),]
counts <- counts[,!(names(counts) %in% exclude.samples)]
```

```{r sanitize-datatable}
sanitize_datatable = function(df, ...) {
 # remove dashes which cause wrapping
 DT::datatable(df, ..., rownames=gsub("-", "_", rownames(df)),
                   colnames=gsub("-", "_", colnames(df)))
}
```

```{r set-random-seed}
# set seed for reproducibility
set.seed(1454944673)
```

## Sample metadata
```{r summary-table, echo=FALSE, cache=FALSE}
sanitize_datatable(summarydata, style='bootstrap')
```

```{r heatmap-function}
get_heatmap_fn = function(summarydata) {
    # return the pheatmap function with or without metadata
    if(ncol(metadata) == 0) {
       return(pheatmap)
    }
    else {
    # rownames(metadata) = summarydata$Name
    heatmap_fn = function(data, ...) {
        pheatmap(data, annotation=metadata, clustering_method = "ward.D2",
        clustering_distance_cols = "correlation", ...)
    }
    return(heatmap_fn)
}}
heatmap_fn = get_heatmap_fn(summarydata)
```

# Quality control metrics

```{r qualimap-run}
qualimap_run = "Mapped" %in% colnames(summarydata)
do_quality = "Mapped.reads" %in% colnames(summarydata)
```

## Mapped reads
```{r mapped-plot, eval=qualimap_run}
ggplot(summarydata, aes(x=Name, y=Mapped)) +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    geom_bar(stat="identity") +
    ylab("mapped reads") + xlab("")
```

## Genomic mapping rate
```{r mapping-rate-plot, eval=qualimap_run}
ggplot(summarydata, aes(x=Name, y=Mapping.Rate)) +
    geom_bar(stat="identity") +
    ylab("mapping rate") + xlab("") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90))
```


## Number of genes detected
```{r genes-detected-plot}
dd = data.frame(Name=colnames(counts), Genes.Detected = colSums(counts > 0))
ggplot(dd, aes(x=Name, y=Genes.Detected)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("")
```

## Gene detection saturation
```{r saturation-plot, eval=do_quality}
col_mapped = ifelse(qualimap_run, "Mapped", "Mapped.reads")
dd = data.frame(Mapped=summarydata[,col_mapped], Genes.Detected = colSums(counts > 0))
ggplot(dd, aes(x=Mapped, y=Genes.Detected)) +
    geom_point() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("reads mapped")
```

## Exonic mapping rate
```{r exonic-mapping-plot, eval="Exonic.Rate" %in% colnames(summarydata)}
ggplot(summarydata, aes(x=Name, y=Exonic.Rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("exonic mapping rate") + xlab("")
```

## rRNA mapping rate
```{r check-rRNA-eval}
eval_rRNA = "rRNA_rate" %in% colnames(summarydata) & !sum(is.na(summarydata$rRNA_rate)) == nrow(summarydata)
```
```{r rRNA-rate-plot, eval=eval_rRNA & do_quality}
ggplot(summarydata, aes(x=Name, y=rRNA_rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("rRNA rate") + xlab("")
```

## 5'->3' bias
```{r 53bias, eval="X5.3.bias" %in% colnames(summarydata) & do_quality}
ggplot(summarydata, aes(x=Name, y=X5.3.bias)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("5'->3' bias") + xlab("")
```

## Boxplot of log10 counts per gene
```{r boxplot-raw}
melted = melt(counts)
colnames(melted) = c("sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Boxplot of log10 TMM-normalized counts per gene
Trimmed mean of M-values (TMM) normalization is described
[here](http://genomebiology.biomedcentral.com/articles/10.1186/gb-2010-11-3-r25)

Robinson, M. D., & Oshlack, A. (2010). A scaling normalization method for differential expression analysis of RNA-seq data. Genome Biology, 11(3). doi:10.1186/gb-2010-11-3-r25

```{r boxplot-normalized}
y = DGEList(counts=counts)
y = calcNormFactors(y)
normalized_counts = cpm(y, normalized.lib.sizes=TRUE)
melted = melt(normalized_counts)
colnames(melted) = c("gene", "sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Density of log10 TMM-normalized counts
```{r density-normalized}
ggplot(melted, aes(x=count, group=sample)) +
    geom_density() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Correlation heatmaps of TMM-normalized counts

```{r newcovars}
metadata$rRNA_rate <- summarydata$rRNA_rate
metadata$Exonic_rate <- summarydata$Exonic.Rate
```

### Tcon cells {.tabset}

#### Correlation (Pearson)
```{r pearson-heatmap-normalized.tcon, eval=ncol(counts) > 1}
celltype="Tcon"
normalized_counts.sub <- normalized_counts[,which(metadata$celltype==celltype)]
heatmap_fn(cor(normalized_counts.sub, method="pearson"))
```

#### Correlation (Spearman)
```{r spearman-heatmap-normalized.tcon, eval=ncol(counts) > 1}
heatmap_fn(cor(normalized_counts.sub, method="spearman"))
```

### Treg cells {.tabset}

#### Correlation (Pearson)
```{r pearson-heatmap-normalized.treg, eval=ncol(counts) > 1}
celltype="Treg"
normalized_counts.sub <- normalized_counts[,which(metadata$celltype==celltype)]
heatmap_fn(cor(normalized_counts.sub, method="pearson"))
```

#### Correlation (Spearman)
```{r spearman-heatmap-normalized.treg, eval=ncol(counts) > 1}
heatmap_fn(cor(normalized_counts.sub, method="spearman"))
```

### CD8 cells {.tabset}

#### Correlation (Pearson)
```{r pearson-heatmap-normalized.cd8, eval=ncol(counts) > 1}
celltype="CD8"
normalized_counts.sub <- normalized_counts[,which(metadata$celltype==celltype)]
heatmap_fn(cor(normalized_counts.sub, method="pearson"))
```

#### Correlation (Spearman)
```{r spearman-heatmap-normalized.cd8, eval=ncol(counts) > 1}
heatmap_fn(cor(normalized_counts.sub, method="spearman"))
```

### NK cells {.tabset}

#### Correlation (Pearson)
```{r pearson-heatmap-normalized.nk, eval=ncol(counts) > 1}
celltype="NK"
normalized_counts.sub <- normalized_counts[,which(metadata$celltype==celltype)]
heatmap_fn(cor(normalized_counts.sub, method="pearson"))
```

#### Correlation (Spearman)
```{r spearman-heatmap-normalized.nk, eval=ncol(counts) > 1}
heatmap_fn(cor(normalized_counts.sub, method="spearman"))
```



## PCA plots

```{r pca-plotter}
pca_plot = function(comps, nc1, nc2, colorby) {
   c1str = paste0("PC", nc1)
   c2str = paste0("PC", nc2)
  ggplot(comps, aes_string(c1str, c2str, color=colorby)) +
    geom_point() + theme_bw() +
    xlab(paste0(c1str, ": ", round(pc$percentVar[nc1] * 100), "% variance")) +
    ylab(paste0(c2str, ": ", round(pc$percentVar[nc2] * 100), "% variance"))
  }
```

### Tcon cells {.tabset}
`r celltype="Tcon"`

```{r pca.tcon, eval=ncol(counts) > 1}
cellindices <- which(metadata$celltype==celltype)
counts.sub <- counts[,cellindices]
summarydata.sub <- summarydata[cellindices,]
dds = DESeqDataSetFromMatrix(countData=counts.sub, colData=summarydata.sub, design=~Name)
vst = varianceStabilizingTransformation(dds)
pca_loadings = function(object, ntop=500) {
  rv <- matrixStats::rowVars(assay(object))
  select <- order(rv, decreasing = TRUE)[seq_len(min(ntop,
      length(rv)))]
  pca <- prcomp(t(assay(object)[select, ]))
  percentVar <- pca$sdev^2/sum(pca$sdev^2)
  names(percentVar) = colnames(pca$x)
  pca$percentVar = percentVar
  return(pca)}
pc = pca_loadings(vst)
comps = data.frame(pc$x)
comps$Name = rownames(comps)
library(dplyr)
comps = comps %>% left_join(summarydata, by=c("Name"="Name"))
colorby = "Name"
```

#### PC1 vs. PC2
```{r PC1-PC2.tcon}
pca_plot(comps, 1, 2, colorby)
```

#### PC3 vs. PC4
```{r PC3-PC4.tcon}
pca_plot(comps, 3, 4, colorby)
```

#### PC5 vs. PC6
```{r PC5-PC6.tcon}
pca_plot(comps, 5, 6, colorby)
```

#### Variance explained by component
```{r component-loadings.tcon}
ggplot(data.frame(component=reorder(names(pc$percentVar), -pc$percentVar),
                  percent_var=pc$percentVar),
       aes(component, percent_var)) + geom_bar(stat='identity') +
  ylab("percent of total variation") + xlab("") + theme_bw()
```


### Treg cells {.tabset}
`r celltype="Treg"`

```{r pca.treg, ref.label='pca.tcon', eval=ncol(counts) > 1}
```

#### PC1 vs. PC2
```{r PC1-PC2.treg, ref.label='PC1-PC2.tcon'}
```

#### PC3 vs. PC4
```{r PC3-PC4.treg, ref.label='PC3-PC4.tcon'}
```

#### PC5 vs. PC6
```{r PC5-PC6.treg, ref.label='PC5-PC6.tcon'}
```

#### Variance explained by component
```{r component-loadings.treg, ref.label='component-loadings.tcon'}
```


### CD8 cells {.tabset}
`r celltype="CD8"`

```{r pca.cd8, ref.label='pca.tcon', eval=ncol(counts) > 1}
```

#### PC1 vs. PC2
```{r PC1-PC2.cd8, ref.label='PC1-PC2.tcon'}
```

#### PC3 vs. PC4
```{r PC3-PC4.cd8, ref.label='PC3-PC4.tcon'}
```

#### PC5 vs. PC6
```{r PC5-PC6.cd8, ref.label='PC5-PC6.tcon'}
```

#### Variance explained by component
```{r component-loadings.cd8, ref.label='component-loadings.tcon'}
```


### NK cells {.tabset}
`r celltype="NK"`

```{r pca.nk, ref.label='pca.tcon', eval=ncol(counts) > 1}
```

#### PC1 vs. PC2
```{r PC1-PC2.nk, ref.label='PC1-PC2.tcon'}
```

#### PC3 vs. PC4
```{r PC3-PC4.nk, ref.label='PC3-PC4.tcon'}
```

#### PC5 vs. PC6
```{r PC5-PC6.nk, ref.label='PC5-PC6.tcon'}
```

#### Variance explained by component
```{r component-loadings.nk, ref.label='component-loadings.tcon'}
```




