---
title: "Cell deconvolutions - Normal lungs and lungs with distal tumors"
author: "`r getOption('author')`"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
params:
    bcbFile: "data/bcb_sub_lung.rda"
    outputDir: "."
---

```{r setup, message=FALSE}
library(bcbioRNASeq)
library(DeconRNASeq)
library(rio)


# Shared R Markdown settings
prepareRNASeqTemplate()
if (file.exists("setup.R")) {
    source("setup.R")
}

# Directory paths
dataDir <- file.path(params$outputDir, "data")

# Load bcbioRNASeq object
bcbName <- load(params$bcbFile)
bcb <- get(bcbName, inherits = FALSE)


kable(as.data.frame(colData(bcb)))
```

```{r header, child="_header.Rmd", eval=file.exists("_header.Rmd")}
```

The idea here is to compare the ratio of cell signatures of N1 an dN2 neutrophils in our tumor, normal lung and lungs with distal tumors to see if there are any shifts.

```{r counts, include=FALSE}
tpm <- tpm(bcb) %>% as.data.frame()
rowData <- rowData(bcb)

annotated <- cbind(as.data.frame(tpm), as.data.frame(rowData))

```

# Coffelt signatures

Import top 50 differentially expressed genes from Coffelt et al., 2015 (Nature). (from extended Table 1).

Subset to genes with pvalues <0.1

```{r signatures}
signatures <- import("~/Orchestra/PIs/sandra_mcallister/lung_metastasis/meta/coffelt_2015_extended_table1_top50DEgenes.csv")

pos_genes <- subset(signatures, pvalue<0.1 & direction=="positive")$ensemblid
neg_genes <- subset(signatures, pvalue<0.1 & direction=="negative")$ensemblid
```

Get the ratio of the geometric means for the signature genes for all samples in the dataset.

```{r geom_means}
gm_mean = function(x, na.rm=TRUE){
  exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))
}

tpm_pos <- tpm[pos_genes,] 
tpm_pos <- tpm_pos[!apply(tpm_pos, 1, function(x) all(is.na(x))),]

tpm_neg <- tpm[neg_genes,] 
tpm_neg <- tpm_neg[!apply(tpm_neg, 1, function(x) all(is.na(x))),]

negmeans <- tpm_neg %>% apply(., 2, function(x) gm_mean(x))
posmeans <- tpm_pos %>% apply(., 2, function(x) gm_mean(x))
ratios <- posmeans/negmeans %>% as.data.frame() 
ratios <-rownames_to_column(ratios, var="sampleID")
names(ratios) <- c("sampleID", "ratio")
```

Add group ids to data for plotting.
- remove anything with a RIN below 3

```{r annotateratios1}
ratios_annot  <- inner_join(ratios, as.data.frame(colData(bcb)))

# remove anything with a RIN below 2.5

ratios_annot$rin <-   as.numeric(as.character(ratios_annot$rin))
ratios_annot <- subset(ratios_annot, rin>3)

p <- ggplot(ratios_annot, aes(x=sampleclass, y=ratio, color=sampleclass))+geom_boxplot()
p +  geom_jitter(shape=16, size=3 , position=position_jitter(0.00))+ggtitle("Coffelt signature")

t.test(subset(ratios_annot, ratios_annot$sampleclass=="lung_distaltumor" & ratios_annot$rin>3)$ratio,subset(ratios_annot, ratios_annot$sampleclass=="lung_notumor" & ratios_annot$rin>3)$ratio)

```

# Shaul signatures

Import all the stats from the GEO2R recalculations and subset to genes with an adjusted pvalue of less than 0.05 and and lfc of `r lfc=2` `r lfc`

```{r importsigsshaul}
stats <- import(file.path(dataDir, "GSE101584_N2_over_N1.stats.csv"))

lfc=2
N1_stats <- subset(stats, adj.P.Val<0.05 & logFC>lfc)
dim(N1_stats)
N2_stats <- subset(stats, adj.P.Val<0.05 & logFC<(-lfc))
dim(N2_stats)
```


Get the ratio of the geometric means for the signature genes for all samples in the dataset.


```{r sub}
N2_tpms <- left_join(N2_stats, annotated, by=c("Gene.symbol"="symbol")) %>% select(contains ("tumor"), contains("lung"))
N2_tpms <- N2_tpms[!apply(N2_tpms, 1, function(x) all(is.na(x))),]

N1_tpms <- left_join(N1_stats, annotated, by=c("Gene.symbol"="symbol")) %>% select(contains ("tumor"), contains("lung"))
N1_tpms <- N1_tpms[!apply(N1_tpms, 1, function(x) all(is.na(x))),]

N2means <- N2_tpms %>% apply(., 2, function(x) gm_mean(x))
N1means <- N1_tpms %>% apply(., 2, function(x) gm_mean(x))
ratios <- N2means/N1means %>% as.data.frame() 
ratios <-rownames_to_column(ratios, var="sampleID")
names(ratios) <- c("sampleID", "ratio")
```

Add group ids to data for plotting.
- remove anything with a RIN below 3

```{r annotateratios2}
ratios_annot  <- inner_join(ratios, as.data.frame(colData(bcb)))

# remove anything with a RIN below 2.5

ratios_annot$rin <-   as.numeric(as.character(ratios_annot$rin))
ratios_annot <- subset(ratios_annot, rin>3)

p <- ggplot(ratios_annot, aes(x=sampleclass, y=ratio, color=sampleclass))+geom_boxplot()
p+   geom_jitter(shape=16, size=3 , position=position_jitter(0.00))+ggtitle("Shaul signature")

t.test(subset(ratios_annot, ratios_annot$sampleclass=="lung_distaltumor" & ratios_annot$rin>3)$ratio,subset(ratios_annot, ratios_annot$sampleclass=="lung_notumor" & ratios_annot$rin>3)$ratio)
  
```

