<!--
Analysis of anthrax infected cells vs. control cells
-->
```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, highlight=TRUE, fig.align='left', fig.show='hold', cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE, prompt=TRUE, comment='',
fig.cap='')
```
# Overview
A macrophage cell line, THP-1 was transduced via lentivral infection with a shRNA library
directed towards a subset of 4000 genes previously identified as important in
macrophage interaction with intracellular bacteria. The transduced, and non-transfected
control cells were exposed to anthrax for 48 hours until 90% of the cells were killed.
The cells were harvested and the DNA sequenced. The files were are looking at have
counts for each shRNA.

If a shRNA shows up less in the anthrax infected cells than in the control cells,
that means that the gene the shRNA targets might confer survivability to the
anthrax infection. Conversely if a shRNA shows up more often in the anthrax infected
cells than the control cells, that means the gene the shRNA targets might be involved
in the death of the cells.
```{r variables-and-libraries}
library(ggplot2)
library(reshape)
library(googleVis)
library(stringr)
library(edgeR)
library(extrafont)
font_import(pattern="Gill Sans", prompt=FALSE)
#project_dir = '/n/home05/kirchner/cache/projects/kobzik_influenza'
#data_dir = file.path(project_dir, 'data')
disease_type = 'rabbit_fever'
num_samples = 6
project_dir = '/Users/rory/cache/projects/kobzik_influenza'
data_dir = file.path(project_dir, 'raw_data')
results_dir = file.path(project_dir, 'results', disease_type)
wd = file.path(project_dir, 'scripts', disease_type)
setwd(wd)
in_file = file.path(data_dir, disease_type, 'mf_55K_THP1.csv')
#load(file.path(data_dir, 'pathways.Hs.RData'))
FDR_CUTOFF = 0.10
ensembl_gene = 'hsapiens_gene_ensembl'
gene_symbol = 'hgnc_symbol'
filter_type = 'refseq_mrna'
dir.create(results_dir, showWarnings=FALSE)
```

# Data cleaning and sanity checking

The initial data needs to be reformatted in a way we can use it:
```{r}
counts = read.table(in_file, header=TRUE, sep=",", comment.char="", stringsAsFactors=FALSE)
colnames(counts)
```

```{r}
refseq_id = unlist(lapply(counts$Info, function(x) strsplit(x, "|", fixed=TRUE)[[1]][2]))
refseq_id = unlist(lapply(refseq_id, function(x) strsplit(x, ".", fixed=TRUE)[[1]][1]))
# use these symbols, excel ruins he formatting of the symbol. column
# (for example Sept8 -> September 8. Nice)
symbols = unlist(lapply(counts$Label, function(x) strsplit(x, "|", fixed=TRUE)[[1]][1]))
head(symbols)
```
There are some errors in the formatting of the file, the label column is not always
consistent, we might have to fix the symbols column later using the Refseq ids.
```{r}
counts$refseq_id = refseq_id
counts$symbol = symbols
counts = counts[,c("Label", "shRNA_N", "Gene_N", "symbol", "refseq_id",
    "THP1_Contr_norm", "THP1_Contr_2_norm", "THP1_Contr_3_norm", "THP1_Ftular_norm",
    "THP1_Ftular_2_norm", "THP1_Ftular_3_norm")]
colnames(counts) = c("id", "shRNA_id", "gene_id", "symbol", "refseq_id",
    "control_1", "control_2", "control_3", "rf_1", "rf_2", "rf_3")
head(counts)
```

Much nicer. It is not all of the way there yet, though, the numeric columns have
been read in as strings, so we need to change that:
```{r}
for(col in c("control_1", "control_2", "control_3", "rf_1", "rf_2", "rf_3")) {
  counts[,col] = as.numeric(str_replace(counts[,col], ',', ''))
}

```

From Lester's description of the project it seems like this dataset should have only
~ 4000 unique genes but it seems like there are about 7000
genes represented, most of which have between 7-8 shRNA that hit them:

```{r}
shRNA_per_gene = table(counts$refseq_id)
table(shRNA_per_gene)
```

Let's only analyze those:

```{r}
keep = shRNA_per_gene == 7 | shRNA_per_gene == 8
keep = shRNA_per_gene[keep]
counts = counts[counts$refseq_id %in% names(keep),]
shRNA_per_gene = table(counts$refseq_id)
table(shRNA_per_gene)
```

## Exploratory data analysis
We want to get the data in a form that we can manipulate easily during the
exploratory analysis, the *reshape* package, with the near-magic *cast* and
 *melt* functions is awesome for that:

```{r}
library(reshape)
head(counts)
id_cols = c("id", "shRNA_id", "gene_id", "symbol", "refseq_id")
melted = melt(counts, id=id_cols)
colnames(melted) = c(id_cols, "sample", "count")
head(melted)
```
Overall count data:
  ```{r}
 ggplot(melted, aes(count)) + geom_histogram(aes(y = ..density..)) + geom_density() +
    scale_x_log10() +
    theme(text=element_text(family="Gill Sans"))
   # theme_tufte(base_family="GillSans")

 ```
shRNA counts have very similar distributions for each replicate:
```{r}
 ggplot(melted, aes(y=count, x=sample)) + geom_boxplot() + scale_y_log10()
```

We can see that, if we compare for mean shRNA counts per gene across samples, in general
the control samples seem to be tightly correlated to each
other, whereas there is more variability when comparing the controls to anthrax
exposed samples and the anthrax exposed samples to themselves:

```{r}
casted = cast(melted, refseq_id ~ sample, value="count", mean)
plotmatrix(casted[2:8]) + theme(text=element_text(family="Gill Sans"))
```

And in correlation form:
```{r}
row_ids = colnames(casted[2:7])
cor_df = cor(casted[2:7])
colnames(cor_df) = row_ids
rownames(cor_df) = row_ids
qplot(X1, X2, data=melt(cor_df), fill=value, geom="tile") +
    theme(axis.text.x=element_text(angle=-90),
    text=element_text(family="Gill Sans"), axis.title=element_blank()) +
    scale_fill_continuous(name="Pearson correlation")
```

However if we look at a histogram of the raw shRNA counts, there seem to be a set of shRNA
with low (zero) counts in the anthrax samples that are not in the control samples:
```{r}
 ggplot(melted, aes(x=count)) + geom_histogram() + scale_x_log10() + facet_grid(. ~ sample) +
    theme(text=element_text(family="Gill Sans"))
```

# differential shRNA expression
```{r construct-edger-dataframe, results='asis'}
edger_df = counts[, 6:11]
rownames(edger_df) = counts$id
group = factor(unlist(lapply(colnames(edger_df),
                function(x) strsplit(x, "_", fixed=TRUE)[[1]][1])))
dge = DGEList(edger_df, group=group)
```
The MDS of the counts shows that the control samples cluster nicely together,
the rabbit fever infected samples are not clustered to each other but they are
separated from the control samples along the first principal component. They are more
similar to each other than to the control samples which is good for our ability to detect
differential expression.
```{r}
mds = function(counts, condition) {
        nprobes <- nrow(counts)
        nsamples <- ncol(counts)
        counts = cpm(counts)
	distances = dist(t(counts))
	fit = cmdscale(distances, eig=TRUE, k=5)
	colnames(fit$points) = c("one", "two", "three", "four", "five")
	df = as.data.frame(fit$points)
	df$label = rownames(df)
        df$condition = condition
	print(ggplot(df, aes(one, two, color=condition))
            + geom_point() +
              theme(text=element_text(family="Gill Sans")))

	eigs = data.frame(variance_explained=fit$eig / sum(fit$eig))
	eigs$component = as.integer(rownames(eigs))
	#print(ggplot(eigs, aes(component, variance_explained)) + geom_point())
}
mds(dge, group)
```
Libraries tend to have the same number of shRNA counts.
```{r simple-design, results='asis'}
dge = calcNormFactors(dge)
print(gvisTable(dge$samples, options=list(width=900)), "chart")
```

We fit a model of this form to the data:

```{r model}
design = model.matrix(~0+group)
design
```

Common dispersion estimation with biological coefficient of variation.

```{r common-dispersion-estimation}
#dge = estimateCommonDisp(dge, verbose=TRUE)
dge = estimateTrendedDisp(dge)
```

Trended dispersion estimation.

```{r trended-dispersion-estimation}
dge = estimateTagwiseDisp(dge)
```

For low count data there is quite a bit of biological variation but this drops as
the tag count increases. We should treat any differentially expressed tags that have
low CPM as suspect. The overall BCV is low, but it is in the range of what you expect
for cell lines. The trended dispersion estimation is a very good fit; a common
assumption of many tagwise DE algorithms is that the dispersion increases as the
number of counts decreases, or that items with low counts tend to be more variable than
items with high counts.

```{r bcv-plot}
plotBCV(dge)
```

The model fits the data reasonably well.
```{r model-fitting}
fit = glmFit(dge, design)
g = gof(fit)
z = zscoreGamma(g$gof.statistics,shape=g$df/2,scale=2)
qqnorm(z)
```

## Differential expression
```{r}
differential_test = function(dge, pair) {
	et = exactTest(dge, pair=pair)
	et_table = et$table
	et_table$DE = as.numeric(decideTestsDGE(et, p.value=FDR_CUTOFF))
	et_table$FDR = p.adjust(et_table$PValue, method="BH")
	return(et_table)
}
et_table = differential_test(dge, c("control", "rf"))
```
Differential expression shows `r table(et_table$FDR < 0.10)["TRUE"]` shRNA that
are quote a few differentiall expressed between the rabbit fever samples and
control samples.
```{r, results='asis'}
de_counts = merge(counts, et_table, by.x="id", by.y=0)
#multiple_hits = subset(data.frame(table(subset(de_counts, FDR < 0.10)$refseq_id)), Freq > 1)
mhits = data.frame(table(subset(de_counts, FDR < 0.10)$refseq_id))
colnames(mhits) = c("refseq_id", "multiple_hits")
de_counts = merge(de_counts, mhits, by="refseq_id")
#print(gvisTable(multiple_hits), "chart")
#de_counts$multiple_hits = de_counts$refseq_id %in% multiple_hits$refseq_id
```
In addition there are numerous genes with very many shRNA as hits:

```{r}
table(de_counts$multiple_hits)
```
Since there are so many, to priorize hits we will add another column displaying the
number of downregulated and upregulated shRNA in each gene.
```{r}
mhits_up = data.frame(table(subset(de_counts, FDR < 0.10 & DE == 1)$refseq_id))
colnames(mhits_up) = c("refseq_id", "mhits_up")
de_counts = merge(de_counts, mhits_up, by="refseq_id")
mhits_down = data.frame(table(subset(de_counts, FDR < 0.10 & DE == -1)$refseq_id))
colnames(mhits_down) = c("refseq_id", "mhits_down")
de_counts = merge(de_counts, mhits_down, by="refseq_id")
table(de_counts$mhits_up)
table(de_counts$mhits_down)
```
A reasonable approach to reducing how many genes we have to look at is to focus on the
genes with many hits in one direction:
  ```{r}
table(de_counts$mhits_up - de_counts$mhits_down)
```
Annotate the input set of counts with metadata about the genes.
```{r}
ensembl_gene = "hsapiens_gene_ensembl"
filter_type = "refseq_mrna"
annotate_df = function(d) {
	require(biomaRt)
	ensembl = useMart('ensembl', dataset = ensembl_gene)
	a = getBM(attributes=c(filter_type, gene_symbol, "description",
		"entrezgene"),
		filters=c(filter_type), values=d[, 'refseq_id'],
		mart=ensembl)
	m = merge(d, a, by.x='refseq_id', by.y=filter_type)
	return(m)
}
de_annotated = annotate_df(de_counts)
```
Display only the genes that have multiple DE shRNA and write a full table and a
DE table of the results. DE = -1 means it is downregulated in the anthrax infected cells,
0 means it is not DE and 1 means it is upregulated in the anthrax infected cells.

```{r, results='asis'}
print(gvisTable(subset(de_annotated, multiple_hits)), "chart")
write.table(de_annotated, "control_vs_anthrax_all.txt", quote=FALSE, row.names=FALSE,
  sep="\t")
write.table(subset(de_annotated, multiple_hits), "control_vs_anthrax_de.txt",
  quote=FALSE, row.names=FALSE, sep="\t")
```
