---
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    highlight: zenburn
    theme: flatly
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",
               cache=TRUE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='')
```

# Overview
```{r check-library-functions, echo=FALSE}
check_install = function(packages) {
   not_installed = setdiff(packages, rownames(installed.packages()))
   if(length(not_installed) > 0) {
      write(paste("The libraries", not_installed, "are not installed, aborting.",
                   sep=" "), stdout())
      stop()
   }
}
```

```{r load-libraries, echo=FALSE}
packages = c("ggplot2", "reshape", "gplots", "edgeR", "CHBUtils", "pheatmap",
              "DESeq2", "tximport", "DT", "DEGreport", "dplyr")
check_install(packages)
installed = lapply(packages, library, character.only = TRUE)
```

```{r qc-setup}
project_summary = "/home/jnh7/group_dir/PIs/sandra_mcallister/jaclyn-tumor-rnaseq/bcbio/final/2017-03-02_bcbio/project-summary.csv"
counts_file = "/home/jnh7/group_dir/PIs/sandra_mcallister/jaclyn-tumor-rnaseq/bcbio/final/2017-03-02_bcbio/combined.counts"
tx2genes_file = "/home/jnh7/group_dir/PIs/sandra_mcallister/jaclyn-tumor-rnaseq/bcbio/final/2017-03-02_bcbio/tx2gene.csv"
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442",
"#0072B2", "#D55E00", "#CC79A7")
summarydata = read.table(project_summary, header=TRUE, sep=",")
summarydata = summarydata[,colSums(is.na(summarydata)) < nrow(summarydata)]
# handle newer bcbio-nextgen runs that use description as the key
if("description" %in% colnames(summarydata)) {
  rownames(summarydata) = summarydata$description
  summarydata$Name = rownames(summarydata)
  summarydata$description = NULL
} else {
  rownames(summarydata) = summarydata$Name
  # summarydata$Name = NULL
}
summarydata = summarydata[order(rownames(summarydata)),]
if (file.exists(tx2genes_file)) {
  sample_dirs = file.path(dirname(project_summary), "..", rownames(summarydata))
  salmon_files = file.path(sample_dirs, "salmon", "quant.sf")
  sailfish_files = file.path(sample_dirs, "sailfish", "quant.sf")
  new_sailfish = file.path(sample_dirs, "sailfish", "quant", "quant.sf")
  new_salmon = file.path(sample_dirs, "salmon", "quant", "quant.sf")
  if (file.exists(salmon_files[1])) {
    sf_files = salmon_files
  } else if (file.exists(sailfish_files[1])) {
    sf_files = sailfish_files
  }
  else if (file.exists(new_sailfish[1])) {
    sf_files = new_sailfish
  }
  else if (file.exists(new_salmon[1])) {
    sf_files = new_salmon
  }
  names(sf_files) = rownames(summarydata)
  tx2gene = read.table(tx2genes_file, sep=",", row.names=NULL, header=FALSE)
  txi.salmon = tximport(sf_files, type="salmon", tx2gene=tx2gene,
                        reader=readr::read_tsv, countsFromAbundance="lengthScaledTPM")
  counts = round(data.frame(txi.salmon$counts, check.names=FALSE))
} else {
  counts = read.table(counts_file, header=TRUE, row.names="id", check.names=FALSE)
}
counts = counts[, order(colnames(counts)), drop=FALSE]
colnames(counts) = gsub(".counts", "", colnames(counts))

# this is a list of all non user-supplied metadata columns that could appear
known_columns = c("Name", "X.GC", "Exonic.Rate", "Sequences.flagged.as.poor.quality",
    "rRNA_rate", "Fragment.Length.Mean", "Intronic.Rate", "Intergenic.Rate",
    "Mapping.Rate", "Quality.format", "Duplication.Rate.of.Mapped", "Mapped",
    "rRNA", "Sequence.length", "Transcripts.Detected", "Mean.Per.Base.Cov.",
    "Genes.Detected", "Unique.Starts.Per.Read", "unique_starts_per_read",
    "complexity", "X5.3.bias", "Duplicates.pct", "Duplicates", "Mapped.reads",
    "Average.insert.size", "Mapped.reads.pct", "Total.reads",
    "avg_coverage_per_region", "Mapped.Reads")
summarydata[,"Fragment.Length.Mean"] = summarydata$Average.insert.size
metadata = summarydata[, !colnames(summarydata) %in% known_columns, drop=FALSE]
metadata = metadata[, colSums(is.na(metadata)) < nrow(metadata), drop=FALSE]
```

```{r sanitize-datatable}
sanitize_datatable = function(df, ...) {
 # remove dashes which cause wrapping
 DT::datatable(df, ..., rownames=gsub("-", "_", rownames(df)),
                   colnames=gsub("-", "_", colnames(df)))
}
```

```{r set-random-seed}
# set seed for reproducibility
set.seed(1454944673)
```

## Sample metadata
```{r summary-table, echo=FALSE, cache=FALSE}
sanitize_datatable(summarydata, style='bootstrap')
```

```{r heatmap-function}
get_heatmap_fn = function(summarydata) {
    # return the pheatmap function with or without metadata
    if(ncol(metadata) == 0) {
       return(pheatmap)
    }
    else {
    # rownames(metadata) = summarydata$Name
    heatmap_fn = function(data, ...) {
        pheatmap(data, annotation=metadata, clustering_method = "ward.D2",
        clustering_distance_cols = "correlation", ...)
    }
    return(heatmap_fn)
}}
heatmap_fn = get_heatmap_fn(summarydata)
```

# Quality control metrics

```{r qualimap-run}
qualimap_run = "Mapped" %in% colnames(summarydata)
do_quality = "Mapped.reads" %in% colnames(summarydata)
```

## Mapped reads
```{r mapped-plot, eval=qualimap_run}
ggplot(summarydata, aes(x=Name, y=Mapped)) +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    geom_bar(stat="identity") +
    ylab("mapped reads") + xlab("")
```

```{r mapped-plot-noqualimap, eval=!qualimap_run & do_quality}
ggplot(summarydata, aes(x=Name, y=Mapped.reads)) +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    geom_bar(stat="identity") +
    ylab("mapped reads") + xlab("")
```

## Genomic mapping rate
```{r mapping-rate-plot, eval=qualimap_run}
ggplot(summarydata, aes(x=Name, y=Mapping.Rate)) +
    geom_bar(stat="identity") +
    ylab("mapping rate") + xlab("") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90))
```

```{r mapping-rate-plot-noqualimap, eval=!qualimap_run & do_quality}
ggplot(summarydata, aes(x=Name, y=Mapped.reads.pct)) +
    geom_bar(stat="identity") +
    ylab("mapping rate") + xlab("") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90))
```

## Number of genes detected
```{r genes-detected-plot}
dd = data.frame(Name=colnames(counts), Genes.Detected = colSums(counts > 0))
ggplot(dd, aes(x=Name, y=Genes.Detected)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("")
```

## Gene detection saturation
```{r saturation-plot, eval=do_quality}
col_mapped = ifelse(qualimap_run, "Mapped", "Mapped.reads")
dd = data.frame(Mapped=summarydata[,col_mapped], Genes.Detected = colSums(counts > 0))
ggplot(dd, aes(x=Mapped, y=Genes.Detected)) +
    geom_point() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("reads mapped")
```

## Exonic mapping rate
```{r exonic-mapping-plot, eval="Exonic.Rate" %in% colnames(summarydata)}
ggplot(summarydata, aes(x=Name, y=Exonic.Rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("exonic mapping rate") + xlab("")
```

## rRNA mapping rate
```{r check-rRNA-eval}
eval_rRNA = "rRNA_rate" %in% colnames(summarydata) & !sum(is.na(summarydata$rRNA_rate)) == nrow(summarydata)
```
```{r rRNA-rate-plot, eval=eval_rRNA & do_quality}
ggplot(summarydata, aes(x=Name, y=rRNA_rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("rRNA rate") + xlab("")
```

## Estimated fragment length of paired-end reads
```{r fragment-length-plot, eval="Fragment.Length.Mean" %in% colnames(summarydata) & do_quality}
ggplot(summarydata, aes(x=Name, y=Fragment.Length.Mean)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("fragment length") + xlab("")
```

## 5'->3' bias
```{r 53bias, eval="X5.3.bias" %in% colnames(summarydata) & do_quality}
ggplot(summarydata, aes(x=Name, y=X5.3.bias)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("5'->3' bias") + xlab("")
```

## Boxplot of log10 counts per gene
```{r boxplot-raw}
melted = melt(counts)
colnames(melted) = c("sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Boxplot of log10 TMM-normalized counts per gene
Trimmed mean of M-values (TMM) normalization is described
[here](http://genomebiology.biomedcentral.com/articles/10.1186/gb-2010-11-3-r25)

Robinson, M. D., & Oshlack, A. (2010). A scaling normalization method for differential expression analysis of RNA-seq data. Genome Biology, 11(3). doi:10.1186/gb-2010-11-3-r25

```{r boxplot-normalized}
y = DGEList(counts=counts)
y = calcNormFactors(y)
normalized_counts = cpm(y, normalized.lib.sizes=TRUE)
melted = melt(normalized_counts)
colnames(melted) = c("gene", "sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Density of log10 TMM-normalized counts
```{r density-normalized}
ggplot(melted, aes(x=count, group=sample)) +
    geom_density() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Correlation heatmap of TMM-normalized counts {.tabset}

### Correlation (Pearson)
```{r pearson-heatmap-normalized, eval=ncol(counts) > 1}
heatmap_fn(cor(normalized_counts, method="pearson"))
```

### Correlation (Spearman)
```{r spearman-heatmap-normalized, eval=ncol(counts) > 1}
heatmap_fn(cor(normalized_counts, method="spearman"))
```

## PCA plots {.tabset}
```{r pca, eval=ncol(counts) > 1}
dds = DESeqDataSetFromMatrix(countData=counts, colData=summarydata, design=~Name)
vst = varianceStabilizingTransformation(dds)
pca_loadings = function(object, ntop=500) {
  rv <- matrixStats::rowVars(assay(object))
  select <- order(rv, decreasing = TRUE)[seq_len(min(ntop,
      length(rv)))]
  pca <- prcomp(t(assay(object)[select, ]))
  percentVar <- pca$sdev^2/sum(pca$sdev^2)
  names(percentVar) = colnames(pca$x)
  pca$percentVar = percentVar
  return(pca)}
pc = pca_loadings(vst)
comps = data.frame(pc$x)
comps$Name = rownames(comps)
library(dplyr)
comps = comps %>% left_join(summarydata, by=c("Name"="Name"))
colorby = "Name"
```

```{r pca-plotter}
pca_plot = function(comps, nc1, nc2, colorby) {
   c1str = paste0("PC", nc1)
   c2str = paste0("PC", nc2)
  ggplot(comps, aes_string(c1str, c2str, color=colorby)) +
    geom_point() + theme_bw() +
    xlab(paste0(c1str, ": ", round(pc$percentVar[nc1] * 100), "% variance")) +
    ylab(paste0(c2str, ": ", round(pc$percentVar[nc2] * 100), "% variance"))
  }
```

### PC1 vs. PC2
```{r PC1-PC2}
pca_plot(comps, 1, 2, colorby)
```

### PC3 vs. PC4
```{r PC3-PC4}
pca_plot(comps, 3, 4, colorby)
```

### PC5 vs. PC6
```{r PC5-PC6}
pca_plot(comps, 5, 6, colorby)
```

### Variance explained by component
```{r component-loadings}
ggplot(data.frame(component=reorder(names(pc$percentVar), -pc$percentVar),
                  percent_var=pc$percentVar),
       aes(component, percent_var)) + geom_bar(stat='identity') +
  ylab("percent of total variation") + xlab("") + theme_bw()
```

```{r shiny-app, echo=F}
# this is to plot data with https://github.com/lpantano/shiny/tree/master/expression
degObj(assay(vst), metadata, "deg.rda")
```