---
output:
  knitrBootstrap::bootstrap_document:
    theme: readable
    highlight: zenburn
    theme.chooser: TRUE
    highlight.chooser: TRUE
  html_document:
    toc: true
    highlight: zenburn
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='', bootstrap.show.code=FALSE)
```
# Basic QC Overview (only 13 samples)

The same graphs as before, but this time with only 13 samples, after removing the 3 problematic samples (**SH-ins-1**, **HepG2-ins-4** and **SH-ins-4**). So, we are down to only 2 replicates for SHS insulin treated samples, and 3 for HepG2 insulin-treated samples.

All 13 samples look pretty great from the standpoint of mapping rate, number of genes, normalization, etc., but there are still some issues with how they are clustering together. **SH-con-1** clusters with the HepG2 samples, and the HepG2 samples are still not clustering very well together, irrespective of treatment groups.

```{r qc-setup}
library(ggplot2)
library(reshape)
library(gplots)
library(edgeR)
library(CHBUtils)
library(pheatmap)
project_summary = "/Users/rkhetani/Dropbox/HBC consults/Flanagan_RNA-Seq/flanagan-bcbio/2015-07-13_flanagan_rnaseq_13samples/project-summary.csv"
counts_file = "/Users/rkhetani/Dropbox/HBC consults/Flanagan_RNA-Seq/flanagan-bcbio/2015-07-13_flanagan_rnaseq_13samples/combined.counts"
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442",
"#0072B2", "#D55E00", "#CC79A7")
summarydata = data.frame(read.table(project_summary, header=TRUE, sep=","), row.names="Name", check.rows=FALSE)
summarydata$Name = rownames(summarydata)
summarydata = summarydata[order(summarydata$Name),]
counts = read.table(counts_file, header=TRUE, row.names="id", check.names=FALSE)
counts = counts[, order(colnames(counts))]
# this is a list of all non user-supplied metadata columns that could appear
known_columns = c("Name", "X.GC", "Exonic.Rate", "Sequences.flagged.as.poor.quality",
    "rRNA.rate", "Fragment.Length.Mean", "Intronic.Rate", "Intergenic.Rate",
    "Mapping.Rate", "Quality.format", "Duplication.Rate.of.Mapped", "Mapped",
    "rRNA", "Sequence.length", "Transcripts.Detected", "Mean.Per.Base.Cov.",
    "Genes.Detected", "Unique.Starts.Per.Read", "unique_starts_per_read",
    "complexity", "X5.3.bias")
```

```{r heatmap-function}
get_heatmap_fn = function(summarydata) {
    # return the pheatmap function with or without metadata
    metadata = summarydata[, !colnames(summarydata) %in% known_columns, drop=FALSE]
    if(ncol(metadata) == 0) {
       return(pheatmap)
    }
    else {
    # rownames(metadata) = summarydata$Name
    heatmap_fn = function(data, ...) {
        pheatmap(data, annotation=metadata, ...)
    }
    return(heatmap_fn)
}}
heatmap_fn = get_heatmap_fn(summarydata)
```

# Quality control metrics

## Mapped reads
```{r mapped-plot}
ggplot(summarydata, aes(x=Name, y=Mapped)) +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    geom_bar(stat="identity") +
    ylab("mapped reads") + xlab("")
```

## Genomic mapping rate
```{r mapping-rate-plot}
ggplot(summarydata, aes(x=Name, y=Mapping.Rate)) +
    geom_bar(stat="identity") +
    ylab("mapping rate") + xlab("") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90))
```

## Unique mapping rate
```{r unique-rate-plot}
dd = data.frame(Name=names(counts), Unique=colSums(counts), Mapped=summarydata[,"Mapped"])
ggplot(dd, aes(x=Name, y=Unique/Mapped)) +
    geom_bar(stat="identity") +
    ylab("unique mapping rate") + xlab("") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90))
```

## Number of genes detected
```{r genes-detected-plot}
dd = data.frame(Name=names(counts), Genes.Detected = colSums(counts > 0))
ggplot(dd, aes(x=Name, y=Genes.Detected)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("")
```

## Gene detection saturation
```{r saturation-plot}
dd = data.frame(Mapped=summarydata$Mapped, Genes.Detected = colSums(counts > 0))
ggplot(dd, aes(x=Mapped, y=Genes.Detected)) +
    geom_point() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("reads mapped")
```

## Exonic mapping rate
```{r exonic-mapping-plot}
ggplot(summarydata, aes(x=Name, y=Exonic.Rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("exonic mapping rate") + xlab("")
```

## rRNA mapping rate
```{r rRNA-rate-plot}
ggplot(summarydata, aes(x=Name, y=rRNA.rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("rRNA rate") + xlab("")
```

## Boxplot of log10 counts per gene
```{r boxplot-raw}
melted = melt(counts)
colnames(melted) = c("sample", "count")
melted$sample = factor(melted$sample)
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Boxplot of log10 TMM-normalized counts per gene
Trimmed mean of M-values (TMM) normalization is described
[here](http://genomebiology.com/2010/11/3/R25)

Robinson, M. D., & Oshlack, A. (2010). A scaling normalization method for differential expression analysis of RNA-seq data. Genome Biology, 11(3). doi:10.1186/gb-2010-11-3-r25

```{r boxplot-normalized}
y = DGEList(counts=counts)
y = calcNormFactors(y)
normalized_counts = cpm(y, normalized.lib.sizes=TRUE)
melted = melt(normalized_counts)
colnames(melted) = c("gene", "sample", "count")
melted$sample = factor(melted$sample)
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Density of log10 TMM-normalized counts
```{r density-normalized}
ggplot(melted, aes(x=count, group=sample)) +
    geom_density() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Correlation (Pearson) heatmap of TMM-normalized counts
```{r pearson-heatmap-normalized}
heatmap_fn(cor(normalized_counts, method="pearson"))
```

## Correlation (Spearman) heatmap of TMM-normalized counts
```{r spearman-heatmap-normalized}
heatmap_fn(cor(normalized_counts, method="spearman"))
```

## MDS plot of TMM-normalized counts
```{r mds-normalized}
mds(normalized_counts, k=length(colnames(normalized_counts)) - 1)
```
