---
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: united
title: "Microarray of RIP-chip samples - exclude WT1"
bibliography: "references.bib"
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(knitcitations)
cleanbib()
options("citation_format" = "pandoc")

clientname="Valerie Schumacher "
clientemail="Valerie.Schumacher@childrens.harvard.edu"
labPI="Schumacher"
lablocation="HMS"
analystname="John Hutchinson"
analystemail="jhutchin@hsph.harvard.edu"


library(knitr)
opts_chunk$set(warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, cache=TRUE, tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120), dev="svg")
options(width=200)
```

---

Array analysis for `r clientname` (`r clientemail`), `r labPI` group at `r lablocation`.  

Contact `r analystname` (`r analystemail`) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots. 

---

# Methods Summary  


---

# Setup

## Variables
Working directories, files and other variables necessary to the analysis.

```{r variables}
## Setup Data and Results directory variables
if(file.exists("/n/hsphS10/hsphfs1/chb/projects/vs_ripchip/")){
  baseDir="/n/hsphS10/hsphfs1/chb/projects/vs_ripchip/"
    } else if (file.exists("/Users/johnhutchinson/Work/projects/vs_ripchip/")){
    baseDir="/Users/johnhutchinson/Work/projects/vs_ripchip/"
    }

dataDir <- file.path(baseDir, "data")
metaDir <- file.path(baseDir, "meta")
resultsDir <- file.path(baseDir, "results")

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7") # colorblind friendly palette
covarsfilename="covars.desc" # tab delimited file describing samples
lowintensity.percentile=0.1
mad.quantile.cutoff=0.1
pvalue.cutoff=0.05
highlight.color="green"
lfc.cutoff=log2(1.5)
# number of genes to show in heatmap
numtop=50
```

## Libraries

[Bioconductor](http://www.bioconductor.org) and [R](http://cran.r-project.org/) libraries used to process and visualize the data.

```{r libraries_variables}
library(oligo) # array utilities
library(pd.mogene.1.0.st.v1)# array layout annotation
library(mogene10sttranscriptcluster.db)
library(arrayQualityMetrics) # array quality control reports
library(limma) # array statistical analyses
library(pheatmap) # pretty heatmaps
library(plyr) # data format utility
library(reshape2) # data format utility
library(devtools) # install libraries from github
install_git("git://github.com/hbc/CHBUtils.git") # misc personal utilities
library(CHBUtils) # some homegrown functions
library(ggplot2) # pretty graphs
library(ggdendro) # for pretty dendrograms
library(RColorBrewer) # more colors
library(gridExtra) # for arranging multiple plots
library(venneuler) # for venm diagrams
library(GO.db) # database of GO terms and genes
library(biomaRt) # large biological database
library(sva)
library(magrittr)
library(lattice)
library(RRHO)
library(biomaRt)
library(dplyr)
```

## Functions

```{r functions}
# for plotting amount of variation explained by principal components
PCAplot.sd.eset <- function(eset=NULL,  title=NULL){
  eset.core <- exprs(eset)
  myPca.core <- prcomp(t(eset.core))
  # SD of components
  sdevdf <- data.frame(cbind(as.numeric(myPca.core$sdev),c(1:length(myPca.core$sdev))))
  sdevdf$prop <-  sdevdf$X1/sum(sdevdf$X1)
  sdevdf$cum <- cumsum(sdevdf$prop)
  ggplot(sdevdf, aes(x=X2, y=prop)) + 
    geom_point(size=4, color="red") + 
    scale_x_continuous('Component') + 
    scale_y_continuous('Standard Deviation') +
    ggtitle(title) +
    geom_line(data=sdevdf, aes(x=X2, y=cum))
}

# used for formatting labels on ggplots
fmt <- function(){ 
  function(x) format(x,nsmall = 1,scientific = FALSE)
}

vcplot <- function(stats, plottitle="Volcano Plot with Marginal Distributions", pval.cutoff=0.05, lfc.cutoff=1, shade.colour="green", shade.alpha=0.25, point.colour="gray", point.alpha=0.75, point.outline.colour="darkgray", line.colour="gray", axistitlefontsize=24, axislabelfontsize=12) {
  # get range of log fold change and p-value values to setup plot borders
  range.lfc <- c(floor(min(stats$logFC)), ceiling(max(stats$logFC)))
  range.pval <- c(floor(min(-log10(stats$adj.P.Val))), ceiling(max(-log10(stats$adj.P.Val))))
  
  #make top plot - density plot with fold changes
  lfcd <- as.data.frame(cbind(density(stats$logFC)$x, density(stats$logFC)$y))
  hist_top <- ggplot(data=stats, aes(x=logFC))+
    geom_density(color=line.colour)+
    geom_ribbon(data=subset(lfcd, V1>lfc.cutoff),aes(x=V1,ymax=V2),ymin=0,fill=shade.colour, alpha=shade.alpha)+
    theme_bw()+ 
    theme(axis.title.x=element_blank())+
    theme(plot.margin=unit(c(3,-5.5,4,3), "mm") )+
    scale_x_continuous(limits = range.lfc, breaks = range.lfc[1]:range.lfc[2], expand = c(.05,.05))+
    scale_y_continuous(labels=fmt()) +
   theme(axis.text=element_text(size=axislabelfontsize), axis.title=element_text(size=axistitlefontsize))
  
  # make blank plot
  empty <- ggplot()+geom_point(aes(1,1), colour="white")+
    theme(panel.grid=element_blank(),
          panel.border=element_blank(),
          axis.ticks=element_blank(), 
          panel.background=element_blank(), 
          axis.text.x=element_blank(),  
          axis.text.y=element_blank(),          
          axis.title.x=element_blank(), 
          axis.title.y=element_blank()
    )
  
  #make scatter volcano plot
  scat.poly.up <- with(stats, data.frame(x=as.numeric(c(lfc.cutoff,  lfc.cutoff, max(range.lfc),max(range.lfc))), y=as.numeric(c(-log10(pval.cutoff), max(range.pval), max(range.pval),-log10(pval.cutoff)))))
  
  scatter <- ggplot(data=stats, aes(x=logFC, y=-log10(adj.P.Val))) +
    geom_point(alpha=point.alpha, pch=21, fill=point.colour, color=point.outline.colour) +
    geom_polygon(data=scat.poly.up, aes(x=x,y=y), fill=shade.colour, alpha=shade.alpha) +
    xlab("log2 fold change") + ylab("-log10(adjusted p-value)") +
    theme_bw()+
    theme(legend.position="none") +
    theme(plot.margin=unit(c(3,-5.5,4,3), "mm") )+
    scale_x_continuous(limits = range.lfc, breaks = range.lfc[1]:range.lfc[2], expand = c(.05,.05))+
    scale_y_continuous(labels=fmt(), limits = range.pval)+
    theme(axis.text=element_text(size=axislabelfontsize), axis.title=element_text(size=axistitlefontsize))    
  
  # make right plot - density plot of adjusted pvalues
  pvald <- as.data.frame(cbind(density(-log10(stats$adj.P.Val))$x, density(-log10(stats$adj.P.Val))$y))
  hist_right <- ggplot(data=stats, aes(x=-log10(adj.P.Val)))+
    geom_density(color=line.colour)+
    geom_ribbon(data=subset(pvald, V1>-log10(pval.cutoff)),aes(x=V1,ymax=V2),ymin=0,fill=shade.colour, alpha=shade.alpha)+
    theme_bw()+coord_flip()+
    scale_x_continuous(limits = range.pval)+
    theme(axis.title.y=element_blank())+ 
    theme(plot.margin=unit(c(3,-5.5,4,3), "mm"))+
    theme(axis.text=element_text(size=axislabelfontsize), axis.title=element_text(size=axistitlefontsize))    
  
  # plot all plots
  pp.logfc <- ggplotGrob(hist_top)
  pp.empty <- ggplotGrob(empty)
  pp.volc <- ggplotGrob(scatter)
  pp.pval  <- ggplotGrob(hist_right)
  
 grid.arrange(top=textGrob(plottitle, gp = gpar(fontsize = axistitlefontsize*1.2, fontface = 'bold'),),
               arrangeGrob(pp.logfc,
                           pp.volc, 
                           heights=c(1,3),
                           ncol=1),
               arrangeGrob(pp.empty,
                           pp.pval,  
                           heights=c(1,3),
                           ncol=1), 
               ncol=2, 
               widths=c(3,1))
} 

```

---

# Import Data and Metadata

## Data

- load in phenotypes and array names from metadata file (covars.desc) in "metadata" directory
  - this file contains the names and descriptions of CEL files contained in the data directory 
- use array names to load in arrays 

```{r dataload, results='hide'}
covars <- read.table(file.path(metaDir, covarsfilename),header=TRUE, sep="\t", row.names=1) # simple tab delimited file with CEL file in first column (no heading for this column) and sample metadata (i.e. sampleID, treatment group, batch etc.) in subsequent columns

celFiles <- file.path(dataDir, row.names(covars))
affyRaw <- read.celfiles(celFiles)
pData(affyRaw) <- covars 
sampleNames(affyRaw) <- pData(affyRaw)$sampleID
validObject(affyRaw)
rm(covars)
```

## Sample metadata

```{r covars, results='asis'}
# Sample information table
kable(pData(affyRaw))
```

---

# PreProcessing 

## Raw Data 

### Quality Control


```{r rawQC, eval=FALSE}
arrayQualityMetrics(expressionset=affyRaw, outdir=file.path(resultsDir, 'report_raw'), force=TRUE, do.logtransform=TRUE, intgroup=c("pulldown", "source", "run"))
```

[Raw Data QC Report](../results/report_raw/index.html)

The arrays look fine, no problems at all.

## RMA Normalized Data

- background correct and normalize data with RMA `r citep("10.1093/bioinformatics/19.2.185")`

- summarize probesets on the gene ('core') level

```{r normalize, results='hide'}
affyNorm.core <- rma(affyRaw, target="core", background=TRUE, normalize=TRUE)
```

### Quality Control
- using arrayQualityMetrics library

```{r normQC, eval=FALSE}
arrayQualityMetrics(expressionset=affyNorm.core, outdir=file.path(resultsDir, paste("report_rma.core", sep=".")), force=TRUE, do.logtransform=FALSE, intgroup=c("pulldown", "source", "run"))
```

[Normalized Data QC Report](../results/report_rma.core/index.html)

The arrays look good, and the clustering looks very good. 

### Unsupervised Clustering of RMA Normalized Data

#### Hierarchical Clustering
The goal of these analyses are to naiively evaluate the variability within the raw data and determine whether this variability can predict the different sample groups

The first method produces a dendrogram by performing  
>  a hierarchical cluster analysis using a set of dissimilarities for the n objects being clustered

Here, I've fist highlighted which input sample was mRNA, and then highlighted the different pulldown categories. 

```{r cluster1, out.width='100%'}
plot_dendro(affyNorm.core, title="", labels.colname="pulldown", colors.colname="source")
plot_dendro(affyNorm.core, title="", labels.colname="pulldown", colors.colname="pulldown")
```

#### Principal Component Analysis (PCA)

This second approach is a dimension reduction and visualisation technique that is used to project the multivariate (i.e.multiple genes) data vector of each array into a lower-dimensional plot, such that the spatial arrangement of the points in the plot reflects the overall data (dis)similarity between the arrays. The data is typically reduced to a small number of dimensions (or components) which explain most of the sample variability. This [Youtube slideshow](https://www.youtube.com/watch?v=BfTMmoDFXyE) gives a pretty good basic explanation of what PCA is doing.

```{r PCAsd1, out.width='75%'}
PCAplot.sd.eset(affyNorm.core, title="")
```

Here, each point depicts the amount of variation explained by each component and the line shows the cumulative amount. For this data set,  very few dimensions (3) can explain >75% of the variation observed in the samples.

As plots with more than 2 dimensions are difficult to visualize, we typically  split up the dimensions/components and plot them pairwise against each other; the plots here show scatterplots of the arrays along all dual combinations of the first three principal components. In the first plot, each sample group is represented by a separate color and in the second plot each sample is represented by a different color. 

You can use these plots to explore if the arrays cluster, find outliers, and determine whether this is according to an intended experimental factor or according to unintended causes such as batch effects. In these plots, I have once again first highlighted the RNA source (mRNA or whole RNA) and then highlighted the individual pulldown categories.

```{r pca1, out.width='100%'}
PCAplot.eset(affyNorm.core, categories="source", title="", colorpalette=cbPalette, numcomponents=3, alpha=0.75)
PCAplot.eset(affyNorm.core, categories="pulldown", title="", colorpalette=cbPalette, numcomponents=3, alpha=0.75)
```

There is  a high degree of clustering by pulldown type. So much so, that when you plot PC1 against PC2, both the input samples and stau2 samples largely overlap their respective sample groups. These results are consistent with a lower level of biological variation, likely resulting from the use of a single cell line for the experiment, making these replicates closer to technical replicates. The single input sample outlier is the mRNA sample.

These unsupervised clustering resutls reveal that the mRNA input sample has clear differences from the whole RNA input samples, more than can be attributed to batch effect, given how closely the *Stau2* pulldown whole RNA repeat sample from the same run clusters with the *Stau2* whole RNA samples from the first run.

---

## Batch Correction

The whole RNA and mRNA input RNA samples were run on different days and may show batch effects. As an identical sample was also run both days, we can try adjusting for batch using ComBat r citep("doi: 10.1093/biostatistics/kxj037") from the sva package and see if it improves the data clustering. 

```{r batchm, dev="png"}
library(sva)

# Create model with batch as factor variable
mod <- model.matrix(~as.factor(pulldown), data=pData(affyNorm.core))

modcombat <- model.matrix(~1, data=pData(affyNorm.core))

batch <- pData(affyNorm.core)$run
# Modify expression matrix
eset.core <- exprs(affyNorm.core)
combat_edata <- ComBat(dat=eset.core,
                       batch=batch,
                       mod=modcombat,
                       par.prior=TRUE, 
                       prior.plots=TRUE)
affyNorm.core.batch <- affyNorm.core
exprs(affyNorm.core.batch) <- combat_edata

PCAplot.eset(affyNorm.core.batch, categories="source", title="", colorpalette=cbPalette, numcomponents=3, alpha=0.75)
PCAplot.eset(affyNorm.core.batch, categories="pulldown", title="", colorpalette=cbPalette, numcomponents=3, alpha=0.75)
```

Based on the low variation within the Stau2 group before batch correction (look back to the first PCA plot) there does not appear to be much need for batch correction. In fact, the PCA plot after batch correction shows an undesirable loss of signal (and averaging betweeen samples), potentially due to the lack of inclusion of a repeated input sample. Given this, I decided to skip batch correction for this analysis.

Continuing forward, I removed the repeated Stau2 whole RNA sample.

```{r removerepeat}
pd <- pData(affyNorm.core)
affyNorm.core <- affyNorm.core[,!(pd$sampleID=="stau2-whole-3" & pd$run==2)]

# drop unused levels from pData as well
pData(affyNorm.core) <- as.data.frame(apply(pData(affyNorm.core), 2, function(x) factor(x)))
```

---

## Annotate

So far we have only been working with the probesets,without reference to the genes they assay. Here we load in metadata about the probesets on the array (feature data), the gene symbols in particular.

```{r features, results='hide'}
featureData(affyNorm.core) <- getNetAffx(affyNorm.core, "transcript") # this will load the Affymetrix annotation, including the probeID, into the fData
# get gene symbols and entrezIDs for all probesets
fData(affyNorm.core)$symbol <- as.character(unlist(mget(featureNames(affyNorm.core), mogene10sttranscriptclusterSYMBOL, ifnotfound=NA))) # curated annotations from Bioconductor 
fData(affyNorm.core)$entrezID <- as.character(unlist(mget(featureNames(affyNorm.core), mogene10sttranscriptclusterENTREZID, ifnotfound=NA))) # curated annotations from Bioconductor 
numprobes <- nrow(fData(affyNorm.core))
```

## Filter Probesets
Reducing the number of genes assayed reduces  the multiple test correction and may allow us to identify more differentially expressed genes.

Starting  with `r numprobes` probes remaining we can filter:

### By Annotation
- remove the control probes and probes without annotated genes

```{r filter1}
affyNorm.core <- affyNorm.core[which(!is.na(fData(affyNorm.core)$symbol) & fData(affyNorm.core)$category=="main"),]
numprobes <- nrow(fData(affyNorm.core))
```

`r numprobes` probes remaining

### By Cross Hybridization
- some probes are annotated as potentially hybridizing to multiple targets

```{r filter2}
affyNorm.core <- affyNorm.core[which(fData(affyNorm.core)$crosshybtype=="1"),]
numprobes <- nrow(fData(affyNorm.core))
```

`r numprobes` probes remaining

### By Low Expression Level
- remove probes with low expression levels (bottom `r lowintensity.percentile*100`% of all expression levels) in all samples

```{r filter3}
eset.core <- exprs(affyNorm.core)
affyNorm.core <- affyNorm.core[!(apply(eset.core, 1, function(x) all(x<quantile(exprs(affyNorm.core), 0.1)))),]
numprobes <- nrow(fData(affyNorm.core))
```

`r numprobes` probes remaining

### By Low Variability
- remove probes with lower variation among all samples (without regard for group status) (dropped the bottom `r mad.quantile.cutoff*100`%) 

```{r filter4}
eset.core <- exprs(affyNorm.core)
rowmads <- apply(eset.core, 1, mad)
mad.cutoff <- as.numeric(quantile(rowmads, mad.quantile.cutoff))
affyNorm.core <- affyNorm.core[rowmads>mad.cutoff,]
numprobes <- nrow(fData(affyNorm.core))
```

`r numprobes` probes remaining

---

# Statistical Analyses

## Limma

A linear model for microarray data analysis (Limma) was performed on the samples to identify differentially expressed genes for comparisons of the sample groups. Limma fits a linear model to the expression data for all samples for each gene and is designed to handle complex experiments involving comparisons between many RNA targets simultaneously.

To perform limma, we construct two matrices. The design matrix provides a representation of the different sample groups which have been analysed. The contrast matrix allows the coefficients defined by the design matrix to be combined into contrasts of interest.

### Design 
- make a matrix with arrays as rows, sample groups as columns
- a one or a zero indicate respectively, that a sample either belongs or does not belong to the sample group 

```{r design2, results="asis"}
design <- model.matrix(~ 0 + group,  data=pData(affyNorm.core))
# make sure the headings match
library(magrittr) # fun with pipes
colnames(design) %>% sub("group", "", .) %>% sub("-", "_", .) -> colnames(design)
kable(design)
```

### Contrasts
- to perform specified pairwise comparisons
- in this table, columns are contrasts/comparisons and rows are sample groups
-  a zero denotes that the sample group is not involved in the contrast, a 1 denotes that it has higher expression in the contrast and a -1 denotes lower expression in the contrast

I  setup four  contrasts, two to select genes that show a significant expression change between the input (both mRNA and whole RNA) and pulldown Stau2 samples and two to select genes that show a significant expression change between the input (both mRNA and whole RNA) and pulldown WT1 samples.

```{r contrastmatrix2, results='asis'}
contrast.matrix <- makeContrasts(stau2_whole=stau2_whole-input_whole, 
                                 stau2_mRNA=stau2_whole-input_mrna,
                                 wt1_whole=wt1_whole-input_whole,
                                 wt1_mRNA=wt1_whole-input_mrna,
                                 levels=colnames(design))
dimnames(contrast.matrix)$Contrasts <- gsub(" " , "", dimnames(contrast.matrix)$Contrasts)
kable(contrast.matrix)
```

These matrices are used to fit a linear model to the data. The linear model is applied and pairwise comparisons are performed to identify differentially expressed genes.

- first fit the linear model based on the design matrix for each gene based on the given series of arrays
- using the contrast matrix, compute estimated coefficients and standard errors for contrasts
- compute moderated t-statistics and log-odds of differential expression by empirical Bayes shrinkage of the standard errors towards a common value

### Linear model

- for each gene based on the given series of arrays

```{r linearmodel}
eset.core <- exprs(affyNorm.core)
fit.core <- lmFit(eset.core, design) 
```

**Compute estimated coefficients and standard errors for contrasts**

```{r contrastfit}
fit2.core <- contrasts.fit(fit.core, contrast.matrix) 
```

### Bayes shrinkage

**Compute moderated t-statistics and log-odds of differential expression**

- by empirical Bayes shrinkage of the standard errors towards a common value

```{r bayes}
fit2.core <- eBayes(fit2.core) 
```

--- 

# Results

## Statistics

- as calculated by Limma

```{r allstats}
all.results <- lapply(seq(1:length(dimnames(contrast.matrix)$Contrasts)), function(num) {
  contrast <- dimnames(contrast.matrix)$Contrasts[num]
  stats <- topTable(fit2.core, coef=num, sort.by="B",adjust.method="BH",number=nrow(fData(affyNorm.core)), genelist=fData(affyNorm.core)[,c("probesetid", "symbol", "entrezID")])
  stats$Passes.FDR.threshold  <-  as.factor(stats$adj.P.Val<pvalue.cutoff)
  eset <- exprs(affyNorm.core)
  eset  <-  eset[match(stats$probesetid, row.names(eset)),]
  stats.eset <- cbind(stats, eset)
  return(list(contrast=contrast, stats.eset=stats.eset))
  })

# output all results to files
lapply(seq(1:length(dimnames(contrast.matrix)$Contrasts)), function(num) {
  contrast <- dimnames(contrast.matrix)$Contrasts[num]
  out.stats=as.data.frame(all.results[[num]]$stats.eset)
  write.table(out.stats, file=file.path(resultsDir, paste("all.genes.stats.exprs", contrast, "xls", sep=".")),  sep ="\t",, row.names=F, col.names=T)
})
```

### Statistics and expression levels of all genes for these comparisons

*Note that for all these files, I have not summarized values for genes assayed by multiple probes (i.e. by taking the median value), so you may see multiple instances of the same gene in the results*

`r x=1`
[Stau2 whole RNA pulldown stats - all genes](../results/all.genes.stats.exprs.`r all.results[[x]]$contrast`.xls)
`r x=x+2`
[WT1 whole RNA pulldown stats - all genes](../results/all.genes.stats.exprs.`r all.results[[x]]$contrast`.xls)

**These summary tables contain the following information:**

- logFC is the log2-fold change
- the AveExpr is the average expression value accross all arrays
- the moderated t-statistic (t) is the logFC to its standard error, the P.Value is the associated p-value
- the adj.P.Value is the p-value adjusted for multiple testing (by FDR) 
- the B-value (B) is the log-odds that a gene is differentially expressed (the-higher-the-better)
- the last 8 columns contain the log-transformed normalized expression levels for these genes in each sample

## Identifying Genes Enriched by RBP pulldown

### Volcano plots

Here we can visulize the relationship between the fold changes in expression observed for the different pulldowns. Our best candidate genes will not only have a statistically significant difference in gene expression between the two sample groups (as measured adjusted pvlaue) but also a large change (as measured by the log2 fold change). We are also only interested in genes that are enriched after pulldown, not those that are higher in the input samples.

**Each of these plots contains 3 subplots:**

1) Bottom left - the volcano plot, a scatter plot with the observed log2fold changes (extremes are better) plotted against the -log10 adjusted pvalues (higher is better). For these contrasts, we are looking for genes that are enriched in the pulldown, genes that have at least an adjusted pvalue of `r pvalue.cutoff` and a log 2 fold change more than `r lfc.cutoff` are highlighted with a green box.   

2) Upper left - a density plot (smoothed histogram) of the log2 fold changes observed for the contrast, the part of the distribution above `r lfc.cutoff` is highlighted under the curve in `r highlight.color`.  

3) Lower right - a density plot (smoothed histogram) of the adjusted pvalued observed for the contrast, the part of the distribution above `r pvalue.cutoff` is highlighted under the curve in `r highlight.color`. Note that for this plot, this highlight also included genes enriched in the input samples.

```{r ggplotexps, out.width='100%', dev="svg"}
vcplot(stats=all.results[[1]]$stats.eset, plottitle="Stau2 whole RNA pulldown vs. input", lfc.cutoff = lfc.cutoff, pval.cutoff = pvalue.cutoff, shade.colour=highlight.color, axistitlefontsize = 28, axislabelfontsize = 18)          
```

Using these pvalue and log2 fold change cutoffs we can identify which genes are showing enrichment in the two pulldowns. The cutoffs I have picked here (pvalue<`r pvalue.cutoff` and log2foldchange>`r lfc.cutoff`) are within accepted range, if a bit stringent, but are arbitrary. 

**If you want to change these cutoffs to be more or less stringent, you can filter the Excel files above by adj.P.Val and logFC in Excel.**

For these cutoffs: 
`r x=1`
- the `r all.results[[x]]$contrast` contrast, has `r nrow(subset(all.results[[x]]$stats.eset, logFC>lfc.cutoff & adj.P.Val<pvalue.cutoff))` enriched probesets probing `r length(unique(subset(all.results[[x]]$stats.eset, logFC>lfc.cutoff & adj.P.Val<pvalue.cutoff)$symbol))` genes.

### Statistics and expression levels for just these differentially expressed genes

```{r ouputtop, results='hide'}
# output top results to files
lapply(seq(1:length(dimnames(contrast.matrix)$Contrasts)), function(num) {
  contrast <- dimnames(contrast.matrix)$Contrasts[num]
  out.stats=as.data.frame(subset(all.results[[num]]$stats.eset, logFC>lfc.cutoff & adj.P.Val<pvalue.cutoff))
  write.table(out.stats, file=file.path(resultsDir, paste("top.enriched.genes.stats.exprs", contrast, "xls", sep=".")),  sep ="\t",, row.names=F, col.names=T)
})
```

*Note that, once again, for all these files I have not summarized values for genes assayed by multiple probes (i.e. by taking the median value), so you may see multiple instances of the same gene in the results*

`r x=1`
[Stau2 whole RNA pulldown stats - top enriched genes](../results/top.enriched.genes.stats.exprs.`r all.results[[x]]$contrast`.xls)

*These summary tables contain the same information as the files above*

### Heatmaps of genes bound by Stau2

#### Top 50 for Stau2

```{r heatmaps2, out.width='100%', dev='png', fig.height=11, fig.width=8.5}
top.results.Stau2 <- subset(all.results[[1]]$stats.eset, logFC>lfc.cutoff & adj.P.Val<pvalue.cutoff)

# microarray intensities fwith rows labelle diwht probeset and gene
top.results.Stau2 <- top.results.Stau2 %>% tbl_df() %>% select(., -contains("wt1"))

# grab top 50 genes by logFC
eset.of.interest.Stau2 <- top.results.Stau2 %>%  
  arrange(., desc(logFC) ) %>% 
  select(.,  contains("symbol"), contains("whole"), logFC) %>%
  .[1:numtop,]
names(eset.of.interest.Stau2) <- gsub("_whole", "", names(eset.of.interest.Stau2))
# cluster with hclust and reorder matrix
eset.of.interest.Stau2 <- select(eset.of.interest.Stau2, -logFC, -symbol) %>% 
  as.matrix(.) %>% 
  dist(., method="euclidean") %>% 
  hclust(., method="centroid") %>%
  .$order %>%
  eset.of.interest.Stau2[.,]


# annotations
pd <- pData(affyNorm.core) %>% tbl_df()

heatmap.annots <- pd %>% filter(., source=="whole_RNA", pulldown!="Wt1") %>% mutate(., sampleID=sub("_whole", "", sampleID)) %>% select(., sampleID, pulldown)
heatmap.annots <- as.data.frame(heatmap.annots)
heatmap.annots <- col2rownames(heatmap.annots, "sampleID")

# annotation colors
pulldown_colors <- c(cbPalette[1:2] )
names(pulldown_colors) <- unique(unlist(heatmap.annots$pulldown))
ann_colors = list(pulldown = pulldown_colors)
## Heatmaps

pheatmap(as.matrix(select(eset.of.interest.Stau2, -symbol, -logFC)),
         annotation=heatmap.annots, 
         color = colorRampPalette(rev(brewer.pal(n = 11, name = "RdBu")))(100),
         #color=rev(brewer.pal(11,"RdBu")), 
         cluster_cols = FALSE, 
         cluster_rows = FALSE,
         main="Genes enriched after pulldown with Stau2", 
         fontsize=12,
         fontsize_row=8,
         annotation_colors=ann_colors,
         show_colnames=TRUE, 
         show_rownames=TRUE,
         labels_row = eset.of.interest.Stau2$symbol,
                 cellwidth=40, annotation_names_col = FALSE)
```

## Gene Ontologies

```{r functional1, fig.align='center', dev='png'}
#top200.Stau2 <- arrange(top.results.Stau2 , desc(logFC)) %>% select(., symbol) %>% as.data.frame() %>% unlist(.) %>% as.vector(.) %>% .[1:200]
top200.Stau2 <- filter(top.results.Stau2, logFC>=1) %>% select(., symbol) %>% as.data.frame() %>% unlist(.) %>% as.vector(.)
library(gProfileR)
background <- all.results[[1]]$stats.eset %>% tbl_df() %>% dplyr::select(., symbol) %>% as.data.frame() %>% unlist(.) %>% as.vector(.)

 gprofiler_results  <-     gprofiler(query = top200.Stau2, 
            organism = "mmusculus",
            ordered_query = T, 
            exclude_iea = F, 
            max_p_value = 0.05, 
            max_set_size = 0,
            correction_method = "gSCS",
            hier_filtering = "none", 
            domain_size = "annotated") 
 
write.csv(gprofiler_results, file.path(resultsDir, "GOresults.csv"))
``` 

Running the all genes with at least a two-fold expression change (logFC>=1) through [g:profiler](http://biit.cs.ut.ee/gprofiler/) reveals significant enrichment of genes involved in cytoskeletal protein binding. 
[File with GO results](../results/GOresults.csv)

```{r gosummaries, dev='png', width="100%"}
library(GOsummaries)
bp1 <- gosummaries(list(`Biological Process`=top200.Stau2), organism = "mmusculus", ordered_query = TRUE, go_branches = "BP")
plot(bp1, panel_height=0, wordcloud_colors=c( "#41B6C4", "#081D58"))

cc1 <- gosummaries(list(`Cellular Component`=top200.Stau2), organism = "mmusculus", ordered_query = TRUE, go_branches = "CC")
plot(cc1, panel_height=0, wordcloud_colors=c( "#41B6C4", "#081D58"))
mf1 <- gosummaries(list(`Molecular Function`=top200.Stau2), organism = "mmusculus", ordered_query = TRUE, go_branches = "MF")
plot(mf1, panel_height=0, wordcloud_colors=c( "#41B6C4", "#081D58"))
```

Gene ontology enrichment analyses can yield an overwhelming number of enriched categories, many with redundant functionality. We can simplify this output by identifying the most representative subset of the terms, using metrics which measure the semantic similarity of the terms. [Revigo](http://revigo.irb.hr/) `r citep("10.1371/journal.pone.0021800")` performs such analyses, using an algortithm which forms

>  groups of highly similar GO terms, where the choice of the groups’ representatives is guided by the p-values

The algorithm takes into account the parent-child structure of the gene onotology database

> If the p-values are quite close and one term is a child node of the other, REVIGO will tend to choose the parent term

The algorithm also tries to find more specific GO terms.

> Very general GO terms, however, are always avoided as cluster representatives ... as they tend to be uninformative

Revigo allows visualization of these representatives and their relations to the terms within their group  as a [treemap](http://en.wikipedia.org/wiki/Treemapping). Here the color depicts a grouping of related terms, the size of a block, it's pvalue from g:profiler and the large text the most representative gene ontology term for the related group.


```{r revigo, dev='png'}
# A treemap R script produced by the REVIGO server at http://revigo.irb.hr/
# If you found REVIGO useful in your work, please cite the following reference:
# Supek F et al. "REVIGO summarizes and visualizes long lists of Gene Ontology
# terms" PLoS ONE 2011. doi:10.1371/journal.pone.0021800

# author: Anton Kratz <anton.kratz@gmail.com>, RIKEN Omics Science Center, Functional Genomics Technology Team, Japan
# created: Fri, Nov 02, 2012  7:25:52 PM
# last change: Fri, Nov 09, 2012  3:20:01 PM

# -----------------------------------------------------------------------------
# If you don't have the treemap package installed, uncomment the following line:
# install.packages( "treemap" );
library(treemap) 								# treemap package by Martijn Tennekes

# Set the working directory if necessary
# setwd("C:/Users/username/workingdir");

# --------------------------------------------------------------------------
# Here is your data from REVIGO. Scroll down for plot configuration options.

bp.pval.revigo.names  <- c("term_ID","description","freqInDbPercent","abslog10pvalue","uniqueness","dispensability","representative");
bp.pval.revigo.data <- rbind(c("GO:0000003","reproduction",0.214,4.4711,1.000,0.000,"reproduction"),
c("GO:0006996","organelle organization",0.932,25.9830,0.796,0.000,"organelle organization"),
c("GO:1902589","single-organism organelle organization",0.635,21.9872,0.716,0.619,"organelle organization"),
c("GO:0044087","regulation of cellular component biogenesis",0.127,2.8210,0.693,0.486,"organelle organization"),
c("GO:0030030","cell projection organization",0.378,3.9393,0.777,0.588,"organelle organization"),
c("GO:0008152","metabolic process",82.183,5.6038,0.999,0.000,"metabolism"),
c("GO:0009987","cellular process",65.994,12.4145,0.998,0.000,"cellular process"),
c("GO:0022414","reproductive process",0.121,4.4815,0.993,0.000,"reproductive process"),
c("GO:0023052","signaling",3.838,3.8447,0.994,0.000,"signaling"),
c("GO:0032259","methylation",2.781,2.0491,0.987,0.000,"methylation"),
c("GO:0032501","multicellular organismal process",0.790,2.6536,0.993,0.000,"multicellular organismal process"),
c("GO:0032502","developmental process",1.387,10.6635,0.993,0.000,"developmental process"),
c("GO:0044699","single-organism process",53.975,11.2692,0.997,0.000,"single-organism process"),
c("GO:0048518","positive regulation of biological process",0.684,10.8125,0.779,0.000,"positive regulation of biological process"),
c("GO:0048523","negative regulation of cellular process",0.942,6.2534,0.693,0.343,"positive regulation of biological process"),
c("GO:0048522","positive regulation of cellular process",0.429,9.3363,0.712,0.317,"positive regulation of biological process"),
c("GO:0048519","negative regulation of biological process",1.243,5.7825,0.766,0.365,"positive regulation of biological process"),
c("GO:0051235","maintenance of location",0.029,3.5513,0.812,0.247,"positive regulation of biological process"),
c("GO:2001251","negative regulation of chromosome organization",0.003,2.2381,0.648,0.579,"positive regulation of biological process"),
c("GO:0016310","phosphorylation",6.300,3.9706,0.922,0.317,"positive regulation of biological process"),
c("GO:0010646","regulation of cell communication",0.380,5.4634,0.724,0.313,"positive regulation of biological process"),
c("GO:0051302","regulation of cell division",0.069,5.6038,0.747,0.270,"positive regulation of biological process"),
c("GO:0050794","regulation of cellular process",13.656,14.0200,0.672,0.578,"positive regulation of biological process"),
c("GO:0044057","regulation of system process",0.025,1.8356,0.693,0.249,"positive regulation of biological process"),
c("GO:0044089","positive regulation of cellular component biogenesis",0.002,1.3325,0.724,0.606,"positive regulation of biological process"),
c("GO:0043547","positive regulation of GTPase activity",0.130,3.9281,0.699,0.284,"positive regulation of biological process"),
c("GO:0065009","regulation of molecular function",0.835,3.6990,0.779,0.331,"positive regulation of biological process"),
c("GO:0065008","regulation of biological quality",2.709,2.7620,0.752,0.405,"positive regulation of biological process"),
c("GO:2000021","regulation of ion homeostasis",0.008,1.6126,0.836,0.399,"positive regulation of biological process"),
c("GO:0051983","regulation of chromosome segregation",0.011,1.3969,0.789,0.234,"positive regulation of biological process"),
c("GO:0097193","intrinsic apoptotic signaling pathway",0.015,1.4168,0.748,0.417,"positive regulation of biological process"),
c("GO:0033160","positive regulation of protein import into nucleus, translocation",0.000,2.0685,0.767,0.559,"positive regulation of biological process"),
c("GO:0050896","response to stimulus",8.818,5.9914,0.994,0.000,"response to stimulus"),
c("GO:0051179","localization",17.864,2.3585,0.995,0.000,"localization"),
c("GO:0051641","cellular localization",1.638,7.2510,0.967,0.000,"cellular localization"),
c("GO:0008104","protein localization",1.716,5.2573,0.953,0.346,"cellular localization"),
c("GO:0033036","macromolecule localization",1.969,4.8697,0.966,0.354,"cellular localization"),
c("GO:0030705","cytoskeleton-dependent intracellular transport",0.009,2.6716,0.922,0.596,"cellular localization"),
c("GO:0051649","establishment of localization in cell",1.489,3.8097,0.936,0.340,"cellular localization"),
c("GO:0065007","biological regulation",14.918,14.2733,0.994,0.000,"biological regulation"),
c("GO:0071840","cellular component organization or biogenesis",5.425,22.1349,0.994,0.000,"cellular component organization or biogenesis"),
c("GO:0044238","primary metabolic process",51.925,3.5850,0.984,0.016,"primary metabolism"),
c("GO:0006281","DNA repair",1.953,15.6696,0.723,0.033,"DNA repair"),
c("GO:0006302","double-strand break repair",0.063,6.4237,0.792,0.680,"DNA repair"),
c("GO:0000278","mitotic cell cycle",0.087,5.5200,0.826,0.684,"DNA repair"),
c("GO:1901699","cellular response to nitrogen compound",0.025,1.8125,0.907,0.424,"DNA repair"),
c("GO:0023051","regulation of signaling",0.379,4.0048,0.744,0.539,"DNA repair"),
c("GO:0006310","DNA recombination",1.840,5.8729,0.838,0.652,"DNA repair"),
c("GO:0018205","peptidyl-lysine modification",0.061,4.1993,0.854,0.628,"DNA repair"),
c("GO:0006366","transcription from RNA polymerase II promoter",0.236,2.2652,0.852,0.256,"DNA repair"),
c("GO:0044770","cell cycle phase transition",0.030,2.3635,0.819,0.676,"DNA repair"),
c("GO:0048583","regulation of response to stimulus",0.690,4.4634,0.739,0.497,"DNA repair"),
c("GO:0006357","regulation of transcription from RNA polymerase II promoter",0.211,1.6308,0.702,0.359,"DNA repair"),
c("GO:0044763","single-organism cellular process",41.995,12.1713,0.899,0.118,"DNA repair"),
c("GO:0035556","intracellular signal transduction",2.718,11.8182,0.628,0.689,"DNA repair"),
c("GO:0016070","RNA metabolic process",13.709,3.6144,0.806,0.539,"DNA repair"),
c("GO:0044237","cellular metabolic process",53.874,3.9508,0.931,0.190,"DNA repair"),
c("GO:0071173","spindle assembly checkpoint",0.005,2.0768,0.834,0.552,"DNA repair"),
c("GO:0044260","cellular macromolecule metabolic process",28.588,8.3893,0.835,0.525,"DNA repair"),
c("GO:0009059","macromolecule biosynthetic process",17.669,5.0650,0.850,0.506,"DNA repair"),
c("GO:0051298","centrosome duplication",0.004,4.7212,0.688,0.548,"DNA repair"),
c("GO:0070647","protein modification by small protein conjugation or removal",0.133,1.3215,0.852,0.677,"DNA repair"),
c("GO:0044249","cellular biosynthetic process",28.205,2.7932,0.858,0.554,"DNA repair"),
c("GO:0043170","macromolecule metabolic process",32.799,6.6402,0.943,0.100,"DNA repair"),
c("GO:0044271","cellular nitrogen compound biosynthetic process",15.794,2.5884,0.845,0.504,"DNA repair"),
c("GO:0006351","transcription, DNA-templated",9.014,5.8539,0.776,0.685,"DNA repair"),
c("GO:0044267","cellular protein metabolic process",8.784,1.6308,0.823,0.614,"DNA repair"),
c("GO:0051301","cell division",1.357,7.5591,0.906,0.122,"DNA repair"),
c("GO:0051304","chromosome separation",0.039,1.9431,0.812,0.688,"DNA repair"),
c("GO:0007265","Ras protein signal transduction",0.101,3.2899,0.706,0.542,"DNA repair"),
c("GO:0051321","meiotic cell cycle",0.023,5.1864,0.818,0.618,"DNA repair"),
c("GO:0007264","small GTPase mediated signal transduction",0.234,6.0757,0.696,0.520,"DNA repair"),
c("GO:1901576","organic substance biosynthetic process",28.886,2.3206,0.886,0.663,"DNA repair"),
c("GO:0046483","heterocycle metabolic process",33.326,6.7959,0.909,0.259,"DNA repair"),
c("GO:0006139","nucleobase-containing compound metabolic process",29.917,7.2907,0.851,0.528,"DNA repair"),
c("GO:0019438","aromatic compound biosynthetic process",15.562,4.6459,0.850,0.502,"DNA repair"),
c("GO:0010564","regulation of cell cycle process",0.066,7.2083,0.664,0.668,"DNA repair"),
c("GO:0043543","protein acylation",0.061,1.5045,0.860,0.629,"DNA repair"),
c("GO:0006725","cellular aromatic compound metabolic process",33.051,7.2573,0.909,0.143,"DNA repair"),
c("GO:0006259","DNA metabolic process",6.339,11.1993,0.827,0.382,"DNA repair"),
c("GO:0006260","DNA replication",1.998,2.7620,0.817,0.658,"DNA repair"),
c("GO:0098534","centriole assembly",0.002,1.4597,0.779,0.609,"DNA repair"),
c("GO:0090304","nucleic acid metabolic process",20.158,9.7375,0.819,0.388,"DNA repair"),
c("GO:0007154","cell communication",4.358,4.4295,0.895,0.142,"DNA repair"),
c("GO:0010452","histone H3-K36 methylation",0.001,2.6038,0.719,0.573,"DNA repair"),
c("GO:0010467","gene expression",16.702,2.2013,0.885,0.442,"DNA repair"),
c("GO:0018130","heterocycle biosynthetic process",16.150,4.3696,0.849,0.507,"DNA repair"),
c("GO:0018105","peptidyl-serine phosphorylation",0.011,1.3449,0.866,0.543,"DNA repair"),
c("GO:0000075","cell cycle checkpoint",0.020,3.9872,0.829,0.611,"DNA repair"),
c("GO:0007049","cell cycle",1.410,14.2676,0.906,0.123,"DNA repair"),
c("GO:0043412","macromolecule modification",5.091,4.9136,0.903,0.217,"DNA repair"),
c("GO:0051052","regulation of DNA metabolic process",0.156,3.5287,0.708,0.497,"DNA repair"),
c("GO:0007059","chromosome segregation",0.373,4.2204,0.915,0.106,"DNA repair"),
c("GO:0009314","response to radiation",0.092,2.5986,0.930,0.405,"DNA repair"),
c("GO:0018022","peptidyl-lysine methylation",0.003,4.4908,0.861,0.493,"DNA repair"),
c("GO:0022402","cell cycle process",0.527,14.2328,0.804,0.110,"DNA repair"),
c("GO:1901362","organic cyclic compound biosynthetic process",16.583,4.3242,0.876,0.497,"DNA repair"),
c("GO:1901360","organic cyclic compound metabolic process",33.913,6.4522,0.942,0.230,"DNA repair"),
c("GO:0034641","cellular nitrogen compound metabolic process",33.428,4.6904,0.880,0.432,"DNA repair"),
c("GO:0006950","response to stress",4.119,4.4855,0.909,0.621,"DNA repair"),
c("GO:0009966","regulation of signal transduction",0.363,3.6073,0.672,0.560,"DNA repair"),
c("GO:0000726","non-recombinational repair",0.013,2.7305,0.813,0.597,"DNA repair"),
c("GO:0000725","recombinational repair",0.042,2.6144,0.798,0.656,"DNA repair"),
c("GO:0006464","cellular protein modification process",2.895,5.4660,0.805,0.239,"DNA repair"),
c("GO:0006928","cellular component movement",0.492,3.7852,0.913,0.109,"DNA repair"),
c("GO:0000724","double-strand break repair via homologous recombination",0.041,2.6144,0.795,0.656,"DNA repair"),
c("GO:2001038","regulation of cellular response to drug",0.000,1.7471,0.814,0.495,"DNA repair"),
c("GO:0034645","cellular macromolecule biosynthetic process",17.200,5.7986,0.793,0.446,"DNA repair"),
c("GO:0009058","biosynthetic process",30.325,2.0066,0.984,0.038,"biosynthesis"),
c("GO:0006807","nitrogen compound metabolic process",38.650,3.4498,0.984,0.044,"nitrogen compound metabolism"),
c("GO:0044702","single organism reproductive process",0.096,4.9914,0.925,0.052,"single organism reproductive process"),
c("GO:0071704","organic substance metabolic process",56.178,2.6968,0.984,0.057,"organic substance metabolism"),
c("GO:0044767","single-organism developmental process",1.354,10.6021,0.773,0.069,"single-organism developmental process"),
c("GO:0061061","muscle structure development",0.035,2.0947,0.820,0.689,"single-organism developmental process"),
c("GO:0043009","chordate embryonic development",0.043,4.2807,0.769,0.699,"single-organism developmental process"),
c("GO:0007569","cell aging",0.006,1.8633,0.817,0.600,"single-organism developmental process"),
c("GO:0060322","head development",0.004,6.1192,0.840,0.590,"single-organism developmental process"),
c("GO:0007420","brain development",0.041,5.0942,0.760,0.697,"single-organism developmental process"),
c("GO:0030029","actin filament-based process",0.077,1.5072,0.924,0.091,"actin filament-based process"),
c("GO:0007017","microtubule-based process",0.172,8.9872,0.920,0.098,"microtubule-based process"));

bp.pval.stuff <- data.frame(bp.pval.revigo.data);
names(bp.pval.stuff) <- bp.pval.revigo.names;

bp.pval.stuff$abslog10pvalue <- as.numeric( as.character(bp.pval.stuff$abslog10pvalue) );
bp.pval.stuff$freqInDbPercent <- as.numeric( as.character(bp.pval.stuff$freqInDbPercent) );
bp.pval.stuff$uniqueness <- as.numeric( as.character(bp.pval.stuff$uniqueness) );
bp.pval.stuff$dispensability <- as.numeric( as.character(bp.pval.stuff$dispensability) );

treemap(
	bp.pval.stuff,
	index = c("representative","description"),
	vSize = "abslog10pvalue",
	type = "categorical",
	vColor = "representative",
	title = "REVIGO Biological Process Gene Ontology treemap - pvalue based",
	inflate.labels = FALSE,      # set this to TRUE for space-filling group labels - good for posters
	lowerbound.cex.labels = 0,   # try to draw as many labels as possible (still, some small squares may not get a label)
	bg.labels = "#CCCCCC00",     # define background color of group labels
												       # "#CCCCCC00" is fully transparent, "#CCCCCCAA" is semi-transparent grey, NA is opaque
	position.legend = "none",
  border.col=c("black","gray35"),
  palette.HCL.options=list(hue_start=80, hue_perm=TRUE),
  border.lwds=c(5,2),
  fontsize.labels=c(36,12),
  fontcolor.labels=c("black", "gray34"),
  fontsize.title=14
  )

################################################################################################################
################################################################################################################

################################################################################################################
################################################################################################################

cc.pval.revigo.names <-  c("term_ID","description","freqInDbPercent","abslog10pvalue","uniqueness","dispensability","representative");
cc.pval.revigo.data <- rbind(c("GO:0005623","cell",64.133,15.2565,0.990,0.000,"cell"),
c("GO:0005856","cytoskeleton",0.714,23.0000,0.574,0.000,"cytoskeleton"),
c("GO:0044464","cell part",64.133,15.3439,0.906,0.262,"cytoskeleton"),
c("GO:0044449","contractile fiber part",0.027,4.7721,0.552,0.360,"cytoskeleton"),
c("GO:0005694","chromosome",0.973,11.0980,0.564,0.495,"cytoskeleton"),
c("GO:0044446","intracellular organelle part",5.949,20.7932,0.471,0.640,"cytoskeleton"),
c("GO:0097431","mitotic spindle pole",0.000,2.8097,0.602,0.538,"cytoskeleton"),
c("GO:0072686","mitotic spindle",0.014,3.4660,0.518,0.700,"cytoskeleton"),
c("GO:0042641","actomyosin",0.008,1.4225,0.548,0.673,"cytoskeleton"),
c("GO:0043228","non-membrane-bounded organelle",8.439,29.6162,0.603,0.493,"cytoskeleton"),
c("GO:0044430","cytoskeletal part",0.579,15.7375,0.430,0.502,"cytoskeleton"),
c("GO:0044427","chromosomal part",0.450,3.6253,0.518,0.503,"cytoskeleton"),
c("GO:0044428","nuclear part",0.645,18.4214,0.528,0.663,"cytoskeleton"),
c("GO:0043232","intracellular non-membrane-bounded organelle",7.682,29.6162,0.492,0.629,"cytoskeleton"),
c("GO:0005737","cytoplasm",38.159,10.8633,0.809,0.465,"cytoskeleton"),
c("GO:0044424","intracellular part",43.664,24.5482,0.821,0.144,"cytoskeleton"),
c("GO:0043229","intracellular organelle",15.790,21.9508,0.551,0.690,"cytoskeleton"),
c("GO:0044422","organelle part",6.517,22.3904,0.613,0.573,"cytoskeleton"),
c("GO:0005634","nucleus",2.809,20.6364,0.589,0.365,"cytoskeleton"),
c("GO:0043292","contractile fiber",0.030,4.1649,0.655,0.363,"cytoskeleton"),
c("GO:0031981","nuclear lumen",0.437,20.9872,0.494,0.491,"cytoskeleton"),
c("GO:0005622","intracellular",46.136,24.6162,0.907,0.199,"cytoskeleton"),
c("GO:0005911","cell-cell junction",0.091,3.4698,0.846,0.000,"cell-cell junction"),
c("GO:0030054","cell junction",0.154,5.4789,0.973,0.000,"cell junction"),
c("GO:0031974","membrane-enclosed lumen",0.517,17.4237,0.973,0.000,"membrane-enclosed lumen"),
c("GO:0032991","macromolecular complex",14.462,4.1494,0.977,0.000,"macromolecular complex"),
c("GO:0043226","organelle",16.716,18.7282,0.978,0.000,"organelle"),
c("GO:0044456","synapse part",0.066,2.7799,0.973,0.000,"synapse part"),
c("GO:0045202","synapse",0.094,3.5638,0.973,0.000,"synapse"),
c("GO:0030427","site of polarized growth",0.013,3.7282,0.943,0.023,"site of polarized growth"),
c("GO:0030496","midbody",0.014,2.8041,0.943,0.023,"midbody"),
c("GO:0031252","cell leading edge",0.034,2.1107,0.941,0.025,"cell leading edge"),
c("GO:0097458","neuron part",0.092,3.5186,0.938,0.027,"neuron part"),
c("GO:0044463","cell projection part",0.617,2.8356,0.886,0.032,"cell projection part"),
c("GO:0001726","ruffle",0.015,2.1018,0.903,0.597,"cell projection part"),
c("GO:0060053","neurofilament cytoskeleton",0.000,1.7033,0.607,0.646,"cell projection part"),
c("GO:0030426","growth cone",0.008,2.9393,0.881,0.573,"cell projection part"),
c("GO:0042995","cell projection",1.485,4.9626,0.928,0.035,"cell projection"),
c("GO:0034708","methyltransferase complex",0.032,1.8013,0.843,0.072,"methyltransferase complex"),
c("GO:0043234","protein complex",8.273,7.1331,0.916,0.573,"methyltransferase complex"),
c("GO:1902494","catalytic complex",3.772,1.4711,0.903,0.404,"methyltransferase complex"));

  
  
  
  
cc.pval.stuff <- data.frame(cc.pval.revigo.data);
names(cc.pval.stuff) <- cc.pval.revigo.names;

cc.pval.stuff$abslog10pvalue <- as.numeric( as.character(cc.pval.stuff$abslog10pvalue) );
cc.pval.stuff$freqInDbPercent <- as.numeric( as.character(cc.pval.stuff$freqInDbPercent) );
cc.pval.stuff$uniqueness <- as.numeric( as.character(cc.pval.stuff$uniqueness) );
cc.pval.stuff$dispensability <- as.numeric( as.character(cc.pval.stuff$dispensability) );

# check the tmPlot command documentation for all possible parameters - there are a lot more
treemap(
  cc.pval.stuff,
	index = c("representative","description"),
	vSize = "abslog10pvalue",
	type = "categorical",
	vColor = "representative",
	title = "REVIGO Cellular Component Gene Ontology treemap - pvalue based",
	inflate.labels = FALSE,      # set this to TRUE for space-filling group labels - good for posters
	lowerbound.cex.labels = 0,   # try to draw as many labels as possible (still, some small squares may not get a label)
	bg.labels = "#CCCCCC00",     # define background color of group labels
	# "#CCCCCC00" is fully transparent, "#CCCCCCAA" is semi-transparent grey, NA is opaque
	position.legend = "none",
	border.col=c("black","gray35"),
	palette.HCL.options=list(hue_start=80, hue_perm=TRUE),
	border.lwds=c(5,2),
	fontsize.labels=c(36,12),
	fontcolor.labels=c("black", "gray34"),
  fontsize.title=14
)

################################################################################################################
################################################################################################################

################################################################################################################
################################################################################################################

mf.pval.revigo.names <-  c("term_ID","description","freqInDbPercent","abslog10pvalue","uniqueness","dispensability","representative");
mf.pval.revigo.data <- rbind(c("GO:0003774","motor activity",0.216,6.1637,0.885,0.000,"motor activity"),
c("GO:0016817","hydrolase activity, acting on acid anhydrides",7.557,5.8153,0.896,0.274,"motor activity"),
c("GO:0004386","helicase activity",1.170,1.7986,0.868,0.465,"motor activity"),
c("GO:0003777","microtubule motor activity",0.060,3.9508,0.894,0.363,"motor activity"),
c("GO:0017111","nucleoside-triphosphatase activity",7.074,6.6108,0.845,0.678,"motor activity"),
c("GO:0003824","catalytic activity",69.295,1.6946,0.992,0.000,"catalytic activity"),
c("GO:0005085","guanyl-nucleotide exchange factor activity",0.061,4.8210,0.975,0.000,"guanyl-nucleotide exchange factor activity"),
c("GO:0005088","Ras guanyl-nucleotide exchange factor activity",0.031,4.4318,0.952,0.000,"Ras guanyl-nucleotide exchange factor activity"),
c("GO:0005488","binding",55.585,17.3316,0.989,0.000,"binding"),
c("GO:0005515","protein binding",2.482,15.8729,0.900,0.000,"protein binding"),
c("GO:0008047","enzyme activator activity",0.115,5.1124,0.916,0.000,"enzyme activator activity"),
c("GO:0030234","enzyme regulator activity",0.437,2.1343,0.975,0.000,"enzyme regulator activity"),
c("GO:0018024","histone-lysine N-methyltransferase activity",0.015,4.6676,0.790,0.015,"histone-lysine N-methyltransferase activity"),
c("GO:0008276","protein methyltransferase activity",0.193,2.2291,0.816,0.589,"histone-lysine N-methyltransferase activity"),
c("GO:0016772","transferase activity, transferring phosphorus-containing groups",9.188,2.1811,0.846,0.403,"histone-lysine N-methyltransferase activity"),
c("GO:0016278","lysine N-methyltransferase activity",0.016,2.4101,0.821,0.613,"histone-lysine N-methyltransferase activity"),
c("GO:0008757","S-adenosylmethionine-dependent methyltransferase activity",1.037,1.8097,0.794,0.689,"histone-lysine N-methyltransferase activity"),
c("GO:0016773","phosphotransferase activity, alcohol group as acceptor",4.038,2.8729,0.834,0.517,"histone-lysine N-methyltransferase activity"),
c("GO:0008170","N-methyltransferase activity",0.452,2.2823,0.806,0.484,"histone-lysine N-methyltransferase activity"),
c("GO:0004402","histone acetyltransferase activity",0.010,3.5186,0.894,0.130,"histone-lysine N-methyltransferase activity"),
c("GO:0034212","peptide N-acetyltransferase activity",0.017,2.9431,0.891,0.495,"histone-lysine N-methyltransferase activity"),
c("GO:0004677","DNA-dependent protein kinase activity",0.001,2.1101,0.893,0.459,"histone-lysine N-methyltransferase activity"),
c("GO:0016741","transferase activity, transferring one-carbon groups",3.193,1.5003,0.859,0.273,"histone-lysine N-methyltransferase activity"),
c("GO:0004672","protein kinase activity",1.875,1.7055,0.836,0.673,"histone-lysine N-methyltransferase activity"),
c("GO:0004674","protein serine/threonine kinase activity",0.565,3.1878,0.848,0.168,"histone-lysine N-methyltransferase activity"),
c("GO:0016740","transferase activity",22.118,1.6003,0.955,0.029,"transferase activity"),
c("GO:0001046","core promoter sequence-specific DNA binding",0.005,1.3188,0.919,0.038,"core promoter sequence-specific DNA binding"),
c("GO:0003677","DNA binding",13.924,12.9281,0.830,0.200,"core promoter sequence-specific DNA binding"),
c("GO:0003676","nucleic acid binding",21.746,9.3125,0.818,0.282,"core promoter sequence-specific DNA binding"),
c("GO:1901265","nucleoside phosphate binding",20.353,4.7144,0.820,0.317,"core promoter sequence-specific DNA binding"),
c("GO:0003682","chromatin binding",0.115,5.9245,0.924,0.050,"chromatin binding"),
c("GO:0008092","cytoskeletal protein binding",0.162,10.1273,0.789,0.052,"cytoskeletal protein binding"),
c("GO:0015631","tubulin binding",0.066,6.6180,0.783,0.524,"cytoskeletal protein binding"),
c("GO:0042393","histone binding",0.016,1.6364,0.815,0.476,"cytoskeletal protein binding"),
c("GO:0032403","protein complex binding",0.146,5.6925,0.790,0.556,"cytoskeletal protein binding"),
c("GO:0019899","enzyme binding",0.221,7.8447,0.785,0.575,"cytoskeletal protein binding"),
c("GO:0017016","Ras GTPase binding",0.029,5.7471,0.771,0.495,"cytoskeletal protein binding"),
c("GO:0005516","calmodulin binding",0.017,2.2798,0.815,0.478,"cytoskeletal protein binding"),
c("GO:0008270","zinc ion binding",3.455,2.7011,0.857,0.074,"zinc ion binding"),
c("GO:0017076","purine nucleotide binding",15.986,6.0132,0.729,0.686,"zinc ion binding"),
c("GO:0035639","purine ribonucleoside triphosphate binding",15.477,6.1746,0.738,0.660,"zinc ion binding"),
c("GO:0046914","transition metal ion binding",7.335,1.7190,0.841,0.694,"zinc ion binding"),
c("GO:0097159","organic cyclic compound binding",42.443,11.6091,0.845,0.304,"zinc ion binding"),
c("GO:1901363","heterocyclic compound binding",42.435,12.2055,0.845,0.124,"zinc ion binding"),
c("GO:0097367","carbohydrate derivative binding",16.547,3.3665,0.870,0.196,"zinc ion binding"),
c("GO:0046872","metal ion binding",15.488,3.9830,0.823,0.594,"zinc ion binding"),
c("GO:0005524","ATP binding",13.885,7.7055,0.683,0.275,"zinc ion binding"),
c("GO:0043168","anion binding",20.956,5.9747,0.829,0.415,"zinc ion binding"),
c("GO:0043167","ion binding",33.309,8.6308,0.852,0.267,"zinc ion binding"),
c("GO:0043169","cation binding",15.811,3.4868,0.838,0.430,"zinc ion binding"),
c("GO:0036094","small molecule binding",21.712,3.1506,0.864,0.219,"zinc ion binding"),
c("GO:0016787","hydrolase activity",21.764,1.5482,0.955,0.074,"hydrolase activity"));

  
  
  
mf.pval.stuff <- data.frame(mf.pval.revigo.data);
names(mf.pval.stuff) <- mf.pval.revigo.names;

mf.pval.stuff$abslog10pvalue <- as.numeric( as.character(mf.pval.stuff$abslog10pvalue) );
mf.pval.stuff$freqInDbPercent <- as.numeric( as.character(mf.pval.stuff$freqInDbPercent) );
mf.pval.stuff$uniqueness <- as.numeric( as.character(mf.pval.stuff$uniqueness) );
mf.pval.stuff$dispensability <- as.numeric( as.character(mf.pval.stuff$dispensability) );

# check the tmPlot command documentation for all possible parameters - there are a lot more
treemap(
  mf.pval.stuff,
  index = c("representative","description"),
	vSize = "abslog10pvalue",
	type = "categorical",
	vColor = "representative",
	title = "REVIGO Molecular Function Gene Ontology treemap - pvalue based",
	inflate.labels = FALSE,      # set this to TRUE for space-filling group labels - good for posters
	lowerbound.cex.labels = 0,   # try to draw as many labels as possible (still, some small squares may not get a label)
	bg.labels = "#CCCCCC00",     # define background color of group labels
	# "#CCCCCC00" is fully transparent, "#CCCCCCAA" is semi-transparent grey, NA is opaque
	position.legend = "none",
	border.col=c("black","gray35"),
	palette.HCL.options=list(hue_start=80, hue_perm=TRUE),
	border.lwds=c(5,2),
	fontsize.labels=c(36,12),
	fontcolor.labels=c("black", "gray34"),
  fontsize.title=14
)
```


### Specific GO terms

- are any of the genes from the interaction comparison involved in cell adhesion?
- working from Gene Ontology term for cellular senescence, I pulled down all genes annotated with this [term](http://amigo.geneontology.org/amigo/term/GO:0007155) (and it's "child" terms)
- here are the top 10 genes associated with cell adhesion, the full list is in the Excel file below

```{r GO_annots}
# pull in probeid to GO term mappings
probeGOmappings <- AnnotationDbi::select(mogene10sttranscriptcluster.db, row.names(all.results[[1]]$stats.eset), "GO", "PROBEID")
# get probes associated with cell adhesion specific GO terms
probeGOmappings.celladhesion <- probeGOmappings[which(probeGOmappings$GO %in% c(as.vector(GOBPCHILDREN$"GO:0007155"), "GO:0007155")),]

# assign probeids to gene symbols
geneGOmappings.celladhesion <- merge(fData(affyNorm.core)[,c("probesetid", "symbol", "entrezID")], probeGOmappings.celladhesion, by.x="probesetid", by.y="PROBEID", all.y=TRUE)
# aggregate all GO terms for individual genes
geneGOmappings.celladhesion <- aggregate(geneGOmappings.celladhesion$GO,geneGOmappings.celladhesion['symbol'],paste,collapse=',')
names(geneGOmappings.celladhesion) <- c("symbol", "GO")

# get ranks for all genes and add to stat dataframe
ranks <- seq(1:nrow(top.results.Stau2))
top.results.Stau2 <- cbind(ranks, top.results.Stau2)

# merge in cell adhesion specific GO term data
top.results.Stau2 <- merge(top.results.Stau2, geneGOmappings.celladhesion, by="symbol", all.x=TRUE)


kable(top.results.Stau2[!is.na(top.results.Stau2$GO),][1:10,], row.names=F)

write.table(top.results.Stau2[!is.na(top.results.Stau2$GO),], file=file.path(resultsDir, "celladhesion.GO.genes.xls"), sep="\t", row.names=F, col.names=T) 
```

[Excel file with Cell Adhesion Genes statistics](celladhesion.GO.genes.xls)

----

# R Session Info

(useful if replicating these results)

```{r sessioninfo}
sessionInfo()
```

---

# References

```{r writebib, results='hide', echo=FALSE, message=FALSE}
write.bibtex(file="references.bib")
```

