---
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: journal
    css: custom.css
title: "NMF analysis of proteomic data - Set1"
theme: united
---

```{r setupstate, echo=FALSE}
library(knitr)
opts_chunk$set(tidy=TRUE, cache=FALSE,  highlight=TRUE, figalign="center", echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120))
options(width=200)
```

# NMF ANALYSES OF PROTEOMIC DATA - 6plex dataset - Set1

## Summary  

Analysis of proteomic data for [Leo Wang](mailto:Leo.Wang@childrens.harvard.edu) in the Wager's group at the Joslin.

Nonnegative matrix factorization [(NMF)][1] analysis was conducted on the samples to explore sample similarity and to identify proteins/genes that might be characteristic for individual isolated cell types. NMF tries to identify 'factors' (or metagenes, a weighted list of genes) that best describe individual samples. This is useful to not only identify how similar the samples are, but also in which way they are similar.

Contact [John Hutchinson](mailto:jhutchin@hsph.harvard.edu) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the included results and plots.

## Setup

### Libraries 

```{r libraries, results='hide', message=FALSE, warning=FALSE, warning=FALSE}
library(knitr)
require(NMF)
library(genefilter)
library(ggdendro)
library(ggplot2)
library(GOstats)
library(biomaRt)
library(AnnotationForge)
library(plyr)
library(googleVis)
library(CHBUtils)
library(dplyr)
library(readr)

ensembl = useMart("ensembl",dataset="mmusculus_gene_ensembl")
attributes <- listAttributes(ensembl)
filters <- listFilters(ensembl)
```

### Functions

```{r functions}
PCAplot.sd <- function(eset=NULL,  title=NULL){
  eset.core <- exprs(eset)
  myPca.core <- prcomp(t(eset.core))
  # SD of components
  sdevdf <- data.frame(cbind(as.numeric(myPca.core$sdev),c(1:length(myPca.core$sdev))))
  sdevdf$prop <-  sdevdf$X1/sum(sdevdf$X1)
  sdevdf$cum <- cumsum(sdevdf$prop)
  ggplot(sdevdf, aes(x=X2, y=prop)) + 
    geom_point(size=4, color="red") + 
    scale_x_continuous('Component') + 
    scale_y_continuous('Standard Deviation') +
    ggtitle(title) +
    geom_line(data=sdevdf, aes(x=X2, y=cum))
}
```

### Variables

```{r variables}
if (file.exists("/n/hsphS10/hsphfs1/chb/projects/aw_NMF_proteomics/Mar2015")) {
  baseDir <- "/n/hsphS10/hsphfs1/chb/projects/aw_NMF_proteomics/Mar2015"
  }  else if (file.exists("/Users/johnhutchinson/projects/aw_NMF_proteomics/Mar2015")) {
    baseDir <- "/Users/johnhutchinson/projects/aw_NMF_proteomics/Mar2015"
    } else {
      baseDir <- "/Volumes/ody/consults/aw_NMF_proteomics/Mar2015"
      }
dataDir <- file.path(baseDir, "data", "Set1")
resultsDir <- file.path(baseDir, "results", "Set1")
metaDir <- file.path(baseDir, "meta", "Set1")
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")

TRAIN=50
RUN=1250
mad.cutoff=0.5

minnumfeatures=25
```

## Data preparation and exploration

### Load in Data

Only the corrected counts  were analysed. All counts  log transformed to stabilize variance. 

```{r dataload, results='hide'}
data <- read.csv(file.path(dataDir, "normalized.data.JH.csv"))
metadata <- read.delim(file.path(metaDir, "metadata.set1.txt"), header=T) 
metadata$sampleID <- paste(metadata$group_short, metadata$replicate, sep="")


# some potentiall  hacky text manipulations occur here to label the results
data$GeneName <- gsub(", ", "/", gsub("\\[|\\]|\\'", "", data$GeneName))

# grab the annotation columns
annot <- data[,as.vector(metadata$columnid[which(metadata$datatype=="descriptor")])]
names(annot)[2] <- "Modifications"
# mod annotation columns to get unique and R-parseable IDs
annot$Modifications <- gsub(": ","_", annot$Modifications)
annot$Modifications <- gsub("; ","/", annot$Modifications)
annot$Modifications[annot$Modifications==""] <- NA
annot$num <- seq(1,nrow(annot))
annot$uniqid <- paste(annot$num,annot$GeneName, annot$Modifications,sep="-")

row.names(annot) <- annot$uniqid
row.names(data) <- annot$uniqid


# grab the columns with corrected/normalized data
data.corr <- data[,as.vector(unlist(metadata$columnid[which(metadata$datatype=="norm")]))]
# trim any rows with only zero values for spectra
annot <- annot[!apply(data.corr, 1, function(x) all(x==0)),]
data.corr <- data.corr[!apply(data.corr, 1, function(x) all(x==0)),]

# log transform data
data.corr <- log2(data.corr)
# check to make sure all the results are finite i.e. no zeros snuck in
length(which(!apply(data.corr, 1, function(s) all(is.finite(s)))))

# subset metadata
metadata.corr <- subset(metadata, datatype=="norm")
metadata.corr <- col2rownames(metadata.corr, "columnid", removecol = FALSE)

# are annots and data the same rowsize?
nrow(data.corr)==nrow(annot)
identical(row.names(data.corr),row.names(annot))
ncol(data.corr)==nrow(metadata.corr)
all(names(data.corr)==row.names(metadata.corr))

# stuff into an ExpressionSEt object
eset.corr <- ExpressionSet(assayData=as.matrix(data.corr), phenoData=new("AnnotatedDataFrame", data = metadata.corr), featureData=new("AnnotatedDataFrame", data=annot))

rm(data.corr, metadata.corr)
```

### Initial Clustering

Initial hierarchical clustering based on the euclidian distance of individual samples.

```{r clustering}
plot_dendro(eset.corr,  title="Normalized Data", labels.colname="sampleID", colors.colname="group_long")
```

The dendrograms look good. There is clear separation based on sample type.

### Principle Component Analyses

```{r PCA}
PCAplot.sd(eset.corr,   title="Proportion of Variance Explained by Components  - Corrected Data")
PCAplot.eset(eset.corr,  categories="group_short", title="PCA - Normalized Data", colorpalette=cbPalette, numcomponents=4)
```

The first four components explain ~87.5% of the sample differences for both approaches to the data, and when compared pairwise,  the first two components easily separate out the cell types.

## NMF Preparation

### Estimating the factorization rank
From ["An introduction to NMF package"][2]

>A critical parameter in NMF is the factorization rank r. It determines the number of metagenes used
to approximate the target matrix. Given a NMF method and the target matrix, a common way of
deciding on r is to try different values, compute some quality measure of the results, and choose
the best value according to this quality criteria.

#### Using metrics:

From ["An introduction to NMF package"][2]

>(Brunet et al. 2004) proposed to take the first value of r for which the cophenetic coefficient starts
decreasing, (Hutchins et al. 2008) suggested to choose the first value where the RSS curve presents
an inflection point, and (Frigyesi et al. 2008) considered the smallest value at which the decrease
in the RSS is lower than the decrease of the RSS obtained from random data.

```{r estimatefactoriziationrank, results='hide', cache=TRUE}
groups.corr <-  as.factor(eset.corr$group_short)
estim.corr <- nmf(eset.corr, 2:5, nrun = TRAIN, seed = 123456, .options='v-p') #disable parallel compute
plot(estim.corr)
```

These results suggest there are at 3-4 metagenes to separate all the classes as  the largest inflection point of the rss curve occurs at the 3rd factorization rank and the cophenetic curve drops at the 4th factorization ranks.

Even for random data, increasing factorization ranks lead to more variables to fit the data, possibly leading to overfitting the data. Too assess this, we can run the same metrics after randomly permuting the data, and comparing the performance of the original and randomized datasets. 

```{r overfitcheck, results='hide',warning=FALSE, message=FALSE, error=FALSE,cache=TRUE}
# shuffle original data to look for overfitting
eset.corr.rand <- randomize(eset.corr)
# estimate quality measures from the shuffled data (use default NMF algorithm)
estim.corr.rand <- nmf(eset.corr.rand, 2:5, nrun = TRAIN, seed = 12345, .options="v-p")
# plot measures on same graph
plot(estim.corr, estim.corr.rand)
```

In the plots above, the solid lines represent the actual data while the dotted lines represent the same data after random shuffling by permuting the rows of each column, using a different permutation each time. These results show that overfitting is not an issue.

#### Qualitative assessment

We can also examine heatmaps of the consensus matrix for each value of the factorization rank to see if the clusters (or consensus blocks) obtained correspond to the known cell types.

```{r estimatefactoriziationrank.qualitative, results='hide'}
consensusmap(estim.corr, annCol=list(samplegroups=as.character(unlist(eset.corr$group_short ))),annColors=list(samplegroups=cbPalette[1:3]), labCol=groups.corr, labRow=groups.corr, scale="row", color='-RdYlBu2:200')
```

We could probably use 2 or 3 metagenes to separate the samples. In the interest of simplicity, we will proceed with Rank=2 (2 metagenes), where the "consensus" clustering  best corresponds to the D1 and D4 samples. 

Going forward for this analysis, I tried looking at factorization ranks of 2 (i.e. 2 metagenes) to see if a deeper analysis could better define the expected groups.

### Choosing an algorithm

To this point, I have been using the default (*brunet*) algorithm for the gross tuning of NMF; we can further fine tune NMF by choosing an alternate algorithm.
 
To do so, I ran NMF on the data using four of the most popular algorithms (with error tracking enabled), and compared the error plots. In these error plots, each track is normalized so that its first value equals one, and stops at the iteration where the method's convergence criterion was fulfilled; algorithms that converge at lower objective values are optimal.

```{r comparealgs, cache=TRUE, results="hide", message=FALSE, error=FALSE,warning=FALSE}
res.multi.method.2 <- nmf(eset.corr, 2, list("brunet", "KL", "lee","nsNMF"), nrun=TRAIN, seed = 123456, .options = "tv-p")
plot(res.multi.method.2, main="NMF residuals - 2 metagenes")
```

Here, the *nsMMF* algorirthm works best. The nsNMF approach has the slight benefit of creating 'sparse' metagenes which are easier to interpret functionally, so I proceeded with this algorithm.

## Complete NMF Algorithm

The NMF analyses so far have used a low number of iterations (<50) so may not be completely accurate,. Here I ran NMF with `r RUN` iterations to allow the algorithm to fully converge and yield as accurate a result as possible.

```{r fullNMF, eval=TRUE}
res.final.2 <- nmf(eset.corr, 2, "nsNMF", nrun=RUN, .options = "tv-p")

# save precomputed NMF values, hack to avoid using caching
save(list="res.final.2", file=file.path(resultsDir,  "RDATA.res.final.2"))
```


```{r fullNMFload, echo=FALSE}
# load precomputed NMF values, hack to avoid using caching
load(file.path(resultsDir,  RDATA.res.final.2))
```

We can look at the final results in various ways:

The consensus matrix plot is similar to the previous plot, just with more iterations. It shows how well the calculated metagenes cluster the samples.

```{r rank2consensumap}
consensusmap(res.final.2,  annCol=list(samplegroups=as.character(unlist(eset.corr$group_short ))),  labCol=groups.corr, labRow=groups.corr, annColors=list(samplegroups=cbPalette[1:2], basis=cbPalette[5:6], consensus=brewer.pal(3, "Spectral")))
```

### NMF Metagene Feature Extraction

The "Metagene Components"" plots indicate the contributions of a gene to the individual metagenes. 

Here I first plotted all genes that contribute to any metagene.

```{r rank2basismap, cache=TRUE}
basismap(res.final.2, scale="r1",annColors=list(basis=cbPalette[5:6], consensus=brewer.pal(2, "Spectral")), main="Metagene Components - All Contributing Genes") # shows metagenes in relation to samples
```

#### All NMF Metagene Features

We can extract the genes that make a contribution to the different metagenes. In the Excel files below, I included all genes that contribute to a metagene.

```{r rank2.metagenes}
fs2 <- featureScore(res.final.2) # get all the feature scores which measure specificity to which a gene contributes to a metagene
f2 <- extractFeatures(res.final.2) # extract the the 50 most features with the highest specifciity for each of the metagenes

# dataframe to present the metagene features that includes both the metagene annotations and the featurescores
metagene2.1 <- cbind(fData(eset.corr)[f2[[1]],], fs2[f2[[1]]]) 
names(metagene2.1)[ncol(metagene2.1)] <- "featurescore"
metagene2.2 <- cbind(fData(eset.corr)[f2[[2]],], fs2[f2[[2]]]) 
names(metagene2.2)[ncol(metagene2.2)] <- "featurescore"

write.table(as.matrix(metagene2.1), file=file.path(resultsDir, paste("metagene", 1, "xls", sep=".")), sep="\t", quote=F)
write.table(as.matrix(metagene2.2), file=file.path(resultsDir, paste("metagene", 2, "xls", sep=".")), sep="\t", quote=F)

## get unique annotations for genes (NMF returns number referenced IDs) in metagenes
unique.metagenesymbols <- lapply(f2, function(x) {
  genenames <- unique(unlist(fData(eset.corr)[x,"GeneName"]))
  return(genenames)
  })
## get number of unique genes in each metagene
numgenes <- unlist(lapply(unique.metagenesymbols, length))
```

[Excel file of metagene number 1 features](../results/6plex_1/metagene.1.xls)  
[Excel file of metagene number 2 features](../results/6plex_1/metagene.2.xls)  

#### Most Specific NMF Metagene Features

As genes can contribute to more than one metagene it can be beneficial to investigate genes that have relatively restricted contribution to a single metagene. This contribution can be measured with the featurescore, as defined by [Kim et al. (2007).][5] and yields a number within the range of 0 to 1; the higher the feature score for a feature of a metagene, the more metagene-specific the feature. 

We can filter for the features that are the most specific to each metagene. Here I filtered for the most specific features of each metagene, with a relative contribution cutoff selected so that there are at least 20 features shown for each metagene. 

```{r specificmetagene}
numfeatures <- sapply(seq(0,1,0.01), function(num) {
  lapply(extractFeatures(res.final.2, num), function(x){
    length(x)
  })
})
rel.basis.contrib.cutoff <- max(seq(0,1,0.01)[apply(numfeatures,2, function(x) all(x>50))])

basismap(res.final.2, scale="r1", subsetRow=rel.basis.contrib.cutoff,annColors=list(basis=cbPalette[5:8], consensus=brewer.pal(4, "Spectral")), main="Metagene Components - Most Specific Genes") # shows metagenes in relation to samples


fs2 <- featureScore(res.final.2) # get all the feature scores which measure specificity to which a gene contributes to a metagene
f2 <- extractFeatures(res.final.2, rel.basis.contrib.cutoff) # extract the the 50 most features with the highest specifciity for each of the metagenes

# dataframe to present the metagene features that includes both the metagene annotations and the featurescores
sp.metagene2.1 <- cbind(fData(eset.corr)[f2[[1]],], fs2[f2[[1]]]) 
names(sp.metagene2.1)[ncol(sp.metagene2.1)] <- "featurescore"
sp.metagene2.2 <- cbind(fData(eset.corr)[f2[[2]],], fs2[f2[[2]]]) 
names(sp.metagene2.2)[ncol(sp.metagene2.2)] <- "featurescore"

write.table(as.matrix(sp.metagene2.1), file=file.path(resultsDir, paste("specific.metagene", 1, "xls", sep=".")), sep="\t", quote=F)
write.table(as.matrix(sp.metagene2.2), file=file.path(resultsDir, paste("specific.metagene", 2, "xls", sep=".")), sep="\t", quote=F)
```

[Excel file of metagene number 1 features](../results/6plex_1/specific.metagene.1.xls)  
[Excel file of metagene number 2 features](../results/6plex_1/specific.metagene.2.xls)  

### Relating Samples to Metagenes

The Mixture coefficients plot show how well the metagenes define the individual samples.

```{r rank3coefmap}
coefmap(res.final.2, scale="c1", labCol=groups.corr, annColors=list(basis=cbPalette[5:6], consensus=brewer.pal(2, "Spectral"))) # shows the individual genes in relation to metagenes and samples
```

Looking at the plot, we see that we can classify the different cell types with different combinations of metagenes. We can try to classify the different cell types with a specific combination of metagenes that defines each cell type uniquely, by clustering on two metagenes (of  `r numgenes[1]` and `r numgenes[2]` genes respectively).
For instance, looking at the coefficient plot, we can classify the RBM cells with one metagene of  `r numgenes[1]` genes.

Expanding on this, 
RBMs = metagene 2
PMs = metagene  1

#### RBM specific metagene

- these are the most specific genes of metagene 2

```{r unmob.metagene.print, results='asis', echo=FALSE}
kable(rbind(sp.metagene2.2,sp.metagene2.2))
```

#### PM specific metagene
- these are the features of metagene 1

```{r mobBM.metagene.print, echo=FALSE, results='asis'}
kable(rbind(sp.metagene2.1,sp.metagene2.1))
```

## R Session Information Summary

```{r savesession, echo=FALSE}
sessionInfo()
save.image(file.path(resultsDir, "RDATA.full"))
```

[1]: http://www.biomedcentral.com/1471-2105/11/367 (NMF)
[2]: http://nmf.r-forge.r-project.org/vignettes/NMF-vignette.pdf (NMF vignette)
[3]: http://www.bioconductor.org/packages/release/bioc/html/GOstats.html (GOstats)
[4]: http://www.geneontology.org/GO.doc.shtml (GO)
[5]: http://www.ncbi.nlm.nih.gov/pubmed/17483501 (Kim et al., 2007)

