
```{r toptable}
stats.top.core <- topTable(fit2.core, coef=1, adjust="fdr", sort.by="p", p.value=1, number=100, genelist=cbind(fit2.core$genes,symbols)) 
# expression levels for these genes in all samples
eset.top.core = exprs(affyNorm.core[stats.top.core[, "ID"], ]) 
colnames(eset.top.core) = pData(affyNorm.core)$Sample
eset.stats.top.core <- cbind(stats.top.core, eset.top.core)
```




### output pvalues and expression levels to files

```{r outputs}
write.table(stats.core, file="./results/all.genes.stats.tab", quote=F, sep="\t", row.names=F, col.names=T)
write.table(eset.stats.top.core, file='./results/KO_vs_WT.top100.pvalues.expression_levels.xls', row.names=F, sep='\t')
```
[results for all genes]("./results/all.genes.stats.tab")
[genes with top100 pvalues]("./results/KO_vs_WT.top100.pvalues.xls")


### heatmaps of top results

```{r heatmaps, fig.cap="KO_vs_WT.top100pvalues.annotated.heatmap"}
## add in gene symbols for each Affy probe
row.names(eset.top.core)=  paste(row.names(eset.top.core), stats.top.core$symbols, sep= "-")
#write out heatmaps with Affy probeset IDs
pheatmap(eset.top.core, main=colnames(fit2.core$coefficients), color=brewer.pal(11, "RdBu"), fontsize_row=8)
```


##PATHPRINT
## Pathprint, can we find any cell types that match well? how do the samples differ in terms of pathways?
mic.data.platform <- "GPL6246"
mic.data.species <- "mouse"
mic.data.names <- pData(affyNorm.core)$Sample

mic.data.fingerprint <- exprs2fingerprint(exprs = eset.core, platform = mic.data.platform, species = mic.data.species, progressBar = TRUE)

## KO v WT
### variable pahtwyas
pdf("./results/pathprint.variable.pathways.pdf")
heatmap(mic.data.fingerprint[apply(mic.data.fingerprint, 1, sd) > 0.5, ], labCol = mic.data.names, mar = c(10,20), col = c("blue", "white", "red"))
dev.off()

consistent.pathways=mic.data.fingerprint[apply(mic.data.fingerprint, 1, sd) > 0.5, ]
pdf("./results/pathprint.consistent.pathways.pdf", height=11, width=8.5)
heatmap(consistent.pathways[which(consistent.pathways[,1]==consistent.pathways[,2] & consistent.pathways[,3]==consistent.pathways[,4] & consistent.pathways[,1]!=consistent.pathways[,4]),1:4], labCol = mic.data.names[1:4], mar = c(10,25), col = c("blue", "white", "red"))
dev.off()

# similar cell types
WT.consensus=consensusFingerprint(mic.data.fingerprint[,1:2], threshold=0.9)
geo.WT.consensus.distance=consensusDistance(WT.consensus, GEO.fingerprint.matrix)
table(GEO.metadata.matrix[match(head(row.names(geo.WT.consensus.distance),50), GEO.metadata.matrix$GSM),"Source"])

KO.consensus=consensusFingerprint(mic.data.fingerprint[,1:2], threshold=0.9)
geo.KO.consensus.distance=consensusDistance(KO.consensus, GEO.fingerprint.matrix)
table(GEO.metadata.matrix[match(head(row.names(geo.KO.consensus.distance),50), GEO.metadata.matrix$GSM),"Source"])
```

[^1]: BioC
[^2]: arrayQualityMetrics
[^3]: RMA
