<!-- Performs DESeq on two UNREPLICATED samples. Uses the blind method when
estimating the dispersions so do not use this if you have replicates as
this will ignore it. Edit the variables in the variables block for
your experiment.

infile: a file with the rownames as ids with a column for each sample
wd: directory to run the script in
design: describes what the two columns represent
gene_symbol: how to look up the gene symbol using biomaRt
ensembl_gene: how to look up the ensembl gene id using biomaRt
filter_type: the type of identifier that your id is

The biomaRt vignette contains documentation that can help you figure what
these should be:
http://www.bioconductor.org/packages/2.11/bioc/html/biomaRt.html
-->

```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, echo=FALSE, highlight=TRUE, figalign="center", fig.height=6, fig.width=6, message=FALSE, error=FALSE, warning=FALSE, cache=TRUE)
```


```{r libraries}
library("DESeq")
library("googleVis")
library("ggplot2")
library("ggthemes")
library("scales")
```

```{r variables}
wd = '/Users/rory/cache/projects/ben_trap_pilot/scripts/'
setwd(wd)
infile = paste(wd, '../data/combined.counts', sep="")
design = c("CLK1", "CLK3")
gene_symbol = 'mgi_symbol'
ensembl_gene = 'mmusculus_gene_ensembl'
filter_type = 'ensembl_gene_id'
```
Differential expression of CLK1 vs CLK3 for a pilot study.

```{r data-load}
design = c("CLK1", "CLK3")
counts = read.table(infile, header=TRUE, row.names=1, col.names=design)
cds = newCountDataSet(counts, design)
```

```{r annotate-df}
annotate_df = function(d) {
	require(biomaRt)
	ensembl = useMart('ensembl', dataset = ensembl_gene)
	a = getBM(attributes=c(filter_type, gene_symbol, "description"),
		filters=c(filter_type), values=d[, 'id'],
		mart=ensembl)
	m = merge(d, a, by.x='id', by.y=filter_type)
	return(m)
}
```

```{r gvistabl-clean}
clean_gvistable = function(df) {
	df[sapply(df, is.infinite)] = NA
	names(df)[match("baseMeanA", names(df))] = c(design[1])
	names(df)[match("baseMeanB", names(df))] = c(design[2])
	return(df)
}
```


## Normalization
Here is a sample of 25 of the non-normalized counts of reads mapping to
genes in the annotation. These are just chosen at random for
display purposes:
```{r counts, results='asis'}
annotated_counts = head(as.data.frame(counts(cds)), 25)
annotated_counts$id = rownames(annotated_counts)
annotated_counts = annotate_df(annotated_counts)
print(gvisTable(annotated_counts, options=list(width=900)), "chart")
```
To normalize for different depth of sequencing, size factors are calculated
for each sample.
Call $k_{ij}$ the count for gene $i$ in sample $j$.
For all samples $j$, assume most of the counts for genes will have
a constant ratio comparing any two samples in $j$, reflecting the assumption
that most genes are not differentially expressed between samples.
A reference set of
counts is created by taking $f_i$, geometric mean of the counts for each
gene $i$, and the median of of $\frac{k_i}{f_i}$ for sample $j$ is called the size
factor for sample $j$. The size factors can be thought of as an attempt to
normalize for the total number
of reads mapping to transcripts, depending on the assumption that most
of the transcripts are not vastly different between samples.

Here are the size factors for each sample in this experiment:
```{r sizefactors, tidy=TRUE}
cds = estimateSizeFactors(cds)
sizeFactors(cds)
```

These are the normalized counts, which are the raw counts divided by
the size factor, for the same 25 genes:
```{r normalized_counts, results='asis'}
annotated_normalized = head(as.data.frame(counts(cds, normalized=TRUE)), 25)
annotated_normalized$id = rownames(annotated_normalized)
annotated_normalized = annotate_df(annotated_normalized)
print(gvisTable(annotated_normalized, options=list(width=900)), "chart")
```

We can see the assumption holds true by plotting the ratio of counts for
both two samples and noting the distribution has a large central peak.
On the left are the raw counts and on the right are the normalized counts.
The effect of normalization is to shift the mean ratio towards 1.

```{r ratio_hist, fig.cap=""}
raw = counts(cds)
countdata = data.frame(ratio=raw[,1] / raw[,2])
c = ggplot(countdata, aes(ratio)) + geom_density() +
	scale_x_log10(breaks=c(0.01, 0.1, 1, 10, 100),
		labels=math_format(format=log10)) +
	theme_grey(base_size=16, base_family="GillSans") +
	opts(title="raw counts")
c
normalized = counts(cds, normalized=TRUE)
normdata = data.frame(ratio=normalized[,1] / normalized[,2])
n = ggplot(normdata, aes(ratio)) + geom_density() +
	scale_x_log10(breaks=c(0.01, 0.1, 1, 10, 100),
		labels=math_format(format=log10)) +
	theme_grey(base_size=16, base_family="GillSans") +
	opts(title="normalized counts")
n
```

## Variance estimation
For samples without replicates it is not possible to get an accurate
repesentation of the biological variability. However if we make an
assumption that genes with counts at a given level all have similar
dispersions we can use the dispersions calculated across all genes with
a given number of counts as a proxy for the dispersion of that gene.
Dispersion describes the fluctuation of the observed counts for a gene
around the mean counts for that gene and can be the square of the
coefficient of biological variation.
For example, if a gene's expression level fluctuates by 20% the
calculated dispersion is
$0.2^2 = 0.04$.

Below is a plot of the estimated dispersion for genes with given counts;
the dispersion increases as the mean count decreases meaning it is
more difficult to accurately measure the expression level of
low-count genes.

```{r test, echo=FALSE, cache=TRUE, fig.cap=""}
cds = estimateDispersions(cds, method='blind',  sharingMode="fit-only")
plotDispEsts = function(cds) {
	estimates = data.frame(means = rowMeans(counts(cds, normalized=TRUE)),
		variance = fitInfo(cds)$perGeneDispEsts)
	xg = 10^seq(-0.5, 5, length.out=300)
	yg = fitInfo(cds)$dispFun(xg)
	fitline = data.frame(xg=xg, yg=yg)
	p = ggplot(estimates, aes(means, variance)) + geom_point(size=1) +
		scale_x_log10() + scale_y_log10() +
		geom_line(data=fitline, aes(xg, yg), color="red") +
		theme_grey(base_size=16, base_family="GillSans") +
		opts(title="dispersion estimation while pooling all samples") +
		xlab("mean number of mapped reads per gene") +
		ylab("estimated dispersion")
	p
}
plotDispEsts(cds)
```

## Differential expression
### All upregulated genes with adjusted p-value < 0.1:
```{r upregulated, echo=FALSE,  results='asis', cache=TRUE}
res = nbinomTest(cds, design[1], design[2])
upreg_sig = subset(res, padj < 0.1 & foldChange > 1)
upreg_sig = clean_gvistable(upreg_sig)
print(gvisTable(annotate_df(upreg_sig), options=list(width=900)), "chart")
```

### All downregulated genes with adjusted p-value < 0.1:
```{r downregulated, echo=FALSE, cache=TRUE, results='asis'}
downreg_sig = subset(res, padj < 0.1 & foldChange < 1)
downreg_sig = clean_gvistable(downreg_sig)
print(gvisTable(annotate_df(downreg_sig), options=list(width=900)), "chart")
```

```{r diffexp_plots, echo=FALSE, cache=TRUE}
plotMvA <- function(res) {
    plot(res$baseMean, res$log2FoldChange, log="x", pch=20,
		cex=.3, col=ifelse(res$padj < .1, "red", "black"))
}
```

On the left is a MvA plot showing the calculated fold change vs. the mean
expression level. On the right is a histogram of the unadjusted p-values
as a diagnostic; samples with no differentially expressed genes should
have an exponential-like p-value distribution without the uptick at the
extreme low p-values.

```{r diagnostic}
plotMvA(res)
hist(res$pval, breaks=100, col="skyblue", border="slateblue", main="")
```

The top 25 most downregulated genes by fold change for genes with reads
in both conditions.
```{r downreg_foldchange, results='asis', cache=TRUE}
#print(gvisTable(annotated_normalized, options=list(width=900)), "chart")
non_inf = subset(res, foldChange < Inf & log2FoldChange > -Inf)
downreg = head(non_inf[order(non_inf$foldChange, -non_inf$baseMean), ], 25)
downreg = clean_gvistable(downreg)
annotated_downreg = annotate_df(downreg)
print(gvisTable(annotated_downreg, options=list(width=900)), "chart")
```
The top 25 most upregulated genes by fold change for genes with reads
in both conditions.
```{r upreg_foldchange, echo=FALSE, results='asis', cache=TRUE}
upreg = head(non_inf[order(-non_inf$foldChange, -non_inf$baseMean), ], 25)
upreg = clean_gvistable(upreg)
annotated_upreg = annotate_df(upreg)
print(gvisTable(annotated_upreg, options=list(width=900)), "chart")
```

```{r write_results}
annotated_res = annotate_df(res)
out_file = paste(infile, ".deseq.annotated.txt", sep="")
write.table(annotated_res, out_file, quote=FALSE, row.names=FALSE, sep="\t")
```
