---
  html_document:
    toc: true
    highlight: zenburn
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='')
```

# Overview

```{r check-library-functions, echo=FALSE}
check_install = function(packages) {
   not_installed = setdiff(packages, rownames(installed.packages()))
   if(length(not_installed) > 0) {
      write(paste("The libraries", not_installed, "are not installed, aborting.",
                   sep=" "), stdout())
      stop()
   }
}
```

```{r load-libraries, echo=FALSE}
packages = c("ggplot2", "reshape", "gplots", "edgeR", "CHBUtils", "pheatmap",
              "DESeq2", "tximport", "logging", "DT")
check_install(packages)
installed = lapply(packages, library, character.only = TRUE)
```

```{r qc-setup}
basicConfig()
project_summary = "/n/data1/cores/bcbio/gu_rnaseq/project_hg19/final/2016-03-21_project_hg19/project-summary.csv"
counts_file = "/n/data1/cores/bcbio/gu_rnaseq/project_hg19/final/2016-03-21_project_hg19/combined.counts"
tx2genes_file = "/n/data1/cores/bcbio/gu_rnaseq/project_hg19/final/2016-03-21_project_hg19/tx2gene.csv"
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442",
"#0072B2", "#D55E00", "#CC79A7")
summarydata = data.frame(read.table(project_summary, header=TRUE, sep=","), row.names="Name", check.rows=FALSE)
summarydata$Name = rownames(summarydata)
summarydata = summarydata[order(summarydata$Name),]
if (file.exists(tx2genes_file)) {
  loginfo("Using gene counts calculated from the Sailfish transcript counts.")
  sf_files = file.path("..", "..", rownames(summarydata), "sailfish",
                      rownames(summarydata), "quant.sf")
  names(sf_files) = rownames(summarydata)
  tx2gene = read.table(tx2genes_file, sep=",", row.names=NULL, header=FALSE)
  txi.salmon = tximport(sf_files, type="salmon", tx2gene=tx2gene,
                        reader=readr::read_tsv, countsFromAbundance="lengthScaledTPM")
  counts = round(data.frame(txi.salmon$counts, check.names=FALSE))
} else {
  loginfo("Using gene counts calculated from featureCounts.")
  counts = read.table(counts_file, header=TRUE, row.names="id", check.names=FALSE)
}
counts = counts[, order(colnames(counts)), drop=FALSE]
colnames(counts) = gsub(".counts", "", colnames(counts))

# this is a list of all non user-supplied metadata columns that could appear
known_columns = c("Name", "X.GC", "Exonic.Rate", "Sequences.flagged.as.poor.quality",
    "rRNA_rate", "Fragment.Length.Mean", "Intronic.Rate", "Intergenic.Rate",
    "Mapping.Rate", "Quality.format", "Duplication.Rate.of.Mapped", "Mapped",
    "rRNA", "Sequence.length", "Transcripts.Detected", "Mean.Per.Base.Cov.",
    "Genes.Detected", "Unique.Starts.Per.Read", "unique_starts_per_read",
    "complexity", "X5.3.bias", "Duplicates.pct", "Duplicates", "Mapped.reads",
    "Median.insert.size", "Mapped.reads.pct", "Total.reads",
    "avg_coverage_per_region", "Mapped.Reads")
metadata = summarydata[, !colnames(summarydata) %in% known_columns, drop=FALSE]
```

```{r sanitize-datatable}
sanitize_datatable = function(df, ...) {
 # remove dashes which cause wrapping
 DT::datatable(df, ..., rownames=gsub("-", "_", rownames(df)),
                   colnames=gsub("-", "_", colnames(df)))
}
```

```{r set-random-seed}
# set seed for reproducibility
set.seed(1454944673)
```

## Sample metadata
```{r summary-table, echo=FALSE, cache=FALSE}
sanitize_datatable(summarydata, style='bootstrap')
```

```{r heatmap-function}
get_heatmap_fn = function(summarydata) {
    # return the pheatmap function with or without metadata
    if(ncol(metadata) == 0) {
       return(pheatmap)
    }
    else {
    # rownames(metadata) = summarydata$Name
    heatmap_fn = function(data, ...) {
        pheatmap(data, annotation=metadata, ...)
    }
    return(heatmap_fn)
}}
heatmap_fn = get_heatmap_fn(summarydata)
```

# Quality control metrics

```{r qualimap-run}
qualimap_run = "Mapped" %in% colnames(summarydata)
```

## Mapped reads
```{r mapped-plot, eval=qualimap_run}
ggplot(summarydata, aes(x=Name, y=Mapped)) +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    geom_bar(stat="identity") +
    ylab("mapped reads") + xlab("")
```

```{r mapped-plot-noqualimap, eval=!qualimap_run}
ggplot(summarydata, aes(x=Name, y=Mapped.reads)) +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    geom_bar(stat="identity") +
    ylab("mapped reads") + xlab("")
```

## Genomic mapping rate
```{r mapping-rate-plot, eval=qualimap_run}
ggplot(summarydata, aes(x=Name, y=Mapping.Rate)) +
    geom_bar(stat="identity") +
    ylab("mapping rate") + xlab("") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90))
```

```{r mapping-rate-plot-noqualimap, eval=!qualimap_run}
ggplot(summarydata, aes(x=Name, y=Mapped.reads.pct)) +
    geom_bar(stat="identity") +
    ylab("mapping rate") + xlab("") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90))
```

## Number of genes detected
```{r genes-detected-plot}
dd = data.frame(Name=colnames(counts), Genes.Detected = colSums(counts > 0))
ggplot(dd, aes(x=Name, y=Genes.Detected)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("")
```

## Gene detection saturation
```{r saturation-plot}
dd = data.frame(Mapped=summarydata$Mapped, Genes.Detected = colSums(counts > 0))
ggplot(dd, aes(x=Mapped, y=Genes.Detected)) +
    geom_point() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("reads mapped")
```

## Exonic mapping rate
```{r exonic-mapping-plot, eval="Exonic.Rate" %in% colnames(summarydata)}
ggplot(summarydata, aes(x=Name, y=Exonic.Rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("exonic mapping rate") + xlab("")
```

## rRNA mapping rate
```{r check-rRNA-eval}
eval_rRNA = "rRNA_rate" %in% colnames(summarydata) & !sum(is.na(summarydata$rRNA_rate)) == nrow(summarydata)
```
```{r rRNA-rate-plot, eval=eval_rRNA}
ggplot(summarydata, aes(x=Name, y=rRNA_rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("rRNA rate") + xlab("")
```

## Estimated fragment length of paired-end reads
```{r fragment-length-plot, eval="Fragment.Length.Mean" %in% colnames(summarydata)}
ggplot(summarydata, aes(x=Name, y=Fragment.Length.Mean)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("fragment length") + xlab("")
```

## Boxplot of log10 counts per gene
```{r boxplot-raw}
melted = melt(counts)
colnames(melted) = c("sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Boxplot of log10 TMM-normalized counts per gene
Trimmed mean of M-values (TMM) normalization is described
[here](http://genomebiology.com/2010/11/3/R25)

Robinson, M. D., & Oshlack, A. (2010). A scaling normalization method for differential expression analysis of RNA-seq data. Genome Biology, 11(3). doi:10.1186/gb-2010-11-3-r25

```{r boxplot-normalized}
y = DGEList(counts=counts)
y = calcNormFactors(y)
normalized_counts = cpm(y, normalized.lib.sizes=TRUE)
melted = melt(normalized_counts)
colnames(melted) = c("gene", "sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Density of log10 TMM-normalized counts
```{r density-normalized}
ggplot(melted, aes(x=count, group=sample)) +
    geom_density() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Correlation (Pearson) heatmap of TMM-normalized counts
```{r pearson-heatmap-normalized, eval=ncol(counts) > 1}
heatmap_fn(cor(normalized_counts, method="pearson"))
```

## Correlation (Spearman) heatmap of TMM-normalized counts
```{r spearman-heatmap-normalized, eval=ncol(counts) > 1}
heatmap_fn(cor(normalized_counts, method="spearman"))
```

## PCA plot
```{r pca, eval=ncol(counts) > 1}
dds = DESeqDataSetFromMatrix(countData=counts, colData=summarydata, design=~Name)
vst = varianceStabilizingTransformation(dds)
plotPCA(vst, intgroup=c("Name"))
```

## Heatmap by concordance correlation coefficient
http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004075

```{r propcor-heatmap, eval=ncol(counts) > 1}
propcor = function(x, y) {
    x = log(x + 0.1)
    y = log(y + 0.1)
    num = 2 * cov(x, y)
    denom = var(x) + var(y)
return(num/denom)}

do_propcor = function(x) {
     mat = list()
     for(i in seq_len(ncol(x))) {
         for(j in seq_len(ncol(x))) {
        x2 = x[, i]
        y2 = x[, j]
        mat = c(mat, propcor(x2, y2)) } }
    mat = unlist(mat)
    mat = matrix(mat, ncol(x), ncol(x))
    colnames(mat) = colnames(x)
    rownames(mat) = colnames(x)
    return(mat)}

heatmap_fn(do_propcor(normalized_counts))
```
```{r deseq2-tximport-helper-functions}
# snagged from development version of DESeq
DESeqDataSetFromTximport <- function(txi, colData, design, ...)
{
  counts <- round(txi$counts)
  mode(counts) <- "integer"
  dds <- DESeqDataSetFromMatrix(countData=counts, colData=colData, design=design, ...)
  stopifnot(txi$countsFromAbundance %in% c("no","scaledTPM","lengthScaledTPM"))
  if (txi$countsFromAbundance %in% c("scaledTPM","lengthScaledTPM")) {
    message("using length scaled TPM counts from tximport")
  } else {
    message("using counts and average transcript lengths from tximport")
    lengths <- txi$length
    dimnames(lengths) <- dimnames(dds)
    assays(dds)[["avgTxLength"]] <- lengths
  }
  return(dds)
}

subset_tximport = function(txi, rows, columns) {
    txi$counts = txi$counts[rows, columns]
    txi$abundance = txi$abundance[rows, columns]
    txi$length = txi$length[rows, columns]
    return(txi)
}
```

```{r de-setup}
library(DEGreport)
library(vsn)
design = ~day+sampletype
condition = "sampletype"
```

# Differential expression

```{r deseq2-expression-analysis, results='asis'}
counts <- counts[rowSums(counts>0)>1,]
if(exists("txi.salmon")) {
   loginfo("Using Sailfish gene counts for the DESeq2 model.")
   txi.salmon = subset_tximport(txi.salmon, rownames(counts), colnames(counts))
   dds = DESeqDataSetFromTximport(txi.salmon, colData=summarydata, design=design)
} else {
  loginfo("Using counts from featureCounts for the DESeq2 model.")
  dds = DESeqDataSetFromMatrix(countData=counts,
      colData=summarydata, design = design  )
}
geoMeans = apply(counts, 1, function(row) if (all(row == 0)) 0 else
                 exp(mean(log(row[row != 0]))))
dds = estimateSizeFactors(dds, geoMeans=geoMeans)
dds = DESeq(dds)
```

## Effect of variance stabilization

```{r deseq-diagnostics, results='asis'}
par(mfrow=c(1,3))
notAllZero <- (rowSums(counts(dds))>0)
rld <- rlog(dds)
vsd <- varianceStabilizingTransformation(dds)
rlogMat <- assay(rld)
vstMat <- assay(vsd)

meanSdPlot(log2(counts(dds,normalized=TRUE)[notAllZero,] + 1))
meanSdPlot(assay(rld[notAllZero,]))
meanSdPlot(assay(vsd[notAllZero,]))
```

## Dispersion estimates

```{r dispersion-estimate}
plotDispEsts(dds)
```

```{r deseq2-handler}
handle_deseq2 = function(dds, summarydata, column) {
  all_combs = combn(levels(summarydata[,column]), 2, simplify=FALSE)
  all_results = list()
  contrast_strings = list()
  for(comb in all_combs) {
    contrast_string = paste(comb, collapse=" vs ")
    contrast = c(column, comb)
    res = results(dds, contrast=contrast)
    res = res[order(res$padj),]
    all_results = c(all_results, res)
    contrast_strings = c(contrast_strings, contrast_string)
  }
  names(all_results) = contrast_strings
  return(all_results)
}
```

## MA-plots

```{r DESeq-output, results='asis'}
all_results = handle_deseq2(dds, summarydata, condition)
len = length(all_results)
nr = ceiling( len / 3 )
nc = ceiling( len / nr )
par(mfrow=c(nr,nc))
for(i in seq(length(all_results))) {
  plotMA(all_results[[i]])
  title(paste("MA plot for contrast", names(all_results)[i]))
}
```

## Volcano-plots

```{r DESeq-volcano}
for(i in seq(length(all_results))) {
  stats = as.data.frame(all_results[[i]][,c(2,6)])
  p = volcano_density_plot(stats, title=names(all_results)[i], lfc.cutoff=1.5)
  print(p)
}
```

## DEGreport

```{r get-groups}
get_groups <- function(d, comp, condition)
{
  g <- unlist(strsplit(comp," "))
  g1 <- d$Name[d[, (names(d)==condition)]==g[1]]
  g2 <- d$Name[d[, (names(d)==condition)]==g[3]]
  list(g1,g2)
}
```

### Pvalues-vs-Mean

Here we plot some information about how the p-values are correlated with the
mean or the standard deviation.

```{r DEGreport-M}
plots = list()
scale_factor = round(1/nr * 14)
for(i in seq(length(all_results))) {
  plots[[i]] = degMean(all_results[[i]]$pvalue, rlogMat) +
  theme_bw(base_size = scale_factor) +
  ggtitle(paste0("Pvalues-vs-Mean for ", names(all_results)[i]))
}
do.call(grid.arrange,plots)
```

### Pvalues-vs-Variation

```{r DEGreport-V}
plots = list()
for(i in seq(length(all_results))) {
  plots[[i]] = degVar(all_results[[i]]$pvalue, rlogMat) +
  theme_bw(base_size = scale_factor) +
  ggtitle(paste0("Pvalues-vs-Variation for ", names(all_results)[i]))
}
do.call(grid.arrange,plots)
```

### Mean-vs-Variation
```{r DEGreport-MV}
plots = list()
for(i in seq(length(all_results))) {
  g <- get_groups(summarydata, names(all_results)[i], condition)
  if(length(g[[1]]) < 2 | length(g[[2]]) < 2) {
     next
   }
  plots[[i]] = degMV(g[[1]], g[[2]], all_results[[i]]$pvalue, counts(dds,normalized=TRUE)) +
  theme_bw(base_size = scale_factor) +
  ggtitle(paste0("Mean-vs-Variation for ", names(all_results)[i]))
}
if(length(plots) > 0) {
    do.call(grid.arrange,plots)
}
```

## Differentially expressed genes

```{r DESeq-tables, results='asis'}
for(i in seq(length(all_results))) {
  cat(paste("Lowest adjusted p-value hits for", names(all_results)[i]))
  out_df = as.data.frame(all_results[[i]])
  out_df$id = rownames(out_df)
  out_df = out_df[, c("id", colnames(out_df)[colnames(out_df) != "id"])]
  write.table(out_df, file=paste(names(all_results)[i], ".tsv", sep=""),
                         quote=FALSE, sep="\t", row.names=FALSE, col.names=TRUE)
  sig_genes = subset(out_df, padj < 0.05)
  DT::datatable(sig_genes)
  cat("\n")
}
```
