<!--
Analysis of anthrax infected cells vs. control cells
-->
```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, highlight=TRUE, fig.align='left', fig.show='hold', cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE, prompt=TRUE, comment='',
fig.cap='')
```
# Overview
A macrophage cell line, THP-1 was transduced via lentivral infection with a shRNA library
directed towards a subset of 4000 genes previously identified as important in
macrophage interaction with intracellular bacteria. The transduced, and non-transfected
control cells were exposed to anthrax for 48 hours until 90% of the cells were killed.
The cells were harvested and the DNA sequenced. The files were are looking at have
counts for each shRNA.

If a shRNA shows up less in the anthrax infected cells than in the control cells,
that means that the gene the shRNA targets might confer survivability to the
anthrax infection. Conversely if a shRNA shows up more often in the anthrax infected
cells than the control cells, that means the gene the shRNA targets might be involved
in the death of the cells.
```{r variables-and-libraries}
library(ggplot2)
library(reshape)
library(googleVis)
library(stringr)
library(edgeR)
library(extrafont)
font_import(pattern="Gill Sans", prompt=FALSE)
#project_dir = '/n/home05/kirchner/cache/projects/kobzik_influenza'
#data_dir = file.path(project_dir, 'data')
project_dir = '/Users/rory/cache/projects/kobzik_influenza'
data_dir = file.path(project_dir, 'raw_data')
results_dir = file.path(project_dir, 'results', 'plague')
wd = file.path(project_dir, 'scripts', 'plague')
setwd(wd)
in_file = file.path(data_dir, 'plague', 'YP-HMAC-BC14_shRNA.csv')
#load(file.path(data_dir, 'pathways.Hs.RData'))
FDR_CUTOFF = 0.10
ensembl_gene = 'hsapiens_gene_ensembl'
gene_symbol = 'hgnc_symbol'
filter_type = 'refseq_mrna'
dir.create(results_dir, showWarnings=FALSE)
```

# Data cleaning and sanity checking
Here is what the header of the files looks like.
```{r}
counts = read.table(in_file, header=TRUE, sep=",", comment.char="", stringsAsFactors=FALSE)
colnames(counts)
```
Samples 1, 4 and 6 are cells infected with plague and samples 2, 3 and 5 are controls.
Let's rename the columns to reflect that bit of information and clean up some of
the column names to be easier to work with.

Ouch. Let's fix that.

```{r}
refseq_id = unlist(lapply(counts$info, function(x) strsplit(x, "|", fixed=TRUE)[[1]][2]))
refseq_id = unlist(lapply(refseq_id, function(x) strsplit(x, ".", fixed=TRUE)[[1]][1]))
# use these symbols, excel ruins he formatting of the symbol. column
# (for example Sept8 -> September 8. Nice)
symbols = unlist(lapply(counts$label, function(x) strsplit(x, "|", fixed=TRUE)[[1]][1]))
counts$refseq_id = refseq_id
counts$symbol = symbols
counts = counts[,c("label", "shRNA.", "gene.", "symbol", "refseq_id",
     "yp1_norm", "yp2_norm", "yp3_norm", "yp4_norm", "yp5_norm",
     "yp6_norm")]
colnames(counts) = c("id", "shRNA_id", "gene_id", "symbol", "refseq_id",
    "plague_1", "control_1", "plague_2", "control_2", "plague_3", "control_3")
head(counts)
```

Much nicer.
I again left out the non-normalized columns because one of the datasets
only has the normalized columns, so we'll use the normalized column for the analysis
across the board. There are also three malformed rows that have NA values, so
those need to be removed as well:

```{r}
dim(counts)
counts = counts[rowSums(is.na(counts)) == 0,]
dim(counts)
```

Most genes have 7-8 shRNA targetting them, so we will only analyze those.

```{r}
shRNA_per_gene = table(counts$refseq_id)
table(shRNA_per_gene)
keep = shRNA_per_gene == 7 | shRNA_per_gene == 8
keep = shRNA_per_gene[keep]
counts = counts[counts$refseq_id %in% names(keep),]
shRNA_per_gene = table(counts$refseq_id)
table(shRNA_per_gene)
```

## Exploratory data analysis
We want to get the data in a form that we can manipulate easily during the
exploratory analysis, the *reshape* package, with the near-magic *cast* and
 *melt* functions is awesome for that:

```{r}
library(reshape)
head(counts)
id_cols = c("id", "shRNA_id", "gene_id", "symbol", "refseq_id")
melted = melt(counts, id=id_cols)
colnames(melted) = c(id_cols, "sample", "count")
head(melted)
```
Overall count data:
  ```{r}
 ggplot(melted, aes(count)) + geom_histogram(aes(y = ..density..)) + geom_density() +
    scale_x_log10() +
    theme(text=element_text(family="Gill Sans"))
   # theme_tufte(base_family="GillSans")

 ```
Unlike the anthrax data, the plague data is missing the tail of the distribution of
counts around zero.

shRNA counts have very similar distributions for each replicate:
```{r}
 ggplot(melted, aes(y=count, x=sample)) + geom_boxplot() + scale_y_log10()
```

We can see that, if we compare for mean shRNA counts per gene across samples, all of
the samples are highly correlated.

```{r}
casted = cast(melted, refseq_id ~ sample, value="count", mean)
plotmatrix(casted[2:7]) + theme(text=element_text(family="Gill Sans"))
```

And in correlation form:
```{r}
row_ids = colnames(casted[2:7])
cor_df = cor(casted[2:7], use='complete.obs')
colnames(cor_df) = row_ids
rownames(cor_df) = row_ids
qplot(X1, X2, data=melt(cor_df), fill=value, geom="tile") +
    theme(axis.text.x=element_text(angle=-90),
    text=element_text(family="Gill Sans"), axis.title=element_blank()) +
    scale_fill_continuous(name="Pearson correlation")
```

Finally looking at the raw counts:
```{r}
 ggplot(melted, aes(x=count)) + geom_histogram() + scale_x_log10() + facet_grid(. ~ sample) +
    theme(text=element_text(family="Gill Sans"))
```
These samples are highly, highly correlated.

# differential shRNA expression
```{r construct-edger-dataframe, results='asis'}
edger_df = counts[, 6:11]
rownames(edger_df) = counts$id
group = factor(unlist(lapply(colnames(edger_df),
                function(x) strsplit(x, "_", fixed=TRUE)[[1]][1])))
dge = DGEList(edger_df, group=group)
```
The MDS of the counts shows that the control samples cluster nicely together, but the anthrax samples are all over the place; they are almost just as different than each other in the first and and second components as they are the control samples.

```{r}
mds = function(counts, condition) {
        nprobes <- nrow(counts)
        nsamples <- ncol(counts)
        counts = cpm(counts)
	distances = dist(t(counts))
	fit = cmdscale(distances, eig=TRUE, k=5)
	colnames(fit$points) = c("one", "two", "three", "four", "five")
	df = as.data.frame(fit$points)
	df$label = rownames(df)
        df$condition = condition
	print(ggplot(df, aes(one, two, color=condition))
            + geom_point() +
              theme(text=element_text(family="Gill Sans")))

	eigs = data.frame(variance_explained=fit$eig / sum(fit$eig))
	eigs$component = as.integer(rownames(eigs))
	#print(ggplot(eigs, aes(component, variance_explained)) + geom_point())
}
mds(dge, group)
```
Libraries tend to have the same number of shRNA counts.
```{r simple-design, results='asis'}
dge = calcNormFactors(dge)
print(gvisTable(dge$samples, options=list(width=900)), "chart")
```

We fit a model of this form to the data:

```{r model}
mds(dge, group)
design = model.matrix(~0+group)
design
```

Common dispersion estimation with biological coefficient of variation.

```{r common-dispersion-estimation}
#dge = estimateCommonDisp(dge, verbose=TRUE)
dge = estimateTrendedDisp(dge)
```

Trended dispersion estimation.

```{r trended-dispersion-estimation}
dge = estimateTagwiseDisp(dge)
```

There is an extremely low amount of biological variation in these samples which
mirrors the seemingly low variation across the samples in general when we
initially looked at the data. These samples almost look like technical replicates:

```{r bcv-plot}
plotBCV(dge)
```

The model fits the data reasonably well.
```{r model-fitting}
fit = glmFit(dge, design)
g = gof(fit)
z = zscoreGamma(g$gof.statistics,shape=g$df/2,scale=2)
qqnorm(z)
```

## Differential expression
```{r}
differential_test = function(dge, pair) {
	et = exactTest(dge, pair=pair)
	et_table = et$table
	et_table$DE = as.numeric(decideTestsDGE(et, p.value=FDR_CUTOFF))
	et_table$FDR = p.adjust(et_table$PValue, method="BH")
	return(et_table)
}
et_table = differential_test(dge, c("control", "plague"))
```
Since these samples are highly similar to each other and their differences do not
seem to be systematic between the samples (from the MDS plots), there are not any
shRNA which are detectable as differentially expressed between the control and
plague infected samples.

<!-- ```{r, results='asis'} -->
<!-- de_counts = merge(counts, et_table, by.x="id", by.y=0) -->
<!-- multiple_hits = subset(data.frame(table(subset(de_counts, FDR < 0.10)$refseq_id)), Freq > 1) -->
<!-- colnames(multiple_hits) = c("refseq_id", "freq") -->
<!-- print(gvisTable(multiple_hits), "chart") -->
<!-- de_counts$multiple_hits = de_counts$refseq_id %in% multiple_hits$refseq_id -->
<!-- ``` -->

<!-- Annotate the input set of counts with metadata about the genes. -->
<!-- ```{r} -->
<!-- ensembl_gene = "hsapiens_gene_ensembl" -->
<!-- filter_type = "refseq_mrna" -->
<!-- annotate_df = function(d) { -->
<!-- 	require(biomaRt) -->
<!-- 	ensembl = useMart('ensembl', dataset = ensembl_gene) -->
<!-- 	a = getBM(attributes=c(filter_type, gene_symbol, "description", -->
<!-- 		"entrezgene"), -->
<!-- 		filters=c(filter_type), values=d[, 'refseq_id'], -->
<!-- 		mart=ensembl) -->
<!-- 	m = merge(d, a, by.x='refseq_id', by.y=filter_type) -->
<!-- 	return(m) -->
<!-- } -->
<!-- de_annotated = annotate_df(de_counts) -->
<!-- ``` -->
<!-- Display only the genes that have multiple DE shRNA and write a full table and a -->
<!-- DE table of the results. DE = -1 means it is downregulated in the anthrax infected cells, -->
<!-- 0 means it is not DE and 1 means it is upregulated in the anthrax infected cells. -->

<!-- ```{r, results='asis'} -->
<!-- print(gvisTable(subset(de_annotated, multiple_hits)), "chart") -->
<!-- write.table(de_annotated, "control_vs_anthrax_all.txt", quote=FALSE, row.names=FALSE, -->
<!--   sep="\t") -->
<!-- write.table(subset(de_annotated, multiple_hits), "control_vs_anthrax_de.txt", -->
<!--   quote=FALSE, row.names=FALSE, sep="\t") -->
<!-- ``` -->
