```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, cache=FALSE, highlight=TRUE, fig.align="center",
               warning=FALSE, error=TRUE, message=FALSE, echo=FALSE,
               fig.height=9, fig.width=9, fig.pos="htbp", fig.show="asis")
```

### GWASeq exploratory plots

Summary distribution plots of GWASeq variant results. Some common conventions 
used:

- Low count variants are defined as present in less than 10 samples. Conversely,
  high count variants are in more than 10.
- bait/offbait specify if regions fall into bait regions or in target regions
  outside of the baits.
  
\FloatBarrier

### Distribution of variant counts

The most common variant calls are low count heterozygous variants.

```{r total_variants_plot, fig.cap="The overall counts of variants per sample, dominated by low-sample count variants."}
p3 <- ggplot(d, aes(x=total_variants)) + geom_histogram() +
     opts(title="Distribution of variant counts: all") +
     xlab("Total variants")
p3
```

```{r total_variants_low, fig.cap="Overall counts zoomed to low frequency variants, by bait and non-bait regions."}
p4 <- ggplot(d.low, aes(x=total_variants)) + geom_histogram() +
      facet_grid(is_bait ~ .) +
      opts(title="Distribution of variant counts: low") +
      xlab("Total variants")
p4
```

```{r het_hom_variants, fig.cap="Low variant count samples are primarily heterozygotes."}
phh <- ggplot(d, aes(x=pct_hets)) + geom_histogram(aes(y=..density..)) +
       facet_grid(is_low ~ is_bait) +
       opts(title="Heterozygous variant distribution") +
       xlab("Percent heterozygotes")
phh
```

\FloatBarrier

### Characteristics of low count variants

Low count variants are primarily found in multiple batches, which is nice
evidence they are likely to be real. The mean coverage is similar between high
and low count variants, but differ between bait and nonbait distributions.

```{r sample_variants_plot, fig.cap="For low count samples, number of variants versus number of batches seen. The most common result is low variants in multiple batches."}
p1 <- ggplot(d.low, aes(total_variants, set)) + geom_point(position="jitter") +
     opts(title="Variants and batches: low") +
     xlab("Sample variants") + ylab("Batches")
p1
```

```{r sample_mean_plot, fig.cap="The mean coverage depth per sample for low and high count variants. Bait/nonbait distributions differ."}
p2 <- ggplot(d, aes(x=DP_sample_mean)) + geom_histogram() + facet_grid(is_low ~ is_bait) +
     opts(title="Depth of coverage") +
     xlab("Mean coverage depth per sample")
p2
```

\FloatBarrier

### Filtering approaches

Some plots of quality metrics for variants, faceted by bait/offbait and variant
sample counts (low, high). These provide some useful ways we could assign
cutoffs for additional variant filtering.

Broad filtered reads using: `FS>200.0, QD<2.0, ReadPosRankSum<-20.0, InbreedingCoeff<-0.8`

```{r qual_by_bait, fig.cap="Variant quality scores capped at 10000, by bait and variant sample count."}
p5 <- ggplot(d, aes(x=qual_trunc)) + geom_histogram() +
      facet_grid(is_low ~ is_bait) +
      opts(title="Quality score distribution") +
      xlab("Quality score")
p5
```

```{r qd_distribution, fig.cap="Variant Confidence by Depth."}
p6 <- ggplot(d, aes(x=QD)) + geom_histogram() +
      facet_grid(is_low ~ is_bait) +
      opts(title="QD Distribution") +
      xlab("Variant Confidence by Depth")
```
