<!--

Analysis of pollutant vs. influenza data for the Kobzik lab
-->
```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, echo=FALSE, highlight=TRUE, fig.align='left', fig.show='hold', cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE,
fig.cap='')
```
# Overview
Severity of influenza infection has been shown to be correlated, in part, to
ambient air pollution (citation needed for this). This experiment used ROFA as stand-in for
air pollutants to attempt to determine if there are separate pathways for cell death
turns on during influenza infection with prior ROFA exposure as compared to
plain influenza infection. The data is from a shRNA experiment performed from a single
batch of cells in culture, with four treatment conditions of the cells done in parallel:
control (no insult), ROFA (ROFA application to the cells), virus (influenza virus
application to the cells) and virus + ROFA (preincubation with ROFA followed by
incubation with influenza). The primary aim of this analysis is to determine if there
are genes involved in the ROFA + virus response that are diffrent than the response to
the virus alone. A secondary aim is to determine which genes are involved in the viral
response compared to the control response alone.
```{r variables-and-libraries}
library(ggplot2)
library(reshape)
library(googleVis)
library(stringr)
library(edgeR)
wd = '/n/home05/kirchner/cache/projects/kobzik_influenza/scripts/'
setwd(wd)
in_file = paste(wd, '../data/ROFAFluScreenKobzikFeb222013.csv', sep='')
load(paste(wd, '../data/pathways.Hs.RData', sep=''))
FDR_CUTOFF = 0.10
ensembl_gene = 'hsapiens_gene_ensembl'
gene_symbol = 'hgnc_symbol'
filter_type = 'refseq_mrna'
results_dir = paste(wd, "../results/", sep='')
```

```{r data-load, cache=TRUE}
counts = read.table(in_file, header=TRUE, sep=",")
```
We will use the order column as the ID column to later merge the
tables together, since those values are unique.

```{r id-column, cache=TRUE}
control_cols = c("control_rep1", "control_rep2", "control_rep3", "control_rep4")
ROFA_cols = c("ROFA_rep1", "ROFA_rep2", "ROFA_rep3", "ROFA_rep4")
virus_cols = c("virus_rep1", "virus_rep2", "virus_rep3", "virus_rep4")
ROFA_virus_cols = c("ROFA_virus_rep1", "ROFA_virus_rep2", "ROFA_virus_rep3", "ROFA_virus_rep4")
small_data = subset(counts, select=c(control_cols, ROFA_cols, virus_cols, ROFA_virus_cols, "order", "RefSeq", "GeneSymbol"))
melted = melt(small_data, id=17:19)
melted$treatment = sapply(melted$variable, function(x) str_split(x, "_rep", n=2)[[1]][1])
names(melted) = c("order", "refseq", "symbol", "replicate", "count", "treatment")
melted$treatment = factor(melted$treatment)
melted$replicate = factor(melted$replicate)
```

## Exploratory data analysis
Counts on the whole are non-normal, with the mode of the counts around a thousand
units. This is different than normal expression data where you would expected the
majority of counts to be zero.

```{r prefiltering-exploratory-analysis}
 ggplot(melted, aes(x=count)) + geom_histogram()
 ```

shRNA counts have very similar distributions for each of the treatment conditions:

```{r logged-counts}
 ggplot(melted, aes(x=count)) + geom_histogram() + scale_x_log10() + facet_grid(. ~ treatment)
```
```{r raw-boxplot-by-treatment}
 ggplot(melted, aes(y=count, x=treatment)) + geom_boxplot()
 ```

In particular the counts for the control and ROFA alone are highly correlated.

```{r pairwise-plots-of-average}
casted = cast(melted, refseq ~ treatment, value="count", mean)
plotmatrix(casted[2:5])
```

Looking individually at each sample in a treatment conditon, the
distribution of shRNA counts is also comparable:

```{r raw-boxplot-to-examine-replicates}
 ggplot(melted, aes(y=count, x=replicate)) + geom_boxplot() + facet_grid(~treatment,
	 drop=TRUE, space="free", scales="free") + opts(axis.text.x=theme_text(angle=-90))
```
```{r pairwise-log-log-plots-for-replicates}
pairwise_plot_of_count = function(df, condition) {
	sub = subset(df, treatment == condition)
	sub$log_count = log(sub$count)
	casted = cast(sub, refseq ~ replicate, value="log_count", mean)
	plotmatrix(casted[2:5])
}
pairwise_plot_of_count(melted, "control")
pairwise_plot_of_count(melted, "ROFA")
pairwise_plot_of_count(melted, "virus")
pairwise_plot_of_count(melted, "ROFA_virus")
```

Each treatment group has higher within-group correlation, with the control and
ROFA only groups having the highest correlation with each other, with ROFA + virus
having the lowest correlation with the two control groups.
```{r high-correlation2}
 cor.m = melt(cor(small_data[1:16]))
 ggplot(cor.m, aes(X1, X2)) + geom_tile(aes(fill=value)) + opts(axis.text.x=theme_text(angle=-90))
```
Some shRNA libraries are designed having a variable number of shRNA for each gene.
This library does not, almost all have 5-6 shRNA designed per refseq ID.
```{r multiple-hits, results='asis'}
multihit_table = table(small_data$RefSeq)
print(gvisTable(data.frame(table(multihit_table))), "chart")
```

# EdgeR preparation
EdgeR was chosen to do this analysis; I wrote the author and he agreed that EdgeR would
be appropriate for this type of analysis, assuming the data looks reasonable. In
order to use EdgeR, the data has to be in a specific form which is shown below:
```{r construct-edger-dataframe, results='asis'}
edger_df = data.frame(order=small_data$order, small_data[1:16])
rownames(edger_df) = small_data$order
print(gvisTable(head(edger_df), options=list(width=900)), "chart")
edger_df$order = NULL
```

```{r load-dge}
group = as.factor(c("control", "control", "control", "control",
	"ROFA", "ROFA", "ROFA", "ROFA",
	"virus", "virus", "virus", "virus",
	"ROFA_virus", "ROFA_virus", "ROFA_virus", "ROFA_virus"))
virus = factor(c("no", "no", "no", "no", "no", "no", "no", "no",
	"yes", "yes", "yes", "yes", "yes", "yes", "yes", "yes"))
rofa = factor(c("no", "no", "no", "no", "yes", "yes", "yes", "yes",
	"no", "no", "no", "no", "yes", "yes", "yes", "yes"))

dge = DGEList(edger_df, group=group)
```
After loading the data into EdgeR, we can look at some different aspects of the data
before deciding what model to fit to the data.

Looking at the first and second principal components of the data we can see a clear clustering
with the first principal component separated by experimental condition and the second
principal component by replicate. From the description of the experiment it doesn't
seem like there should be a systematic bias with replicate; the experiments were done in
a single batch and while the libraries were sequenced in batches, the batches don't
correlate to replicate.

The second component is relateively small compared to the first component though, so it
might not be necessary to correct for it. Since it was not part of the experimental design
we will not correct for it now and instead correct later if it turns out to explain
a good portion of the variance.

```{r rolled-own-mds-plot}
rolled_mds = function(counts) {
	distances = dist(counts)
	fit = cmdscale(distances, eig=TRUE, k=6)
	colnames(fit$points) = c("one", "two", "three", "four", "five", "six")
	df = as.data.frame(fit$points)
	df$label = rownames(df)
	df$treatment = sapply(df$label, function(x) str_split(x, "_rep", n=2)[[1]][1])
	df$replicate = sapply(df$label, function(x) str_split(x, "_rep", n=2)[[1]][2])
	print(ggplot(df, aes(one, two, color=treatment, shape=replicate)) + geom_point())
	print(ggplot(df, aes(three, four, color=treatment, shape=replicate)) + geom_point())
	print(ggplot(df, aes(five, six, color=treatment, shape=replicate)) + geom_point())
	eigs = data.frame(variance_explained=fit$eig / sum(fit$eig))
	eigs$component = as.integer(rownames(eigs))
	print(ggplot(eigs, aes(component, variance_explained)) + geom_point())
	return(df)
}
mds_data = rolled_mds(t(dge$counts))
```
Libraries tend to have the same number of shRNA counts.
```{r simple-design, results='asis'}
dge = calcNormFactors(dge)
print(gvisTable(dge$samples, options=list(width=900)), "chart")
```

## Exploratory model fitting

We fit a model of this form to the data:

```{r model}
design = model.matrix(~0+group)
design
```

Common dispersion estimation with biological coefficient of variation.

```{r common-dispersion-estimation}
dge = estimateCommonDisp(dge, verbose=TRUE)
```

Trended dispersion estimation.

```{r trended-dispersion-estimation}
dge = estimateTagwiseDisp(dge)
```

For low count data there is quite a bit of biological variation but this drops as
the tag count increases. We should treat any differentially expressed tags that have
low CPM as suspect. The overall BCV is low, but it is in the range of what you expect
for cell lines.

```{r bcv-plot}
plotBCV(dge)
```

We fit the model to the data and it fits reasonably well, with some deviation at the
higher quantiles.
```{r model-fitting}
fit = glmFit(dge, design)
g = gof(fit)
z = zscoreGamma(g$gof.statistics,shape=g$df/2,scale=2)
qqnorm(z)
```

Fitting the model results in separating out the treatments based on their
first and second principal components instead of just the first. The cumulative
effect of treatment swamps out the replicate correlation we were seeing earlier,
the first two components explain over 90% of the data, supporting our decision
to not correct for the minor replicate effect.
```{r model-fit-effect-on-MDS}
mds_data = rolled_mds(t(fit$fitted.values))
```

## Differential expression
```{r differential-test-function}
differential_test = function(dge, pair) {
	et = exactTest(dge, pair=pair)
	et_table = et$table
	et_table$DE = as.numeric(decideTestsDGE(et, p.value=FDR_CUTOFF))
	et_table$FDR = p.adjust(et_table$PValue, method="BH")
	return(et_table)
}
```
```{r tabulate-multiple-hits-function}
multiple_hits_table = function(df, field="id") {
	table(table(df$field))
}
```
```{r annotate-edgeR-de-df-function}
annotate_edgeR_de_df = function(df) {
	m = merge(df, counts[, c("order", "RefSeq")], by.x=0, by.y="order")
	colnames(m) = c("order", "logFC", "logCPM", "pvalue", "DE", "FDR", "id")
	m = annotate_df(m)
	return(m)
}
```

```{r diffexp-functions}
write_diffexp_table = function(df, name, direction, results_dir="") {
	out_file = paste(results_dir, name, "_", direction, ".txt", sep="")
	write.table(df, out_file, quote=FALSE, row.names=FALSE, sep="\t")
}
process_downregulated = function(dge, pair, cname) {
	et_table = differential_test(dge, pair)
	down = subset(et_table, DE == -1, drop=TRUE)
	if(dim(down)[1] != 0) {
		print(sprintf("Total number of %s shRNA in %s: %d", "downregulated", cname, dim(down)[1]))
		annotated = annotate_edgeR_de_df(down)
		return(annotated)
	}
	else {
		print(sprintf("There are 0 dowregulated shRNA for %s", cname))
		return(data.frame())
	}
}
process_upregulated = function(dge, pair, cname) {
	et_table = differential_test(dge, pair)
	up = subset(et_table, DE == 1, drop=TRUE)
	if(dim(up)[1] != 0) {
		print(sprintf("Total number of %s shRNA in %s: %d", "upregulated", cname, dim(up)[1]))
		annotated = annotate_edgeR_de_df(up)
		return(annotated)
	}
	else {
		print(sprintf("There are 0 upregulated shRNA for %s", cname))
		return(data.frame())
	}
}
annotate_df = function(d) {
	require(biomaRt)
	ensembl = useMart('ensembl', dataset = ensembl_gene)
	a = getBM(attributes=c(filter_type, gene_symbol, "description",
		"entrezgene"),
		filters=c(filter_type), values=d[, 'id'],
		mart=ensembl)
	m = merge(d, a, by.x='id', by.y=filter_type)
	return(m)
}
annotate_df_with_uniprot = function(d) {
	require(biomaRt)
	ensembl = useMart('ensembl', dataset = ensembl_gene)
	a = getBM(attributes=c(filter_type, "uniprot_swissprot_accession"),
		filters=c(filter_type), values=d[, 'id'],
		mart=ensembl)
	m = merge(d, a, by.x='id', by.y=filter_type)
	return(m)
}
wrap_gvistable = function(table) {
	to_display = c("id", "hgnc_symbol", "description")
	if(dim(table)[1] > 0) {
		print(gvisTable(table[, to_display], options=list(width=900)), "chart")
	}
}
subset_multiple_hits = function(df, field="id") {
	multiple_hits_table(df, field)
	multihit_table = table(df[, field])
	multihit_ids = names(multihit_table[multihit_table > 1])
	return(droplevels(df[df$id %in% multihit_ids,]))
}
process_multiple_hits = function(df, name, direction) {
	multihits = subset_multiple_hits(df)
	#multihits = annotate_df_with_uniprot(multihits)
	write_diffexp_table(multihits, name, direction, results_dir)
	return(multihits)
}
```
```{r process-regulation-function}
process_regulation = function(df, pair, name) {
	down = process_downregulated(df, pair, name)
	down = annotate_df_with_uniprot(down)
	write_diffexp_table(down, name, "down", results_dir)
	wrap_gvistable(down)
	multi_down = process_multiple_hits(down, name, "multihits-down")
	up = process_upregulated(df, pair, name)
	up = annotate_df_with_uniprot(up)
	write_diffexp_table(up, name, "up", results_dir)
	wrap_gvistable(up)
	multi_up = process_multiple_hits(up, name, "multihits-up")
	return(rbind(down, up))
}
tableCat <- function(inFrame) {
    outText <- paste(names(inFrame), collapse = " | ")
    outText <- c(outText, paste(rep("---", ncol(inFrame)), collapse = " | "))
    invisible(apply(inFrame, 1, function(inRow) {
        outText <<- c(outText, paste(inRow, collapse = " | "))
    }))
    return(outText)
}
```
```{r get-counts-of-de}
get_counts_from_de = function(df, counts) {
	de_counts = subset(counts, order %in% df$order)
	return(de_counts)
}
```
```{r block-of-go-helper-functions}
```
```{r setup_chunk, echo=FALSE}
opts_chunk$set(tidy=TRUE, echo=FALSE, highlight=TRUE, figalign="center", fig.height=6, fig.width=6, cache=FALSE, highlight=TRUE, autodep=TRUE)
```
### ROFA vs. control
There are no ROFA vs. control genes that pass the FDR cutoff; this is expected from
the preliminary data in the grant.
```{r results='asis'}
down_and_up = process_regulation(dge, c(1, 2), "rofa-vs-control")
```
### Virus vs. control
```{r virus-vs-control, results='asis'}
down_and_up = process_regulation(dge, c(1, 4), "virus-vs-control")
```
### ROFA + virus vs virus
```{r results='asis'}
down_and_up = process_regulation(dge, c(4, 3), "ROFA+virus-vs-virus")
```
## Multiple hits
It is SOP for shRNA experiments to restrict the genes looked at by only considering
genes with more than one shRNA hit called differentially expressed. I subsetted
the above tables for genes that had multiple hits.

### Virus vs. control
```{r results='asis'}
df = read.csv('virus-vs-control_multihits-down.txt', header=TRUE, sep="\t")
print(gvisTable(df), "chart")
df = read.csv('virus-vs-control_multihits-up.txt', header=TRUE, sep="\t")
print(gvisTable(df), "chart")
```
### ROFA + virus vs virus
```{r results='asis'}
df = read.csv('ROFA+virus-vs-virus_multihits-down.txt', header=TRUE, sep="\t")
print(gvisTable(df), "chart")
df = read.csv('ROFA+virus-vs-virus_multihits-up.txt', header=TRUE, sep="\t")
print(gvisTable(df), "chart")
```

## Drug targets
Using http://targetmine.nibio.go.jp/ I annotated the hits with compounds known to
interact with the multiple-hit genes. I only kept compounds that target more than one of
the multiply hit genes.

### Virus vs. control
Only the downregulated genes had hits.
```{r results='asis'}
df = read.csv('../results/drugs/virus_vs_control_down_multidrug_hits.tsv',
	header=TRUE, sep="\t")
print(gvisTable(df), "chart")
```
### ROFA + virus vs virus
```{r results='asis'}
df = read.csv('../results/drugs/rofa_and_virus_vs_virus_down_multidrug_hits.tsv',
	header=TRUE, sep="\t")
print(gvisTable(df), "chart")
df = read.csv('../results/drugs/rofa_and_virus_vs_virus_up_multidrug_hits.tsv',
	header=TRUE, sep="\t")
print(gvisTable(df), "chart")
```

## Possible improvements
I tried fitting a GLM to the data and doing a differential expression on that; the
idea was to fit a model of the form: counts ~ rofa + virus + rofa:virus, and then
query what is differentially expressed considering the interaction between rofa and
virus. This gave around ten hits, but none of them had double hits. I left the code
for that in the addendum in case it becomes useful.

It also might be useful to include the single gene hits, but combine them together
somehow; for example if there are many single hits with a common pathway, that might
be more indicative of a real functional hit than if there are more than one shRNA that
target an individual gene in a pathway.


# Extra pathway analysis
```{r pathway-enrichment}
pathEnrich = function (Genelist, geneset=pathways.Hs, Nchip=28546)
{
    genelist = unique(Genelist[!is.na(Genelist)])
	Nsig <- length(genelist)
    hyper <- as.data.frame(matrix(nrow = length(geneset), ncol = 1))
    colnames(hyper) <- c("p-value")
	hyper[,1] = as.numeric(lapply(geneset,function(x)
		{
			if(length(intersect(genelist,x))<1) return(1)
			else return(sum(dhyper(length(intersect(genelist,x)):Nsig,length(x),Nchip - length(x), Nsig)))
		}))
    hyper[,2] <- p.adjust(hyper[, 1], method = "BH")
	overlap = lapply(geneset,function(x)
		{
			return(as.list(intersect(genelist,x)))
		})
    hyper[,3] = as.numeric(lapply(overlap,function(x) return(length(x))))
    hyper[,4] = as.numeric(lapply(geneset,function(x) return(length(x))))
    hyper[,5] <- names(geneset)
	genes = lapply(overlap, function(x) return(as.numeric(x)))
	hyper$genes = I(genes)
    colnames(hyper) <- c("p.value","FDR", "nGenes","nPathway","Name", "Genes")
    hyper = hyper[with(hyper, order(FDR)),]
    return(hyper)
	}
```

### Virus vs. control
```{r results='asis'}
to_display = c("Name", "p.value", "FDR", "nGenes", "nPathway")
down_and_up = process_regulation(dge, c(1, 4), "virus-vs-control")
pathwayde = pathEnrich(down_and_up$entrezgene, Nchip=length(unique(counts$RefSeq)))
print(gvisTable(pathwayde[,to_display]), "chart")
out_file = paste(results_dir, "virus_vs_control_pathways.txt", sep="")
write.table(pathwayde, out_file, quote=FALSE, row.names=FALSE, sep="\t")
#cat(tableCat(pathwayde[1:10,]), sep = "\n")
#print(xtable(pathwayde[1:10,]))
```


### ROFA + virus vs virus
```{r results='asis'}
to_display = c("Name", "p.value", "FDR", "nGenes", "nPathway")
down_and_up = process_regulation(dge, c(4, 3), "ROFA+virus-vs-virus")
pathwayde = pathEnrich(down_and_up$entrezgene, Nchip=length(unique(counts$RefSeq)))
print(gvisTable(pathwayde[,to_display]), "chart")
out_file = paste(results_dir, "rofa_and_virus_vs_virus_pathways.txt", sep="")
write.table(pathwayde, out_file, quote=FALSE, row.names=FALSE, sep="\t")
#cat(tableCat(pathwayde[1:10,]), sep = "\n")
#print(xtable(pathwayde[1:10,]))
```

## Wrapup
```{r wrapup}
```

## Citations
The pathway analysis code was snagged from Emmanuel Dimont (edimont2@gmail.com).

# Addendum
What if we fit a slightly different model to the data?

```{r glm-test}
melted$virus = (melted$treatment == "virus") | (melted$treatment == "ROFA_virus")
melted$ROFA = (melted$treatment == "ROFA") | (melted$treatment == "ROFA_virus")
design_df = data.frame(group=group, rofa, virus)
design = model.matrix(~rofa + virus)
design
dge <- estimateGLMCommonDisp(dge, design, verbose=TRUE)
dge <- estimateGLMTrendedDisp(dge, design)
dge <- estimateGLMTagwiseDisp(dge, design)
fit <- glmFit(dge, design)
lrt <- glmLRT(fit)
de_table = lrt$table
de_table$DE = decideTests(de_table)
de_table$FDR = p.adjust(de_table$PValue, method="BH")
x = annotate_edgeR_de_df(de_table)
```
