---
output:
  html_document:
    code_folding: hide
    theme: yeti
    toc: yes
    toc_float: true
    toc_depth: 6
    number_sections: true
    fig_height: 6 
title: "Church - Cell Anucleation RNAseq Analysis"
bibliography: references.bib
author: "John N Hutchinson"
---

[R]: https://www.r-project.org
[RStudio]: https://www.rstudio.com
[`bcbio-rnaseq`]: https://github.com/roryk/bcbio.rnaseq
[`bcbio]: https://github.com/chapmanb/bcbio-nextgen
[`DESeq2`]: https://bioconductor.org/packages/release/bioc/html/DESeq2.html
[`salmon`]: https://combine-lab.github.io/salmon/
[`tximport`]: https://bioconductor.org/packages/release/bioc/html/tximport.html
[Bioconductor]: https://bioconductor.org
[Ensembl]: http://useast.ensembl.org/Mus_musculus/Info/Index



```{r setup1, echo=FALSE}
# Setup report details
clientname="Stan Wang"
clientemail="Stan_Wang@hms.harvard.edu"
lablocation="HMS" 
analystname="John Hutchinson"
analystemail="jhutchin@hsph.harvard.edu"
```

> RNA-Seq analysis for `r clientname` (`r clientemail`) at `r lablocation`. Contact `r analystname` (`r analystemail`) for additional details. 
> This document was generated on `r date()` using [R][] and [RStudio][]
 
```{r setup2, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",
               cache=TRUE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='')
```

# Overview

Fastq files were processed in [bcbio][] and pseudo-count data generated by [salmon][]. Various RNAseq metrics were summarized with [bcbio-rnaseq][].

> The project has 8 samples, 4 controls and 4 cell specific Kcnd1 knockouts.

Here I do some basic QC to see if the data looks good and if whether any samples look like they have issues.

```{r check-library-functions, echo=FALSE}
check_install = function(packages) {
   not_installed = setdiff(packages, rownames(installed.packages()))
   if(length(not_installed) > 0) {
      write(paste("The libraries", not_installed, "are not installed, aborting.",
                   sep=" "), stderr())
      stop()
   }
}
```

```{r load-libraries, echo=FALSE}
packages = c("ggplot2", "reshape", "gplots", "edgeR", "CHBUtils", "pheatmap",
              "DESeq2", "tximport", "DT", "DEGreport", "dplyr")
check_install(packages)
installed = lapply(packages, library, character.only = TRUE)
```

```{r qc-setup, results='hide'}
project_summary = "/n/data1/cores/bcbio/PIs/george_church/church_anucleated_rnaseq/bcbio/final/2017-04-06_bcbio/project-summary.csv"
counts_file = "/n/data1/cores/bcbio/PIs/george_church/church_anucleated_rnaseq/bcbio/final/2017-04-06_bcbio/combined.counts"
tx2genes_file = "/n/data1/cores/bcbio/PIs/george_church/church_anucleated_rnaseq/bcbio/final/2017-04-06_bcbio/tx2gene.csv"
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442",
"#0072B2", "#D55E00", "#CC79A7")
summarydata = read.table(project_summary, header=TRUE, sep=",")
summarydata = summarydata[,colSums(is.na(summarydata)) < nrow(summarydata)]
# handle newer bcbio-nextgen runs that use description as the key
if("description" %in% colnames(summarydata)) {
  rownames(summarydata) = summarydata$description
  summarydata$Name = rownames(summarydata)
  summarydata$description = NULL
} else {
  rownames(summarydata) = summarydata$Name
  # summarydata$Name = NULL
}
summarydata = summarydata[order(rownames(summarydata)),]
colnames(summarydata) = gsub(".", "_", colnames(summarydata), fixed=TRUE)
if (file.exists(tx2genes_file)) {
  sf_files = list.files("../../", pattern="quant.sf", recursive=TRUE, full.names=TRUE)
  if(length(sf_files) > 0) {
    sf_files = sf_files[order(sf_files)]
    names(sf_files) = rownames(summarydata)
    tx2gene = read.table(tx2genes_file, sep=",", row.names=NULL, header=FALSE)
    txi.salmon = tximport(sf_files, type="salmon", tx2gene=tx2gene,
             reader=readr::read_tsv, countsFromAbundance="lengthScaledTPM")
    counts = round(data.frame(txi.salmon$counts, check.names=FALSE))
} else {
  counts = read.table(counts_file, header=TRUE, row.names="id", check.names=FALSE)
}}
counts = counts[, order(colnames(counts)), drop=FALSE]
colnames(counts) = gsub(".counts", "", colnames(counts))

# this is a list of all non user-supplied metadata columns that could appear
known_columns = c("Name", "X_GC", "Exonic_Rate", "Sequences_flagged_as_poor_quality",
    "rRNA_rate", "Fragment_Length_Mean", "Intronic_Rate", "Intergenic_Rate",
    "Mapping_Rate", "Quality_format", "Duplication_Rate_of_Mapped", "Mapped",
    "rRNA", "Sequence_length", "Transcripts_Detected", "Mean_Per_Base_Cov_",
    "Genes_Detected", "Unique_Starts_Per_Read", "unique_starts_per_read",
    "complexity", "X5_3_bias", "Duplicates_pct", "Duplicates", "Mapped_reads",
    "Average_insert_size", "Mapped_reads_pct", "Total_reads",
    "avg_coverage_per_region", "Mapped_Reads")
summarydata[,"Fragment_Length_Mean"] = summarydata$Average_insert_size
metadata = summarydata[, !colnames(summarydata) %in% known_columns, drop=FALSE]
metadata = metadata[, colSums(is.na(metadata)) < nrow(metadata), drop=FALSE]
```

```{r sanitize-datatable}
sanitize_datatable = function(df, ...) {
 # remove dashes which cause wrapping
 DT::datatable(df, ..., rownames=gsub("-", "_", rownames(df)),
                   colnames=gsub("-", "_", colnames(df)))
}
```

```{r set-random-seed}
# set seed for reproducibility
set.seed(1454944673)
```

## Sample metadata
```{r summary-table, echo=FALSE, cache=FALSE}
sanitize_datatable(summarydata, style='bootstrap')
```

```{r heatmap-function}
get_heatmap_fn = function(summarydata) {
    # return the pheatmap function with or without metadata
    if(ncol(metadata) == 0) {
       return(pheatmap)
    }
    else {
    # rownames(metadata) = summarydata$Name
    heatmap_fn = function(data, ...) {
        pheatmap(data, annotation=metadata, clustering_method = "ward.D2",
        clustering_distance_cols = "correlation", ...)
    }
    return(heatmap_fn)
}}
heatmap_fn = get_heatmap_fn(summarydata)
```

# Quality control metrics

```{r qualimap-run}
qualimap_run = "Exonic_Rate" %in% colnames(summarydata)
do_quality = "Total_reads" %in% colnames(summarydata)
```

## Mapped reads
```{r mapped-plot, eval=do_quality}
ggplot(summarydata, aes(x=Name, y=Mapped)) +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    geom_bar(stat="identity") +
    ylab("mapped reads") + xlab("")
```

## Genomic mapping rate
```{r mapping-rate-plot, eval=do_quality}
mapping_string = ifelse("Mapped_Reads" %in% colnames(summarydata),
                        "Mapped_Reads", "Mapped_reads_pct")
ggplot(summarydata, aes_string(x="Name", y=mapping_string)) +
    geom_bar(stat="identity") +
    ylab("mapping rate") + xlab("") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90))
```

## Number of genes detected
```{r genes-detected-plot}
dd = data.frame(Name=colnames(counts), Genes_Detected = colSums(counts > 0))
ggplot(dd, aes(x=Name, y=Genes_Detected)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("")
```

## Gene detection saturation
```{r saturation-plot, eval=do_quality}
col_mapped = ifelse(qualimap_run, "Mapped", "Mapped_reads")
dd = data.frame(Mapped=summarydata[,col_mapped], Genes_Detected = colSums(counts > 0))
ggplot(dd, aes(x=Mapped, y=Genes_Detected)) +
    geom_point() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("reads mapped")
```

## Exonic mapping rate
```{r exonic-mapping-plot, eval="Exonic_Rate" %in% colnames(summarydata)}
ggplot(summarydata, aes(x=Name, y=Exonic_Rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("exonic mapping rate") + xlab("")
```

## rRNA mapping rate
```{r check-rRNA-eval}
eval_rRNA = "rRNA_rate" %in% colnames(summarydata) & !sum(is.na(summarydata$rRNA_rate)) == nrow(summarydata)
```
```{r rRNA-rate-plot, eval=eval_rRNA & do_quality}
ggplot(summarydata, aes(x=Name, y=rRNA_rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("rRNA rate") + xlab("")
```

## Estimated fragment length of paired-end reads
```{r fragment-length-plot, eval="Fragment_Length_Mean" %in% colnames(summarydata) & do_quality}
ggplot(summarydata, aes(x=Name, y=Fragment_Length_Mean)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("fragment length") + xlab("")
```

## 5'->3' bias
```{r 53bias, eval="X5_3_bias" %in% colnames(summarydata) & do_quality}
ggplot(summarydata, aes(x=Name, y=X5_3_bias)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("5'->3' bias") + xlab("")
```

## Boxplot of log10 counts per gene
```{r boxplot-raw}
melted = melt(counts)
colnames(melted) = c("sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Boxplot of log10 TMM-normalized counts per gene
Trimmed mean of M-values (TMM) normalization is described
[here](http://genomebiology.biomedcentral.com/articles/10.1186/gb-2010-11-3-r25)

Robinson, M. D., & Oshlack, A. (2010). A scaling normalization method for differential expression analysis of RNA-seq data. Genome Biology, 11(3). doi:10.1186/gb-2010-11-3-r25

```{r boxplot-normalized}
y = DGEList(counts=counts)
y = calcNormFactors(y)
normalized_counts = cpm(y, normalized.lib.sizes=TRUE)
melted = melt(normalized_counts)
colnames(melted) = c("gene", "sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Density of log10 TMM-normalized counts
```{r density-normalized}
ggplot(melted, aes(x=count, group=sample)) +
    geom_density() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Correlation heatmap of TMM-normalized counts {.tabset}

### Correlation (Pearson)
```{r pearson-heatmap-normalized, eval=ncol(counts) > 1}
heatmap_fn(cor(normalized_counts, method="pearson"))
```

### Correlation (Spearman)
```{r spearman-heatmap-normalized, eval=ncol(counts) > 1}
heatmap_fn(cor(normalized_counts, method="spearman"))
```

## PCA plots {.tabset}
```{r pca, eval=ncol(counts) > 1}
dds = DESeqDataSetFromMatrix(countData=counts, colData=summarydata, design=~Name)
vst = varianceStabilizingTransformation(dds)
pca_loadings = function(object, ntop=500) {
  rv <- matrixStats::rowVars(assay(object))
  select <- order(rv, decreasing = TRUE)[seq_len(min(ntop,
      length(rv)))]
  pca <- prcomp(t(assay(object)[select, ]))
  percentVar <- pca$sdev^2/sum(pca$sdev^2)
  names(percentVar) = colnames(pca$x)
  pca$percentVar = percentVar
  return(pca)}
pc = pca_loadings(vst)
comps = data.frame(pc$x)
comps$Name = rownames(comps)
library(dplyr)
comps = comps %>% left_join(summarydata, by=c("Name"="Name"))
colorby = "Name"
```

```{r pca-plotter}
pca_plot = function(comps, nc1, nc2, colorby) {
   c1str = paste0("PC", nc1)
   c2str = paste0("PC", nc2)
  if(!(c1str %in% colnames(comps) && c2str %in% colnames(comps))) {
    warning("Higher order components not found, skipping plotting.")
    return(NA)
  }
  ggplot(comps, aes_string(c1str, c2str, color=colorby)) +
    geom_point() + theme_bw() +
    xlab(paste0(c1str, ": ", round(pc$percentVar[nc1] * 100), "% variance")) +
    ylab(paste0(c2str, ": ", round(pc$percentVar[nc2] * 100), "% variance"))
  }
```

### PC1 vs. PC2
```{r PC1-PC2}
pca_plot(comps, 1, 2, colorby)
```

### PC3 vs. PC4
```{r PC3-PC4}
pca_plot(comps, 3, 4, colorby)
```

### PC5 vs. PC6
```{r PC5-PC6}
pca_plot(comps, 5, 6, colorby)
```

### Variance explained by component
```{r component-loadings}
ggplot(data.frame(component=reorder(names(pc$percentVar), -pc$percentVar),
                  percent_var=pc$percentVar),
       aes(component, percent_var)) + geom_bar(stat='identity') +
  ylab("percent of total variation") + xlab("") + theme_bw()
```

```{r shiny-app, echo=F}
# this is to plot data with https://github.com/lpantano/shiny/tree/master/expression
degObj(assay(vst), metadata, "deg.rda")
```

# TPM matrix
```{r write-tpm-matrix}
library(readr)
tpm = txi.salmon$abundance %>%
  as.data.frame() %>%
  tibble::rownames_to_column()
write_csv(tpm, "tpm.csv")
``````{r deseq2-tximport-helper-functions}
# snagged from development version of DESeq
DESeqDataSetFromTximport <- function(txi, colData, design, ...)
{
  counts <- round(txi$counts)
  mode(counts) <- "integer"
  dds <- DESeqDataSetFromMatrix(countData=counts, colData=colData, design=design, ...)
  stopifnot(txi$countsFromAbundance %in% c("no","scaledTPM","lengthScaledTPM"))
  if (txi$countsFromAbundance %in% c("scaledTPM","lengthScaledTPM")) {
    message("using length scaled TPM counts from tximport")
  } else {
    message("using counts and average transcript lengths from tximport")
    lengths <- txi$length
    dimnames(lengths) <- dimnames(dds)
    assays(dds)[["avgTxLength"]] <- lengths
  }
  return(dds)
}

subset_tximport = function(txi, rows, columns) {
    txi$counts = txi$counts[rows, columns]
    txi$abundance = txi$abundance[rows, columns]
    txi$length = txi$length[rows, columns]
    return(txi)
}
```

```{r deseq2-helper-functions}
deseq2resids = function(dds) {
  # calculate residuals for a deseq2 fit
  fitted = t(t(assays(dds)[["mu"]]) / sizeFactors(dds))
  return(counts(dds, normalized=TRUE) - fitted)
}
```

```{r de-setup}
library(DEGreport)
library(vsn)
design = ~treatment
condition = "treatment"
```

# Differential expression

```{r deseq2-expression-analysis, results='asis'}
counts <- counts[rowSums(counts>0)>1,]
if(exists("txi.salmon")) {
   txi.salmon = subset_tximport(txi.salmon, rownames(counts), colnames(counts))
   dds = DESeqDataSetFromTximport(txi.salmon, colData=summarydata, design=design)
} else {
  dds = DESeqDataSetFromMatrix(countData=counts,
      colData=summarydata, design = design  )
}
geoMeans = apply(counts, 1, function(row) if (all(row == 0)) 0 else
                 exp(mean(log(row[row != 0]))))
dds = estimateSizeFactors(dds, geoMeans=geoMeans)
dds = DESeq(dds, betaPrior=FALSE)
```

## Effect of variance stabilization

```{r deseq-diagnostics, results='asis'}
par(mfrow=c(1,3))
notAllZero <- (rowSums(counts(dds))>0)
rld <- rlog(dds)
vsd <- varianceStabilizingTransformation(dds)
rlogMat <- assay(rld)
vstMat <- assay(vsd)

meanSdPlot(log2(counts(dds,normalized=TRUE)[notAllZero,] + 1))
meanSdPlot(assay(rld[notAllZero,]))
meanSdPlot(assay(vsd[notAllZero,]))
```

## Dispersion estimates

```{r dispersion-estimate}
plotDispEsts(dds)
```

```{r deseq2-handler}
handle_deseq2 = function(dds, summarydata, column) {
  all_combs = combn(levels(summarydata[,column]), 2, simplify=FALSE)
  all_results = list()
  contrast_strings = list()
  for(comb in all_combs) {
    contrast_string = paste(comb, collapse=" vs ")
    contrast = c(column, comb)
    res = results(dds, contrast=contrast, addMLE=TRUE)
    res = res[order(res$padj),]
    all_results = c(all_results, res)
    contrast_strings = c(contrast_strings, contrast_string)
  }
  names(all_results) = contrast_strings
  return(all_results)
}
```

## MA-plots

```{r nicer-ma-plot}
plotMA = function(res, contrast_name=NULL) {
  res = data.frame(res)
  res = subset(res, !is.na(padj))
  p = ggplot(res, aes(baseMean, log2FoldChange, color=padj < 0.05)) +
    geom_point(size=0.8) +
    scale_x_log10(
      breaks = scales::trans_breaks("log10", function(x) 10^x),
      labels = scales::trans_format("log10", scales::math_format(10^.x))) +
    annotation_logticks(sides='b') +
    xlab("mean expression across all samples") +
    ylab(expression(log[2]*" fold change")) +
    scale_color_manual(values=c("black", "red", "green")) +
    guides(color=FALSE)
  if(!is.null(contrast_name)) {
    p = p +
      ggtitle(paste("MA-plot for contrast ", contrast_name))
  }
  return(p)
}
```

```{r DESeq-output, results='asis'}
all_results = handle_deseq2(dds, summarydata, condition)
len = length(all_results)
nr = ceiling( len / 3 )
nc = ceiling( len / nr )
par(mfrow=c(nr,nc))
for(i in seq(length(all_results))) {
  res = all_results[[i]]
  ymax = max(res$log2FoldChange, na.rm=TRUE)
  ymin = min(res$log2FoldChange, na.rm=TRUE)
  plotMA(all_results[[i]], names(all_results)[i])
}
```

## Volcano-plots

```{r DESeq-volcano}
for(i in seq(length(all_results))) {
  stats = as.data.frame(all_results[[i]][,c(2,6)])
  p = volcano_density_plot(stats, title=names(all_results)[i], lfc.cutoff=1.5)
  print(p)
}
```

## DEGreport

```{r get-groups}
get_groups <- function(d, comp, condition)
{
  g <- unlist(strsplit(comp," "))
  g1 <- d$Name[d[, (names(d)==condition)]==g[1]]
  g2 <- d$Name[d[, (names(d)==condition)]==g[3]]
  list(g1,g2)
}
```

### Pvalues-vs-Mean

Here we plot some information about how the p-values are correlated with the
mean or the standard deviation.

```{r DEGreport-M}
plots = list()
scale_factor = round(1/nr * 14)
for(i in seq(length(all_results))) {
  plots[[i]] = degMean(all_results[[i]]$pvalue, rlogMat) +
  theme_bw(base_size = scale_factor) +
  ggtitle(paste0("Pvalues-vs-Mean for ", names(all_results)[i]))
}
do.call(grid.arrange,plots)
```

### Pvalues-vs-Variation

```{r DEGreport-V}
plots = list()
for(i in seq(length(all_results))) {
  plots[[i]] = degVar(all_results[[i]]$pvalue, rlogMat) +
  theme_bw(base_size = scale_factor) +
  ggtitle(paste0("Pvalues-vs-Variation for ", names(all_results)[i]))
}
do.call(grid.arrange,plots)
```

### Mean-vs-Variation
```{r DEGreport-MV}
plots = list()
for(i in seq(length(all_results))) {
  g <- get_groups(summarydata, names(all_results)[i], condition)
  if(length(g[[1]]) < 2 | length(g[[2]]) < 2) {
     next
   }
   g = c(summarydata[g[[1]], condition], summarydata[g[[2]], condition])
  plots[[i]] = degMV(g, all_results[[i]]$pvalue, counts(dds,normalized=TRUE)) +
  theme_bw(base_size = scale_factor) +
  ggtitle(paste0("Mean-vs-Variation for ", names(all_results)[i]))
}
if(length(plots) > 0) {
    do.call(grid.arrange,plots)
}
```

## Differentially expressed genes

```{r DESeq-tables, cache=FALSE, echo=FALSE}
for(i in seq(length(all_results))) {
  cat(paste("Lowest adjusted p-value hits for", names(all_results)[i]))
  out_df = as.data.frame(all_results[[i]])
  out_df$id = rownames(out_df)
  out_df = out_df[, c("id", colnames(out_df)[colnames(out_df) != "id"])]
  write.table(out_df, file=paste(names(all_results)[i], ".tsv", sep=""),
                         quote=FALSE, sep="\t", row.names=FALSE, col.names=TRUE)
  sig_genes = subset(out_df, padj < 0.05)
  sanitize_datatable(sig_genes, style='bootstrap')
  cat("\n")
}
```
