---
output:
  html_document:
    code_folding: hide
    theme: yeti
    toc: yes
    toc_float: true
    toc_depth: 6
    number_sections: true
    fig_height: 6 
title: "Church - Cell Anucleation RNAseq Analysis"
bibliography: references.bib
author: "John N Hutchinson"
---

[R]: https://www.r-project.org
[RStudio]: https://www.rstudio.com
[bcbio-rnaseq]: https://github.com/roryk/bcbio.rnaseq
[bcbio]: https://github.com/chapmanb/bcbio-nextgen
[DESeq2]: https://bioconductor.org/packages/release/bioc/html/DESeq2.html
[salmon]: https://combine-lab.github.io/salmon/
[tximport]: https://bioconductor.org/packages/release/bioc/html/tximport.html
[Bioconductor]: https://bioconductor.org
[Ensembl]: http://useast.ensembl.org/Mus_musculus/Info/Index


```{r setup1, echo=FALSE}
# Setup report details
clientname="Stan Wang"
clientemail="Stan_Wang@hms.harvard.edu"
lablocation="HMS" 
analystname="John Hutchinson"
analystemail="jhutchin@hsph.harvard.edu"
```

> RNA-Seq analysis for `r clientname` (`r clientemail`) at `r lablocation`. Contact `r analystname` (`r analystemail`) for additional details. 
> This document was generated on `r date()` using [R][] and [RStudio][]
 
```{r setup2, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",
               cache=TRUE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='')
```

# Overview

Fastq files were processed in [bcbio][] and pseudo-count data generated by [salmon][]. Various RNAseq metrics were summarized with [bcbio-rnaseq][].

There are 3 biologic replicates of HEK293 cells for each different treatment:
  
  - **DMSO** - untreated samples, 
  
  - **enucleate** - treated and sorted for lack of a nucleus
  
  - **nuclei** - treated samples sorted for nuclei alone
  
  - **nucleated** - treated samples sorted for cells still containing nuclei 
  
All samples received the lowest dilution of ERCC spike-in mix 1 as the input RNA for library prep was 20ng per sample.

Samples were sequenced over 4 lanes, as oppposed to the [intial QC aanlysis][https://www.dropbox.com/s/pn6yz4lbtrf0iik/20170421-qc_summary.zip?dl=0], here I combine them before proceeding with the QC.

Here I will continue to look at some some QC metrics and unsupervised clustering after merging lanes, dropping failed sample and removing ribosomal RNA.


```{r load-libraries, echo=FALSE}
check_install = function(packages) {
   not_installed = setdiff(packages, rownames(installed.packages()))
   if(length(not_installed) > 0) {
      write(paste("The libraries", not_installed, "are not installed, aborting.",
                   sep=" "), stderr())
      stop()
   }
}
packages = c("ggplot2", "reshape", "gplots", "edgeR", "CHBUtils", "pheatmap","DESeq2", "tximport", "DT", "DEGreport", "dplyr", "janitor", "rio", "rdrop2")
check_install(packages)
installed = lapply(packages, library, character.only = TRUE)
```


```{r qc-setup, results='hide'}
if (file.exists("/Volumes/orch/PIs/george_church/church_anucleated_rnaseq/bcbio/final/2017-04-06_bcbio/")){
  baseDir <- "/Volumes/orch/PIs/george_church/church_anucleated_rnaseq/bcbio/final/2017-04-06_bcbio/"
} else if (file.exists("/n/data1/cores/bcbio/PIs/george_church/church_anucleated_rnaseq/")) {
  baseDir <- "/n/data1/cores/bcbio/PIs/george_church/church_anucleated_rnaseq/bcbio/final/2017-04-06_bcbio/"
} else {
  print("no base directory found")
}

project_summary = file.path(baseDir, "project-summary.csv")
counts_file = file.path(baseDir, "combined.counts")
tx2genes_file = file.path(baseDir, "tx2gene.csv")
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442","#0072B2", "#D55E00", "#CC79A7")
summarydata = read.table(project_summary, header=TRUE, sep=",")
summarydata = summarydata[,colSums(is.na(summarydata)) < nrow(summarydata)]
# handle newer bcbio-nextgen runs that use description as the key
if("description" %in% colnames(summarydata)) {
  rownames(summarydata) = summarydata$description
  summarydata$Name = rownames(summarydata)
  summarydata$description = NULL
} else {
  rownames(summarydata) = summarydata$Name
  # summarydata$Name = NULL
}
summarydata = summarydata[order(rownames(summarydata)),]
summarydata <- clean_names(summarydata)

if (file.exists(tx2genes_file)) {
  sf_files = list.files(file.path(dirname(tx2genes_file), "../../"), pattern="quant.sf", recursive=TRUE, full.names=TRUE)
  if(length(sf_files) > 0) {
    sf_files = sf_files[order(sf_files)]
    names(sf_files) = rownames(summarydata)
    tx2gene = read.table(tx2genes_file, sep=",", row.names=NULL, header=FALSE)
    txi.salmon = tximport(sf_files, type="salmon", tx2gene=tx2gene,
             reader=readr::read_tsv, countsFromAbundance="lengthScaledTPM")
    counts = round(data.frame(txi.salmon$counts, check.names=FALSE))
} else {
  counts = read.table(counts_file, header=TRUE, row.names="id", check.names=FALSE)
}}
counts = counts[, order(colnames(counts)), drop=FALSE]


# this is a list of all non user-supplied metadata columns that could appear
known_columns = c("rrna_rate", "duplication_rate_of_mapped", "mapped_reads", "mapped_paired_reads", "quality_format", "intergenic_rate", "sequence_length", "exonic_rate", "total_reads", "percentgc", "rrna", "duplicates", "name", "sequences_flagged_as_poor_quality", "average_insert_size", "intronic_rate", "x5_3_bias", "fragment_length_mean","x_gc", "mapped_reads_pct")
summarydata[,"mapped_reads_pct"] <- (summarydata$mapped_reads/summarydata$total_reads)*100

metadata = summarydata[, !colnames(summarydata) %in% known_columns, drop=FALSE]
metadata = metadata[, colSums(is.na(metadata)) < nrow(metadata), drop=FALSE]
metadata <- metadata[, c( "treatment", "lane", "replicate")] 
```

# Merge lanes

```{r mergelanes}
counts.t <- t(counts)

counts.t <- cbind(counts.t, summarydata)


counts.t <- counts.t %>% dplyr::select(contains("ENSG"), sampleid)

counts.t <- counts.t %>% group_by(sampleid)

temp <- summarize_each(counts.t, funs(sum))



counts.summed <- counts.t %>% tbl_df() %>% group_by(sampleid) %>% summarize_each(funs(sum))


```

# Drop failed samples

Sample nucleated_3 appears to have failed, as shown in the previous report [here][https://www.dropbox.com/s/pn6yz4lbtrf0iik/20170421-qc_summary.zip?dl=0], so will be dropped.


```{r dropfailures}


```


```{r sanitize-datatable}
sanitize_datatable = function(df, ...) {
 # remove dashes which cause wrapping
 DT::datatable(df, ..., rownames=gsub("-", "_", rownames(df)),
                   colnames=gsub("-", "_", colnames(df)))
}
```

```{r set-random-seed}
# set seed for reproducibility
set.seed(1454944673)
```

## Sample metadata
```{r summary-table, echo=FALSE, cache=FALSE}
sanitize_datatable(summarydata, style='bootstrap')
```

```{r heatmap-function}
get_heatmap_fn = function(summarydata) {
    # return the pheatmap function with or without metadata
    if(ncol(metadata) == 0) {
       return(pheatmap)
    }
    else {
    # rownames(metadata) = summarydata$Name
    heatmap_fn = function(data, ...) {
        pheatmap(data, annotation=metadata, clustering_method = "ward.D2",
        clustering_distance_cols = "correlation", ...)
    }
    return(heatmap_fn)
}}
heatmap_fn = get_heatmap_fn(summarydata)
```


# Quality control metrics

```{r qualimap-run}
qualimap_run = "exonic_rate" %in% colnames(summarydata)
do_quality = "total_reads" %in% colnames(summarydata)
```

## Mapped reads

The number of mapped reads are low (we'd prefer to see >20 million reads mapped), nucleated_3 is below an acceptable range.

```{r mapped-plot, eval=do_quality}
ggplot(summarydata, aes(x=name, y=mapped_reads, fill=treatment)) +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    geom_bar(stat="identity") +
    ylab("mapped reads") + xlab("")
```

## Genomic mapping rate

The genomic mapping rate represents the percentage of reads mapping to the reference genome. Low mapping rates are indicative of sample contamination, poor sequencing quality or other artifacts. For the majority of samples, we are seeing a lower mapping rate than we'd like (i.e. >90%). Once again, nucleated_3 is outside of the acceptable range. 

```{r mapping-rate-plot, eval=do_quality}
mapping_string = "mapped_reads_pct"
ggplot(summarydata, aes_string(x="name", y=mapping_string, fill="treatment")) +
    geom_bar(stat="identity") +
    ylab("mapping rate") + xlab("") +ylim(c(0,100))+
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90))
```

## Number of genes detected

The number of genes detected for each sample is a bit low  for most samples, with over nucliei_3 showing levels that are lower than recommended. As gene detection can strongly affect sample variance, we will keep an eye on this factor to see if the samples cluster by it. 

```{r genes-detected-plot}
dd = data.frame(Name=colnames(counts), genes_detected = colSums(counts > 0), treatment=summarydata$treatment)
ggplot(dd, aes(x=Name, y=genes_detected, fill=treatment)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("")
```

## Gene detection saturation

Gene detection saturation was nearly reached for the majority of the samples, which means that most of the samples were able to detect the majority of expressed genes. This is good to see. There is no clear linear trend here of genes detected increasing with the number of reads mapped ine the samples, which may indicate gene detection saturation. The low gene detection with nuclei_3 appears to suggest that more reads are picking up non-genic nucleic acids, potentially DNA?

```{r saturation-plot, eval=do_quality}
col_mapped =  "mapped_reads"
dd = data.frame(mapped=summarydata[,col_mapped], genes_detected = colSums(counts > 0), treatment=summarydata$treatment, replicate=as.character(summarydata$replicate))
ggplot(dd, aes(x=mapped, y=genes_detected, color=treatment, shape=replicate)) +
    geom_point(size=5) +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("reads mapped")
```

## Exonic mapping rate

The exonic mapping rates are good for the non-nuclear samples (excepting nucleated_3, which looks like a failure), with over 60% of total reads mapping to exons. The lower exonic mapping rates in the nuclei may be due to a higher percentage of non-processed transcripts in those samples.

```{r exonic-mapping-plot, eval="exonic_rate" %in% colnames(summarydata)}
ggplot(summarydata, aes(x=name, y=exonic_rate, fill=treatment)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("exonic mapping rate") + xlab("")
```

## rRNA mapping rate

Antyhing above 10% of reads being rRNA is unacceptable. The particularly high levels in nuclei_3 probably explain why we are picking up fewer genes in that sample.

```{r check-rRNA-eval}
eval_rRNA = "rrna_rate" %in% colnames(summarydata) & !sum(is.na(summarydata$rrna_rate)) == nrow(summarydata)
```
```{r rRNA-rate-plot, eval=eval_rRNA & do_quality}
ggplot(summarydata, aes(x=name, y=rrna_rate, fill=treatment)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("rRNA rate") + xlab("")
```

These values are fine. Anything below 5% is reasonable, the KO_2 sample is relatively higher but in absolute terms, sill quite low.

## 5'->3' bias

There is some bias, but it is fairly consistent among samples (excepting, once again, nucleated_3).

```{r 53bias, eval="x5_3_bias" %in% colnames(summarydata) & do_quality}
ggplot(summarydata, aes(x=name, y=x5_3_bias, fill=treatment)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("5'->3' bias") + xlab("")
```

## Boxplot of log10 counts per gene

Generally, we expect similar count spreads for all genes between samples unless the library sizes or total RNA expression are different. The count spreads have a few outliers for every sample, liely reflecting the presence of rRNA in the samples.

```{r boxplot-raw}

melted = melt(counts)
colnames(melted) = c("sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Boxplot of log10 TMM-normalized counts per gene

Trimmed mean of M-values (TMM) normalization is described 
[here](http://genomebiology.biomedcentral.com/articles/10.1186/gb-2010-11-3-r25) [@Robinson2010]

Robinson, M. D., & Oshlack, A. (2010). A scaling normalization method for differential expression analysis of RNA-seq data. Genome Biology, 11(3). doi:10.1186/gb-2010-11-3-r25


Generally, we expect similar count spreads for all genes between samples unless the total expressed RNA per sample is different. With the exception of nucleated_3 and nuclei_3, the samples look good, with similar log<sub>10</sub> DESeq2-normalized counts per gene densities:


```{r boxplot-normalized}
y = DGEList(counts=counts)
y = calcNormFactors(y)
normalized_counts = cpm(y, normalized.lib.sizes=TRUE)
melted = melt(normalized_counts)
colnames(melted) = c("gene", "sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Density of normalized counts

Nucleated_3, nuclei_3 and maybe nuclei_1 look like outliers here.

```{r density-normalized}

melted.annot <- merge(melted, summarydata, by.x="sample", by.y="name")
ggplot(melted.annot, aes(x=count, group=sample, color=sampleid)) +
    geom_density() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

# Sample Similarity Analyses (Unsupervised clustering)

## Correlation heatmap of TMM-normalized counts {.tabset}


### Correlation (Spearman)

Perhaps unsurprisingly, nucleated_3 is a complete outlier.

```{r spearman-heatmap-normalized, eval=ncol(counts) > 1}
heatmap_fn(cor(normalized_counts, method="spearman"))
```

## PCA plots {.tabset}

Sample nucleated_3 is a complete outlier again. 

```{r pca, eval=ncol(counts) > 1}
dds = DESeqDataSetFromMatrix(countData=counts, colData=summarydata, design=~name)
vst = varianceStabilizingTransformation(dds)
pca_loadings = function(object, ntop=500) {
  rv <- matrixStats::rowVars(assay(object))
  select <- order(rv, decreasing = TRUE)[seq_len(min(ntop,
      length(rv)))]
  pca <- prcomp(t(assay(object)[select, ]))
  percentVar <- pca$sdev^2/sum(pca$sdev^2)
  names(percentVar) = colnames(pca$x)
  pca$percentVar = percentVar
  return(pca)}
pc = pca_loadings(vst)
comps = data.frame(pc$x)
comps$name = rownames(comps)
library(dplyr)
comps = comps %>% left_join(summarydata, by=c("name"="name"))
colorby = "name"
```

```{r pca-plotter}
#pca-plotter
pca_plot = function(comps, nc1, nc2, colorby, shapeby) {
   c1str = paste0("PC", nc1)
   c2str = paste0("PC", nc2)
  if(!(c1str %in% colnames(comps) && c2str %in% colnames(comps))) {
    warning("Higher order components not found, skipping plotting.")
    return(NA)
  }
  ggplot(comps, aes_string(c1str, c2str, color=colorby, shape=shapeby)) +
    geom_point(size=5) + theme_bw() +
    xlab(paste0(c1str, ": ", round(pc$percentVar[nc1] * 100), "% variance")) +
    ylab(paste0(c2str, ": ", round(pc$percentVar[nc2] * 100), "% variance"))
  }
```

### PC1 vs. PC2
```{r PC1-PC2}
comps$replicate <- as.character(comps$replicate)
pca_plot(comps, 1, 2, colorby="treatment", shapeby="replicate")
```

### PC3 vs. PC4
```{r PC3-PC4}
pca_plot(comps, 3, 4, colorby, shapeby="replicate")
```

### PC5 vs. PC6
```{r PC5-PC6}
pca_plot(comps, 5, 6, colorby, shapeby="replicate")
```

### Variance explained by component
```{r component-loadings}
ggplot(data.frame(component=reorder(names(pc$percentVar), -pc$percentVar),
                  percent_var=pc$percentVar),
       aes(component, percent_var)) + geom_bar(stat='identity') +
  ylab("percent of total variation") + xlab("") + theme_bw()
```

---

# SessionInfo

```{r sessioninfo}
sessionInfo()
```


# References